### 2020-09-01 README.Rmd
Initial commit

---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "README-"
)
# Please put your title here to include it in the file below.
Title <- "Species & genus plant classification with NEON hyperspectral data"
```

# neonHScompendium

[![Binder](https://mybinder.org/badge_logo.svg)](https://mybinder.org/v2/gh///master?urlpath=rstudio)

This repository contains the data and code for our paper:

> Authors, (YYYY). _`r Title`_. Name of journal/book <https://doi.org/xxx/xxx>

Our pre-print is online here:

> Authors, (YYYY). _`r Title`_. Name of journal/book, Accessed `r format(Sys.Date(), "%d %b %Y")`. Online at <https://doi.org/xxx/xxx>

### How to cite

Please cite this compendium as:

> Authors, (`r format(Sys.Date(), "%Y")`). _Compendium of R code and data for `r Title`_. Accessed `r format(Sys.Date(), "%d %b %Y")`. Online at <https://doi.org/xxx/xxx>

## Contents

The **analysis** directory contains:

  - [:file\_folder: paper](/analysis/paper): R Markdown source document
    for manuscript. Includes code to reproduce the figures and tables
    generated by the analysis. It also has a rendered version,
    `paper.docx`, suitable for reading (the code is replaced by figures
    and tables in this file)
  - [:file\_folder: data](/analysis/data): Data used in the analysis.
  - [:file\_folder: figures](/analysis/figures): Plots and other
    illustrations
  - [:file\_folder:
    supplementary-materials](/analysis/supplementary-materials):
    Supplementary materials including notes and other documents
    prepared and collected during the analysis.

## How to run in your broswer or download and run locally

This research compendium has been developed using the statistical programming
language R. To work with the compendium, you will need
installed on your computer the [R software](https://cloud.r-project.org/)
itself and optionally [RStudio Desktop](https://rstudio.com/products/rstudio/download/).

The simplest way to explore the text, code and data is to click on
[binder](https://mybinder.org/v2/gh///master?urlpath=rstudio)
to open an instance of RStudio in your browser, which will have the
compendium files ready to work with. Binder uses rocker-project.org
Docker images to ensure a consistent and reproducible computational
environment. These Docker images can also be used locally.

You can download the compendium as a zip from from this URL:
[master.zip](/archive/master.zip). After unzipping:
- open the `.Rproj` file in RStudio
- run `devtools::install()` to ensure you have the packages this analysis depends on (also listed in the
[DESCRIPTION](/DESCRIPTION) file).
- finally, open `analysis/paper/paper.Rmd` and knit to produce the `paper.docx`, or run `rmarkdown::render("analysis/paper/paper.Rmd")` in the R console

### Licenses

**Text and figures :**  [CC-BY-4.0](http://creativecommons.org/licenses/by/4.0/)

**Code :** See the [DESCRIPTION](DESCRIPTION) file

**Data :** [CC-0](http://creativecommons.org/publicdomain/zero/1.0/) attribution requested in reuse

### Contributions

We welcome contributions from everyone. Before you get started, please see our [contributor guidelines](CONTRIBUTING.md). Please note that this project is released with a [Contributor Code of Conduct](CONDUCT.md). By participating in this project you agree to abide by its terms.

### 2020-09-01 analysis/paper/paper.Rmd
Initial commit

---
title: "Species & genus plant classification with NEON hyperspectral data"
author:
  - Victoria Scholl\thanks{maxwell.b.joseph@colorado.edu, Earth Lab, University of Colorado, Boulder, CO 80303, USA}
  - Maxwell B. Joseph\thanks{maxwell.b.joseph@colorado.edu, Earth Lab, University of Colorado, Boulder, CO 80303, USA}

#date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  bookdown::pdf_document2:
    keep_tex: true
    toc: false
    includes:
      in_header: header.sty
bibliography: references.bib
csl: "ecology.csl" # Insert path for the bib-style
abstract: |
  Text of abstract
keywords: |
  keyword 1; keyword 2; keyword 3
highlights: |
  These are the highlights. 
---

<!-- This is the format for text comments that will be ignored during renderings. Do not put R code in these comments because it will not be ignored. -->

<!-- With the following code you can access and display values from the yml header above. -->

Keywords: `r rmarkdown::metadata$keywords`

Highlights: `r rmarkdown::metadata$highlights`

<!-- The following code chunk defines some general settings how code chunks should behave. -->

```{r setup, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  warning = FALSE,
  message = FALSE,
  echo = FALSE,
  comment = "#>",
  fig.path = "../figures/",
  dpi = 300
)
```

<!-- The actual document text starts here: -->

# Introduction

Here is a citation [@Marwick2017]

# Background

# Methods

# Results

<!-- Here's some example analysis code: -->

```{r get-data, eval = FALSE}
# Note the path that we need to use to access our data files when rendering this document
my_data <- read.csv(here::here('analysis/data/raw_data/my_csv_file.csv'))
```

```{r demo-plot, fig.cap="A plot of random numbers"}
plot(rnorm(10))
```

Figure \@ref(fig:demo-plot) shows how we can have a caption and cross-reference for a plot

```{r demo-inline-code}
x <- round(pi, 2)
```

Here is an example of inline code `r x` in the middle of a sentence.

# Discussion

# Conclusion

# Acknowledgements

<!-- The following line inserts a page break  -->

\newpage

# References

<!-- The following line ensures the references appear here for the MS Word or HTML output files, rather than right at the end of the document (this will not work for PDF files):  -->

<div id="refs"></div>

\newpage

### Colophon

This report was generated on `r Sys.time()` using the following computational environment and dependencies:

```{r colophon, cache = FALSE}
# which R packages and versions?
if ("devtools" %in% installed.packages()) devtools::session_info()
```

The current Git commit details are:

```{r}
# what commit is this file at? 
if ("git2r" %in% installed.packages() & git2r::in_repository(path = ".")) git2r::repository(here::here())  
```

### 2020-09-01 analysis/paper/paper.Rmd
adding gitignores to remove certain files from github

@@
 author:
-  - Victoria Scholl\thanks{maxwell.b.joseph@colorado.edu, Earth Lab, University of Colorado, Boulder, CO 80303, USA}
+  - Victoria Scholl\thanks{victoria.scholl@colorado.edu, Earth Lab and Geography Dept., University of Colorado, Boulder, CO 80303, USA}
   - Maxwell B. Joseph\thanks{maxwell.b.joseph@colorado.edu, Earth Lab, University of Colorado, Boulder, CO 80303, USA}
@@
 
-Here is a citation [@Marwick2017]
+The National Ecological Observatory Network (NEON) is a valuable source of publicly available ecological data across the United States [@keller2008continental].
+
+We have extracted all NEON AOP hyperspectral data for every mapped stem in NEON plots where field data and airborne remote sensing data were collected during the same year.
+
+Here's the code to extract the spectra: [https://gist.github.com/mbjoseph/5c18781e508460e14f64193571b98b7d](https://gist.github.com/mbjoseph/5c18781e508460e14f64193571b98b7d)
+
+We used a machine learning approach to evaluate plant identification potential using the hyperspectral data at multiple taxonomic resolutions: species and genus levels.
+
+Species classification is an active research area [@scholl2020integrating].
 
@@
 # Methods
+
+## Data cleaning
+
+## Species classification
+
+## Genus classification
+
+Same setup as species classification, but used the genus as each class label.
 
@@
 
+Max has cool ideas and knows many R tricks
+
 <!-- The following line inserts a page break  -->
@@
 
-### Colophon
+\clearpage
 
-This report was generated on `r Sys.time()` using the following computational environment and dependencies:
+# Figure legends {-}
 
-```{r colophon, cache = FALSE}
-# which R packages and versions?
-if ("devtools" %in% installed.packages()) devtools::session_info()
+## Figure 1 {-}
+
+Figure 1 info...
+
+## Figure 2 {-}
+
+It ain't much...
+
+# Figures {-}
+
+## Figure 1 {-}
+
+```{r spectra, fig.cap = "", echo=FALSE,  out.width = "450px", message = FALSE, warning=FALSE}
+library(here)
+knitr::include_graphics(here::here("analysis", "figures", "spectra_figure_test.png"))
 ```
 
-The current Git commit details are:
+\clearpage
 
-```{r}
-# what commit is this file at? 
-if ("git2r" %in% installed.packages() & git2r::in_repository(path = ".")) git2r::repository(here::here())  
+## Figure 2 {-}
+
+```{r excel, fig.cap = "", echo=FALSE,  out.width = "450px", message = FALSE, warning=FALSE}
+knitr::include_graphics(here::here("analysis", "figures", "honest_work.jpeg"))
 ```
+
+\clearpage

### 2020-09-01 analysis/paper/paper.Rmd
updated pdf formatting
line numbers and removed the kaywords and highlights

@@
 ---
-title: "Species & genus plant classification with NEON hyperspectral data"
+title: \textbf{Species and genus plant classification with NEON hyperspectral remote sensing data}
+
 author:
@@
 
-#date: "`r format(Sys.time(), '%d %B, %Y')`"
 output:
@@
 bibliography: references.bib
-csl: "ecology.csl" # Insert path for the bib-style
-abstract: |
-  Text of abstract
-keywords: |
-  keyword 1; keyword 2; keyword 3
-highlights: |
-  These are the highlights. 
+csl: ecology.csl 
 ---
-
-<!-- This is the format for text comments that will be ignored during renderings. Do not put R code in these comments because it will not be ignored. -->
-
-<!-- With the following code you can access and display values from the yml header above. -->
-
-Keywords: `r rmarkdown::metadata$keywords`
-
-Highlights: `r rmarkdown::metadata$highlights`
-
-<!-- The following code chunk defines some general settings how code chunks should behave. -->
 
@@
   message = FALSE,
-  echo = FALSE,
+  echo = TRUE,
   comment = "#>",
-  fig.path = "../figures/",
   dpi = 300
@@
 
-# Introduction
+\linenumbers
+
+\begin{abstract}
+Abstract text here
+\end{abstract}
+
+```{r, results='hide', message=FALSE, include=FALSE}
+# Run the main analysis
+```
+
+# Introduction {-}
 
@@
 
-# Background
+# Background {-}
 
-# Methods
+# Methods {-}
 
-## Data cleaning
+## Data cleaning {-}
 
-## Species classification
+## Species classification {-}
 
-## Genus classification
+## Genus classification {-}
 
@@
 
-# Results
+# Results {-}
 
@@
 
-# Discussion
+# Discussion {-}
 
-# Conclusion
-
-# Acknowledgements
+# Acknowledgements {-}
 
@@
 
-\newpage
+\clearpage
 
-# References
-
-<!-- The following line ensures the references appear here for the MS Word or HTML output files, rather than right at the end of the document (this will not work for PDF files):  -->
+# References {-}
 
@@
 
-Figure 1 info...
+Figure 1 description...
 

### 2020-09-02 R/spectral-lib.Rmd
adding Max's gist code for extracting the spectral data

---
title: "Extracting all spectra for NEON mapped stems"
author: "Max Joseph"
date: "2020-06-11"
output: html_document
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r load-libraries, message=FALSE}
library(tidyverse)
library(neonUtilities)
#library(devtools)
#install_github('NEONScience/NEON-geolocation/geoNEON', dependencies=TRUE)
library(geoNEON)
library(sp)
#devtools::install_github('earthlab/neonhs')
library(neonhs)
library(Rtsne)
library(ggthemes)
```

Acquire vegetation data:

```{r get-veg-data}
dir.create('data')
zipsByProduct(dpID = "DP1.10098.001",
              savepath = "data",
              check.size = FALSE)
stackByTable("data/filesToStack10098", folder = TRUE)
```

Read the mapping data and apparent individual data:

```{r read-map}
vegmap <- read_csv("data/filesToStack10098/stackedFiles/vst_mappingandtagging.csv") %>%
  filter(!is.na(stemAzimuth), !is.na(pointID))
vegind <- read_csv("data/filesToStack10098/stackedFiles/vst_apparentindividual.csv")
```

Get precise locations for tagged plants and merge the mapping and individual data:

```{r get-precise-locs, warning=FALSE}
vegmap <- getLocTOS(vegmap, "vst_mappingandtagging")

samples_to_keep <- vegind %>%
  filter(plantStatus == "Live") %>%
  mutate(year = lubridate::year(date)) %>%
  group_by(individualID) %>%
  summarize(year = list(seq(min(year), max(year), by = 1))) %>%
  ungroup %>%
  unnest(year)

samples_to_keep %>%
  count(individualID) %>%
  arrange(-n)

veg <- as_tibble(vegmap) %>%
  group_by(individualID) %>%
  # some records are revised at later dates with better coordinates
  filter(!is.na(adjDecimalLatitude),
         date == max(date)) %>%
  ungroup %>%
  select(-date) %>%
  right_join(samples_to_keep) %>%
  distinct(individualID, year, .keep_all = TRUE)

veg %>%
  count(individualID, year) %>%
  arrange(-n)

veg %>%
  count(individualID) %>%
  arrange(-n)

veg %>%
  filter(individualID == "NEON.PLA.D03.JERC.02221") %>%
  View

nrow(vegmap)
nrow(veg)
```

Save output file:

```{r write-veg}
write_csv(veg, 'neon-veg.csv')
```

Visualize the imbalance of species identity:

```{r make-veg-tibble}
species_counts <- veg %>%
  count(scientificName) %>%
    separate(scientificName, into = c('genus', 'species'), sep = ' ',
           extra = 'drop', remove = FALSE) %>%
  mutate(name = paste(genus, species)) %>%
  arrange(-n) %>%
  mutate(scientific_name = fct_reorder(name, n),
         order = 1:n()) %>%
  dplyr::select(scientific_name, order, n)

species_counts %>%
  ggplot(aes(x = n, y = order)) +
  scale_x_log10() +
  geom_point(color = 'dodgerblue') +
  xlab('Number of mapped stem locations') +
  ylab('Order: commonest to rarest') +
  theme_minimal() +
  theme(panel.grid.minor = element_blank())
ggsave('~/Desktop/veg-freq.png', width = 4.5, height = 3)

# number of mapped stems
sum(species_counts$n)

# number of singletons
species_counts %>%
  filter(n == 1)

# number of species with fewer than 10 records
species_counts %>%
  filter(n < 10)

# number of species with fewer than 100 records
species_counts %>%
  filter(n < 100)
```

Split the veg data into different datasets for sites and year:

```{r split-veg}
split_veg <- veg %>%
  mutate(siteID_year = paste(siteID, year, sep = '_')) %>%
  split(.$siteID_year)
```

For each site-year combination, extract hyperspectral data for all plants:

```{r extract-data}
spectra <- vector(mode = 'list', length = length(split_veg))
dir.create('hs_data', showWarnings = FALSE)
csv_dir <- file.path("hs_data", "tabular")
dir.create(csv_dir, recursive = TRUE, showWarnings = FALSE)

try_extract <- function(path, pts) {
  out <- tryCatch(hs_extract_pts(path, pts, bands = 1:426),
                  error = function(e) NULL)
  return(out)
}

for (i in seq_along(split_veg)) {
  print(paste("processing year/site", i, "of", length(split_veg)))
  subd <- split_veg[[i]]
  identifier <- names(split_veg)[i]
  outfile <- file.path(csv_dir, paste0(identifier, ".csv"))
  if (!file.exists(outfile)) {
    h5_dir <- file.path('hs_data', identifier)
    dir.create(h5_dir, showWarnings = FALSE)
    siteyear <- identifier %>% str_split('_') %>% unlist
    site_code <- siteyear[1]
    year <- siteyear[2]

    # download data
    try(byTileAOP(dpID = "DP3.30006.001", site = site_code, year = year,
            easting = subd$adjEasting, northing = subd$adjNorthing,
            check.size = FALSE, savepath = h5_dir))
    hs_paths <- list.files(path = h5_dir, pattern = 'reflectance.h5',
                           recursive = TRUE, full.names = TRUE)
    if (length(hs_paths) > 0) {
      # extract spectra
      spdf <- SpatialPointsDataFrame(subd[, c('adjEasting', 'adjNorthing')],
                                 data = subd,
                                 proj4string = CRS(hs_proj4string(hs_paths[1])))
      out <- list()
      for (j in seq_along(hs_paths)) {
        res <- try_extract(hs_paths[j], spdf)
        first_band <- grep('^band1', names(res), value = TRUE)[1]
        if (!is.na(first_band)) {
          na_vals <- is.na(res[[first_band]])
          out[[i]] <- res[!na_vals, ]
        }
      }
      hs_df <- lapply(out, as.data.frame) %>%
        bind_rows %>%
        as_tibble %>%
        select(individualID, namedLocation,
               year, uid, namedLocation,
               adjEasting, adjNorthing,
               taxonID, taxonRank, scientificName,
               starts_with('band')) %>%
        distinct
      write_csv(hs_df, outfile)
      unlink(h5_dir, recursive = TRUE, force = TRUE)
    }
  }
}
```

Now, join all of the data together, and create one big data frame.

```{r join-df}
all_spectra <- list.files(csv_dir, full.names = TRUE) %>%
  lapply(function(f) {
    read_csv(f) %>%
      pivot_longer(starts_with("band")) %>%
      rename(reflectance = value) %>%
      separate(name, into = c("band_idx", "wavelength_nm")) %>%
      mutate(wavelength_nm = parse_number(wavelength_nm))
  })

nrows <- lapply(all_spectra, nrow) %>%
  unlist

all_spectra <- all_spectra[nrows > 0] %>%
  bind_rows %>%
  mutate(mask = case_when(
  .$wavelength_nm > 1340 & .$wavelength_nm < 1445 ~ TRUE,
  .$wavelength_nm > 1790 & .$wavelength_nm < 1955 ~ TRUE,
  .$wavelength_nm > 2400 ~ TRUE,
  TRUE ~ FALSE),
  spectraID = paste(individualID, year, sep = "_")) %>%
  select(spectraID, everything())

write_csv(all_spectra, "all_spectra.csv")
```

```{r plot-spectra}
all_spectra <- read_csv("all_spectra.csv")

p <- all_spectra %>%
  group_by(scientificName) %>%
  mutate(n = n()) %>%
  ggplot(aes(wavelength_nm, ifelse(mask, NA, reflectance), group = individualID)) +
  geom_path(alpha = .2) +
  xlab('Wavelength (nm)') +
  ylab('Reflectance') +
  facet_wrap(~reorder(scientificName, -n), labeller = label_wrap_gen()) +
  theme_minimal()
ggsave("spectra.png", plot = p, width = 15, height = 15)
```

Look at the sample sizes per species

```{r ss}
all_spectra %>%
  distinct(scientificName, spectraID) %>%
  count(scientificName) %>%
  ggplot(aes(x = n, y = reorder(scientificName, -n))) +
  geom_point() +
  ylab("Species") +
  xlab("Number of spectra") +
  theme(axis.text.y = element_text(size = 7)) +
  scale_x_log10()
```

Plot repeat samples of individuals:

```{r check-dups}
all_spectra %>%
  count(individualID) %>%
  arrange(-n) %>%
  top_n(1) %>%
  `[`(1:56, ) %>%
  left_join(all_spectra) %>%
  mutate(reflectance = ifelse(mask, NA, reflectance)) %>%
  ggplot(aes(x = wavelength_nm, y = reflectance, color = factor(year))) +
  geom_line() +
  facet_wrap(~individualID)
```

### 2020-09-02 R/spectral-lib.Rmd
setting a directory for the spectral lib gist
to avoid having tons of files living in the R folder, keep them in the raw_data folder for now

@@
 ```{r get-veg-data}
-dir.create('data')
+# set output directory within the analysis/data/raw_data path 
+out_dir <- here::here("analysis","data","raw_data","spectral-lib")
+dir.create(out_dir)
+dir.create(file.path(out_dir, 'data'))
+
 zipsByProduct(dpID = "DP1.10098.001",
-              savepath = "data",
+              savepath = file.path(out_dir, 'data'),
               check.size = FALSE)
-stackByTable("data/filesToStack10098", folder = TRUE)
+stackByTable(file.path(out_dir, "data","filesToStack10098"), folder = TRUE)
 ```

### 2020-09-02 R/spectral-lib.Rmd
updated the file paths
to keep all the spectral lib files in the analysis data directory

@@
 library(ggthemes)
+library(dplyr)
 ```
@@
 ```{r get-veg-data}
-# set output directory within the analysis/data/raw_data path 
+# VS-NOTE: set output directory within the analysis/data/raw_data path 
 out_dir <- here::here("analysis","data","raw_data","spectral-lib")
@@
 ```{r read-map}
-vegmap <- read_csv("data/filesToStack10098/stackedFiles/vst_mappingandtagging.csv") %>%
+vegmap <- read_csv(file.path(out_dir, "data", "filesToStack10098", "stackedFiles", 
+                            "vst_mappingandtagging.csv")) %>%
   filter(!is.na(stemAzimuth), !is.na(pointID))
-vegind <- read_csv("data/filesToStack10098/stackedFiles/vst_apparentindividual.csv")
+vegind <- read_csv(file.path(out_dir, "data", "filesToStack10098", "stackedFiles",
+                             "vst_apparentindividual.csv"))
 ```
@@
 ```{r write-veg}
-write_csv(veg, 'neon-veg.csv')
+write_csv(veg, file.path(out_dir, "neon-veg.csv"))
 ```
@@
   theme(panel.grid.minor = element_blank())
-ggsave('~/Desktop/veg-freq.png', width = 4.5, height = 3)
+ggsave(file.path(out_dir, "veg-freq.png"), width = 4.5, height = 3)
 
@@
 spectra <- vector(mode = 'list', length = length(split_veg))
-dir.create('hs_data', showWarnings = FALSE)
-csv_dir <- file.path("hs_data", "tabular")
+dir.create(file.path(out_dir, "hs_data"), showWarnings = FALSE)
+csv_dir <- file.path(out_dir, "hs_data", "tabular")
 dir.create(csv_dir, recursive = TRUE, showWarnings = FALSE)

### 2020-09-09 R/spectral-lib.Rmd

@@
 library(neonUtilities)
-#library(devtools)
-#install_github('NEONScience/NEON-geolocation/geoNEON', dependencies=TRUE)
+#install.packages('remotes')
+#remotes::install_github('NEONScience/NEON-geolocation/geoNEON', dependencies=TRUE)
 library(geoNEON)
 library(sp)
-#devtools::install_github('earthlab/neonhs')
+#remotes::install_github('earthlab/neonhs')
 library(neonhs)
-library(Rtsne)
-library(ggthemes)
 library(dplyr)
+library(here)
 ```
@@
 ```{r get-veg-data}
-# VS-NOTE: set output directory within the analysis/data/raw_data path 
 out_dir <- here::here("analysis","data","raw_data","spectral-lib")
-dir.create(out_dir)
-dir.create(file.path(out_dir, 'data'))
+dir.create(out_dir, recursive = TRUE, showWarnings = FALSE)
 
 zipsByProduct(dpID = "DP1.10098.001",
-              savepath = file.path(out_dir, 'data'),
+              savepath = out_dir,
               check.size = FALSE)
-stackByTable(file.path(out_dir, "data","filesToStack10098"), folder = TRUE)
+stackByTable(file.path(out_dir, "filesToStack10098"), folder = TRUE)
 ```
@@
 ```{r read-map}
-vegmap <- read_csv(file.path(out_dir, "data", "filesToStack10098", "stackedFiles", 
+vegmap <- read_csv(file.path(out_dir, "filesToStack10098", "stackedFiles", 
                             "vst_mappingandtagging.csv")) %>%
   filter(!is.na(stemAzimuth), !is.na(pointID))
-vegind <- read_csv(file.path(out_dir, "data", "filesToStack10098", "stackedFiles",
+vegind <- read_csv(file.path(out_dir, "filesToStack10098", "stackedFiles",
                              "vst_apparentindividual.csv"))
@@
   arrange(-n)
-
-veg %>%
-  filter(individualID == "NEON.PLA.D03.JERC.02221") %>%
-  View
 
@@
 
-write_csv(all_spectra, "all_spectra.csv")
+write_csv(all_spectra, file.path(out_dir, "all_spectra.csv"))
 ```
-
-```{r plot-spectra}
-all_spectra <- read_csv("all_spectra.csv")
-
-p <- all_spectra %>%
-  group_by(scientificName) %>%
-  mutate(n = n()) %>%
-  ggplot(aes(wavelength_nm, ifelse(mask, NA, reflectance), group = individualID)) +
-  geom_path(alpha = .2) +
-  xlab('Wavelength (nm)') +
-  ylab('Reflectance') +
-  facet_wrap(~reorder(scientificName, -n), labeller = label_wrap_gen()) +
-  theme_minimal()
-ggsave("spectra.png", plot = p, width = 15, height = 15)
-```
-
-Look at the sample sizes per species
-
-```{r ss}
-all_spectra %>%
-  distinct(scientificName, spectraID) %>%
-  count(scientificName) %>%
-  ggplot(aes(x = n, y = reorder(scientificName, -n))) +
-  geom_point() +
-  ylab("Species") +
-  xlab("Number of spectra") +
-  theme(axis.text.y = element_text(size = 7)) +
-  scale_x_log10()
-```
-
-Plot repeat samples of individuals:
-
-```{r check-dups}
-all_spectra %>%
-  count(individualID) %>%
-  arrange(-n) %>%
-  top_n(1) %>%
-  `[`(1:56, ) %>%
-  left_join(all_spectra) %>%
-  mutate(reflectance = ifelse(mask, NA, reflectance)) %>%
-  ggplot(aes(x = wavelength_nm, y = reflectance, color = factor(year))) +
-  geom_line() +
-  facet_wrap(~individualID)
-```

### 2020-09-10 analysis/paper/paper.Rmd
adding Max's gist code for extracting the spectral data
added outline for paper

@@
 ```{r, results='hide', message=FALSE, include=FALSE}
+# Extract hyperspectral reflectance data for each individual plant
+#R/spectral-lib.Rmd
+
 # Run the main analysis
+
 ```
@@
 # Introduction {-}
+
+1 - motivation for work, why bother?
+  - unprecedented access to unique hyperspectral airborne imagery from NEON (cite)
+  - acknowledge additional data to add info (lidar, rgb) (cite) 
+  - how meaningful can the HS data alone be to predict species? accuracies? 
+  - we expect there to be meaningful information in the HS data for species/genus
+2 - knowledge gaps in lit/field?
+  - how useful are HS data across many species, ecosystems, geographies? 
+  - where is the most important information contained? which wavelengths most important for species discrimination? differ across sites? (cite) 
+  - generally more site-specific studies? 
+  - prediction of species vs genus? How much easier is genus? Or is this difficult too? (cite hierarchical taxonomic classification recent study)
+3 - how we addressed the knowledge gaps
+  - the need to generalize methods across locations? --> NEON
+  - which wavelengths are most important
+  - data harmonization across NEON sites (why are the wavelengths different? Ask NEON person?)
 
@@
 # Methods {-}
+NEON data collection protocols
+AOP flights
+Plant mapping 
+Network of sites and observations (how many? ecoclimatic domains? ecosystem coverage?
+
+Figure - NEON sites and domains, which sites do we have data from? # individuals from each site? 
+Acknowledge imbalance spatially and temporally...
+Any sites not represented? Like Hawaii
+
+## Data extraction {-}
+
+Extracted hyperspectral data for all mapped stems at NEON sites
+Assumed plant locations don't change over time 
+Filtered plants to keep only "live" ones
+NEON data products (cite)
+Spectrometer orthorectified surface directional reflectance mosaic
+Woody plant vegetation structure field based measurements
+Mask created for water absorption bands (cite)
+Packages neonUtilities, geoNEON, neonhs (cite)
 
@@
 
+? Filtered taxonomic levels, labels species & genus
+Wavelength interpolation - filtered out the PUUM site in Hawaii because of wavelengths
+Filtering based on # samples
+Train/validate split 80/20
+
+Figure - histogram of species counts, show which ones were used to train
+Figure - cleaned data, make it legible
+
 ## Species classification {-}
+
+Random forest classifier
+Hyperparameter optimization (mtry, ntree)
+Overall accuracy, confusion matrices, variable importance
 
@@
 # Results {-}
+
+Classification accuracy was higher at the genus level (60%) compared to the species level (55%)
+
+Figure - confusion matrices
+Figure - variable importance vs wavelength
 
@@
 # Discussion {-}
+
+Restate findings at high level
+
+How do our findings compare to other similar studies. Consistent? Differences? (cite IDTReeS) Other ranges of accuracy using NEON HS data? How many sites too?
+
+Potential limitations
+  - class imbalance - we tried to filter based on # samples, but our model has then only seen common species and would be unable to identify rare ones
+  - filtering shadows, non-vegetation pixels.... (cite a paper from last IDTReeS comp)
+
+Future directions, next steps
+  - targeted species/genus for increased sampling to improve representation in the data set?
+   - more advanced models. ML DL
+   - RF treats each feature as independent
+   - spectra are sequences of wavelengths - explore DL methods for sequences (i.e. recurrent neural networks)
+   - hierarchical taxonomic approach - borrow information from closely related species (i.e. common and rare species > use common species observations)
+   - (cite our other paper) intelligent crown geometry processing 
+   - (ben weinstein) tree crown detection - extract multiple pixels per individual using rectangular crown boundaries
+   - herbarium / hyperspectral data measurements at other scales. handheld spectrometer. other spectral libraries. 
+   - combine HS with RGB, lidar
+
+We conclude that HS data allow you to get __ accuracy
+Hopes for outcomes
+
+# Appendix / Supplement ideas {-}
+
+- Species histogram to show which ones were excluded for few # samples 
+- Wavelength interpolation figures?
 

### 2020-09-10 R/clean_spectral_data.Rmd
added scripts for data cleaning and classifying

---
title: "Clean spectral data"
author: "Victoria Scholl"
date: "09/10/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(here)
library(dplyr)
library(ggplot2)
library(stringr)
library(tidyr)
library(tidyverse)
library(kableExtra)
library(reshape2)
library(pbapply)
```

### Explore data

Read original spectra, extracted using NEON data (spectral-lib.Rmd).

Columns:

* **individualID** - unique identifier for each individual plant. Individuals can have multiple measurements over time.
* **uid** - unique identifier for the observation/record
* **spectraID** - unique identifier for each reflectance spectrum (a combination of individualID and year, YYYY )
* **namedLocation** - name of measurement location in NEON database
* **utmZone, adjEasting, adjNorthing** - UTM zone and coordinates of mapped stem
* **taxonID** - species code
* **scientificName** - scientific name, associated with taxonID, lowest taxonomic rank that can be determined
* **taxonRank** - lowest level taxonomic rank that can be determined for the individual (genus, species, variety)
* **band_idx** - number of the spectral band (band1, band 2, ... band 426)
* **wavelength_nm** - band wavelength in units of nm
* **reflectance** - decimal reflectance [0,1] extracted from NEON NIS hyperspectral imagery collected during same year
* **mask** - TRUE/LFALSE to identify water absorption bands

```{r}
# read_csv --> no factors, creates a tibble
all_spectra <- read_csv(here::here("analysis","data","raw_data",
                                   "spectral-lib","all_spectra.csv"))

# quick look at the values in each column
glimpse(all_spectra)
```

Number of unique **individualID**, **uid**, **spectraID** values:

```{r, message=TRUE}
message("# of unique individualID values:")
length(unique(all_spectra$individualID))

message("# of unique uid values:")
length(unique(all_spectra$uid))

message("# of unique spectraID (uid + year) values:")
length(unique(all_spectra$spectraID))
```

Number of unique **scientificName** & **taxonID** values (these should match):

```{r}
message("# of unique taxonID values:")
length(unique(all_spectra$taxonID))

message("# of unique scientificName values:")
length(unique(all_spectra$scientificName))
```

How many taxonomic identifications are at the genus vs species level? (Using the NEON-provided taxon Rank data)

```{r}
species_count <- all_spectra %>% 
  dplyr::distinct(spectraID, .keep_all = TRUE) %>%
  dplyr::group_by(taxonRank) %>% 
  dplyr::tally() %>% 
  kableExtra::kable() %>%
  kableExtra::kable_styling(bootstrap_options = "striped", 
                            full_width = F, 
                            position = "left")

species_count
```

Filter the spectra to keep only those classified at the genus, species, and variety taxon ranks. Add new columns for genus and species by separating the **scientificName** field.

```{r}
# remove entries with taxonRank kingdom, family (maintain only genus, species, variety ranks)
all_spectra <- all_spectra %>% 
  dplyr::filter(taxonRank %in% c("genus", "species", "variety"))

# create new columns for genus and species (disregard variations within species) 
all_spectra <- all_spectra %>% 
  tidyr::separate(scientificName, into = c('genus', 'species'), sep = ' ', 
           extra = 'drop', remove = FALSE) %>% 
  # add a column combining genus and species for classification
  dplyr::mutate(genusSpecies = paste(genus,species))

message("# of unique genus values:")
length(unique(all_spectra$genus))

message("# of unique species values:")
length(unique(all_spectra$species))

message("# of unique genusSpecies values")
length(unique(all_spectra$genusSpecies))
```

```{r}
# example where there are multiple variations of a single genusSpecies value
all_spectra %>% 
  filter(genus == "Pseudotsuga") %>% 
  select(taxonID, genusSpecies, scientificName) %>% 
  distinct()
```

### Visualize spectra

Based on Max's Gist code: https://gist.github.com/mbjoseph/5c18781e508460e14f64193571b98b7d

Plot all spectra; color the spectra based on genus/species taxonomic level.

```{r, warning=FALSE, fig.height=16, fig.width=14}
all_spectra %>%
  ggplot(aes(wavelength_nm, ifelse(mask, NA, reflectance), group = spectraID,
             # color each facet plot based on taxonRank
             color = taxonRank)) +
  geom_path(alpha = .2) +
  labs(x = '\n Wavelength (nm)',
       y = 'Reflectance \n',
       color = "Taxon Rank") + 
  facet_wrap(~genusSpecies
             # wrap long names
             ,labeller = label_wrap_gen(20)) + 
  scale_color_manual(values = c("#e6550d","#1c9099", "#a8ddb5")) + 
  theme_minimal() + 
  # Adjust the text size and rotation angle 
  theme(text = element_text(size=11),
        axis.text.x = element_text(angle=90, hjust=1),
        axis.title.y = element_text(size = 14),
        axis.title.x = element_text(size = 14),
        legend.text = element_text(size = 14),
        legend.title = element_text(size = 14))
```

### Explore the bands

Look at the starting wavelength ("band1") for each reflectance spectrum:

```{r}
# how many different starting wavelength values in the data set? 
count_min_wavelengths <- all_spectra %>%
  dplyr::filter(band_idx == "band1") %>% 
  dplyr::group_by(wavelength_nm) %>%
  dplyr::count(wavelength_nm)

kableExtra::kable(count_min_wavelengths) %>%
  kableExtra::kable_styling(bootstrap_options = "striped", 
                            full_width = F, 
                            position = "left")
```

Filter spectra with starting wavelengths <380nm for now, since they complicate the wavelength interpolation.
All spectra with a starting wavelength of 347nm come from the [PUUM](https://www.neonscience.org/field-sites/field-sites-map/PUUM) site in Hawaii - why? TBD.

```{r}
# identify spectraID values that start at 347nm
spectra_to_remove <- all_spectra %>%
  dplyr::filter(band_idx == "band1") %>% 
  dplyr::filter(wavelength_nm == 347) %>% 
  dplyr::select(spectraID)

# remove these spectra from the data set
all_spectra <- all_spectra %>% 
  dplyr::filter(!(spectraID %in% spectra_to_remove$spectraID))

# how many different starting wavelength values in the filtered data set? 
count_min_wavelengths <- all_spectra %>%
  dplyr::filter(band_idx == "band1") %>% 
  dplyr::group_by(wavelength_nm) %>%
  dplyr::count(wavelength_nm)

kableExtra::kable(count_min_wavelengths) %>%
  kableExtra::kable_styling(bootstrap_options = "striped", 
                            full_width = F, 
                            position = "left")
```

Spectra have different starting wavelengths (i.e. 381, 383, 385nm...) and increase in increments of 5nm. Most of them start at 384nm.

```{r}
set.seed(44) #for obama

# find individualIDs that start with each of these wavelength band1 values
for(wl in unique(count_min_wavelengths$wavelength_nm)){

  tmp <- all_spectra %>% 
    dplyr::filter(band_idx == "band1") %>% 
    dplyr::filter(wavelength_nm == wl) %>% 
    dplyr::select(spectraID) %>% 
    dplyr::sample_n(1)

  message(paste("spectraID starting at", wl, "nm: ", as.character(tmp$spectraID)))

}
```

```{r warning=FALSE, fig.width=12, include=FALSE, eval=FALSE}
# plot spectra to see their starting/ending wavelengths
all_spectra %>% 
  dplyr::filter(spectraID %in% c(#"NEON.PLA.D20.PUUM.00954_2019",
                                 "NEON.PLA.D11.CLBJ.00496_2018", 
                                 "NEON.PLA.D01.HARV.09041_2018",
                                 "NEON.PLA.D08.DELA.03903_2017")) %>%
  ggplot(aes(x = wavelength_nm, y = reflectance, group = spectraID)) +
  geom_line(aes(color = spectraID), 
            alpha = 0.4) +
  labs(title = "Spectra with different starting wavelengths") + 
  theme_bw()
```

Beginning of wavelength range:

```{r warning=FALSE, fig.width=12}
# plot spectra to see their starting wavelengths
all_spectra %>% 
  dplyr::filter(spectraID %in% c(#"NEON.PLA.D20.PUUM.00954_2019",
                                 "NEON.PLA.D11.CLBJ.00496_2018", 
                                 "NEON.PLA.D01.HARV.09041_2018",
                                 "NEON.PLA.D08.DELA.03903_2017")) %>%
  ggplot(aes(x = wavelength_nm, y = reflectance, group = spectraID)) +
  geom_line(aes(color = spectraID), 
            alpha = 0.5, size = 1) +
  labs(title = "Spectra with different wavelengths: zoom in on beginning of spectra") + 
  # zoom in on beginning of spectra
  lims(x = c(380,400), y = c(0,0.1)) +
  theme_bw()
```

End of wavelength range:

```{r warning=FALSE, fig.width=12}
# plot spectra to see their ending wavelengths
all_spectra %>% 
  dplyr::filter(spectraID %in% c(#"NEON.PLA.D20.PUUM.00954_2019",
                                 "NEON.PLA.D11.CLBJ.00496_2018", 
                                 "NEON.PLA.D01.HARV.09041_2018",
                                 "NEON.PLA.D08.DELA.03903_2017")) %>%
  ggplot(aes(x = wavelength_nm, y = reflectance, group = spectraID)) +
  geom_line(aes(color = spectraID), 
            alpha = 0.5, size = 1) +
  labs(title = "Spectra each with different wavelengths: zoom in on end of spectra") + 
  # zoom in on end of spectra
  lims(x = c(2455,2515), y = c(0, 0.3)) + 
  theme_bw()
```

### Standardize the bands: Interpolate wavelengths

Interpolate wavelength values to have the same 426 bands starting at 384 nm, ending at 2512 nm, incrementing by 5 nm. To avoid data gaps on the ends, let's use a larger range.

```{r wavelengths_interp}
# define the target wavelengths.
# use a larger range to get data on the ends.
# with a 5nm spacing, add 10nm to each end, and 4 to the total number of wavelengths. 
min_wl <- 384 
max_wl <- 2512 
num_wl <- 426
wavelength_vals <- seq(min_wl - 10, max_wl + 10,
                       length.out = num_wl + 4)
```

```{r}
# get bad bands for the target wavelengths

# identify a spectraID starting at the target wavelengths
tmp_spectrum <- all_spectra %>% 
    dplyr::filter(band_idx == "band1") %>% 
    dplyr::filter(wavelength_nm == min_wl) %>% 
    dplyr::select(spectraID) %>% 
    dplyr::sample_n(1)

# list the wavelengths that are bad bands, indicated by mask == TRUE
bad_bands <- all_spectra %>% 
  dplyr::filter(spectraID == tmp_spectrum$spectraID) %>%
  dplyr::filter(mask==TRUE) %>%
  dplyr::select(wavelength_nm, mask) 
```

```{r, warning=FALSE}
# interpolate all spectra efficiently 
interp_wl <- all_spectra %>%
  split(.$spectraID) %>%
  pblapply(function(df) {

    interpolated_reflectance <- approx(x= df$wavelength_nm, 
                                       y = df$reflectance, 
                                       xout = wavelength_vals)
    tibble(spectraID = unique(df$spectraID), 
           band_idx = seq_along(wavelength_vals), 
           # set wavelength values as integers
           wavelength_nm = round(interpolated_reflectance$x), 
           reflectance = interpolated_reflectance$y)

  }) %>%
  bind_rows %>% 
  # remove rows with wavelength values outside the range of interest 
  dplyr::filter(wavelength_nm >= min_wl) %>% 
  dplyr::filter(wavelength_nm <= max_wl) %>% 
  # update the band_idx column to be band1, ... band426
  dplyr::group_by(spectraID) %>% 
  dplyr::arrange(wavelength_nm) %>%
  dplyr::mutate(band_idx = paste0("band", 1:n())) %>%
  dplyr::arrange(spectraID)
```

```{r}
# set reflectance values to NA for bad bands 
interp_wl$reflectance[interp_wl$wavelength_nm %in% bad_bands$wavelength_nm] <- NA
```

Now, each spectrum (designated by a unique spectraID value) has the same 426 wavelength values. Reflectance values are set to NA for the bad band wavelength values. 
Merge the interpolated wavelengths and reflectance values into a copy of the data frame.

```{r, message=FALSE}
# make a copy of the original spectra 
cleaned_spectra <- all_spectra %>% 
  dplyr::select(-c(wavelength_nm, reflectance)) %>% 
  # merge in the interpolated wavelengths
  dplyr::left_join(interp_wl)
```

```{r, fig.width=12}
# plot interpolated spectra to see their starting wavelengths
interp_wl %>% 
  dplyr::filter(spectraID %in% c(#"NEON.PLA.D20.PUUM.00954_2019",
                                 "NEON.PLA.D11.CLBJ.00496_2018", 
                                 "NEON.PLA.D01.HARV.09041_2018",
                                 "NEON.PLA.D08.DELA.03903_2017")) %>% 
  ggplot(aes(x = wavelength_nm, y = reflectance, group = spectraID)) +
  geom_line(aes(color = spectraID), 
            alpha = 0.5, size = 1) +
  labs(title = "Spectra w interpolated wavelengths: Zoom in on beginning of spectra") + 
  # zoom in on beginning of spectra
  lims(x = c(380,400), y = c(0,0.1)) +
  theme_bw()
```

### Partition data into train, evaluation, & test sets

How many samples per species? Filter out species with just a few samples.

```{r filter_species_N_samples, fig.height=15, fig.width = 10}
# threshold to keep species with at more than this # of samples
thresh = 19

# count number of samples (unique spectra) per species 
taxonCounts <- cleaned_spectra %>% 
    distinct(spectraID, .keep_all = TRUE) %>% 
    #group_by(taxonID) %>% 
    group_by(genusSpecies) %>% 
    tally() %>% 
    # add a new column to indicate whether species count exceeds threshold
    mutate(countGT = case_when(
      n > thresh ~ TRUE,
      n <= thresh ~ FALSE
    ))

# count total number of samples across all species
samplesCountTotal <- taxonCounts %>% 
  summarise(sum(n))

speciesCountTotal <- nrow(taxonCounts)

# count total number of species with # samples greater than threshold
samplesCountGT <- taxonCounts %>% 
  filter(countGT == TRUE) %>% 
  summarise(sum(n))
speciesCountGT <- nrow(taxonCounts %>% 
  filter(countGT == TRUE))

# plot histogram with count per genusSpecies. color bars based on whether
# # samples exceeds the threshold
ggplot(taxonCounts, aes(y = reorder(genusSpecies, n), x = n)) + 
                          #reorder(taxonID, n), x = n)) + 
  geom_bar(stat = "identity", aes(fill = countGT)) + 
  labs(title = paste("Histogram of samples per species.",
                     speciesCountTotal, "total species;",
                     speciesCountGT, "species with greater than",
                     thresh, "samples"),
       y = "Genus Species",
       fill = paste("count greater than threshold", thresh)) + 
  theme_bw() + 
  scale_fill_manual(values = c("#d95f0e", "#7bccc4"))
```

```{r}
# Filter out species with less than N samples 
taxonCounts <- cleaned_spectra %>% 
    distinct(spectraID, .keep_all = TRUE) %>% 
    group_by(genusSpecies) %>% 
    tally() %>% 
  dplyr::filter(n>thresh)

filtered_spectra <- cleaned_spectra %>%
  dplyr::filter(genusSpecies %in% taxonCounts$genusSpecies)
```

Split multiple observations of individuals across sets. Add a column to the data (“train”, “eval”, “test”). The 80/20 ratio (based on the [Pareto Principle](https://en.wikipedia.org/wiki/Pareto_principle)) for training/testing is common, where the 20% can be split into 10/10 for testing/validation. Andrew NG recommends 60/20/20 for training/testing/validation in his online machine learning course.

```{r split_train_test_valid}
set.seed(44)
# efficiently split into train, test, valid groups 
splits <- filtered_spectra %>%
	distinct(spectraID) %>%
	mutate(group = sample(c("train", "valid", "test"), n(), 
	                      replace=TRUE, prob = c(.6, .2, .2)))

# check the percentage of each group
splits %>%
  group_by(group) %>% 
  tally() %>% 
  mutate(percent = n / sum(n))

# join these group labels into the spectra data frame 
filtered_spectra <- filtered_spectra %>%
	left_join(splits, by = "spectraID")
```

```{r}
# are all genusSpecies values present in each group?
filtered_spectra %>% 
  filter(group == "train") %>%
  select(genusSpecies) %>%
  unique() %>% 
  nrow()

filtered_spectra %>% 
  filter(group == "test") %>%
  select(genusSpecies) %>%
  unique() %>% 
  nrow()

filtered_spectra %>% 
  filter(group == "valid") %>%
  select(genusSpecies) %>%
  unique() %>% 
  nrow()
```

### Output "clean dataset"

Write spectra to csv file. 
```{r}
write_csv(x = filtered_spectra,
          path = here::here("data", "cleaned_spectra.csv"))
```

Visualize the spectra, facet by scientific name, color by taxon rank. 
```{r, warning=FALSE, fig.height=12, fig.width=16}
filtered_spectra %>%
  ggplot(aes(wavelength_nm, ifelse(mask, NA, reflectance), group = spectraID,
             # color each facet plot based on taxonRank
             color = taxonRank)) +
  geom_path(alpha = .2) +
  labs(x = '\n Wavelength (nm)',
       y = 'Reflectance \n',
       color = "Taxon Rank") + 
  facet_wrap(~genusSpecies
             # wrap long scientific names
             ,labeller = label_wrap_gen(20)
             ,ncol=10) + 
  scale_color_manual(values = c("#e6550d","#1c9099", "#a8ddb5")) + 
  theme_minimal() + 
  # Adjust the text size and rotation angle 
  theme(text = element_text(size=11),
        axis.text.x = element_text(angle=90, hjust=1),
        axis.title.y = element_text(size = 14),
        axis.title.x = element_text(size = 14),
        legend.text = element_text(size = 14),
        legend.title = element_text(size = 14))
```

### 2020-09-15 R/clean_spectral_data.Rmd
writing figures to files
noticed that the all_spectra file is smaller and has fewer samples than it should... guess I'll run the spectral-lib code again and try to see what's up

@@
                                    "spectral-lib","all_spectra.csv"))
-
-# quick look at the values in each column
-glimpse(all_spectra)
 ```
@@
 
-Based on Max's Gist code: https://gist.github.com/mbjoseph/5c18781e508460e14f64193571b98b7d
-
 Plot all spectra; color the spectra based on genus/species taxonomic level.
 
-```{r, warning=FALSE, fig.height=16, fig.width=14}
+```{r, warning=FALSE, fig.height=16, fig.width=14, eval=FALSE}
 all_spectra %>%
@@
   dplyr::filter(spectraID %in% c(#"NEON.PLA.D20.PUUM.00954_2019",
-                                 "NEON.PLA.D11.CLBJ.00496_2018", 
-                                 "NEON.PLA.D01.HARV.09041_2018",
-                                 "NEON.PLA.D08.DELA.03903_2017")) %>%
+                                 "NEON.PLA.D10.RMNP.04210_2018", 
+                                 "NEON.PLA.D03.DSNY.00878_2018",
+                                 "NEON.PLA.D01.BART.03618_2017")) %>%
   ggplot(aes(x = wavelength_nm, y = reflectance, group = spectraID)) +
@@
             alpha = 0.5, size = 1) +
-  labs(title = "Spectra with different wavelengths: zoom in on beginning of spectra") + 
+  labs(title = "Spectra with different wavelengths: zoom in on beginning of spectra",
+       x = "wavelength [nm]",
+       color = "spectrum ID") + 
   # zoom in on beginning of spectra
@@
   theme_bw()
+
+# save figure to file
+ggsave(filename = here::here("analysis","figures", "wl_before_interp_380nm.pdf"), width = 9, height = 4)
 ```
@@
             alpha = 0.5, size = 1) +
-  labs(title = "Spectra each with different wavelengths: zoom in on end of spectra") + 
+  labs(title = "Spectra each with different wavelengths: zoom in on end of spectra",
+        x = "wavelength [nm]",
+       color = "spectrum ID") + 
   # zoom in on end of spectra
@@
   theme_bw()
+
+# save figure to file
+ggsave(filename = here::here("analysis","figures", "wl_before_interp_2500nm.pdf"), width = 9, height = 4)
 ```
@@
   dplyr::filter(spectraID %in% c(#"NEON.PLA.D20.PUUM.00954_2019",
-                                 "NEON.PLA.D11.CLBJ.00496_2018", 
-                                 "NEON.PLA.D01.HARV.09041_2018",
-                                 "NEON.PLA.D08.DELA.03903_2017")) %>% 
+                                 "NEON.PLA.D10.RMNP.04210_2018", 
+                                 "NEON.PLA.D03.DSNY.00878_2018",
+                                 "NEON.PLA.D01.BART.03618_2017")) %>% 
   ggplot(aes(x = wavelength_nm, y = reflectance, group = spectraID)) +
@@
             alpha = 0.5, size = 1) +
-  labs(title = "Spectra w interpolated wavelengths: Zoom in on beginning of spectra") + 
+  labs(title = "Spectra w interpolated wavelengths: Zoom in on beginning of spectra",
+       x = "wavelength [nm]",
+       color = "spectrum ID") + 
   # zoom in on beginning of spectra
@@
   theme_bw()
+
+# save figure to file
+ggsave(filename = here::here("analysis","figures", "wl_after_interp_380nm.pdf"), width = 9, height = 4)
 ```
@@
                      thresh, "samples"),
-       y = "Genus Species",
-       fill = paste("count greater than threshold", thresh)) + 
+       y = "Scientific Name (Genus + Species)",
+       fill = paste("Count >", thresh)) + 
   theme_bw() + 
   scale_fill_manual(values = c("#d95f0e", "#7bccc4"))
+
+# save figure to file
+ggsave(filename = here::here("analysis","figures", "hist_samples_per_species.pdf"), width = 10, height = 12)
 ```

### 2020-09-15 R/spectral-lib.Rmd
added timer code
to see how long the spec lib code takes

@@
 ```{r get-veg-data}
+# start timer 
+start_time <- Sys.time()
+
 out_dir <- here::here("analysis","data","raw_data","spectral-lib")
@@
 ```
+
+```{r how_long_did_this_take}
+# stop timer
+end_time <- Sys.time()
+
+elapsed_time <- end_time - start_time
+
+print(difftime(end_time, start_time, 
+         units = c("auto", "secs", "mins", "hours",
+                   "days", "weeks")))
+```

### 2020-09-15 R/clean_spectral_data.Rmd
running through the clean data script again
looks like there is more data than before - I wonder if that's because there is newer data from this past season

@@
                                    "spectral-lib","all_spectra.csv"))
+
+# quick look at the values in each column
+glimpse(all_spectra)
 ```

### 2020-09-16 R/clean_spectral_data.Rmd
write histogram figure to file
noticed that i had to change the spectraID of spectra to plot during the wavelength plots... the spectral data appears to be different each time I run the spectral-lib code

@@
 all_spectra %>% 
-  dplyr::filter(spectraID %in% c(#"NEON.PLA.D20.PUUM.00954_2019",
-                                 "NEON.PLA.D11.CLBJ.00496_2018", 
-                                 "NEON.PLA.D01.HARV.09041_2018",
-                                 "NEON.PLA.D08.DELA.03903_2017")) %>%
+  dplyr::filter(spectraID %in% c("NEON.PLA.D07.MLBS.00412_2018", 
+                                 "NEON.PLA.D01.HARV.04350_2018",
+                                 "NEON.PLA.D08.LENO.08750_2015",
+                                 "NEON.PLA.D01.BART.06298_2019")) %>%
   ggplot(aes(x = wavelength_nm, y = reflectance, group = spectraID)) +
@@
 all_spectra %>% 
-  dplyr::filter(spectraID %in% c(#"NEON.PLA.D20.PUUM.00954_2019",
-                                 "NEON.PLA.D10.RMNP.04210_2018", 
-                                 "NEON.PLA.D03.DSNY.00878_2018",
-                                 "NEON.PLA.D01.BART.03618_2017")) %>%
+  dplyr::filter(spectraID %in% c("NEON.PLA.D07.MLBS.00412_2018", 
+                                 "NEON.PLA.D01.HARV.04350_2018",
+                                 "NEON.PLA.D08.LENO.08750_2015",
+                                 "NEON.PLA.D01.BART.06298_2019")) %>%
   ggplot(aes(x = wavelength_nm, y = reflectance, group = spectraID)) +
@@
 interp_wl %>% 
-  dplyr::filter(spectraID %in% c(#"NEON.PLA.D20.PUUM.00954_2019",
-                                 "NEON.PLA.D10.RMNP.04210_2018", 
-                                 "NEON.PLA.D03.DSNY.00878_2018",
-                                 "NEON.PLA.D01.BART.03618_2017")) %>% 
+  dplyr::filter(spectraID %in% c("NEON.PLA.D07.MLBS.00412_2018", 
+                                 "NEON.PLA.D01.HARV.04350_2018",
+                                 "NEON.PLA.D08.LENO.08750_2015",
+                                 "NEON.PLA.D01.BART.06298_2019")) %>% 
   ggplot(aes(x = wavelength_nm, y = reflectance, group = spectraID)) +
@@
 # save figure to file
-ggsave(filename = here::here("analysis","figures", "hist_samples_per_species.pdf"), width = 10, height = 12)
+ggsave(filename = here::here("analysis","figures", "hist_samples_per_species.pdf"), width = 9, height = 12)
 ```

### 2020-10-28 R/create_figures.Rmd
adding the figure script to the repo

---
title: "Create additional figures"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(here)
library(readr)
library(dplyr)
library(sf)
library(ggplot2)
library(ggthemes)
library(spData)
library(tidyverse)
library(mapview)
library(RColorBrewer)
library(ggspatial)
library(rnaturalearth)
library(rnaturalearthdata)
```

## Study area

```{r prepare_data, message=FALSE, warning=FALSE}
# read NEON domain polygons
neon_domains <- sf::st_read("data/shp/NEON_Domains.shp", quiet = TRUE)

# read spectral data
all_spectra <- readr::read_csv(here::here("data","cleaned_spectra.csv"))
spectra_locations <- all_spectra %>%  
  # grab the latitude and longitude coordinates for each spectrum 
  select(latitude = adjDecimalLatitude, longitude = adjDecimalLongitude) %>% 
  # remove redundant rows
  distinct() %>%
  # convert each to a SF point with geographic coordinates 
  st_as_sf(coords = c("longitude", "latitude"), crs = 4326)

# filter the NEON domains, keep only those containing spectra locations
domains_containing_spectra <- neon_domains %>%
  filter(lengths(st_intersects(neon_domains, spectra_locations)) > 0)

# transform domains and spectra points to Albers Equal Area, CRS = 2163 for the map 
domains_filtered_2163 <- sf::st_transform(x = domains_containing_spectra, crs = 2163)
domains_all_2163 <- sf::st_transform(x = neon_domains, crs = 2163)
spectra_2163 <- sf::st_transform(x = spectra_locations, crs = 2163)

# load and transform additional geospatial data for mapping 
data("us_states", package = "spData")
us_states_2163 <- st_transform(spData::us_states, crs = 2163)
#world_2163 <- st_transform(spData::world, crs = 2163)
world_2163 <- ne_countries(scale = "medium", returnclass = "sf") %>% st_transform(crs = 2163)

# set color palette for NEON domains 
colourCount = length(unique(domains_filtered_2163$DomainID))
getPalette = colorRampPalette(brewer.pal(colourCount, "Spectral"))
colors_domains <- data.frame(names = sort(unique(domains_filtered_2163$DomainName)),
                        values = getPalette(colourCount)) #colorRampPalette
```

```{r make_map, fig.height=4, fig.width=6}
# make the map 
ggplot() + 
  # US States
  geom_sf(data = us_states_2163, fill="grey80", color="gray70", alpha=0.3) + 
  # continent boundaries for context 
  geom_sf(data = world_2163, fill="grey80", color="gray50", alpha=0.5) +
  # NEON domains containing spectra
  geom_sf(data = domains_filtered_2163, aes(fill = factor(DomainName)), 
          alpha = 0.7, color = "gray40") + 
  # fill label in legend 
  labs(fill = "NEON Domain") + 
  # apply fill color scale for neon domains 
  scale_fill_manual(breaks = colors_domains$names, values = colors_domains$values) + 
  # add the other NEON domains, just as outlines + 
  geom_sf(data = domains_all_2163, alpha = 0, color = "gray40") +
  # add NEON spectra locations as points
  geom_sf(data = spectra_2163) + 
  # set the extent based on NEON domains where we have spectral data 
  coord_sf(crs = st_crs(2163),
           expand = FALSE,
           xlim = c(as.numeric(st_bbox(domains_filtered_2163)[1]),    # xmin
                    as.numeric(st_bbox(domains_filtered_2163)[3])),   # xmax 
           ylim = c(as.numeric(st_bbox(domains_filtered_2163)[2]),    # ymin
                    as.numeric(st_bbox(domains_filtered_2163)[4]))) + # ymax
  # add scale bar and north arrow 
  annotation_scale(location = "bl", width_hint = 0.4, style = "bar") +
  # add north arrow
  annotation_north_arrow(location = "bl", which_north = "true", 
                         pad_x = unit(0.15, "in"), pad_y = unit(0.25, "in"),
                         style = north_arrow_fancy_orienteering) +
  # set theme 
  theme_bw() 
```

```{r make_map_with_alaska_inset, fig.height=4, fig.width=6}
# make the map WITH ALASKA INSET
main_map <- ggplot() + 
  # US States
  geom_sf(data = us_states_2163, fill="grey80", color="gray70", alpha=0.3) + 
  # continent boundaries for context 
  geom_sf(data = world_2163, fill="grey80", color="gray50", alpha=0.5) +
  # NEON domains containing spectra
  geom_sf(data = domains_filtered_2163, aes(fill = factor(DomainName)), 
          alpha = 0.7, color = "gray40") + 
  # fill label in legend 
  labs(fill = "NEON Domain") + 
  # apply fill color scale for neon domains 
  scale_fill_manual(breaks = colors_domains$names, values = colors_domains$values) + 
  # add the other NEON domains, just as outlines + 
  geom_sf(data = domains_all_2163, alpha = 0, color = "gray40") +
  # add NEON spectra locations as points
  geom_sf(data = spectra_2163) + 
  # set the extent based on NEON domains where we have spectral data 
    coord_sf(crs = st_crs(2163), 
           xlim = c(-2500000, 2500000), 
           ylim = c(-2300000, 730000))  + 
  # get rid of legend
  theme(legend.position = "none")

alaska_inset <- ggplot() + 
  # add Alaska boundaries
  geom_sf(data = us_states_2163, fill="grey", color="gray70", alpha=0.3) + 
  # continent boundaries for context 
  geom_sf(data = world_2163, fill="grey", color="gray70", alpha=0.3) + 
  # Add NEON domains 
  geom_sf(data = domains_filtered_2163, aes(fill = factor(DomainName)), 
          alpha = 0.7, color = "gray40") + 
  # add NEON sites as points 
  geom_sf(data = spectra_2163) + 
  # set extents to zoom in just on Alaska
  coord_sf(crs = st_crs(3467), 
           xlim = c(-2400000, 1600000), 
           ylim = c(200000, 2500000), 
           expand = FALSE, datum = NA) + 
  theme_clean() +
  # get rid of legend for now
  theme(legend.position = "none") + 
  # fill for neon domains in the legend 
  scale_fill_manual(breaks = colors_domains$names, values = colors_domains$values) + 
  # add a border around the alaska inset 
  theme(panel.border = element_rect(colour = "black", fill = NA))

# combine main map with alaska as an inset
main_map +
  annotation_custom(grob = ggplotGrob(alaska_inset),
                    xmin = -2750000,
                    xmax = -2700000 + (1600000 - (-2400000))/2.5,
                    ymin = -2470000,
                    ymax = -2450000 + (2500000 - 20000)/2.5) + 
  # set theme to make background white, subtle grid lines, outline map
  theme_bw()  + 
  # apply color scale for domains across the contiguous US and Alaska
  scale_fill_manual(breaks = colors_domains$names, values = colors_domains$values) + 
  # titles and labels
  labs(fill = "NEON Domain") + 
  # add scale bar
  annotation_scale(location = "br", width_hint = 0.5, style = "bar") +
  # add north arrow
  annotation_north_arrow(location = "tl", which_north = "true", 
                         pad_x = unit(0.15, "in"), pad_y = unit(0.5, "in"),
                         style = north_arrow_fancy_orienteering) 
```

```{r}
ggsave(filename = here::here("output", "study_area.pdf"),
       width = 12, height = 8)
```

### 2020-10-29 R/create_figures.Rmd
trying to get paper to knit
but it depends on a lot of things, not working yet

@@
 ```{r prepare_data, message=FALSE, warning=FALSE}
+
+# download and unzip NEON domain data
+url <- "https://www.neonscience.org/sites/default/files/NEONDomains_0.zip" 
+dest_dir <- here::here("analysis", "data", "shp_data")
+if(!dir.exists(dest_dir)){
+  dir.create(dest_dir)
+}
+dest_file <- here::here("analysis", "data", "shp_data", "NEONDomains.zip")
+download.file(url, dest_file)
+unzip(dest_file, exdir = dest_dir)
+
 # read NEON domain polygons
-neon_domains <- sf::st_read("data/shp/NEON_Domains.shp", quiet = TRUE)
+neon_domains <- sf::st_read(here::here("analysis", "data", "shp_data", "NEON_Domains.shp"), 
+                            quiet = TRUE)
 
 # read spectral data
-all_spectra <- readr::read_csv(here::here("data","cleaned_spectra.csv"))
-spectra_locations <- all_spectra %>%  
+cleaned_spectra <- readr::read_csv(here::here("analysis", "data", "derived_data", 
+                                              "cleaned_spectra.csv"))
+spectra_locations <- cleaned_spectra %>%  
   # grab the latitude and longitude coordinates for each spectrum 
@@
 ```{r}
-ggsave(filename = here::here("output", "study_area.pdf"),
+message("saving study area figure...")
+ggsave(filename = here::here("analysis", "figures", "study_area.pdf"),
        width = 12, height = 8)

### 2020-10-29 analysis/paper/paper.Rmd
trying to get paper to knit
but it depends on a lot of things, not working yet

@@
 Network of sites and observations (how many? ecoclimatic domains? ecosystem coverage?
+Figure \@ref(fig:fig_studyArea)
 
@@
 Extracted hyperspectral data for all mapped stems at NEON sites
+Figure \@ref(fig:fig_spectraExtracted)
 Assumed plant locations don't change over time 
@@
 
-```{r spectra, fig.cap = "", echo=FALSE,  out.width = "450px", message = FALSE, warning=FALSE}
+```{r fig_studyArea, fig.cap = "Map of the NEON sampling locations and domains with data that we used in this study.", echo=FALSE,  out.width = "450px", message = FALSE, warning=FALSE}
 library(here)
-knitr::include_graphics(here::here("analysis", "figures", "spectra_figure_test.png"))
+knitr::include_graphics(here::here("analysis", "figures", "study_area.pdf"))
 ```
@@
 
-```{r excel, fig.cap = "", echo=FALSE,  out.width = "450px", message = FALSE, warning=FALSE}
+```{r fig_honestWork, fig.cap = "", echo=FALSE,  out.width = "450px", message = FALSE, warning=FALSE}
 knitr::include_graphics(here::here("analysis", "figures", "honest_work.jpeg"))
@@
 \clearpage
+
+## Figure 3 {-}
+
+```{r fig_spectraExtracted, fig.cap = "", echo=FALSE,  out.width = "450px", message = FALSE, warning=FALSE}
+library(here)
+knitr::include_graphics(here::here("analysis", "figures", "spectra_figure_test.png"))
+```
+
+\clearpage

### 2020-10-29 analysis/paper/paper.Rmd
paper is knitting now

--- 2020-09-10 analysis/paper/paper.Rmd
@@
 Network of sites and observations (how many? ecoclimatic domains? ecosystem coverage?
+Figure \@ref(fig:fig-studyArea)
 
@@
 Extracted hyperspectral data for all mapped stems at NEON sites
+Figure \@ref(fig:fig-spectraExtracted)
 Assumed plant locations don't change over time 
@@
 
+Figure \@ref(fig:fig-honestWork)
+
 <!-- Here's some example analysis code: -->
@@
 ```
-
-```{r demo-plot, fig.cap="A plot of random numbers"}
-plot(rnorm(10))
-```
-
-Figure \@ref(fig:demo-plot) shows how we can have a caption and cross-reference for a plot
 
@@
 
-```{r spectra, fig.cap = "", echo=FALSE,  out.width = "450px", message = FALSE, warning=FALSE}
+```{r fig-studyArea, fig.cap ="Caption", echo=FALSE,  out.width = "450px", message = FALSE, warning=FALSE}
 library(here)
-knitr::include_graphics(here::here("analysis", "figures", "spectra_figure_test.png"))
+knitr::include_graphics(here::here("analysis", "figures", "study_area.pdf"))
 ```
@@
 
-```{r excel, fig.cap = "", echo=FALSE,  out.width = "450px", message = FALSE, warning=FALSE}
+```{r fig-honestWork, fig.cap = "Caption", echo=FALSE,  out.width = "450px", message = FALSE, warning=FALSE}
 knitr::include_graphics(here::here("analysis", "figures", "honest_work.jpeg"))
@@
 \clearpage
+
+## Figure 3 {-}
+
+```{r fig-spectraExtracted, fig.cap = "Caption", echo=FALSE,  out.width = "450px", message = FALSE, warning=FALSE}
+library(here)
+knitr::include_graphics(here::here("analysis", "figures", "spectra_figure_test.png"))
+```
+
+\clearpage

### 2020-10-29 analysis/paper/paper.Rmd
added citations for R and packages

@@
 #R/spectral-lib.Rmd
-
 # Run the main analysis
-
 ```
@@
 
-1 - motivation for work, why bother?
-  - unprecedented access to unique hyperspectral airborne imagery from NEON (cite)
+1
+* motivation for work, why bother?
+* unprecedented access to unique hyperspectral airborne imagery from NEON (cite)
   - acknowledge additional data to add info (lidar, rgb) (cite) 
@@
 The National Ecological Observatory Network (NEON) is a valuable source of publicly available ecological data across the United States [@keller2008continental].
-
 We have extracted all NEON AOP hyperspectral data for every mapped stem in NEON plots where field data and airborne remote sensing data were collected during the same year.
-
-Here's the code to extract the spectra: [https://gist.github.com/mbjoseph/5c18781e508460e14f64193571b98b7d](https://gist.github.com/mbjoseph/5c18781e508460e14f64193571b98b7d)
 
@@
 Mask created for water absorption bands (cite)
-Packages neonUtilities, geoNEON, neonhs (cite)
+We performed our analysis in R [@R] with packages including neonUtilities [@neonUtilities], geoNEON [@geoNEON], neonhs [@neonhs]
+
+The code to do this is provided in the research compendium here:[https://github.com/vscholl/neonHScompendium](https://github.com/vscholl/neonHScompendium).
 
@@
 
-```{r fig-studyArea, fig.cap ="Caption", echo=FALSE,  out.width = "450px", message = FALSE, warning=FALSE}
+```{r fig-studyArea, fig.cap ="Study Area", echo=FALSE,  out.width = "450px", message = FALSE, warning=FALSE}
 library(here)

### 2020-10-30 analysis/paper/paper.Rmd
working on paper text

@@
 
-Extracted hyperspectral data for all mapped stems at NEON sites
+We extracted hyperspectral data for all mapped stems at NEON sites using the neonhs [@neonhs] package. 
+We assumed that plant locations did not change over time.
+
 Figure \@ref(fig:fig-spectraExtracted)
-Assumed plant locations don't change over time 
+
 Filtered plants to keep only "live" ones

### 2020-10-30 analysis/paper/paper.Rmd
methods - NEON data products and AOP

@@
 # Methods {-}
+The NEON data that we used for our study include both *in-situ* plant measurements and airborne remote sensing image data [@neonDataProducts].
+The Woody Plant Vegetation Structure (data product code NEON.DP1.10098) includes individual plant locations along with species and genus observations that are collected every 1 to 3 years. 
+The NEON Airborne Observation Platform (AOP) collects airborne remote sensing data regularly at NEON sites.
+There are three AOP payloads, each carrying three types of sensors: (1) a high-resolution digital camera to capture red, green, blue (RGB) true-color images, (2) a light detection and ranging (lidar) system to capture discrete and waveform lidar data with approximately 1-4 points/waveforms per square meter, and (3) a pushbroom-style imaging spectrometer (AVIRIS next-gen) to capture hyperspectral data.
+The hyperspectral data contains 426 narrow bands spanning the visible to shortwave infrared wavelengths, from 380 - 2500nm, and the image data products derived have a spatial resolution of 1m
+The AOP flies at an altitude of 1000m above ground level and the flight season typically runs from May to October.
+
+NEON sampling design
 NEON data collection protocols 
@@
 
-We extracted hyperspectral data for all mapped stems at NEON sites using the neonhs [@neonhs] package. 
+We extracted hyperspectral data for all mapped stems at NEON sites.
+
 We assumed that plant locations did not change over time.

### 2020-10-30 analysis/paper/paper.Rmd
methods text
some of the references are formatting weirdly

@@
 The NEON data that we used for our study include both *in-situ* plant measurements and airborne remote sensing image data [@neonDataProducts].
-The Woody Plant Vegetation Structure (data product code NEON.DP1.10098) includes individual plant locations along with species and genus observations that are collected every 1 to 3 years. 
+The Woody Plant Vegetation Structure (data product code DP1.10098) includes individual plant locations along with species and genus observations that are collected every 1 to 3 years. 
 The NEON Airborne Observation Platform (AOP) collects airborne remote sensing data regularly at NEON sites. 
 There are three AOP payloads, each carrying three types of sensors: (1) a high-resolution digital camera to capture red, green, blue (RGB) true-color images, (2) a light detection and ranging (lidar) system to capture discrete and waveform lidar data with approximately 1-4 points/waveforms per square meter, and (3) a pushbroom-style imaging spectrometer (AVIRIS next-gen) to capture hyperspectral data.
-The hyperspectral data contains 426 narrow bands spanning the visible to shortwave infrared wavelengths, from 380 - 2500nm, and the image data products derived have a spatial resolution of 1m
+The hyperspectral data contains 426 narrow bands spanning the visible to shortwave infrared wavelengths, from 380 - 2500nm, and the image data products derived have a spatial resolution of 1m. 
+We worked with the Spectrometer orthorectified surface directional reflectance mosaic (data product code DP3.30006). 
 The AOP flies at an altitude of 1000m above ground level and the flight season typically runs from May to October.

### 2020-10-30 R/create_figures.Rmd
workflow figure?
trying to make one using DiagrammeR but can't figure out how to export it as pdf

@@
 ```
+
+## Workflow figure
+
+```{r}
+library(DiagrammeR)
+library(rsvg)
+
+workflow <- DiagrammeR::grViz("digraph {
+
+graph [layout = dot, rankdir = LR]
+
+# define the global styles of the nodes. We can override these in box if we wish
+node [shape = rectangle, style = filled, fillcolor = Linen]
+
+data1 [label = 'In-situ plant \nmeasurements', shape = folder, fillcolor = LightSteelBlue3]
+data2 [label = 'Hyperspectral \nreflectance', shape = folder, fillcolor = LightSteelBlue3]
+process1 [label =  'Extract \n Reflectance', fillcolor = Gainsboro]
+process2 [label = 'Data cleaning & \nPreparation', fillcolor = Gainsboro]
+classify1 [label = 'Classify Species', fillcolor = AntiqueWhite3]
+classify2[label = 'Classify Genus', fillcolor = AntiqueWhite3]
+results [label= 'Results', shape = oval, fillcolor = Honeydew3]
+
+# edge definitions with the node IDs
+{data1 data2}  -> process1 -> process2 -> {classify1 classify2} -> results
+}")
+
+workflow
+
+export_graph(workflow,
+  file_name = here::here("analysis", "figures", "workflow.pdf"),
+  file_type = "pdf")
+
+DiagrammeR::export_graph(workflow) %>%
+  charToRaw() %>%
+  rsvg::rsvg_pdf(here::here("analysis", "figures", "workflow.pdf"))
+
+#grViz(workflow) %>%
+#    export_svg %>% charToRaw %>% rsvg_pdf(here::here("analysis", "figures", "workflow.pdf"))
+```

### 2020-10-30 analysis/paper/paper.Rmd
methods text

@@
 # Methods {-}
+
+## NEON data {-}
+
+The NEON project monitors ecosystems at 20 domains, eaach with similar climate and ecological characteristics, across the United States.
 The NEON data that we used for our study include both *in-situ* plant measurements and airborne remote sensing image data [@neonDataProducts].
@@
 The hyperspectral data contains 426 narrow bands spanning the visible to shortwave infrared wavelengths, from 380 - 2500nm, and the image data products derived have a spatial resolution of 1m.
+The AOP flies at an altitude of 1000m above ground level and the flight season typically runs from May to October.
 We worked with the Spectrometer orthorectified surface directional reflectance mosaic (data product code DP3.30006).
-The AOP flies at an altitude of 1000m above ground level and the flight season typically runs from May to October.
 
-NEON sampling design
-NEON data collection protocols 
-AOP flights
-Plant mapping 
-Network of sites and observations (how many? ecoclimatic domains? ecosystem coverage?
-Figure \@ref(fig:fig-studyArea)
-
-Figure - NEON sites and domains, which sites do we have data from? # individuals from each site? 
-Acknowledge imbalance spatially and temporally...
-Any sites not represented? Like Hawaii
-
-## Data extraction {-}
-
-We extracted hyperspectral data for all mapped stems at NEON sites.
-
-We assumed that plant locations did not change over time.
-
-Figure \@ref(fig:fig-spectraExtracted)
+We performed our analysis in R [@R] with packages including neonUtilities [@neonUtilities], geoNEON [@geoNEON], neonhs [@neonhs].
+In June 2020, we extracted hyperspectral data for all mapped stems at NEON sites that had an AOP flyover during the same year as when *in-situ* measurements were collected. 
+This included 5972 observations of individual plants. Some individuals had multiple observations across years.
 
@@
 Mask created for water absorption bands (cite)
-We performed our analysis in R [@R] with packages including neonUtilities [@neonUtilities], geoNEON [@geoNEON], neonhs [@neonhs]
 
@@
 
-## Data cleaning {-}
+## Data cleaning and preparation {-}
 
-? Filtered taxonomic levels, labels species & genus
+**? Filtered taxonomic levels, labels species & genus**
 Wavelength interpolation - filtered out the PUUM site in Hawaii because of wavelengths
@@
 
-Figure - histogram of species counts, show which ones were used to train
-Figure - cleaned data, make it legible
+**Figure - histogram of species counts, show which ones were used to train**
+**Figure - cleaned data, make it legible**
+
+After the data cleaning steps, we were left with reflectance data for *____* individuals at sites across 14 of the NEON eco-climatic domains (Figure \@ref(fig:fig-studyArea)). 
+There is a noticeable imbalance of data across space and time; some NEON domains and US states (such as Hawaii) are not represented in our analysis. 
+Years that we found when both *in-situ* and AOP data were collected were 2017, 2018, and 2019. 
+Note that covid-19 postponed NEON data collection as well as the AOP flight season during 2020. 
+We assumed that plant locations did not change over time.
+
+Figure \@ref(fig:fig-spectraExtracted)
 

### 2020-10-30 analysis/paper/paper.Rmd
methods text

@@
 
-Filtered plants to keep only "live" ones
-NEON data products (cite)
-Spectrometer orthorectified surface directional reflectance mosaic
-Woody plant vegetation structure field based measurements
-Mask created for water absorption bands (cite)
+**Filtered plants to keep only "live" ones**
+**Mask created for water absorption bands (cite)**
 
-The code to do this is provided in the research compendium here:[https://github.com/vscholl/neonHScompendium](https://github.com/vscholl/neonHScompendium).
+We provide all of our code in a research compendium here:[https://github.com/vscholl/neonHScompendium](https://github.com/vscholl/neonHScompendium).
 

### 2020-10-31 analysis/paper/paper.Rmd
figure 1 caption
added methods text

@@
 The AOP flies at an altitude of 1000m above ground level and the flight season typically runs from May to October.
+
 We worked with the Spectrometer orthorectified surface directional reflectance mosaic (data product code DP3.30006).
 
+## Data cleaning and preparation {-}
+
 We performed our analysis in R [@R] with packages including neonUtilities [@neonUtilities], geoNEON [@geoNEON], neonhs [@neonhs].
-In June 2020, we extracted hyperspectral data for all mapped stems at NEON sites that had an AOP flyover during the same year as when *in-situ* measurements were collected. 
-This included 5972 observations of individual plants. Some individuals had multiple observations across years.
+In June 2020, we extracted hyperspectral reflectance data for all mapped stems at NEON sites that had an AOP flyover during the same year as when *in-situ* measurements were collected. 
+This included spectral reflectance spectra for 5972 observations of individual plants. 
+Some individuals had multiple observations across years. 
+The individual plants were identified taxon ranks of kingdom, family, genus, species, and variety. 
+Most of them were identified at the species rank.
+We filtered the plants by taxonomic rank to retain only those identified at the genus, species, and variety ranks. 
+Varities were removed to focus on genus and species identifications. 
+There were 74 unique genus values, 118 unique species values, and 151 unique genus-species combinations.
 
-**Filtered plants to keep only "live" ones**
-**Mask created for water absorption bands (cite)**
+The hyperspectral reflectance spectra extracted from the locations of the individual plants
 
-We provide all of our code in a research compendium here:[https://github.com/vscholl/neonHScompendium](https://github.com/vscholl/neonHScompendium).
+```{r, message = FALSE}
+library(knitr)
 
-## Data cleaning and preparation {-}
+df <- data.frame(wavelength = c(347, 381, 382, 384),
+                 number = c(94, 1220, 1733, 2922))
+
+kable(df)
+```
 
@@
 We assumed that plant locations did not change over time.
+
+**Filtered plants to keep only "live" ones**
+**Mask created for water absorption bands (cite)**
+
+We provide all of our code in a research compendium here:[https://github.com/vscholl/neonHScompendium](https://github.com/vscholl/neonHScompendium).
 
@@
 Figure - variable importance vs wavelength
+Figure - multi-year observations of individual plants?
 
@@
 
-Figure 1 description...
+Map to illustrate the NEON sites and domains with data that we used in this study.
 

### 2020-10-31 R/create_figures.Rmd
dont try to make the workflow diagram
knits on laptop now

@@
 library(tidyverse)
-library(mapview)
 library(RColorBrewer)
 library(ggspatial)
+library(rgeos)
 library(rnaturalearth)
@@
 data("us_states", package = "spData")
-us_states_2163 <- st_transform(spData::us_states, crs = 2163)
+us_states_2163 <- sf::st_transform(spData::us_states, crs = 2163)
 #world_2163 <- st_transform(spData::world, crs = 2163)
-world_2163 <- ne_countries(scale = "medium", returnclass = "sf") %>% st_transform(crs = 2163)
+world_2163 <- rnaturalearth::ne_countries(scale = "medium", returnclass = "sf") %>% st_transform(crs = 2163)
 
 # set color palette for NEON domains 
-colourCount = length(unique(domains_filtered_2163$DomainID))
-getPalette = colorRampPalette(brewer.pal(colourCount, "Spectral"))
+colourCount <- length(unique(domains_filtered_2163$DomainID))
+
+# colorbrewer palette can only be up to 11 colors though
+getPalette <- colorRampPalette(brewer.pal(colourCount, "Spectral"))
 colors_domains <- data.frame(names = sort(unique(domains_filtered_2163$DomainName)),
                         values = getPalette(colourCount)) #colorRampPalette
+
+# trying to use a non-color brewer palette to get 14
+#getPalette <- c(brewer.pal(name="Spectral", n = 8), brewer.pal(name="Spectral", n = colourCount-8))
+#colors_domains <- data.frame(names = sort(unique(domains_filtered_2163$DomainName)),
+#                        values = getPalette) 
 ```
@@
   # add scale bar and north arrow 
-  annotation_scale(location = "bl", width_hint = 0.4, style = "bar") +
+  ggspatial::annotation_scale(location = "bl", width_hint = 0.4, style = "bar") +
   # add north arrow
-  annotation_north_arrow(location = "bl", which_north = "true", 
+  ggspatial::annotation_north_arrow(location = "bl", which_north = "true", 
                          pad_x = unit(0.15, "in"), pad_y = unit(0.25, "in"),
@@
 
-```{r}
+```{r include=FALSE, eval=FALSE}
 library(DiagrammeR)

### 2020-10-31 analysis/paper/paper.Rmd
knits on laptop now

@@
 ```{r setup, echo = FALSE}
+library(bookdown)
 knitr::opts_chunk$set(
@@
 
-Figure \@ref(fig:fig-spectraExtracted)
-
 ## Species classification {-}
@@
 Figure - multi-year observations of individual plants?
-
-Figure \@ref(fig:fig-honestWork)
 
@@
 
-```{r fig-honestWork, fig.cap = "Caption", echo=FALSE,  out.width = "450px", message = FALSE, warning=FALSE}
-knitr::include_graphics(here::here("analysis", "figures", "honest_work.jpeg"))
-```
-
 \clearpage
@@
 
-```{r fig-spectraExtracted, fig.cap = "Caption", echo=FALSE,  out.width = "450px", message = FALSE, warning=FALSE}
-library(here)
-knitr::include_graphics(here::here("analysis", "figures", "spectra_figure_test.png"))
-```
-
 \clearpage

### 2020-10-31 R/clean_spectral_data.Rmd
trying to get paper to knit
but it depends on a lot of things, not working yet
working on methods text

@@
 title: "Clean spectral data"
-author: "Victoria Scholl"
-date: "09/10/2020"
 output: html_document
@@
 write_csv(x = filtered_spectra,
-          path = here::here("data", "cleaned_spectra.csv"))
+          path = here::here("analysis", "data", "derived_data", "cleaned_spectra.csv"))
 ```

### 2020-11-01 analysis/paper/paper.Rmd
working on methods text
methods text

@@
 
-The NEON project monitors ecosystems at 20 domains, eaach with similar climate and ecological characteristics, across the United States.
+The NEON project monitors diverse ecosystems at 81 sites across 20 domains, each with similar climate and ecological characteristics, across the United States.
 The NEON data that we used for our study include both *in-situ* plant measurements and airborne remote sensing image data [@neonDataProducts].
@@
 The AOP flies at an altitude of 1000m above ground level and the flight season typically runs from May to October.
-
 We worked with the Spectrometer orthorectified surface directional reflectance mosaic (data product code DP3.30006).
@@
 We performed our analysis in R [@R] with packages including neonUtilities [@neonUtilities], geoNEON [@geoNEON], neonhs [@neonhs].
-In June 2020, we extracted hyperspectral reflectance data for all mapped stems at NEON sites that had an AOP flyover during the same year as when *in-situ* measurements were collected. 
+In June 2020, we extracted hyperspectral reflectance data for all individual mapped stems at NEON sites that had an AOP flyover during the same year as when *in-situ* measurements were collected. 
+We only kept individuals with a "live" plant status (as opposed to other plant status classifications such as "dead", "standing dead", "downed", "insect/disease damaged", and "no longer qualifies").
 This included spectral reflectance spectra for 5972 observations of individual plants. 
@@
 We filtered the plants by taxonomic rank to retain only those identified at the genus, species, and variety ranks. 
-Varities were removed to focus on genus and species identifications. 
+We removed varieties from scientific names to label each individual spectrum with its respective genus and species (when available). 
 There were 74 unique genus values, 118 unique species values, and 151 unique genus-species combinations.
+The hyperspectral reflectance spectra extracted from the locations of the individual plants have different starting wavelengths:
 
-The hyperspectral reflectance spectra extracted from the locations of the individual plants
-
-```{r, message = FALSE}
+```{r, message = FALSE, echo = FALSE}
 library(knitr)
@@
 
-kable(df)
+knitr::kable(x = df, format = "simple")
 ```
 
-**? Filtered taxonomic levels, labels species & genus**
-Wavelength interpolation - filtered out the PUUM site in Hawaii because of wavelengths
-Filtering based on # samples
-Train/validate split 80/20
+Each series of wavelengths increments by 5nm from each starting wavelength to yield 426 spectral bands. 
+Most of them started at 384nm, so we linearly interpolated each spectrum to have the same 426 bands starting at 384nm and ending at 2512nm with 5nm increments.
+We filtered out any spectra with a starting wavelength of 347nm, which were all collected at the Pu'u Maka'ala Natural Area Reserve (PUUM) NEON site, Pacific Tropical Domain 20, Hawaii, since the interpolation would require too much unavailable data outside the available range.
+We flagged "bad bands" from 1340-1445nm, 1790-1995nm, and greater than 2400nms o they could be excluded from later analysis.
+These wavelength ranges feature noisy reflectance data due to high water absoprtion.
 
-**Figure - histogram of species counts, show which ones were used to train**
-**Figure - cleaned data, make it legible**
+Next, we removed samples belonging to species classes with few samples. 
+We figured that a small sample size would lead to a poor representation in model training. 
+From the initial collection of 146 species, 50 species had at least 20 samples. 
+We randomly assigned each spectrum to the training or validation sets, using an 80/20 split. All 50 species (and therefore, all genus values) were represented in each set.
 
-After the data cleaning steps, we were left with reflectance data for *____* individuals at sites across 14 of the NEON eco-climatic domains (Figure \@ref(fig:fig-studyArea)). 
+After these data cleaning steps, we were left with reflectance data for 5266 individuals at sites across 14 of the NEON eco-climatic domains (Figure \@ref(fig:fig-studyArea)). 
 There is a noticeable imbalance of data across space and time; some NEON domains and US states (such as Hawaii) are not represented in our analysis. 
@@
 
-**Filtered plants to keep only "live" ones**
-**Mask created for water absorption bands (cite)**
-
-We provide all of our code in a research compendium here:[https://github.com/vscholl/neonHScompendium](https://github.com/vscholl/neonHScompendium).
+**Figure - histogram of species counts? maybe in appendix, make a table or something to show species info for which ones were used after cleaning/filtering**
+**Figure - cleaned data, make it legible, how to color if at all (the species/genus/variety color doesn't make sense anymore at this stage)...**
 
@@
 Same setup as species classification, but used the genus as each class label.
+
+We provide all of our code in a research compendium here:[https://github.com/vscholl/neonHScompendium](https://github.com/vscholl/neonHScompendium).
 

### 2020-11-01 analysis/paper/paper.Rmd
added reference for bad bands

@@
 The NEON Airborne Observation Platform (AOP) collects airborne remote sensing data regularly at NEON sites. 
-There are three AOP payloads, each carrying three types of sensors: (1) a high-resolution digital camera to capture red, green, blue (RGB) true-color images, (2) a light detection and ranging (lidar) system to capture discrete and waveform lidar data with approximately 1-4 points/waveforms per square meter, and (3) a pushbroom-style imaging spectrometer (AVIRIS next-gen) to capture hyperspectral data.
+There are three AOP payloads, each carrying three types of sensors: (1) a high-resolution digital camera to capture red, green, blue (RGB) true-color images, (2) a light detection and ranging (lidar) system to capture discrete and waveform lidar data with approximately 1-4 points/waveforms per square meter, and (3) a pushbroom-style AVIRIS next-gen NEON Imaging Spectrometer (NIS) to capture hyperspectral data.
 The hyperspectral data contains 426 narrow bands spanning the visible to shortwave infrared wavelengths, from 380 - 2500nm, and the image data products derived have a spatial resolution of 1m.
 The AOP flies at an altitude of 1000m above ground level and the flight season typically runs from May to October.
-We worked with the Spectrometer orthorectified surface directional reflectance mosaic (data product code DP3.30006).
+We worked with the NIS-derived Spectrometer orthorectified surface directional reflectance mosaic (data product code DP3.30006).
 
@@
 
-knitr::kable(x = df, format = "simple")
+knitr::kable(x = df, 
+             # set column names
+             col.names = c("Starting wavelength [nm]",
+                           "Number of spectra"),
+             # center align text in the columns
+             align = c("c", "c"),
+             format = "simple")
 ```
@@
 We filtered out any spectra with a starting wavelength of 347nm, which were all collected at the Pu'u Maka'ala Natural Area Reserve (PUUM) NEON site, Pacific Tropical Domain 20, Hawaii, since the interpolation would require too much unavailable data outside the available range.
-We flagged "bad bands" from 1340-1445nm, 1790-1995nm, and greater than 2400nms o they could be excluded from later analysis.
-These wavelength ranges feature noisy reflectance data due to high water absoprtion.
+We flagged "bad bands" from 1340-1445nm, 1790-1995nm, and greater than 2400nm so they could be excluded from later analysis.
+These wavelength ranges feature noisy reflectance data due to high water absorption [@bajorski2011].
 

### 2020-11-01 R/clean_spectral_data.Rmd
working towards remaking the spectra plot

@@
 all_spectra %>% 
-  dplyr::filter(spectraID %in% c("NEON.PLA.D07.MLBS.00412_2018", 
-                                 "NEON.PLA.D01.HARV.04350_2018",
-                                 "NEON.PLA.D08.LENO.08750_2015",
-                                 "NEON.PLA.D01.BART.06298_2019")) %>%
+  dplyr::filter(spectraID %in% c("NEON.PLA.D11.CLBJ.00496_2018", 
+                                 "NEON.PLA.D01.HARV.09041_2018",
+                                 "NEON.PLA.D08.DELA.03903_2017")) %>%
   ggplot(aes(x = wavelength_nm, y = reflectance, group = spectraID)) +

### 2020-11-02 R/clean_spectral_data.Rmd
histogram figure spectra per species
need to rearrange the figures though in the text so study area comes first

@@
 interp_wl %>% 
-  dplyr::filter(spectraID %in% c("NEON.PLA.D07.MLBS.00412_2018", 
-                                 "NEON.PLA.D01.HARV.04350_2018",
-                                 "NEON.PLA.D08.LENO.08750_2015",
-                                 "NEON.PLA.D01.BART.06298_2019")) %>% 
+  dplyr::filter(spectraID %in% c("NEON.PLA.D11.CLBJ.00496_2018", 
+                                 "NEON.PLA.D01.HARV.09041_2018",
+                                 "NEON.PLA.D08.DELA.03903_2017")) %>% 
   ggplot(aes(x = wavelength_nm, y = reflectance, group = spectraID)) +
@@
   scale_fill_manual(values = c("#d95f0e", "#7bccc4"))
+```
+
+```{r}
+# Histogram with just the taxons that have enough samples 
+samplesCountGT <- taxonCounts %>% 
+  filter(countGT == TRUE)
+
+# plot histogram with count per genusSpecies. color bars based on whether
+# # samples exceeds the threshold
+ggplot(samplesCountGT, aes(y = reorder(genusSpecies, n), x = n)) + 
+  geom_bar(stat = "identity", fill = "#7E9E8F") + 
+  labs(y = "Scientific name\n",
+       x = "\nNumber of spectra") + 
+  #theme(axis.text = element_text(size = 12)) +
+  theme_bw(base_size=18)
 
@@
 ggsave(filename = here::here("analysis","figures", "hist_samples_per_species.pdf"), width = 9, height = 12)
+```
+
+```{r}
+# table with species info and counts
+
+library(taxize)
+
+taxon_info <- all_spectra %>% 
+  select(genusSpecies, taxonID) %>% 
+  distinct()
+
+taxons_GT <- taxonCounts %>% 
+  # keep taxons that have enough samples 
+  filter(countGT == TRUE) %>% 
+  # remove the column that indicates if the taxon has enough samples
+  select(-countGT) %>% 
+  # merge in the taxonID information 
+  left_join(taxon_info, by="genusSpecies")
+
+# for(i in 1:nrow(taxons_GT)){
+#   print(i)
+#   taxize::sci2comm(sci = taxons_GT[i]$genusSpecies)
+# }
+
 ```

### 2020-11-02 analysis/paper/paper.Rmd
histogram figure spectra per species
need to rearrange the figures though in the text so study area comes first

@@
 Most of them started at 384nm, so we linearly interpolated each spectrum to have the same 426 bands starting at 384nm and ending at 2512nm with 5nm increments.
-We filtered out any spectra with a starting wavelength of 347nm, which were all collected at the Pu'u Maka'ala Natural Area Reserve (PUUM) NEON site, Pacific Tropical Domain 20, Hawaii, since the interpolation would require too much unavailable data outside the available range.
+Spectra from the Pu'u Maka'ala Natural Area Reserve (PUUM) NEON site in Hawaii had a starting wavelength of 347nm. 
+This is because NEON sub-contracted the PUUM flights to the Global Airborne Observatory, which has a spectrometer with a different wavelength range compared to the NIS payloads *([cite?](https://www.neonscience.org/observatory/observatory-blog/remote-sensing-partnership-made-paradise))*.
+We filtered out any spectra with a starting wavelength of 347nm, since the interpolation would require too much unavailable data outside the available range.
 We flagged "bad bands" from 1340-1445nm, 1790-1995nm, and greater than 2400nm so they could be excluded from later analysis.
@@
 We figured that a small sample size would lead to a poor representation in model training. 
-From the initial collection of 146 species, 50 species had at least 20 samples. 
+From the initial collection of 146 species, 50 species had at least 20 samples (Figure \@ref(fig:fig-histSpecies)). 
 We randomly assigned each spectrum to the training or validation sets, using an 80/20 split. All 50 species (and therefore, all genus values) were represented in each set.
@@
 
+```{r fig-histSpecies, fig.cap ="Histogram: spectra count per species", echo=FALSE,  out.width = "450px", message = FALSE, warning=FALSE}
+library(here)
+knitr::include_graphics(here::here("analysis","figures", "hist_samples_per_species.pdf"))
+```
+
 \clearpage

### 2020-11-04 R/clean_spectral_data.Rmd
updated some wording to reflect species vs genus
since the genus taxonRanks have a species of sp, they needed to be exluded from the initial count of unique species values. and I changed the wording later from species to "species and genus" or "scientific name label", recognizing that one of the classes was Larix sp, which is a genus taxon rank

@@
 
+# not including spectra with "sp" as the species, when taxonRank is genus
 message("# of unique species values:") 
-length(unique(all_spectra$species))
+tmp <- all_spectra %>% filter(taxonRank != "genus") %>% select(species, taxonRank)
+length(unique(tmp$species))
 
@@
 
-```{r}
+```{r fig.width=9, fig.height=12}
 # Histogram with just the taxons that have enough samples 

### 2020-11-04 analysis/paper/paper.Rmd
updated some wording to reflect species vs genus
since the genus taxonRanks have a species of sp, they needed to be exluded from the initial count of unique species values. and I changed the wording later from species to "species and genus" or "scientific name label", recognizing that one of the classes was Larix sp, which is a genus taxon rank

@@
 We removed varieties from scientific names to label each individual spectrum with its respective genus and species (when available). 
-There were 74 unique genus values, 118 unique species values, and 151 unique genus-species combinations.
+There were 74 unique genus values, 113 unique species values, and 151 unique genus-species combinations.
 The hyperspectral reflectance spectra extracted from the locations of the individual plants have different starting wavelengths:
@@
 
-Next, we removed samples belonging to species classes with few samples. 
+Next, we removed samples belonging to species and genus classes with few samples. 
 We figured that a small sample size would lead to a poor representation in model training. 
-From the initial collection of 146 species, 50 species had at least 20 samples (Figure \@ref(fig:fig-histSpecies)). 
-We randomly assigned each spectrum to the training or validation sets, using an 80/20 split. All 50 species (and therefore, all genus values) were represented in each set.
+From the initial collection of 151 unique scientific names identified at the genus or species level, 50 of them had at least 20 samples and were kept for the classification analysis (Figure \@ref(fig:fig-histSpecies)).
+We randomly assigned each spectrum to the training or validation sets, using an 80/20 split. All 50 scientific name labels were represented in each set.
 
@@
 
-**Figure - histogram of species counts? maybe in appendix, make a table or something to show species info for which ones were used after cleaning/filtering**
+**Maybe in appendix, make a table or something to show species info for which ones were used after cleaning/filtering**
+
 **Figure - cleaned data, make it legible, how to color if at all (the species/genus/variety color doesn't make sense anymore at this stage)...**
@@
 
-It ain't much...
+Histogram showing which scientific names have at least 20 samples to be included in the classification analysis. Note that the "Larix sp." invididuals are classified at the genus level.
 

### 2020-11-04 R/clean_spectral_data.Rmd
added figure with spectra
made it taller vertically and make all the spectra one color for now

@@
 Visualize the spectra, facet by scientific name, color by taxon rank. 
-```{r, warning=FALSE, fig.height=12, fig.width=16}
+```{r, eval = FALSE, include = "FALSE", warning=FALSE, fig.height=12, fig.width=16}
 filtered_spectra %>%
@@
 ```
+
+Color all the spectra with a single color
+
+```{r, eval = FALSE, include = "FALSE", warning=FALSE, fig.height=12, fig.width=16}
+filtered_spectra %>%
+  ggplot(aes(wavelength_nm, ifelse(mask, NA, reflectance), group = spectraID)) +
+  geom_path(alpha = .2, color = "#1c9099") +
+  labs(x = '\n Wavelength (nm)',
+       y = 'Reflectance \n') + 
+  facet_wrap(~genusSpecies
+             # wrap long scientific names
+             ,labeller = label_wrap_gen(20)
+             ,ncol=5) +
+  theme_minimal() + 
+  # Adjust the text size and rotation angle 
+  theme(text = element_text(size=11),
+        axis.text.x = element_text(angle=90, hjust=1),
+        axis.title.y = element_text(size = 14),
+        axis.title.x = element_text(size = 14))
+
+# save figure to file
+ggsave(filename = here::here("analysis","figures", "spectra_per_species.pdf"), width = 10, height = 12)
+```
+
+Color the spectra by the year they were collected?
+
+```{r, eval=FALSE, include=FALSE}
+filtered_spectra %>%
+  ggplot(aes(wavelength_nm, ifelse(mask, NA, reflectance), group = spectraID, color = as.factor(year))) +
+  geom_path(alpha = .2) +
+  labs(x = '\n Wavelength (nm)',
+       y = 'Reflectance \n',
+       color = "Year collected") + 
+  facet_wrap(~genusSpecies
+             # wrap long scientific names
+             ,labeller = label_wrap_gen(20)
+             ,ncol=10) + 
+  scale_color_manual(values = c("#e6550d","#1c9099", "#a8ddb5")) + 
+  theme_minimal() + 
+  # Adjust the text size and rotation angle 
+  theme(text = element_text(size=11),
+        axis.text.x = element_text(angle=90, hjust=1),
+        axis.title.y = element_text(size = 14),
+        axis.title.x = element_text(size = 14),
+        legend.text = element_text(size = 14),
+        legend.title = element_text(size = 14))
+```

### 2020-11-04 analysis/paper/paper.Rmd
added figure with spectra
made it taller vertically and make all the spectra one color for now

@@
 
+Show the spectra for each scientific name class (Figure \@ref(fig:fig-plotSpectra)).
+
 **Maybe in appendix, make a table or something to show species info for which ones were used after cleaning/filtering**
@@
 
-```{r fig-studyArea, fig.cap ="Study Area", echo=FALSE,  out.width = "450px", message = FALSE, warning=FALSE}
+```{r fig-studyArea, fig.cap ="Map", echo=FALSE,  out.width = "450px", message = FALSE, warning=FALSE}
 library(here)
@@
 
-```{r fig-histSpecies, fig.cap ="Histogram: spectra count per species", echo=FALSE,  out.width = "450px", message = FALSE, warning=FALSE}
+```{r fig-histSpecies, fig.cap ="Histogram", echo=FALSE,  out.width = "450px", message = FALSE, warning=FALSE}
 library(here)
@@
 
+```{r fig-plotSpectra, fig.cap ="Spectra", echo=FALSE,  out.width = "450px", message = FALSE, warning=FALSE}
+library(here)
+knitr::include_graphics(here::here("analysis","figures", "spectra_per_species.pdf"))
+```
+
 \clearpage

### 2020-11-04 R/clean_spectral_data.Rmd
decreased fig height to make them fit on 1 page each

@@
 # save figure to file
-ggsave(filename = here::here("analysis","figures", "hist_samples_per_species.pdf"), width = 9, height = 12)
+ggsave(filename = here::here("analysis","figures", "hist_samples_per_species.pdf"), width = 9, height = 10)
 ```
@@
 # save figure to file
-ggsave(filename = here::here("analysis","figures", "spectra_per_species.pdf"), width = 10, height = 12)
+ggsave(filename = here::here("analysis","figures", "spectra_per_species.pdf"), width = 10, height = 10)
 ```

### 2020-11-04 R/clean_spectral_data.Rmd
added more methods text
and made font smaller in histogram fig

@@
   #theme(axis.text = element_text(size = 12)) +
-  theme_bw(base_size=18)
+  theme_bw(base_size=16)
 
 # save figure to file
-ggsave(filename = here::here("analysis","figures", "hist_samples_per_species.pdf"), width = 9, height = 10)
+ggsave(filename = here::here("analysis","figures", "hist_samples_per_species.pdf"), width = 9, height = 9)
 ```
@@
 # save figure to file
-ggsave(filename = here::here("analysis","figures", "spectra_per_species.pdf"), width = 10, height = 10)
+ggsave(filename = here::here("analysis","figures", "spectra_per_species.pdf"), width = 10, height = 9)
 ```

### 2020-11-04 analysis/paper/paper.Rmd
added more methods text
and made font smaller in histogram fig

@@
 
-**Figure - cleaned data, make it legible, how to color if at all (the species/genus/variety color doesn't make sense anymore at this stage)...**
+## Species and genus classification {-}
 
-## Species classification {-}
+We trained a random forest (RF) classifier to predict species based on NEON hyperspectral reflectance data.
+RF is a machine learning approach that utilizes an ensemble of decision trees and iterative subsets of the training data.
+We performed hyperparameter optimization of ntree, the number of decision trees in the forest, and mtry, the number of variables randomly sampled at each node in the tree.
+We used the optimal values of 27 for mtry and 1000 for ntree in our final RF model.
+We computed overall accuracy, confusion matrices, and variable importance plots.
 
-Random forest classifier
-Hyperparameter optimization (mtry, ntree)
-Overall accuracy, confusion matrices, variable importance
-
-## Genus classification {-}
-
-Same setup as species classification, but used the genus as each class label.
-
+To evaluate genus classification accuracy, we followed the same step as described for species classiciation, but we used each spectrum's associated genus as each class label. 
 We provide all of our code in a research compendium here:[https://github.com/vscholl/neonHScompendium](https://github.com/vscholl/neonHScompendium).

### 2020-11-05 R/classify_species.Rmd
working on species classification result table

---
title: "Classify veg"
author: "Victoria Scholl"
date: "08/20/2020 Earth Lab GRA"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
library(tidyr)
library(dplyr)
library(randomForest)
library(caret)
library(rfPermute)
library(parallel)
library(doParallel)

dir.create("output")
```

## Data preparation

```{r read_data, message=FALSE}
# start timer
start_time <- Sys.time()

# read the cleaned data, labeled for train, test, 
cleaned_data <- readr::read_csv(here::here("analysis", "data", "derived_data", 
                                              "cleaned_spectra.csv"))
```

Format the data so each row is a sample (reflectance spectrum) and each column is a descriptive feature (reflectance value at each wavelength). Each row must have a label to predict for the classification, such as genusSpecies (the first two pieces of each scientific name excluding any variety info. Example: Pseudotsuga menziesii. Genus is Pseudotsuga, species is menziesii).

```{r format_data}
# columns to keep for training the classifier
cols_to_keep <- c("spectraID", "band_idx", "reflectance", "genusSpecies")

# training set -------------------------------------------------
# select and format the training data 
train_data <- cleaned_data %>% 
  # keep only the spectra marked as part of the "train" or "test" set
    # train 60% + test 20% = 80% for training with k-fold cross validation
  dplyr::filter(group == "train" | group == "test") %>% 
  # remove all entries with NA reflectance values, indicative of bad bands
  dplyr::filter(!is.na(reflectance)) %>% 
  # filter spectra identified at the genus level, indicated by species == "sp."
  dplyr::filter(species != "sp.") %>% 
  # keep only the columns needed for classification
  dplyr::select(all_of(cols_to_keep)) %>%
  # reshape the data from long to wide so each row is a spectrum, 
    # and each col contains reflectance per band.
  tidyr::pivot_wider(names_from = band_idx, 
                     values_from = reflectance) %>% 
  # remove spectraID since it's not a descriptive feature 
  dplyr::select(-spectraID) 
  # VS-NOTE: subset small # of samples for testing!!!
  #dplyr::sample_frac(0.5)

# convert genusSpecies to factor, so R does classification instead of regression.
# reassign the factor levels after subsetting the training data
train_data$genusSpecies <- factor(train_data$genusSpecies)

# validation set -----------------------------------------------
valid_data <- cleaned_data %>% 
  #  keep only the spectra marked as part of the independent validation set
  dplyr::filter(group == "valid") %>% 
  # remove all entries with NA reflectance values, indicative of bad bands
  dplyr::filter(!is.na(reflectance)) %>% 
  # filter spectra identified at the genus level, indicated by species == "sp."
  dplyr::filter(species != "sp.") %>%
  # keep only the columns needed for classification
  dplyr::select(all_of(cols_to_keep)) %>%
  # reshape the data from long to wide so each row is a spectrum, 
    # and each col contains reflectance per band.
  tidyr::pivot_wider(names_from = band_idx, 
                     values_from = reflectance) %>% 
  # remove spectraID since it's not a descriptive feature 
  dplyr::select(-spectraID)

valid_data$genusSpecies <- factor(valid_data$genusSpecies)
```

## Parameter tuning

Two hyperparameters to tune in randomForest classification: mtry and ntree

```{r caret_parallel, eval=FALSE}
start.time = Sys.time()

cl <- makePSOCKcluster(parallel::detectCores() -1)

registerDoParallel(cl)

## All subsequent models are then run in parallel
model <- train(factor(genusSpecies) ~ ., 
               data = train_data, 
               tuneGrid = expand.grid(mtry = seq(from = 6, to = 30, by = 3)),
               method = "rf",
               metric = "Accuracy")

print(model)

## When you are done:
stopCluster(cl)

end.time = Sys.time()

print(end.time - start.time)

# took 18 minutes to test 8 values of mtry with 60% subset of training data

# save the output model to file 
saveRDS(model, "output/model_mtry_tuning.rds")
```

```{r plot_mtry_tuning, eval=FALSE}
# load mtry tuning model
model <- readRDS("output/model_mtry_tuning.rds")

# save plot to file
pdf("output/mtry_tuning_plot.pdf",
    width = 6,
    height = 4)

plot(model)

dev.off()

plot(model)
```

```{r optimize_ntree, eval=FALSE}
# https://rpubs.com/phamdinhkhanh/389752

start.time = Sys.time()

set.seed(44)

# start parallel cluster
cl <- makePSOCKcluster(parallel::detectCores() -1) 
registerDoParallel(cl)

# iteratively check ntree values 
modellist <- list()
for (ntree in c(100, 200, 500, 1000, 2000)){
  print(ntree)
  fit <- train(factor(genusSpecies) ~ ., 
                data = train_data, 
                method = "rf",
                metric = "Accuracy",
                ntree = ntree,
                tuneGrid=expand.grid(mtry = 27), 
                verbose = TRUE)
  key <- toString(ntree)
  modellist[[key]] <- fit
}

# when you are done with parallel
stopCluster(cl)

# save model list to file 
saveRDS(modellist, "output/modelList_ntree_tuning.rds")

end.time = Sys.time()
end.time - start.time

# 43 minutes
```

```{r ntree_tuning_results, eval=FALSE}
# load ntree tuning model list
modellist <- readRDS("output/modelList_ntree_tuning.rds")

#Compare results
results <- resamples(modellist)
summary(results)

dotplot(results)
```

## Classify species

We classified species using the optimal values for mtry (27) and ntree (1000)

```{r randomForest_species_train, eval=FALSE}
set.seed(44)

start.time = Sys.time()

# set up parallel 
cl <- makePSOCKcluster(parallel::detectCores() -1) 
registerDoParallel(cl)

# CARET train RF model using the optimal parameters 
#rf_classifier_species <- train(factor(genusSpecies) ~ ., 
#                              data = train_data, 
#                              method = "rf",
#                              ntree = 1000,
#                              tuneGrid=expand.grid(mtry = 27),
#                              metric = "Accuracy")

# RANDOMFOREST train RF model using optimal parameters
rf_classifier_species <- randomForest::randomForest(factor(genusSpecies) ~ ., 
                                           data=train_data, 
                                           importance=TRUE,
                                            mtry = 27,
                                            ntree = 1000)

# when you are done with parallel
stopCluster(cl)

end.time = Sys.time()
end.time - start.time

# show classification results
print(rf_classifier_species)

# save classifier to file 
saveRDS(object = rf_classifier_species, file = "output/rf_classifier_species_mtry27_ntree100.rds")
```

```{r load_rf}
rf_classifier_species <- readRDS(here::here("R","output","rf_classifier_species_mtry27_ntree100.rds"))
```

```{r randomForest_species_valid}
# predict species using independent validation set
rf_valid_preds_species <- predict(rf_classifier_species, valid_data, type="class")
```

### Overall Accuracy

```{r randomForest_OA}
# combine observed and predicted values into a single data frame
rf_train_results <- tibble(obs = as.character(train_data$genusSpecies),
                           pred = as.character(as.vector(rf_classifier_species$pred)))

rf_valid_results <- tibble(obs = as.character(valid_data$genusSpecies),
                              pred = as.character(as.vector(rf_valid_preds_species)))

# calculate overall accuracy for training set predictions 
OA_randomForest_train <- sum(rf_train_results$obs == rf_train_results$pred) / nrow(rf_train_results)
print(paste("Overall Accuracy of TRAINING set SPECIES predictions:", round(OA_randomForest_train, 4)))

# calculate overall accuracy for validation set predictions 
OA_randomForest_valid <- sum(rf_valid_results$obs == rf_valid_results$pred) / nrow(rf_valid_results)
print(paste("Overall Accuracy of VALIDATION set SPECIES predictions:", round(OA_randomForest_valid, 4)))
```

### Confusion Matrices

Visualize classification results across species for the TRAINING SET using ggplot

```{r randomForest_train_confMatrix, fig.width=14, fig.height=14, message=FALSE, eval=FALSE}
# list all combinations of true and predicted species 
all_combos <- expand.grid(obs = unique(rf_train_results$obs), 
				    pred = unique(rf_train_results$obs))

# format the observed and predicted values into a conf matrix

rf_train_results %>% 
  # count number of each obs/pred combination
  count(obs,pred) %>% 
  # fill in combinations with counts of zero
  complete(obs, pred, fill = list(n = 0)) %>% 
  # add rows witth species combinations not in predicted set
  right_join(all_combos) %>% 
  # set NA values to 0 in the n column
  mutate(n = ifelse(is.na(n), 0, n)) %>% 
  # set raster cell fill color based on count
  ggplot(aes(x = obs, y = fct_rev(pred), fill = n)) + 
  geom_raster() + 
  # add boxes around diagonal cells
  geom_tile(aes(color = obs == pred, width = 0.95, height = 0.95), size = 0.5) + 
  scale_color_manual(values = c(NA, "black")) + 
  # rotate x axis text 
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) + 
  labs(x = "Truth", y = "Predicted", title = "RandomForest Classification Confusion Matrix: Predicted vs. True species for Training Set") + 
  scale_fill_gradient(low = "white", high = "#43a2ca") + 
  # add count in each cell
  geom_text(aes(label = n, alpha = n > 0)) + 
  # remove the color legend
  #guides(color = "none")
  # remove legend entirely
  theme(legend.position = "none")

# save confusion matrix to file
ggsave(filename = here::here("analysis","figures", "confMat_species.pdf"), width = 14, height = 14)
```

Use the randomForest model to predict validation set species. Visualize classification results across species for the VALIDATION SET using ggplot:

```{r randomForest_valid_confMatrix, fig.width=14, fig.height=14, message=FALSE}

# list all combinations of true and predicted species 
all_combos <- expand.grid(obs = unique(rf_valid_results$obs), 
				    pred = unique(rf_valid_results$obs))

# format the observed and predicted values into a conf matrix

rf_valid_results %>% 
  # count number of each obs/pred combination
  count(obs,pred) %>% 
  # fill in combinations with counts of zero
  complete(obs, pred, fill = list(n = 0)) %>% 
  # add rows witth species combinations not in predicted set
  right_join(all_combos) %>% 
  # set NA values to 0 in the n column
  mutate(n = ifelse(is.na(n), 0, n)) %>% 
  # set raster cell fill color based on count
  ggplot(aes(x = obs, y = fct_rev(pred), fill = n)) + 
  geom_raster() + 
  # add boxes around diagonal cells
  geom_tile(aes(color = obs == pred, width = 0.95, height = 0.95), size = 0.5) + 
  scale_color_manual(values = c(NA, "black")) + 
  # rotate x axis text 
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) + 
  labs(x = "Truth", y = "Predicted", title = "RandomForest Classification Confusion Matrix: Predicted vs. True SPECIES for Validation Set") + 
  scale_fill_gradient(low = "white", high = "#43a2ca") + 
  # add count in each cell
  geom_text(aes(label = n, alpha = n > 0)) + 
  # remove legend entirely
  theme(legend.position = "none")

```

### Variable Importance

```{r varImp_rf_species, results="hide"}

# format the randomForest variable importance for ggplot
imp <- as.data.frame(varImpPlot(rf_classifier_species))
imp$varnames <- rownames(imp) # row names to column
rownames(imp) <- NULL

# isolate the band#'s and wavelength values
band_wl_lut <- cleaned_data %>% select(band_idx, wavelength_nm) %>% 
  # keep only rows with unique combinations of band_idx and wavelength_nm
  distinct()

# add a new column with wavelength_nm values
imp <- merge(imp, band_wl_lut, by.x="varnames", by.y="band_idx")
```

```{r varImp_byWavelength_plot, fig.height=6, fig.width=10}
ggplot(imp, aes(x=wavelength_nm,  
                y=MeanDecreaseAccuracy)) + 
  # add bars where each height is equal to the variable importance metric
  geom_col(aes(fill = MeanDecreaseAccuracy), color = NA) +
  theme_minimal() + 
  labs(x = "Wavelength (nm)", 
       y = "Variable importance (MDA)",
       title = "Variable Importance: Mean Decrease Accuracy (MDA) vs. Wavelength",
       fill = "MDA")  + 
  scale_fill_gradient(low = "#d9d9d9", high = "#43a2ca")
```

```{r varImp_Gini, fig.height=6, fig.width=10}
ggplot(imp, aes(x = wavelength_nm,  
                y = MeanDecreaseGini)) + 
  # add bars where each height is equal to the variable importance metric
  geom_col(aes(fill = MeanDecreaseGini), color = NA) +
  theme_minimal() + 
  labs(x = "Wavelength (nm)", 
       y = "Variable importance (MDG)",
       title = "Variable Importance: Mean Decrease Gini (MDG) vs. Wavelength",
       fill = "MDG")  + 
  scale_fill_gradient(low = "#d9d9d9", high = "#43a2ca")
```

### Add wavelengths of interest

For interpretation of the variable importance plots, wavelengths of interest for the red, green, blue, and NIR bands are added here. The wavelength ranges are based on [Landsat 8 designations](https://www.neonscience.org/hyper-spec-intro).

```{r vertical_dashed_lines}
# plot variable importance with wavelengths of interest marked by dashed vertical lines

ggplot(imp, aes(x = wavelength_nm,  
                y = MeanDecreaseGini)) +

  # add vertical dashed lines to indicate notable wavelengths
  geom_vline(xintercept=480, linetype="dashed", color="blue", size=1, alpha=0.5) + # BLUE 
  geom_vline(xintercept=550, linetype="dashed", color="green", size=1, alpha=0.5) + # GREEN
  geom_vline(xintercept=650, linetype="dashed", color="red", size=1, alpha=0.5) + # RED
  geom_vline(xintercept=870, linetype="dashed", color="grey30", size=1, alpha=0.5) + # NIR 
  geom_vline(xintercept=1600, linetype="dashed", color="chocolate4", size=1, alpha=0.5) + # SWIR

  # add bars where each height is equal to the variable importance metric
  geom_col(aes(fill = MeanDecreaseGini), color = NA) +
  theme_minimal() + 
  labs(x = "Wavelength (nm)", 
       y = "Variable importance (MDG)",
       title = "Variable Importance: Mean Decrease Gini (MDG) vs. Wavelength",
       fill = "MDG")  + 
  scale_fill_gradient(low = "#d9d9d9", high = "#43a2ca")

```
```{r vertical_shaded_regions}
# plot variable importance with wavelengths of interest marked by dashed vertical lines

ggplot(imp, aes(x = wavelength_nm,  
                y = MeanDecreaseGini)) +

  # add vertical shaded regions to indicate notable wavelengths
  annotate(geom = "rect", xmin = 450, xmax = 510, ymin = -5, ymax = Inf, fill = "blue", alpha = 0.1) + # BLUE
  annotate(geom = "rect", xmin = 530, xmax = 590, ymin = -5, ymax = Inf, fill = "green", alpha = 0.1) + # GREEN
  annotate(geom = "rect", xmin = 640, xmax = 670, ymin = -5, ymax = Inf, fill = "red", alpha = 0.1) + # RED
  annotate(geom = "rect", xmin = 850, xmax = 880, ymin = -5, ymax = Inf, fill = "grey30", alpha = 0.1)  + # NIR 
  annotate(geom = "rect", xmin = 1570, xmax = 1650, ymin = -5, ymax = Inf, fill = "chocolate4", alpha = 0.1)  + # SWIR

  # add bars where each height is equal to the variable importance metric
  geom_col(aes(fill = MeanDecreaseGini), color = NA) +
  theme_minimal() + 
  labs(x = "Wavelength (nm)", 
       y = "Variable importance (MDG)",
       title = "Variable Importance: Mean Decrease Gini (MDG) vs. Wavelength",
       fill = "MDG")  + 
  scale_fill_gradient(low = "#d9d9d9", high = "#43a2ca")

```

## Classify genus

```{r set_up_data_genus}
# columns to keep for training the classifier
cols_to_keep_genus <- c("spectraID", "band_idx", "reflectance", "genus")

# training set -------------------------------------------------
# select and format the training data 
train_data_genus <- cleaned_data %>% 
  # keep only the spectra marked as part of the "train" or "test" set
    # train 60% + test 20% = 80% for training with k-fold cross validation
  dplyr::filter(group == "train" | group == "test") %>% 
  # remove all entries with NA reflectance values, indicative of bad bands
  dplyr::filter(!is.na(reflectance)) %>% 
  # keep only the columns needed for classification
  dplyr::select(all_of(cols_to_keep_genus)) %>%
  # reshape the data from long to wide so each row is a spectrum, 
    # and each col contains reflectance per band.
  tidyr::pivot_wider(names_from = band_idx, 
                     values_from = reflectance) %>% 
  # remove spectraID since it's not a descriptive feature 
  dplyr::select(-spectraID)

# convert label to factor, so R does classification instead of regression
train_data_genus$genus <- factor(train_data_genus$genus)

# validation set -----------------------------------------------
valid_data_genus <- cleaned_data %>% 
  #  keep only the spectra marked as part of the independent validation set
  dplyr::filter(group == "valid") %>% 
  # remove all entries with NA reflectance values, indicative of bad bands
  dplyr::filter(!is.na(reflectance)) %>% 
  # keep only the columns needed for classification
  dplyr::select(all_of(cols_to_keep_genus)) %>%
  # reshape the data from long to wide so each row is a spectrum, 
    # and each col contains reflectance per band.
  tidyr::pivot_wider(names_from = band_idx, 
                     values_from = reflectance) %>% 
  # remove spectraID since it's not a descriptive feature 
  dplyr::select(-spectraID)

valid_data_genus$genus <- factor(valid_data_genus$genus)
```

```{r randomForest_train_genus, eval=FALSE}
set.seed(44) # for obama

start.time = Sys.time()

# set up parallel 
cl <- makePSOCKcluster(parallel::detectCores() -1) 
registerDoParallel(cl)

# train random forest classifiers: predict genus
rf_classifier_genus <- randomForest::randomForest(factor(genus) ~ ., 
                                           data=train_data_genus, 
                                           importance=TRUE,
                                           mtry=27,
                                           ntree=1000)
# when you are done with parallel
stopCluster(cl)
end.time = Sys.time()
end.time - start.time

# show classification results
print(print(rf_classifier_genus))

# save classifier to file 
saveRDS(object = rf_classifier_genus, file = "output/rf_classifier_genus_mtry27_ntree100.rds")
```

```{r}
rf_classifier_genus <- readRDS("output/rf_classifier_genus_mtry27_ntree100.rds")
```

### Overall Accuracy

Predict genus for the validation set and assess overall accuracy for the training and validation sets.

```{r randomForest_OA_genus}
# predict species for validation set
p_genus <- predict(rf_classifier_genus, valid_data_genus, type = "class")

# combine observed and predicted values into a single data frame
rf_train_results_genus <- tibble(obs = as.character(train_data_genus$genus),
                              pred = as.character(as.vector(rf_classifier_genus$predicted)))

rf_valid_results_genus <- tibble(obs = as.character(valid_data_genus$genus),
                              pred = as.character(as.vector(p_genus)))

# calculate overall accuracy for training set predictions 
OA_randomForest_train_genus <- sum(rf_train_results_genus$obs == rf_train_results_genus$pred) / nrow(rf_train_results_genus)
print(paste("Overall Accuracy of training set genus predictions:", round(OA_randomForest_train_genus, 4)))

# calculate overall accuracy for validation set predictions 
OA_randomForest_valid_genus <- sum(rf_valid_results_genus$obs == rf_valid_results_genus$pred) / nrow(rf_valid_results)
print(paste("Overall Accuracy of validation set genus predictions:", round(OA_randomForest_valid_genus, 4)))
```

### Confusion Matrices

```{r randomForest_valid_confMatrix_genus, fig.width=12, fig.height=12, message=FALSE}

# list all combinations of true and predicted species 
all_combos_genus <- expand.grid(obs = unique(rf_valid_results_genus$obs), 
				    pred = unique(rf_valid_results_genus$obs))

# format the observed and predicted values into a conf matrix

rf_valid_results_genus %>% 
  # count number of each obs/pred combination
  count(obs,pred) %>% 
  # fill in combinations with counts of zero
  complete(obs, pred, fill = list(n = 0)) %>% 
  # add rows witth species combinations not in predicted set
  right_join(all_combos_genus) %>% 
  # set NA values to 0 in the n column
  mutate(n = ifelse(is.na(n), 0, n)) %>% 
  # set raster cell fill color based on count
  ggplot(aes(x = obs, y = fct_rev(pred), fill = n)) + 
  geom_raster() + 
  # add boxes around diagonal cells
  geom_tile(aes(color = obs == pred, width = 0.95, height = 0.95), size = 0.5) + 
  scale_color_manual(values = c(NA, "black")) + 
  # rotate x axis text 
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) + 
  labs(x = "Truth", y = "Predicted", title = "RandomForest Classification Confusion Matrix: Predicted vs. True GENUS for Validation Set") + 
  scale_fill_gradient(low = "white", high = "#fc8d59") + 
  # add count in each cell
  geom_text(aes(label = n, alpha = n > 0)) + 
  # remove the color legend
  #guides(color = "none")
  # remove legend entirely
  theme(legend.position = "none")
```

```{r randomForest_varImpPlot_ggplot_genus, results="hide"}
imp_genus <- as.data.frame(varImpPlot(rf_classifier_genus))
imp_genus$varnames <- rownames(imp_genus) # row names to column
rownames(imp_genus) <- NULL  
# add a new column with wavelength_nm values
imp_genus <- merge(imp_genus, band_wl_lut, by.x="varnames", by.y="band_idx")
```

```{r varImp_MDA_genus, fig.height=6, fig.width=10}
ggplot(imp_genus, aes(x=wavelength_nm,  
                y=MeanDecreaseAccuracy)) + 
  # add bars where each height is equal to the variable importance metric
  geom_col(aes(fill = MeanDecreaseAccuracy), color = NA) +
  theme_minimal() + 
  labs(x = "Wavelength (nm)", 
       y = "Variable importance (MDA)",
       title = "GENUS classification variable importance: Mean Decrease Accuracy (MDA) vs. Wavelength",
       fill = "MDA")  + 
  scale_fill_gradient(low = "#d9d9d9", high = "#fc8d59")
```

```{r varImp_Gini_genus, fig.height=6, fig.width=10}
ggplot(imp_genus, aes(x = wavelength_nm,  
                y = MeanDecreaseGini)) + 
  # add bars where each height is equal to the variable importance metric
  geom_col(aes(fill = MeanDecreaseGini), color = NA) +
  theme_minimal() + 
  labs(x = "Wavelength (nm)", 
       y = "Variable importance (MDG)",
       title = "GENUS classification variable importance: Mean Decrease Gini (MDG) vs. Wavelength",
       fill = "MDG")  + 
  scale_fill_gradient(low = "#d9d9d9", high = "#fc8d59")
```

### Add reflectance spectrum to interpret variable importance?

```{r varImp_with_refl_spectrum}
imp_and_refl <- imp

```

```{r randomForest_varImpPlot_ggplot, eval=FALSE, include=FALSE}

# default varImpPlot visualization in randomForest
#varImpPlot(rf_classifier)

# format the randomForest variable importance for ggplot
imp <- as.data.frame(varImpPlot(rf_classifier))
imp$varnames <- rownames(imp) # row names to column
rownames(imp) <- NULL

# keep the n features with the highest MeanDecreaseAccuracy values 
imp_top30_mda <- arrange(imp, desc(MeanDecreaseAccuracy)) %>% 
  select(c(MeanDecreaseAccuracy, varnames)) %>% 
  slice_head(n = 40)

ggplot(imp_top30_mda, aes(x=reorder(varnames, MeanDecreaseAccuracy), 
                y=MeanDecreaseAccuracy
                #,color=MeanDecreaseAccuracy
                )) + 
  geom_point() +
  geom_segment(aes(x=varnames,xend=varnames,y=0,yend=MeanDecreaseAccuracy)) +
  #scale_color_continuous(type = "viridis") +
  ylab("MDA") +
  xlab("Variable Name") +
  ggtitle("RandomForest variable importance: Mean Decrease Accuracy (MDA)") +
  labs(color = "MDA") + 
  coord_flip() + 
  theme_minimal()
```

```{r how_long_did_this_take}
# stop timer
end_time <- Sys.time()

elapsed_time <- end_time - start_time

print(difftime(end_time, start_time, 
         units = c("auto", "secs", "mins", "hours",
                   "days", "weeks")))
```

### 2020-11-05 analysis/paper/paper.Rmd
working on species classification result table

@@
 We used the optimal values of 27 for mtry and 1000 for ntree in our final RF model.
+There were 49 species classes and 
 We computed overall accuracy, confusion matrices, and variable importance plots.
@@
 # Results {-}
+
+We report overall classification accuracy of the species and genus classification:
+
+```{r, message = FALSE, echo = FALSE}
+library(knitr)
+
+df <- data.frame(oa = c("species", "genus", 382, 384),
+                 species = c(94, 1220, 1733, 2922),
+                 genus = )
+
+knitr::kable(x = df, 
+             # set column names
+             col.names = c("Starting wavelength [nm]",
+                           "Number of spectra"),
+             # center align text in the columns
+             align = c("c", "c"),
+             format = "simple")
+```
 

### 2020-11-05 R/classify_species.Rmd
working on confusion matrix figures

@@
   theme(legend.position = "none")
-
-# save confusion matrix to file
-ggsave(filename = here::here("analysis","figures", "confMat_species.pdf"), width = 14, height = 14)
 ```
 
-Use the randomForest model to predict validation set species. Visualize classification results across species for the VALIDATION SET using ggplot:
+Use the randomForest model to predict VALIDATION set SPECIES. Visualize classification results across species for the VALIDATION SET using ggplot:
 
@@
   theme(axis.text.x = element_text(angle = 45, hjust = 1)) + 
-  labs(x = "Truth", y = "Predicted", title = "RandomForest Classification Confusion Matrix: Predicted vs. True SPECIES for Validation Set") + 
+  labs(x = "Truth", 
+       y = "Predicted"
+       #, title = "RandomForest Classification Confusion Matrix: Predicted vs. True SPECIES for Validation Set"
+       ) + 
   scale_fill_gradient(low = "white", high = "#43a2ca") + 
@@
   # remove legend entirely
-  theme(legend.position = "none")
+  theme(legend.position = "none",
+        axis.text = element_text(size = 16),
+        axis.title = element_text(size = 16))
 
+ggsave(filename = here::here("analysis","figures", "confMat_species_valid.pdf"), width = 15, height = 15)
 ```
@@
 ```{r}
-rf_classifier_genus <- readRDS("output/rf_classifier_genus_mtry27_ntree100.rds")
+rf_classifier_genus <- readRDS(here::here("R","output","rf_classifier_genus_mtry27_ntree100.rds"))
 ```
@@
   theme(legend.position = "none")
+
+# save figure to file 
+ggsave(filename = here::here("analysis","figures", "confMat_genus_valid.pdf"), width = 10, height = 10)
 ```

### 2020-11-05 analysis/paper/paper.Rmd
working on confusion matrix figures

@@
 
-Figure - confusion matrices
+Confusion matrix for species predictions (Figure \@ref(fig:fig-confMatSpecies))
+Confusion matrix for genus predictions (Figure \@ref(fig:fig-confMatGenus))
+
 Figure - variable importance vs wavelength
@@
 \clearpage
+
+## Figure 4 {-}
+
+```{r fig-confMatSpecies, fig.cap ="Confusion matrix Species", echo=FALSE,  out.width = "450px", message = FALSE, warning=FALSE}
+library(here)
+knitr::include_graphics(here::here("analysis","figures", "confMat_species_valid.pdf"))
+```
+
+\clearpage
+
+## Figure 5 {-}
+
+```{r fig-confMatGenus, fig.cap ="Confusion matrix Species", echo=FALSE,  out.width = "450px", message = FALSE, warning=FALSE}
+library(here)
+knitr::include_graphics(here::here("analysis","figures", "confMat_genus_valid.pdf"))
+```
+
+\clearpage

### 2020-11-05 analysis/paper/paper.Rmd
overall accuracy table

@@
 
-df <- data.frame(oa = c("species", "genus", 382, 384),
-                 species = c(94, 1220, 1733, 2922),
-                 genus = )
+df <- data.frame(oa = c("Out-of-bag (OOB)", "Independent validation"),
+                 species = c(54.6, 54.9),
+                 genus = c(60.5, 60.0))
 
@@
              # set column names
-             col.names = c("Starting wavelength [nm]",
-                           "Number of spectra"),
+             col.names = c("Overall Accuracy (%):",
+                           "Species Classification",
+                           "Genus Classification"),
              # center align text in the columns
-             align = c("c", "c"),
+             align = c("c", "c", "c"),
              format = "simple")

### 2020-11-05 R/clean_spectral_data.Rmd
results OA table and spectra figure

@@
 
-```{r, eval = FALSE, include = "FALSE", warning=FALSE, fig.height=12, fig.width=16}
+```{r, eval = FALSE, include = "FALSE", warning=FALSE, fig.height=8, fig.width=10}
 filtered_spectra %>%
@@
 # save figure to file
-ggsave(filename = here::here("analysis","figures", "spectra_per_species.pdf"), width = 10, height = 9)
+ggsave(filename = here::here("analysis","figures", "spectra_per_species.pdf"), width = 10, height = 8)
 ```

### 2020-11-05 analysis/paper/paper.Rmd
results OA table and spectra figure

@@
              # set column names
-             col.names = c("Overall Accuracy (%):",
+             col.names = c("Overall Accuracy [%]",
                            "Species Classification",

### 2020-11-05 R/classify_species.Rmd
experimentin with variable importance figures
combining MDA and MDG into a single figure?

@@
 
+# save confusion matrix to file 
 ggsave(filename = here::here("analysis","figures", "confMat_species_valid.pdf"), width = 15, height = 15)
@@
 ```{r varImp_byWavelength_plot, fig.height=6, fig.width=10}
-ggplot(imp, aes(x=wavelength_nm,  
+varImp_MDA_species <- ggplot(imp, aes(x=wavelength_nm,  
                 y=MeanDecreaseAccuracy)) + 
@@
   theme_minimal() + 
-  labs(x = "Wavelength (nm)", 
-       y = "Variable importance (MDA)",
-       title = "Variable Importance: Mean Decrease Accuracy (MDA) vs. Wavelength",
+  labs(x = "Wavelength (nm)\n", 
+       y = "Variable importance \nMean Decrease in Accuracy (MDA)\n",
+       #title = "Variable Importance: Mean Decrease Accuracy (MDA)",
        fill = "MDA")  + 
-  scale_fill_gradient(low = "#d9d9d9", high = "#43a2ca")
+  scale_fill_gradient(low = "#d9d9d9", high = "#43a2ca") + 
+   theme(axis.text = element_text(size = 14),
+        axis.title = element_text(size = 16))
+
+varImp_MDA_species
 ```
@@
 ```{r varImp_Gini, fig.height=6, fig.width=10}
-ggplot(imp, aes(x = wavelength_nm,  
+varImp_Gini_species <- ggplot(imp, aes(x = wavelength_nm,  
                 y = MeanDecreaseGini)) + 
@@
   labs(x = "Wavelength (nm)", 
-       y = "Variable importance (MDG)",
-       title = "Variable Importance: Mean Decrease Gini (MDG) vs. Wavelength",
+       y = "Variable importance \nMean Decrease Gini (MDG)\n",
+       #title = "Variable Importance: Mean Decrease Gini (MDG)",
        fill = "MDG")  + 
-  scale_fill_gradient(low = "#d9d9d9", high = "#43a2ca")
+  scale_fill_gradient(low = "#d9d9d9", high = "#43a2ca") + 
+  theme(axis.text = element_text(size = 14),
+        axis.title = element_text(size = 16),
+        legend.text = element_text(size = 14))
+
+varImp_Gini_species
+```
+
+Single figure with both variable importance plots:
+
+```{r varmImp_species}
+library(ggpubr)
+
+ggpubr::ggarrange(varImp_MDA_species, varImp_Gini_species, nrow = 2)
+
+# save confusion matrix to file 
+ggsave(filename = here::here("analysis","figures", "varImp_species.pdf"), width = 10, height = 10)
 ```

### 2020-11-05 R/clean_spectral_data.Rmd
experimentin with variable importance figures
combining MDA and MDG into a single figure?

@@
 
-```{r, eval = FALSE, include = "FALSE", warning=FALSE, fig.height=8, fig.width=10}
+```{r, warning=FALSE, fig.height=8, fig.width=10}
 filtered_spectra %>%
@@
 # save figure to file
-ggsave(filename = here::here("analysis","figures", "spectra_per_species.pdf"), width = 10, height = 8)
+ggsave(filename = here::here("analysis","figures", "spectra_per_species.pdf"), width = 10, height = 11)
 ```

### 2020-11-05 analysis/paper/paper.Rmd
experimentin with variable importance figures
combining MDA and MDG into a single figure?

@@
              # set column names
-             col.names = c("Overall Accuracy [%]",
-                           "Species Classification",
-                           "Genus Classification"),
+             col.names = c("Overall Classification Accuracy [%]",
+                           "Species",
+                           "Genus"),
              # center align text in the columns
@@
 
-Confusion matrix for species predictions (Figure \@ref(fig:fig-confMatSpecies))
-Confusion matrix for genus predictions (Figure \@ref(fig:fig-confMatGenus))
+Confusion matrix for species classification (Figure \@ref(fig:fig-confMatSpecies))
 
-Figure - variable importance vs wavelength
+Confusion matrix for genus classification (Figure \@ref(fig:fig-confMatGenus))
+
+Variable importance for species classification (Figure \@ref(fig:varImpSpecies))
+
 Figure - multi-year observations of individual plants?
@@
 
-```{r fig-confMatGenus, fig.cap ="Confusion matrix Species", echo=FALSE,  out.width = "450px", message = FALSE, warning=FALSE}
+```{r fig-confMatGenus, fig.cap ="Confusion matrix Genus", echo=FALSE,  out.width = "450px", message = FALSE, warning=FALSE}
 library(here)
@@
 \clearpage
+
+## Figure 6 {-}
+
+```{r fig-varImpSpecies, fig.cap ="Variable Importance Species", echo=FALSE,  out.width = "450px", message = FALSE, warning=FALSE}
+library(here)
+knitr::include_graphics(here::here("analysis","figures", "varImp_species.pdf"))
+```
+
+\clearpage

### 2020-11-05 analysis/paper/paper.Rmd
results figures

@@
 
-Variable importance for species classification (Figure \@ref(fig:varImpSpecies))
+Variable importance for species classification (Figure \@ref(fig:fig-varImpSpecies))
 
@@
 
-```{r fig-confMatSpecies, fig.cap ="Confusion matrix Species", echo=FALSE,  out.width = "450px", message = FALSE, warning=FALSE}
+```{r fig-confMatSpecies, fig.cap ="Confusion matrix for species classification", echo=FALSE,  out.width = "450px", message = FALSE, warning=FALSE}
 library(here)
@@
 
-```{r fig-confMatGenus, fig.cap ="Confusion matrix Genus", echo=FALSE,  out.width = "450px", message = FALSE, warning=FALSE}
+```{r fig-confMatGenus, fig.cap ="Confusion matrix for genus classification", echo=FALSE,  out.width = "450px", message = FALSE, warning=FALSE}
 library(here)
@@
 
-```{r fig-varImpSpecies, fig.cap ="Variable Importance Species", echo=FALSE,  out.width = "450px", message = FALSE, warning=FALSE}
+```{r fig-varImpSpecies, fig.cap ="Variable importance for species classification", echo=FALSE,  out.width = "450px", message = FALSE, warning=FALSE}
 library(here)

### 2020-11-05 R/classify_species.Rmd
combined var imp plots into one figure
MDA for both species and genus

@@
 library(doParallel)
+library(ggpubr)
 
@@
   theme(axis.text.x = element_text(angle = 45, hjust = 1)) + 
-  labs(x = "Truth", 
-       y = "Predicted"
+  labs(x = "True Species", 
+       y = "Predicted Species"
        #, title = "RandomForest Classification Confusion Matrix: Predicted vs. True SPECIES for Validation Set"
@@
   labs(x = "Wavelength (nm)\n", 
-       y = "Variable importance \nMean Decrease in Accuracy (MDA)\n",
-       #title = "Variable Importance: Mean Decrease Accuracy (MDA)",
+       y = "Variable importance \n",
+       title = "Variable Importance: Mean Decrease Accuracy (MDA) \n\nSpecies classification",
        fill = "MDA")  + 
@@
    theme(axis.text = element_text(size = 14),
-        axis.title = element_text(size = 16))
+        axis.title = element_text(size = 14),
+        title = element_text(size = 16))
 
@@
 # save confusion matrix to file 
-ggsave(filename = here::here("analysis","figures", "varImp_species.pdf"), width = 10, height = 10)
+#ggsave(filename = here::here("analysis","figures", "varImp_species.pdf"), width = 10, height = 10)
 ```
@@
   theme(axis.text.x = element_text(angle = 45, hjust = 1)) + 
-  labs(x = "Truth", y = "Predicted", title = "RandomForest Classification Confusion Matrix: Predicted vs. True GENUS for Validation Set") + 
+  labs(x = "True Genus", 
+       #title = "RandomForest Classification Confusion Matrix: Predicted vs. True GENUS for Validation Set",
+       y = "Predicted Genus") + 
   scale_fill_gradient(low = "white", high = "#fc8d59") + 
@@
   # remove legend entirely
-  theme(legend.position = "none")
+  theme(legend.position = "none",
+        axis.text = element_text(size = 14),
+        axis.title = element_text(size = 14))
 
@@
 ```{r varImp_MDA_genus, fig.height=6, fig.width=10}
-ggplot(imp_genus, aes(x=wavelength_nm,  
+varImp_MDA_genus <- ggplot(imp_genus, aes(x=wavelength_nm,  
                 y=MeanDecreaseAccuracy)) + 
@@
   theme_minimal() + 
-  labs(x = "Wavelength (nm)", 
-       y = "Variable importance (MDA)",
-       title = "GENUS classification variable importance: Mean Decrease Accuracy (MDA) vs. Wavelength",
+  labs(x = "Wavelength (nm)\n", 
+       y = "Variable importance\n",
+       title = "Genus classification",
        fill = "MDA")  + 
-  scale_fill_gradient(low = "#d9d9d9", high = "#fc8d59")
+  scale_fill_gradient(low = "#d9d9d9", high = "#fc8d59") + 
+   theme(axis.text = element_text(size = 14),
+        axis.title = element_text(size = 14),
+        title = element_text(size = 16))
+
+varImp_MDA_genus
+
 ```
@@
        fill = "MDG")  + 
-  scale_fill_gradient(low = "#d9d9d9", high = "#fc8d59")
+  scale_fill_gradient(low = "#d9d9d9", high = "#fc8d59") + 
+   theme(axis.text = element_text(size = 14),
+        axis.title = element_text(size = 16))
+```
+
+Single figure with MDA variable importance plots for both species and genus classifiers:
+
+```{r varmImp_species}
+ggpubr::ggarrange(varImp_MDA_species, varImp_MDA_genus, nrow = 2)
+
+# save confusion matrix to file 
+ggsave(filename = here::here("analysis","figures", "varImp.pdf"), width = 10, height = 10)
 ```
@@
 ### Add reflectance spectrum to interpret variable importance?
-
-```{r varImp_with_refl_spectrum}
-imp_and_refl <- imp
-
-```
-
-```{r randomForest_varImpPlot_ggplot, eval=FALSE, include=FALSE}
-
-# default varImpPlot visualization in randomForest
-#varImpPlot(rf_classifier)
-
-# format the randomForest variable importance for ggplot
-imp <- as.data.frame(varImpPlot(rf_classifier))
-imp$varnames <- rownames(imp) # row names to column
-rownames(imp) <- NULL
-
-# keep the n features with the highest MeanDecreaseAccuracy values 
-imp_top30_mda <- arrange(imp, desc(MeanDecreaseAccuracy)) %>% 
-  select(c(MeanDecreaseAccuracy, varnames)) %>% 
-  slice_head(n = 40)
-
-ggplot(imp_top30_mda, aes(x=reorder(varnames, MeanDecreaseAccuracy), 
-                y=MeanDecreaseAccuracy
-                #,color=MeanDecreaseAccuracy
-                )) + 
-  geom_point() +
-  geom_segment(aes(x=varnames,xend=varnames,y=0,yend=MeanDecreaseAccuracy)) +
-  #scale_color_continuous(type = "viridis") +
-  ylab("MDA") +
-  xlab("Variable Name") +
-  ggtitle("RandomForest variable importance: Mean Decrease Accuracy (MDA)") +
-  labs(color = "MDA") + 
-  coord_flip() + 
-  theme_minimal()
-```
 

### 2020-11-05 analysis/paper/paper.Rmd
combined var imp plots into one figure
MDA for both species and genus

@@
 
-Variable importance for species classification (Figure \@ref(fig:fig-varImpSpecies))
+Variable importance for species and genus classifiers (Figure \@ref(fig:fig-varImp))
 
@@
 
-```{r fig-varImpSpecies, fig.cap ="Variable importance for species classification", echo=FALSE,  out.width = "450px", message = FALSE, warning=FALSE}
+```{r fig-varImp, fig.cap ="Variable importance", echo=FALSE,  out.width = "450px", message = FALSE, warning=FALSE}
 library(here)
-knitr::include_graphics(here::here("analysis","figures", "varImp_species.pdf"))
+knitr::include_graphics(here::here("analysis","figures", "varImp.pdf"))
 ```

### 2020-11-05 R/classify_species.Rmd
var imp figure formatting

@@
   theme_minimal() + 
-  labs(x = "Wavelength (nm)\n", 
-       y = "Variable importance \n",
-       title = "Variable Importance: Mean Decrease Accuracy (MDA) \n\nSpecies classification",
+  labs(x = "",  # x = "Wavelength (nm)\n", # remove x axis label for the combined plot
+       y = "Species\nVariable importance\n",
+       title = "Variable Importance: Mean Decrease Accuracy (MDA)", # \n\nSpecies classification",
        fill = "MDA")  + 
@@
    theme(axis.text = element_text(size = 14),
-        axis.title = element_text(size = 14),
+        axis.title = element_text(size = 16),
         title = element_text(size = 16))
@@
 varImp_Gini_species
-```
-
-Single figure with both variable importance plots:
-
-```{r varmImp_species}
-library(ggpubr)
-
-ggpubr::ggarrange(varImp_MDA_species, varImp_Gini_species, nrow = 2)
-
-# save confusion matrix to file 
-#ggsave(filename = here::here("analysis","figures", "varImp_species.pdf"), width = 10, height = 10)
 ```
@@
   theme_minimal() + 
-  labs(x = "Wavelength (nm)\n", 
-       y = "Variable importance\n",
-       title = "Genus classification",
+  labs(x = "\nWavelength (nm)", 
+       y = "Genus\nVariable importance\n",
+       #title = "Genus classification",
        fill = "MDA")  + 

### 2020-11-05 analysis/paper/paper.Rmd
figure legend text

@@
 
-**Maybe in appendix, make a table or something to show species info for which ones were used after cleaning/filtering**
-
 ## Species and genus classification {-}
@@
 We used the optimal values of 27 for mtry and 1000 for ntree in our final RF model.
-There were 49 species classes and 
+There were 49 species classes and 27 genus classes for each classification. 
 We computed overall accuracy, confusion matrices, and variable importance plots.
@@
 
-Classification accuracy was higher at the genus level (60%) compared to the species level (55%)
+Classification accuracy was higher at the genus level (60%) compared to the species level (55%).
 
-Confusion matrix for species classification (Figure \@ref(fig:fig-confMatSpecies))
+Confusion matrix for species classification (Figure \@ref(fig:fig-confMatSpecies)).
 
-Confusion matrix for genus classification (Figure \@ref(fig:fig-confMatGenus))
+Confusion matrix for genus classification (Figure \@ref(fig:fig-confMatGenus)).
 
-Variable importance for species and genus classifiers (Figure \@ref(fig:fig-varImp))
-
-Figure - multi-year observations of individual plants?
-
-<!-- Here's some example analysis code: -->
-
-```{r get-data, eval = FALSE}
-# Note the path that we need to use to access our data files when rendering this document
-my_data <- read.csv(here::here('analysis/data/raw_data/my_csv_file.csv'))
-```
-
-```{r demo-inline-code}
-x <- round(pi, 2)
-```
-
-Here is an example of inline code `r x` in the middle of a sentence.
+Variable importance for species and genus classifiers (Figure \@ref(fig:fig-varImp)).
 
@@
 
-<!-- The following line inserts a page break  -->
-
 \clearpage
@@
 
-Histogram showing which scientific names have at least 20 samples to be included in the classification analysis. Note that the "Larix sp." invididuals are classified at the genus level.
+Histogram showing which scientific names have at least 20 samples to be included in the classification analysis. Note that the "Larix sp." individuals are classified at the genus level.
+
+## Figure 3 {-}
+
+Visualizing the hyperspectral reflectance spectra for each of the 50 genus and species classes utilized in our study.
+
+## Figure 4 {-}
+
+Confusion matrix for the species classifier.
+True and predicted species labels are shown here for each sample in the validation set.
+Counts along the diagonal represent correct species predictions.
+Counts outside of the diagonal represent incorrect species predictions.
+Each cell in the matrix is colored based on the number of counts to ease interpretation.
+Low counts are white and high counts are blue, with a gradient in between.
+
+## Figure 5 {-}
+
+Confusion matrix for the genus classifier.
+True and predicted genus labels are shown here for each sample in the validation set.
+Counts along the diagonal represent correct genus predictions.
+Counts outside of the diagonal represent incorrect genus predictions.
+Each cell in the matrix is colored based on the number of counts to ease interpretation.
+Low counts are white and high counts are orange, with a gradient in between.
+
+## Figure 6 {-}
+
+Variable importance plots for the species and genus classification.
+We used hyperspectral reflectance band values as descriptive features for each classifier.
+The Mean Decrease in Accuracy (MDA) importance metrics is plotted for each hyperspectral band.
+Each importance bar is colored based on importance, with low importance as white and follows a gradient to high importance as blue (for species variable importance) or orange (for genus variable importance).
 

### 2020-11-06 analysis/paper/paper.Rmd
added ref for NIS wavelengths

@@
 # Introduction {-}
-
 1
@@
 Each series of wavelengths increments by 5nm from each starting wavelength to yield 426 spectral bands. 
+These starting wavelengths vary depending on which AOP payload was used to collect the hyperspectral data.
+The wavelength values also change slightly each year when the NEON AOP re-calibrates their payload sensors [@kampe2010advances]. 
 Most of them started at 384nm, so we linearly interpolated each spectrum to have the same 426 bands starting at 384nm and ending at 2512nm with 5nm increments.
 Spectra from the Pu'u Maka'ala Natural Area Reserve (PUUM) NEON site in Hawaii had a starting wavelength of 347nm.
-This is because NEON sub-contracted the PUUM flights to the Global Airborne Observatory, which has a spectrometer with a different wavelength range compared to the NIS payloads *([cite?](https://www.neonscience.org/observatory/observatory-blog/remote-sensing-partnership-made-paradise))*.
+This is because NEON sub-contracted the PUUM flights to the Global Airborne Observatory, which has a spectrometer with a different wavelength range compared to the NIS payloads *([how to cite?](https://www.neonscience.org/observatory/observatory-blog/remote-sensing-partnership-made-paradise))*.
 We filtered out any spectra with a starting wavelength of 347nm, since the interpolation would require too much unavailable data outside the available range.
@@
 
-Visualizing the hyperspectral reflectance spectra for each of the 50 genus and species classes utilized in our study.
+Visualizing the extracted hyperspectral reflectance spectra for each of the 50 genus and species classes that we used in our classifications.
 
@@
 Low counts are white and high counts are blue, with a gradient in between.
+Counts of zero are shown with a gray font color, to make non-zero counts (shown with a black font color) easier to identify.
 
@@
 Low counts are white and high counts are orange, with a gradient in between.
+Counts of zero are shown with a gray font color, to make non-zero counts (shown with a black font color) easier to identify.
 
@@
 We used hyperspectral reflectance band values as descriptive features for each classifier.
-The Mean Decrease in Accuracy (MDA) importance metrics is plotted for each hyperspectral band.
+The Mean Decrease in Accuracy (MDA) importance metric value is plotted for each hyperspectral band.
 Each importance bar is colored based on importance, with low importance as white and follows a gradient to high importance as blue (for species variable importance) or orange (for genus variable importance).

### 2020-11-06 analysis/paper/paper.Rmd
starting a table with species and taxon info and counts

@@
 )
+library(knitr)
 ```
@@
 \end{abstract}
-
-```{r, results='hide', message=FALSE, include=FALSE}
-# Extract hyperspectral reflectance data for each individual plant
-#R/spectral-lib.Rmd
-# Run the main analysis
-```
 
@@
 
-```{r, message = FALSE, echo = FALSE}
-library(knitr)
-
+```{r wavelengthTable, message = FALSE, echo = FALSE}
 df <- data.frame(wavelength = c(347, 381, 382, 384),
@@
 We randomly assigned each spectrum to the training or validation sets, using an 80/20 split. All 50 scientific name labels were represented in each set.
+
+```{r taxonTable, message = FALSE, echo = FALSE}
+df <- data.frame(taxonCode = c("ABBA", "ABLAL", "ACRU, ACRUR", "ACSAS, ACSA3", 
+                               "AMLA", "ARTR2", "BELE", "BENE4"),
+                 scientificName = c("Abies balsamea", 
+                                    "Abies lasiocarpa",
+                                    "Acer rubrum",
+                                    "Acer saccharum", 
+                                    "Amelanchier laevis",
+                                    "Artemisia tridentata",
+                                    "Betula lenta",
+                                    "Betula neoalaskana"),
+                 commonName = c("balsam fir", "subalpine fir", "red maple", "sugar maple",
+                                "Allegheny serviceberry", "sagebrush", "sweet birch", "resin birch"),
+                 count = c(47,117, 489, 79, 37, 494, 28, 35))
+
+knitr::kable(x = df, 
+             # set column names
+             col.names = c("Taxon code",
+                           "Scientific name",
+                           "Common name",
+                           "Number of spectra"),
+             # center align text in the columns
+             align = c("c", "c", "c", "c"),
+             format = "simple")
+```
 

### 2020-11-06 analysis/paper/paper.Rmd
working on table with taxon info

--- 2020-11-06 analysis/paper/paper.Rmd
@@
 We randomly assigned each spectrum to the training or validation sets, using an 80/20 split. All 50 scientific name labels were represented in each set.
+
+\clearpage
+
+```{r taxonTable, message = FALSE, echo = FALSE}
+df <- data.frame(taxonCode = c("ABBA", "ABLAL", "ACRU, ACRUR", "ACSAS, ACSA3", 
+                               "AMLA", "ARTR2", "BELE", "BENE4", "CACA18", "CAOV2",
+                               "CATO6", "CELA", "CODR", "FRAM2", "FRPE"),
+                 scientificName = c("*Abies balsamea*", "*Abies lasiocarpa*", "*Acer rubrum*",
+                                    "*Acer saccharum*", "*Amelanchier laevis*", "*Artemisia tridentata*",
+                                    "*Betula lenta*", "*Betula neoalaskana*", "*Carpinus caroliniana*",
+                                    "*Carya ovata*", "*Carya tomentosa*", "*Celtis laevigata*", 
+                                    "*Cornus drummondii*", "*Fraxinus americana*", "*Fraxinus pennsylvanica*"),
+                 commonName = c("balsam fir", "subalpine fir", "red maple", "sugar maple",
+                                "Allegheny serviceberry", "sagebrush", "sweet birch", "resin birch",
+                                "American hornbeam", "shagbark hickory", "mockernut hickory",
+                                "sugarberry", "roughleaf dogwood", "white ash", "green ash"),
+                 count = c(47,117, 489, 79, 37, 494, 28, 35, 170, 46, 134, 179, 31, 20, 92))
+
+knitr::kable(x = df, 
+             # set column names
+             col.names = c("Taxon code",
+                           "Scientific name *(Genus species)*",
+                           "Common name",
+                           "Number of spectra"),
+             # center align text in the columns
+             align = c("l", "l", "l", "l"),
+             format = "simple")
+```
 

### 2020-11-06 R/clean_spectral_data.Rmd
results figures
hist with taxonID instead of genusSpecies

@@
 ```
+Histogram using scientific name
 
-```{r fig.width=9, fig.height=12}
+```{r fig.width=9, fig.height=12, eval=FALSE, include=FALSE}
 # Histogram with just the taxons that have enough samples 
@@
 # table with species info and counts 
-
-library(taxize)
-
 taxon_info <- all_spectra %>% 
@@
   left_join(taxon_info, by="genusSpecies")
+```
 
-# for(i in 1:nrow(taxons_GT)){
-#   print(i)
-#   taxize::sci2comm(sci = taxons_GT[i]$genusSpecies)
-# }
+Histogram using taxon code?
 
+```{r fig.width=9, fig.height=12}
+# Histogram with just the taxons that have enough samples 
+samplesCountGT <- taxonCounts %>% 
+  filter(countGT == TRUE)
+
+ggplot(distinct(taxons_GT, genusSpecies, .keep_all=TRUE), 
+       aes(y = reorder(taxonID, n), x = n)) + 
+  geom_bar(stat = "identity", fill = "#7E9E8F") + 
+  labs(y = "Taxon code\n",
+       x = "\nNumber of spectra") + 
+  #theme(axis.text = element_text(size = 12)) +
+  theme_bw(base_size=16)
+
+# save figure to file
+ggsave(filename = here::here("analysis","figures", "hist_samples_per_species.pdf"), width = 9, height = 10)
 ```
@@
 # save figure to file
-ggsave(filename = here::here("analysis","figures", "spectra_per_species.pdf"), width = 10, height = 11)
+ggsave(filename = here::here("analysis","figures", "spectra_per_species.pdf"), width = 10, height = 10)
 ```

### 2020-11-07 R/clean_spectral_data.Rmd
species info table

@@
   #theme(axis.text = element_text(size = 12)) +
-  theme_bw(base_size=16)
+  theme_bw(base_size=14)
 
 # save figure to file
-ggsave(filename = here::here("analysis","figures", "hist_samples_per_species.pdf"), width = 9, height = 10)
+ggsave(filename = here::here("analysis","figures", "hist_samples_per_species.pdf"), width = 8, height = 9)
 ```

### 2020-11-07 analysis/paper/paper.Rmd
species info table

--- 2020-11-06 analysis/paper/paper.Rmd
@@
 We randomly assigned each spectrum to the training or validation sets, using an 80/20 split. All 50 scientific name labels were represented in each set.
+
+\clearpage
+
+```{r taxonTable, message = FALSE, echo = FALSE}
+library(knitr)
+library(dplyr)
+
+df <- data.frame(taxonCode = c("ABBA", "ABLAL", "ACRU, ACRUR", "ACSAS, ACSA3", 
+                               "AMLA", "ARTR2", "BELE", "BENE4", "CACA18", "CAOV2",
+                               "CATO6", "CELA", "CODR", "FRAM2", "FRPE", "JUNI", 
+                               "JUVIV, JUVI", "LARIX", "LATR2", "LIST2", "LITU",
+                               "LOMA6", "MORU2", "NYBI", "NYSY", "PIEN", "PIMA",
+                               "PICOL, PICO", "PIEC2", "PIFL2"),
+                 scientificName = c("*Abies balsamea*", "*Abies lasiocarpa*", "*Acer rubrum*",
+                                    "*Acer saccharum*", "*Amelanchier laevis*", "*Artemisia tridentata*",
+                                    "*Betula lenta*", "*Betula neoalaskana*", "*Carpinus caroliniana*",
+                                    "*Carya ovata*", "*Carya tomentosa*", "*Celtis laevigata*", 
+                                    "*Cornus drummondii*", "*Fraxinus americana*", 
+                                    "*Fraxinus pennsylvanica*", "*Juglans nigra*", 
+                                    "*Juniperus virginiana*", "*Larix sp.* *",
+                                    "*Larrea tridentata*", "*Liquidambar styraciflua*", 
+                                    "*Liriodendron tulipifera*", "*Lonicera maackii*",
+                                    "*Morus rubra*", "*Nyssa biflora*", "*Nyssa sylvatica*",
+                                    "*Picea engelmannii*", "*Picea mariana*", "*Pinus contorta*",
+                                    "*Pinus echinata*", "*Pinus flexilis*"),
+                 commonName = c("balsam fir", "subalpine fir", "red maple", "sugar maple",
+                                "Allegheny serviceberry", "sagebrush", "sweet birch", "resin birch",
+                                "American hornbeam", "shagbark hickory", "mockernut hickory",
+                                "sugarberry", "roughleaf dogwood", "white ash", "green ash",
+                                "black walnut", "eastern redcedar", "larch", "creosote bush",
+                                "sweetgum", "tuliptree", "Amur honeysuckle", "red mulberry",
+                                "swamp tupelo", "blackgum", "Engelmann spruce", "black spruce",
+                                "lodgepole pine", "shortleaf pine", "limber pine"),
+                 count = c(47,117, 489, 79, 37, 494, 28, 35, 170, 46, 134, 179, 31, 20, 92, 42, 
+                           28, 23, 161, 209, 40, 23, 30, 24, 77, 24, 32, 302, 72, 78))
+
+knitr::kable(x = df, 
+             # set column names
+             col.names = c("Taxon code",
+                           "Scientific name (*species*)",
+                           "Common name",
+                           "Number of spectra"),
+             # center align text in the columns
+             align = c("l", "l", "l", "l"),
+             format = "simple") 
+  # show which rows are grouped together for genus classes visibly in table?
+
+```
 

### 2020-11-08 analysis/paper/paper.Rmd
species info table
added all rows

--- 2020-11-06 analysis/paper/paper.Rmd
@@
 )
-library(knitr)
 ```
@@
 We filtered the plants by taxonomic rank to retain only those identified at the genus, species, and variety ranks. 
-We removed varieties from scientific names to label each individual spectrum with its respective genus and species (when available). 
+We removed varieties from scientific names to label each individual spectrum with its respective genus and species (when available), represented by the first 2 words in the scientific name. 
 There were 74 unique genus values, 113 unique species values, and 151 unique genus-species combinations.
@@
 
+\clearpage
+
 ```{r taxonTable, message = FALSE, echo = FALSE}
+library(knitr)
+library(dplyr)
+
 df <- data.frame(taxonCode = c("ABBA", "ABLAL", "ACRU, ACRUR", "ACSAS, ACSA3", 
-                               "AMLA", "ARTR2", "BELE", "BENE4"),
-                 scientificName = c("Abies balsamea", 
-                                    "Abies lasiocarpa",
-                                    "Acer rubrum",
-                                    "Acer saccharum", 
-                                    "Amelanchier laevis",
-                                    "Artemisia tridentata",
-                                    "Betula lenta",
-                                    "Betula neoalaskana"),
+                               "AMLA", "ARTR2", "BELE", "BENE4", "CACA18", "CAOV2",
+                               "CATO6", "CELA", "CODR", "FRAM2", "FRPE", "JUNI", 
+                               "JUVIV, JUVI", "LARIX", "LATR2", "LIST2", "LITU",
+                               "LOMA6", "MORU2", "NYBI", "NYSY", "PIEN", "PIMA",
+                               "PICOL, PICO", "PIEC2", "PIFL2", "PIPA2", "PITA",
+                               "POGR4", "POTR5", "PSMEM, PSME", "QUAL", "QUCO2",
+                               "QUHE2", "QULA2", "QUMA13", "QUMA3", "QUNI", "QUPA5",
+                               "QURU", "QUST", "RHDA", "TSCA", "TSHE", "ULAM", "ULCR"),
+                 scientificName = c("*Abies balsamea*", "*Abies lasiocarpa*", "*Acer rubrum*",
+                                    "*Acer saccharum*", "*Amelanchier laevis*", "*Artemisia tridentata*",
+                                    "*Betula lenta*", "*Betula neoalaskana*", "*Carpinus caroliniana*",
+                                    "*Carya ovata*", "*Carya tomentosa*", "*Celtis laevigata*", 
+                                    "*Cornus drummondii*", "*Fraxinus americana*", 
+                                    "*Fraxinus pennsylvanica*", "*Juglans nigra*", 
+                                    "*Juniperus virginiana*", "*Larix sp.* *",
+                                    "*Larrea tridentata*", "*Liquidambar styraciflua*", 
+                                    "*Liriodendron tulipifera*", "*Lonicera maackii*",
+                                    "*Morus rubra*", "*Nyssa biflora*", "*Nyssa sylvatica*",
+                                    "*Picea engelmannii*", "*Picea mariana*", "*Pinus contorta*",
+                                    "*Pinus echinata*", "*Pinus flexilis*","*Pinus palustris*", 
+                                    "*Pinus taeda*", "*Populus grandidentata*", "*Populus tremuloides*",
+                                    "*Pseudotsuga menziesii*", "*Quercus alba*", "*Quercus coccinea*",
+                                    "*Quercus hemisphaerica*", "*Quercus laevis*", "*Quercus margaretta*",
+                                    "*Quercus marilandica*", "*Quercus nigra*", "*Quercus pagoda*",
+                                    "*Quercus rubra*", "*Quercus stellata*", "*Rhamnus davurica*",
+                                    "*Tsuga canadensis*", "*Tsuga heterophylla*", "*Ulmus americana*", 
+                                    "*Ulmus crassifolia*"),
                  commonName = c("balsam fir", "subalpine fir", "red maple", "sugar maple",
-                                "Allegheny serviceberry", "sagebrush", "sweet birch", "resin birch"),
-                 count = c(47,117, 489, 79, 37, 494, 28, 35))
+                                "Allegheny serviceberry", "sagebrush", "sweet birch", "resin birch",
+                                "American hornbeam", "shagbark hickory", "mockernut hickory",
+                                "sugarberry", "roughleaf dogwood", "white ash", "green ash",
+                                "black walnut", "eastern redcedar", "larch", "creosote bush",
+                                "sweetgum", "tuliptree", "Amur honeysuckle", "red mulberry",
+                                "swamp tupelo", "blackgum", "Engelmann spruce", "black spruce",
+                                "lodgepole pine", "shortleaf pine", "limber pine", "longleaf pine",
+                                "loblolly pine", "bigtooth aspen", "quaking aspen", "Douglas-fir",
+                                "white oak", "scarlet oak", "Darlington oak", "turkey oak",
+                                "sand post oak", "blackjack oak", "water oak", "cherrybark oak",
+                                "northern red oak", "post oak", "Dahurian buckthorn", 
+                                "eastern hemlock", "western hemlock", "American elm", "cedar elm"),
+                 count = c(47,117, 489, 79, 37, 494, 28, 35, 170, 46, 134, 179, 31, 20, 92, 42, 
+                           28, 23, 161, 209, 40, 23, 30, 24, 77, 24, 32, 302, 72, 78, 374, 87,
+                           54, 148, 88, 72, 29, 69, 132, 68, 53, 60, 30, 148, 151, 224, 161, 59,
+                           70, 26))
 
@@
              col.names = c("Taxon code",
-                           "Scientific name",
+                           "Scientific name (*Genus species*)",
                            "Common name",
@@
              # center align text in the columns
-             align = c("c", "c", "c", "c"),
+             align = c("l", "l", "l", "l"),
              format = "simple") 
+  # show which rows are grouped together for genus classes visibly in table?
+
 ```

### 2020-11-09 analysis/paper/paper.Rmd
added percent to species table
not sure if its helpful but might be

@@
 
+df$count_percent <- round((df$count / sum(df$count)) * 100, digits = 1)
+
 knitr::kable(x = df, 
@@
                            "Common name",
-                           "Number of spectra"),
+                           "Number of spectra",
+                           "Percent (%) of total"),
              # center align text in the columns
-             align = c("l", "l", "l", "l"),
+             align = c("l", "l", "l", "c", "c"),
              format = "simple") 
@@
 
-We conclude that HS data allow you to get __ accuracy
+We conclude that NEON hyperspectral data allow you to achieve 55% or 60% accuracy when classifying species or genus classes.
+
 Hopes for outcomes
@@
 
-Map to illustrate the NEON sites and domains with data that we used in this study.
+Map to illustrate where NEON data were collected that we used in this study.
+Black points represent NEON site locations where *in-situ* plant observations were collected along with airborne remote sensing flights.
+NEON domains containing each respective site are
 

### 2020-11-09 R/clean_spectral_data.Rmd
changed color of hist and spectra plot
and adjusted the histogram size to make it fit on one page for the figures section

@@
   theme_bw() + 
-  scale_fill_manual(values = c("#d95f0e", "#7bccc4"))
+  scale_fill_manual(values = c("#36454f", "#7E9E8F"))
 ```
@@
 # save figure to file
-ggsave(filename = here::here("analysis","figures", "hist_samples_per_species.pdf"), width = 9, height = 9)
+#ggsave(filename = here::here("analysis","figures", "hist_samples_per_species.pdf"), width = 9, height = 9)
 ```
@@
        aes(y = reorder(taxonID, n), x = n)) + 
-  geom_bar(stat = "identity", fill = "#7E9E8F") + 
+  geom_bar(stat = "identity", fill = "#36454f") + 
   labs(y = "Taxon code\n",
@@
 # save figure to file
-ggsave(filename = here::here("analysis","figures", "hist_samples_per_species.pdf"), width = 8, height = 9)
+ggsave(filename = here::here("analysis","figures", "hist_samples_per_species.pdf"), width = 9, height = 9)
 ```
@@
 
-Visualize the spectra, facet by scientific name, color by taxon rank. 
-```{r, eval = FALSE, include = "FALSE", warning=FALSE, fig.height=12, fig.width=16}
-filtered_spectra %>%
-  ggplot(aes(wavelength_nm, ifelse(mask, NA, reflectance), group = spectraID,
-             # color each facet plot based on taxonRank
-             color = taxonRank)) +
-  geom_path(alpha = .2) +
-  labs(x = '\n Wavelength (nm)',
-       y = 'Reflectance \n',
-       color = "Taxon Rank") + 
-  facet_wrap(~genusSpecies
-             # wrap long scientific names
-             ,labeller = label_wrap_gen(20)
-             ,ncol=10) + 
-  scale_color_manual(values = c("#e6550d","#1c9099", "#a8ddb5")) + 
-  theme_minimal() + 
-  # Adjust the text size and rotation angle 
-  theme(text = element_text(size=11),
-        axis.text.x = element_text(angle=90, hjust=1),
-        axis.title.y = element_text(size = 14),
-        axis.title.x = element_text(size = 14),
-        legend.text = element_text(size = 14),
-        legend.title = element_text(size = 14))
-```
-
 Color all the spectra with a single color
@@
   ggplot(aes(wavelength_nm, ifelse(mask, NA, reflectance), group = spectraID)) +
-  geom_path(alpha = .2, color = "#1c9099") +
+  geom_path(alpha = .2, color = "#36454f") +
   labs(x = '\n Wavelength (nm)',
@@
 ```
-
-Color the spectra by the year they were collected?
-
-```{r, eval=FALSE, include=FALSE}
-filtered_spectra %>%
-  ggplot(aes(wavelength_nm, ifelse(mask, NA, reflectance), group = spectraID, color = as.factor(year))) +
-  geom_path(alpha = .2) +
-  labs(x = '\n Wavelength (nm)',
-       y = 'Reflectance \n',
-       color = "Year collected") + 
-  facet_wrap(~genusSpecies
-             # wrap long scientific names
-             ,labeller = label_wrap_gen(20)
-             ,ncol=10) + 
-  scale_color_manual(values = c("#e6550d","#1c9099", "#a8ddb5")) + 
-  theme_minimal() + 
-  # Adjust the text size and rotation angle 
-  theme(text = element_text(size=11),
-        axis.text.x = element_text(angle=90, hjust=1),
-        axis.title.y = element_text(size = 14),
-        axis.title.x = element_text(size = 14),
-        legend.text = element_text(size = 14),
-        legend.title = element_text(size = 14))
-```

### 2020-11-10 analysis/paper/paper.Rmd
working on intro text

@@
 # Introduction {-}
-1
-* motivation for work, why bother?
-* unprecedented access to unique hyperspectral airborne imagery from NEON (cite)
-  - acknowledge additional data to add info (lidar, rgb) (cite) 
-  - how meaningful can the HS data alone be to predict species? accuracies? 
-  - we expect there to be meaningful information in the HS data for species/genus
-2 - knowledge gaps in lit/field?
-  - how useful are HS data across many species, ecosystems, geographies? 
-  - where is the most important information contained? which wavelengths most important for species discrimination? differ across sites? (cite) 
-  - generally more site-specific studies? 
-  - prediction of species vs genus? How much easier is genus? Or is this difficult too? (cite hierarchical taxonomic classification recent study)
-3 - how we addressed the knowledge gaps
-  - the need to generalize methods across locations? --> NEON
-  - which wavelengths are most important
-  - data harmonization across NEON sites (why are the wavelengths different? Ask NEON person?)
+
+Mapping forests is crucial to understanding the effects and feedbacks of changes in climate, land use, and disturbance events across diverse landscapes [@saladin2020rapid; @grossiord2020having; @fassnacht2016review].
+An important aspect of forest mapping efforts is species composition, which describes the amount and distribution of plant species across landscapes. 
+Species composition has been shown to be a fundamental metric for evaluating biodiversity [@williams2020remote], identifying habitats [@nagendra2013remote], providing essential inputs to many ecological models [@pecchi2019species], and informing conservation and restoration efforts [@camarretta2020monitoring].
+Remote sensing technologies enable us to estimate species composition more effectively than using field-based approaches alone. 
+We currently have unprecedented access to publicly available hyperspectral airborne imagery at sites around the United States from the National Ecological Observatory Network (NEON) [@keller2008continental].
+The NEON Airborne Observation Platform (AOP) payload consists of an imaging spectrometer, high resolution digital camera, and light detection and ranging (LiDAR) sensors to capture both spectral and structural measurements of ecosystems [@kampe2010advances].
 
@@
 Years that we found when both *in-situ* and AOP data were collected were 2017, 2018, and 2019. 
-Note that covid-19 postponed NEON data collection as well as the AOP flight season during 2020. 
+Note that covid-19 **(cite?)** postponed NEON data collection as well as the AOP flight season during 2020. 
 We assumed that plant locations did not change over time.
@@
 
-We trained a random forest (RF) classifier to predict species based on NEON hyperspectral reflectance data.
+We trained a random forest (RF) classifier to predict species based on NEON hyperspectral reflectance data as the descriptive features.
 RF is a machine learning approach that utilizes an ensemble of decision trees and iterative subsets of the training data.
-We performed hyperparameter optimization of ntree, the number of decision trees in the forest, and mtry, the number of variables randomly sampled at each node in the tree.
-We used the optimal values of 27 for mtry and 1000 for ntree in our final RF model.
+We performed hyperparameter optimization of *ntree*, the number of decision trees in the forest, and *mtry*, the number of variables randomly sampled at each node in the tree.
+We used the optimal values of 27 for *mtry* and 1000 for *ntree* in our final RF model.
 There were 49 species classes and 27 genus classes for each classification. 
@@
 
-To evaluate genus classification accuracy, we followed the same step as described for species classiciation, but we used each spectrum's associated genus as each class label. 
+To evaluate genus classification accuracy, we followed the same step as described for species classification, but we used each spectrum's associated genus as each class label. 
+We used the randomForest R package [@randomForest] to implement the analysis. 
 We provide all of our code in a research compendium here:[https://github.com/vscholl/neonHScompendium](https://github.com/vscholl/neonHScompendium).
@@
 
-Classification accuracy was higher at the genus level (60%) compared to the species level (55%).
+Classification accuracy was higher at the genus level (60%) compared to the species level (55%) based on both the internal OOB error of the RF model and the independent validation set.
 

### 2020-11-10 analysis/paper/paper.Rmd
working on intro text

@@
 Species composition has been shown to be a fundamental metric for evaluating biodiversity [@williams2020remote], identifying habitats [@nagendra2013remote], providing essential inputs to many ecological models [@pecchi2019species], and informing conservation and restoration efforts [@camarretta2020monitoring].
-Remote sensing technologies enable us to estimate species composition more effectively than using field-based approaches alone. 
+Although remote sensing technologies enable us to estimate species composition more effectively than using field-based approaches alone, species classification is still a challenging task.
+
 We currently have unprecedented access to publicly available hyperspectral airborne imagery at sites around the United States from the National Ecological Observatory Network (NEON) [@keller2008continental].
-The NEON Airborne Observation Platform (AOP) payload consists of an imaging spectrometer, high resolution digital camera, and light detection and ranging (LiDAR) sensors to capture both spectral and structural measurements of ecosystems [@kampe2010advances].
+NEON-monitored sites are spread across ecoclimatic domains with a diverse array of ecosystems and land cover types including deciduous and evergreen forests, shrub/scrub and woody wetlands, mixed forest and grasslands.
+Field technicians collect field-based vegetation measurements including species observations on a semi-annual basis. 
+In addition, NEON’s Airborne Observation Platform (AOP) payload carries a suite of sensors to collect both spectral and structural measurements of ecosystems [@kampe2010advances]. 
+The AOP includes an imaging spectrometer which captures hyperspectral reflectance data in over 400 bands across the visible and near infrared wavelengths. 
+Hyperspectral imagery is frequently used to classify species with accuracies ranging widely anywhere from 45% to 98% [@fassnacht2016review]. 
+Hyperspectral data are capable of capturing reflectance variations and features with hundreds of spectral bands better than multispectral remotely sensed data with a handful to tens of bands [@awad2018forest].
+Many factors influence tree species classification results including the number of species being discerned, the homogeneity of species, the density and structural complexity of the forest, and the taxonomic level of classification [@sothe2019tree].
+Methods are often performed on small scales or in simple ecosystems, and are not successful at scaling to larger areas or different forest types **(cite)**.
+Although the demand for species mapping is justified in many cases, there is also potential for both using and finding value in classifying individuals at other taxonomic levels, as explored in a recent bioacoustics study [@cramer2020chirping].
+In the plant realm, identifying taxons at the genus level had been shown to be easier than at the species level [@tuominen2018assessment].
+[@surovy2019acquisition] report that species are more successfully classified into broader categories such as spruce, pine, and broadleaf, compared to identifying individual species in mixed arrangements.
 
-The National Ecological Observatory Network (NEON) is a valuable source of publicly available ecological data across the United States [@keller2008continental].
+NEON provides a unique opportunity to test, develop, and tune tree classification methods over a continental scale and at multiple taxonomic resolutions. 
 We have extracted all NEON AOP hyperspectral data for every mapped stem in NEON plots where field data and airborne remote sensing data were collected during the same year.
-
 We used a machine learning approach to evaluate plant identification potential using the hyperspectral data at multiple taxonomic resolutions: species and genus levels. 
-
-Species classification is an active research area [@scholl2020integrating].
+Our study contributes to the active field of tree species classification research using freely available NEON data and open, reproducible code [@marconi2019data; @scholl2020integrating]
 

### 2020-11-10 analysis/paper/paper.Rmd
working on intro text

@@
 In the plant realm, identifying taxons at the genus level had been shown to be easier than at the species level [@tuominen2018assessment].
-[@surovy2019acquisition] report that species are more successfully classified into broader categories such as spruce, pine, and broadleaf, compared to identifying individual species in mixed arrangements.
+In their review of over 109 publications, [@surovy2019acquisition] report that species are more successfully classified into broader categories such as spruce, pine, and broadleaf, compared to identifying individual species in mixed arrangements.
 

### 2020-11-10 analysis/paper/paper.Rmd
added some papers to intro text

@@
 Hyperspectral imagery is frequently used to classify species with accuracies ranging widely anywhere from 45% to 98% [@fassnacht2016review]. 
-Hyperspectral data are capable of capturing reflectance variations and features with hundreds of spectral bands better than multispectral remotely sensed data with a handful to tens of bands [@awad2018forest].
+Hyperspectral data are capable of capturing reflectance variations and features with hundreds of spectral bands better than multispectral remotely sensed data with a handful to tens of bands [@awad2018forest; @maschler2018individual].
 Many factors influence tree species classification results including the number of species being discerned, the homogeneity of species, the density and structural complexity of the forest, and the taxonomic level of classification [@sothe2019tree].
@@
 We used a machine learning approach to evaluate plant identification potential using the hyperspectral data at multiple taxonomic resolutions: species and genus levels. 
-Our study contributes to the active field of tree species classification research using freely available NEON data and open, reproducible code [@marconi2019data; @scholl2020integrating]
+Our study contributes to the active field of tree species classification research using freely available NEON data and open, reproducible code [@marconi2019data; @fricker2019convolutional; @scholl2020integrating]
 

### 2020-11-10 analysis/paper/paper.Rmd
intro text

@@
 We have extracted all NEON AOP hyperspectral data for every mapped stem in NEON plots where field data and airborne remote sensing data were collected during the same year.
-We used a machine learning approach to evaluate plant identification potential using the hyperspectral data at multiple taxonomic resolutions: species and genus levels. 
+We used a machine learning approach to evaluate plant identification potential using the hyperspectral data at the species and genus taxonomic levels. 
 Our study contributes to the active field of tree species classification research using freely available NEON data and open, reproducible code [@marconi2019data; @fricker2019convolutional; @scholl2020integrating]
@@
 Spectra from the Pu'u Maka'ala Natural Area Reserve (PUUM) NEON site in Hawaii had a starting wavelength of 347nm.
-This is because NEON sub-contracted the PUUM flights to the Global Airborne Observatory, which has a spectrometer with a different wavelength range compared to the NIS payloads *([how to cite?](https://www.neonscience.org/observatory/observatory-blog/remote-sensing-partnership-made-paradise))*.
+This is because NEON sub-contracted the PUUM flights to the Global Airborne Observatory, which has a spectrometer with a different wavelength range compared to the NIS payloads [@kampe2010advances]. 
 We filtered out any spectra with a starting wavelength of 347nm, since the interpolation would require too much unavailable data outside the available range.
@@
 
-Next, we removed samples belonging to species and genus classes with few samples. 
-We figured that a small sample size would lead to a poor representation in model training. 
+Next, we removed samples belonging to species and genus classes with few samples, since extreme class imbalance is a known problem that can lead to biased classification results [@bader2018biased]. 
 From the initial collection of 151 unique scientific names identified at the genus or species level, 50 of them had at least 20 samples and were kept for the classification analysis (Figure \@ref(fig:fig-histSpecies)).
@@
 We trained a random forest (RF) classifier to predict species based on NEON hyperspectral reflectance data as the descriptive features.
-RF is a machine learning approach that utilizes an ensemble of decision trees and iterative subsets of the training data.
+RF is a machine learning approach that utilizes an ensemble of decision trees and iterative subsets of the training data [@breiman2001random].
 We performed hyperparameter optimization of *ntree*, the number of decision trees in the forest, and *mtry*, the number of variables randomly sampled at each node in the tree.

### 2020-11-11 R/classify_species.Rmd
added vertical shading to variable importance plots
to help interpret which wavelengths were important

@@
                 y=MeanDecreaseAccuracy)) +
+
+  # add vertical shaded regions to indicate notable wavelengths
+  annotate(geom = "rect", xmin = 450, xmax = 510, ymin = 0, 
+           ymax = max(imp$MeanDecreaseAccuracy), fill = "blue", alpha = 0.1) + # BLUE
+  annotate(geom = "rect", xmin = 530, xmax = 590, ymin = 0, 
+           ymax = max(imp$MeanDecreaseAccuracy), fill = "green", alpha = 0.1) + # GREEN
+  annotate(geom = "rect", xmin = 640, xmax = 700, ymin = 0, 
+           ymax = max(imp$MeanDecreaseAccuracy), fill = "red", alpha = 0.1) + # RED
+  annotate(geom = "rect", xmin = 850, xmax = 910, ymin = 0, 
+           ymax = max(imp$MeanDecreaseAccuracy), fill = "grey30", alpha = 0.1)  + # NIR 
+  annotate(geom = "rect", xmin = 1580, xmax = 1640, 
+           ymin = 0, ymax = max(imp$MeanDecreaseAccuracy), fill = "chocolate4", alpha = 0.1) +   # SWIR
+
   # add bars where each height is equal to the variable importance metric
@@
 
-```{r varImp_Gini, fig.height=6, fig.width=10}
+```{r varImp_Gini, fig.height=6, fig.width=10, eval=FALSE, include=FALSE}
 varImp_Gini_species <- ggplot(imp, aes(x = wavelength_nm,  
@@
 The wavelength ranges are based on [Landsat 8 designations](https://www.neonscience.org/hyper-spec-intro).
-
-```{r vertical_dashed_lines}
-# plot variable importance with wavelengths of interest marked by dashed vertical lines
-
-ggplot(imp, aes(x = wavelength_nm,  
-                y = MeanDecreaseGini)) +
-
-  # add vertical dashed lines to indicate notable wavelengths
-  geom_vline(xintercept=480, linetype="dashed", color="blue", size=1, alpha=0.5) + # BLUE 
-  geom_vline(xintercept=550, linetype="dashed", color="green", size=1, alpha=0.5) + # GREEN
-  geom_vline(xintercept=650, linetype="dashed", color="red", size=1, alpha=0.5) + # RED
-  geom_vline(xintercept=870, linetype="dashed", color="grey30", size=1, alpha=0.5) + # NIR 
-  geom_vline(xintercept=1600, linetype="dashed", color="chocolate4", size=1, alpha=0.5) + # SWIR
-
-  # add bars where each height is equal to the variable importance metric
-  geom_col(aes(fill = MeanDecreaseGini), color = NA) +
-  theme_minimal() + 
-  labs(x = "Wavelength (nm)", 
-       y = "Variable importance (MDG)",
-       title = "Variable Importance: Mean Decrease Gini (MDG) vs. Wavelength",
-       fill = "MDG")  + 
-  scale_fill_gradient(low = "#d9d9d9", high = "#43a2ca")
-
-```
-```{r vertical_shaded_regions}
-# plot variable importance with wavelengths of interest marked by dashed vertical lines
-
-ggplot(imp, aes(x = wavelength_nm,  
-                y = MeanDecreaseGini)) +
-
-  # add vertical shaded regions to indicate notable wavelengths
-  annotate(geom = "rect", xmin = 450, xmax = 510, ymin = -5, ymax = Inf, fill = "blue", alpha = 0.1) + # BLUE
-  annotate(geom = "rect", xmin = 530, xmax = 590, ymin = -5, ymax = Inf, fill = "green", alpha = 0.1) + # GREEN
-  annotate(geom = "rect", xmin = 640, xmax = 670, ymin = -5, ymax = Inf, fill = "red", alpha = 0.1) + # RED
-  annotate(geom = "rect", xmin = 850, xmax = 880, ymin = -5, ymax = Inf, fill = "grey30", alpha = 0.1)  + # NIR 
-  annotate(geom = "rect", xmin = 1570, xmax = 1650, ymin = -5, ymax = Inf, fill = "chocolate4", alpha = 0.1)  + # SWIR
-
-  # add bars where each height is equal to the variable importance metric
-  geom_col(aes(fill = MeanDecreaseGini), color = NA) +
-  theme_minimal() + 
-  labs(x = "Wavelength (nm)", 
-       y = "Variable importance (MDG)",
-       title = "Variable Importance: Mean Decrease Gini (MDG) vs. Wavelength",
-       fill = "MDG")  + 
-  scale_fill_gradient(low = "#d9d9d9", high = "#43a2ca")
-
-```
 
@@
                 y=MeanDecreaseAccuracy)) +
+
+  # add vertical shaded regions to indicate notable wavelengths
+  annotate(geom = "rect", xmin = 450, xmax = 510, ymin = 0, 
+           ymax = max(imp_genus$MeanDecreaseAccuracy), fill = "blue", alpha = 0.1) + # BLUE
+  annotate(geom = "rect", xmin = 530, xmax = 590, ymin = 0, 
+           ymax = max(imp_genus$MeanDecreaseAccuracy), fill = "green", alpha = 0.1) + # GREEN
+  annotate(geom = "rect", xmin = 640, xmax = 700, ymin = 0, 
+           ymax = max(imp_genus$MeanDecreaseAccuracy), fill = "red", alpha = 0.1) + # RED
+  annotate(geom = "rect", xmin = 850, xmax = 910, ymin = 0, 
+           ymax = max(imp_genus$MeanDecreaseAccuracy), fill = "grey30", alpha = 0.1)  + # NIR 
+  annotate(geom = "rect", xmin = 1580, xmax = 1640, ymin = 0, 
+           ymax = max(imp_genus$MeanDecreaseAccuracy), fill = "chocolate4", alpha = 0.1) +   # SWIR
+
   # add bars where each height is equal to the variable importance metric
@@
 
-```{r varImp_Gini_genus, fig.height=6, fig.width=10}
+```{r varImp_Gini_genus, fig.height=6, fig.width=10, eval=FALSE, include=FALSE}
 ggplot(imp_genus, aes(x = wavelength_nm,  
@@
 
-```{r varmImp_species}
+```{r varmImp_species_and_genus}
 ggpubr::ggarrange(varImp_MDA_species, varImp_MDA_genus, nrow = 2)

### 2020-11-11 analysis/paper/paper.Rmd
adding text and refs

@@
 
-Mapping forests is crucial to understanding the effects and feedbacks of changes in climate, land use, and disturbance events across diverse landscapes [@saladin2020rapid; @grossiord2020having; @fassnacht2016review].
-An important aspect of forest mapping efforts is species composition, which describes the amount and distribution of plant species across landscapes. 
+Mapping forests is crucial to understanding the effects and feedbacks of changes in climate, land use, invasive species, and disturbance events across diverse landscapes [@saladin2020rapid; @grossiord2020having; @fassnacht2016review; @rocchini2015potential].
+An important aspect of forest mapping efforts is species composition, which describes the amount and distribution of plant species across landscapes [@rocchini2016satellite]. 
 Species composition has been shown to be a fundamental metric for evaluating biodiversity [@williams2020remote], identifying habitats [@nagendra2013remote], providing essential inputs to many ecological models [@pecchi2019species], and informing conservation and restoration efforts [@camarretta2020monitoring].
@@
 We currently have unprecedented access to publicly available hyperspectral airborne imagery at sites around the United States from the National Ecological Observatory Network (NEON) [@keller2008continental].
-NEON-monitored sites are spread across ecoclimatic domains with a diverse array of ecosystems and land cover types including deciduous and evergreen forests, shrub/scrub and woody wetlands, mixed forest and grasslands.
+NEON-monitored sites are spread across ecoclimatic domains with a diverse array of ecosystems and land cover types [@homer2015completion] including deciduous and evergreen forests, shrub/scrub and woody wetlands, mixed forest and grasslands [@schimel2007neon].
 Field technicians collect field-based vegetation measurements including species observations on a semi-annual basis. 
@@
 Hyperspectral imagery is frequently used to classify species with accuracies ranging widely anywhere from 45% to 98% [@fassnacht2016review]. 
-Hyperspectral data are capable of capturing reflectance variations and features with hundreds of spectral bands better than multispectral remotely sensed data with a handful to tens of bands [@awad2018forest; @maschler2018individual].
+Hyperspectral data are more capable of capturing subtle reflectance variations and features using hundreds of narrow spectral bands compared to multispectral remotely sensed data with few to tens of bands [@[@chang2003hyperspectral]; @awad2018forest; @maschler2018individual].
 Many factors influence tree species classification results including the number of species being discerned, the homogeneity of species, the density and structural complexity of the forest, and the taxonomic level of classification [@sothe2019tree].
@@
 
-The NEON project monitors diverse ecosystems at 81 sites across 20 domains, each with similar climate and ecological characteristics, across the United States.
+The NEON project monitors diverse ecosystems at 81 sites across 20 domains, each with similar climate and ecological characteristics, across the United States for a span of 30 years [http://www.neonscience.org](http://www.neonscience.org).
 The NEON data that we used for our study include both *in-situ* plant measurements and airborne remote sensing image data [@neonDataProducts].
-The Woody Plant Vegetation Structure (data product code DP1.10098) includes individual plant locations along with species and genus observations that are collected every 1 to 3 years. 
+The Woody Plant Vegetation Structure (data product code DP1.10098) includes individual plant locations along with species and genus observations that are collected every 1 to 3 years based on NEON's Terrestrial Observation System sampling design [@thorpe2016introduction].
 The NEON Airborne Observation Platform (AOP) collects airborne remote sensing data regularly at NEON sites. 
@@
 From the initial collection of 151 unique scientific names identified at the genus or species level, 50 of them had at least 20 samples and were kept for the classification analysis (Figure \@ref(fig:fig-histSpecies)).
-We randomly assigned each spectrum to the training or validation sets, using an 80/20 split. All 50 scientific name labels were represented in each set.
+We randomly assigned each spectrum to the training or validation sets, using an 80/20 split as per the Pareto Principle [@dunford2014pareto].
+All 50 scientific name labels were represented in each set.
 
@@
 
-After these data cleaning steps, we were left with reflectance data for 5266 individuals at sites across 14 of the NEON eco-climatic domains (Figure \@ref(fig:fig-studyArea)). 
+After these data cleaning steps, we were left with reflectance data in 350 bands (excluding "bad bands") for 5266 individuals at sites across 14 of the NEON eco-climatic domains (Figure \@ref(fig:fig-studyArea)).
 There is a noticeable imbalance of data across space and time; some NEON domains and US states (such as Hawaii) are not represented in our analysis. 
@@
 We trained a random forest (RF) classifier to predict species based on NEON hyperspectral reflectance data as the descriptive features.
-RF is a machine learning approach that utilizes an ensemble of decision trees and iterative subsets of the training data [@breiman2001random].
+RF is a commonly used machine learning approach that utilizes an ensemble of decision trees and iterative subsets of the training data [@breiman2001random; @ballanti2016tree].
 We performed hyperparameter optimization of *ntree*, the number of decision trees in the forest, and *mtry*, the number of variables randomly sampled at each node in the tree.
@@
 There were 49 species classes and 27 genus classes for each classification. 
-We computed overall accuracy, confusion matrices, and variable importance plots.
+We computed overall accuracy based on both the Out of Bag (OOB) evaluation internal to the RF classifier and an independent validation set composed of a random 20% of the available data that was withheld during model training.
+We also computed confusion matrices to evaluate classification performance for each species and genus class.
+Additionally, we reported the Mean Decrease in Accuracy (MDA) variable importance metric to quantify how useful each hyperspectral wavelength band was for classification.
 
@@
 
-We report overall classification accuracy of the species and genus classification:
+We found that classification accuracy was higher at the genus taxonomic level (60%) compared to the species level (55%) based on both the internal OOB error of the RF model and the independent validation set:
 
@@
 
-Classification accuracy was higher at the genus level (60%) compared to the species level (55%) based on both the internal OOB error of the RF model and the independent validation set.
-
-Confusion matrix for species classification (Figure \@ref(fig:fig-confMatSpecies)).
-
-Confusion matrix for genus classification (Figure \@ref(fig:fig-confMatGenus)).
+The classification accuracies varied across species (Figure \@ref(fig:fig-confMatSpecies)) and genus classes (Figure \@ref(fig:fig-confMatGenus)).
 
@@
 
-Restate findings at high level
+**How do our findings compare to other similar studies. Consistent? Differences? (cite IDTReeS) Other ranges of accuracy using NEON HS data? How many sites too?**
 
-How do our findings compare to other similar studies. Consistent? Differences? (cite IDTReeS) Other ranges of accuracy using NEON HS data? How many sites too?
+Our achieved accuracies of 55% and 60% for species and genus classification, respectively, fall within the range of 45% to 98% for hyperspectral-based studies as reported by [@fassnacht2016review].
+Our accuracy values are lower than the 68% - 92% range achieved by submissions to a recent NEON data tree species classification competition [@marconi2019data], which provided participants with remote sensing data  from all three AOP sensors to classify 9 species at a single NEON site.
 
-Potential limitations
-  - class imbalance - we tried to filter based on # samples, but our model has then only seen common species and would be unable to identify rare ones
+**Potential limitations**
+We acknowledge the presence of a class imbalance in both the species and genus training sets. 
+Although we filtered out species and genus classes with less than 20 samples, there is still a range of 20 to almost 500 spectra per taxonomic class. 
+Incorporating additional samples as NEON's database grows over time, or implementing data augmentation techniques [@wong2016understanding] are two ways to reduce this class imbalance for future classification analyses.
+
+e tried to filter based on # samples, but our model has then only seen common species and would be unable to identify rare ones
   - filtering shadows, non-vegetation pixels.... (cite a paper from last IDTReeS comp)
@@
 
-Max has cool ideas and knows many R tricks
+Max has cool ideas and knows many R tricks!
+
+Funding for this work was provided by Earth Lab, through CU Boulder’s Grand Challenge Initiative, and the CIRES at CU Boulder.
 
@@
 Black points represent NEON site locations where *in-situ* plant observations were collected along with airborne remote sensing flights.
-NEON domains containing each respective site are
+Delineated regions depict NEON ecoclimatic domains that contain these sites.
 

### 2020-11-12 analysis/paper/paper.Rmd
paper text

@@
 
-Variable importance for species and genus classifiers (Figure \@ref(fig:fig-varImp)).
+Wavelength variable importance was similar for both the species and genus classifiers (Figure \@ref(fig:fig-varImp)).
 
@@
 
-**Potential limitations**
+**How does the wavelength importance compare to other studies? genus vs species findings?**
+
+**Potential limitations and future improvements**
 We acknowledge the presence of a class imbalance in both the species and genus training sets. 
@@
 
-e tried to filter based on # samples, but our model has then only seen common species and would be unable to identify rare ones
   - filtering shadows, non-vegetation pixels.... (cite a paper from last IDTReeS comp)
-
-Future directions, next steps
-  - targeted species/genus for increased sampling to improve representation in the data set?
    - more advanced models. ML DL
@@
    - hierarchical taxonomic approach - borrow information from closely related species (i.e. common and rare species > use common species observations)
-   - (cite our other paper) intelligent crown geometry processing 
+   - (cite our other paper) intelligent crown geometry processing and classifying crowns instead of just pixels
    - (ben weinstein) tree crown detection - extract multiple pixels per individual using rectangular crown boundaries
    - herbarium / hyperspectral data measurements at other scales. handheld spectrometer. other spectral libraries.
-   - combine HS with RGB, lidar
+   - combine HS with RGB, lidar. this has been shown to improve accracy (cite)
 
@@
 Variable importance plots for the species and genus classification.
-We used hyperspectral reflectance band values as descriptive features for each classifier.
+We used 350 hyperspectral reflectance band values as descriptive features for each classifier.
 The Mean Decrease in Accuracy (MDA) importance metric value is plotted for each hyperspectral band.
 Each importance bar is colored based on importance, with low importance as white and follows a gradient to high importance as blue (for species variable importance) or orange (for genus variable importance).
+Vertical shaded regions indicate approximate wavelengths of blue, green, red, near infrared, and shortwave infrared regions interest for interpretation.
 

### 2021-02-23 R/spectral-lib.Rmd
reinstalling / updating all packages, R, Rstudio, adding comments and working on spectral lib creation for consistency

@@
 library(geoNEON)
-library(sp)
-#remotes::install_github('earthlab/neonhs')
+#install.packages('devtools')
+#devtools::install_github('earthlab/neonhs')
 library(neonhs)
@@
 
-zipsByProduct(dpID = "DP1.10098.001",
+# Download NEON observational data, Woody plant vegetation structure
+neonUtilities::zipsByProduct(dpID = "DP1.10098.001",
               savepath = out_dir,

### 2021-02-25 R/spectral-lib.Rmd
added package name
load sp library

@@
 library(here)
+library(sp)
 ```
@@
 ```{r get-precise-locs, warning=FALSE}
-vegmap <- getLocTOS(vegmap, "vst_mappingandtagging")
+vegmap <- geoNEON::getLocTOS(vegmap, "vst_mappingandtagging")
 

### 2021-02-25 R/clean_spectral_data.Rmd
show initial spectral plot

@@
 
-```{r, warning=FALSE, fig.height=16, fig.width=14, eval=FALSE}
+```{r, warning=FALSE, fig.height=16, fig.width=14, eval=TRUE}
 all_spectra %>%

### 2021-02-25 R/clean_spectral_data.Rmd
print dimensions of spectra since itts cut off from the glimpse command output

@@
 ```{r, message=TRUE}
+# dimensions of the spectra file 
+message("dimensions of all_spectra.csv file: rows   cols")
+dim(all_spectra)
+
 message("# of unique individualID values:")

### 2021-02-25 R/classify_species.Rmd
added boolean variable to train or load classifier

@@
 dir.create("output")
+```
+
+```{r run_or_read_models}
+
+# TRUE or FALSE
+# TRUE = train and tune the classification model 
+# or FALSE =  read existing files, do not train a new model each time.
+run_models <- TRUE
 ```
@@
 
-```{r caret_parallel, eval=FALSE}
+```{r caret_parallel, eval=run_models}
 start.time = Sys.time()
@@
 
-```{r plot_mtry_tuning, eval=FALSE}
+```{r plot_mtry_tuning, eval=run_models}
 # load mtry tuning model
@@
 
-```{r optimize_ntree, eval=FALSE}
+```{r optimize_ntree, eval=run_models}
 # https://rpubs.com/phamdinhkhanh/389752
@@
 
-```{r ntree_tuning_results, eval=FALSE}
+```{r ntree_tuning_results, eval=run_models}
 # load ntree tuning model list
@@
 
-```{r randomForest_species_train, eval=FALSE}
+```{r randomForest_species_train, eval=run_models}
 set.seed(44)
@@
 
-```{r randomForest_train_confMatrix, fig.width=14, fig.height=14, message=FALSE, eval=FALSE}
+```{r randomForest_train_confMatrix, fig.width=14, fig.height=14, message=FALSE, eval=run_models}
 # list all combinations of true and predicted species 
@@
 
-```{r randomForest_train_genus, eval=FALSE}
+```{r randomForest_train_genus, eval=run_models}
 set.seed(44) # for obama

### 2021-03-05 R/clean_spectral_data.Rmd
write a cleaned spectral file before applying filtering steps and labels for classification

--- 2020-11-09 R/clean_spectral_data.Rmd
@@
 
+```{r write_all_cleaned_spectra}
+message("# of unique taxonID values:")
+length(unique(cleaned_spectra$genusSpecies))
+
+message("# of unique scientificName values:")
+length(unique(cleaned_spectra$genus))
+
+message("# of unique spectraID (uid + year) values:")
+length(unique(cleaned_spectra$spectraID))
+
+# save this cleaned data set to a csv file 
+write_csv(x = cleaned_spectra,
+          path = here::here("analysis", "data", "derived_data", "cleaned_spectra.csv"))
+```
+
 ### Partition data into train, evaluation, & test sets
@@
 write_csv(x = filtered_spectra,
-          path = here::here("analysis", "data", "derived_data", "cleaned_spectra.csv"))
+          path = here::here("analysis", "data", "derived_data", "cleaned_filtered_spectra.csv"))
 ```

### 2021-03-05 R/classify_species.Rmd
run script again, using freshly named file that is cleaned and filtered for classification

--- 2020-11-11 R/classify_species.Rmd
@@
 cleaned_data <- readr::read_csv(here::here("analysis", "data", "derived_data", 
-                                              "cleaned_spectra.csv"))
+                                              "cleaned_filtered_spectra.csv"))
 ```

### 2021-03-09 R/add_plant_functional_type.Rmd
working on script to add PFT to spectral lib

---
title: "Add Plant Functional Type labels"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## USDA Plants Database

USDA Plants Database R API tutorial: https://recology.info/2016/10/usda-plants-database-r/

The USDA maintains a database of plant information, some of it trait data, some of it life history. Check it out at http://plants.usda.gov/java/

```{r install_load_packages, message=FALSE}
#install.packages("request")
#library(devtools)
#devtools::install_github("sckott/request")
library('request')
library('tibble')
```

```{r}
# Set the base url
root <- api("https://plantsdb.xyz")

root %>% api_path(heartbeat)
```

^^ My results here do not match those that show up on the tutorial page above. Maybe the API is down or the URL has changed?

## BIEN

RBIEN tutorial: https://cmerow.github.io/RDataScience/3_3_RBIEN_tutorial.html

```{r install_load_packages_bien, message=FALSE}
library(BIEN)
library(ape) # Package for working with phylogenies in R
library(maps) # Useful for making quick maps of occurrences
library(sp) # A package for spatial data
```

Let’s say we’re interested in the species Xanthium strumarium and we’d like some occurrence data.

```{r}
Xanthium_strumarium <- BIEN_occurrence_species(species = "Xanthium strumarium")
#View(Xanthium_strumarium)
head(Xanthium_strumarium)
str(Xanthium_strumarium)
```
The default data that is returned consists of the latitude, longitude and date collected, along with a set of attribution data. If we want more information on these occurrences, we just need to change the arguments:

```{r}
Xanthium_strumarium_full <- BIEN_occurrence_species(species = "Xanthium strumarium",cultivated = T,only.new.world = T,all.taxonomy = T,native.status = T,observation.type = T,political.boundaries = T)

str(Xanthium_strumarium_full)
```

Let’s take a quick look at where those occurrences are.

```{r}
# Make a quick map to plot our points on
map('world',fill=T , col= "grey", bg="light blue")

# Plot the points from the full query in red
points(cbind(Xanthium_strumarium_full$longitude,Xanthium_strumarium_full$latitude),col="red",pch=20,cex=1)

# Plot the points from the default query in blue
points(cbind(Xanthium_strumarium$longitude,Xanthium_strumarium$latitude),col="blue",pch=20,cex=1) 
```

BIEN traits and taxonomic data

Let’s say we’re interested in the genus Abies, and we’d like to get an idea of how many species there are in this genus and what higher taxa it falls within.

*Abies* genus is fir trees (coniferous trees). The "higher_plant_group" returned by the BIEN taxonomy function is "gymnosperms (conifers)"

```{r}
Abies_taxonomy<-BIEN_taxonomy_genus(genus = "Abies")

#View(Abies_taxonomy)
Abies_taxonomy[1:10,]
```

*Acer* genus is maple trees (deciduous trees). The "higher_plant_group" returned by the BIEN taxonomy is "flowering plants".

```{r}
Acer_taxonomy<-BIEN_taxonomy_genus(genus = "Acer")

#View(Acer_taxonomy)
Acer_taxonomy[1:10,]
```
*Acer rubrum* species is the red maple (deciduous trees). The "higher_plant_group" returned by the BIEN taxonomy is "flowering plants".

```{r}
Acer_rubrum_taxonomy <- BIEN_taxonomy_species(species = "Acer rubrum")
Acer_rubrum_taxonomy[1:10,]
```
```{r}
Artemisia_tridentata_taxonomy <- BIEN_taxonomy_species(species = "Artemisia tridentata")
Artemisia_tridentata_taxonomy[1:10,]
```
```{r}
Carpinus_caroliniana_taxonomy <- BIEN_taxonomy_species(species = "Carpinus caroliniana")
Carpinus_caroliniana_taxonomy[1:5,]
```

```{r}
Juniperus_virginiana_taxonomy <- BIEN_taxonomy_species(species = "Juniperus virginiana")
Juniperus_virginiana_taxonomy[1:nrow(Juniperus_virginiana_taxonomy),]
```
```{r}
Quercus_laevis_taxonomy <- BIEN_taxonomy_species(species = "Quercus laevis")
Quercus_laevis_taxonomy[1:nrow(Quercus_laevis_taxonomy),]
```

### 2021-03-09 R/add_plant_functional_type.Rmd
exploring packages for PFT labels

@@
 #devtools::install_github("sckott/request")
-library('request')
-library('tibble')
+library(request)
+library(tibble)
+library(tidyverse)
+library(here)
 ```
@@
 ^^ My results here do not match those that show up on the tutorial page above. Maybe the API is down or the URL has changed?
+
+Based on the GitHub issue discussion (https://github.com/USDA/USDA-APIs/issues/7), someone uploaded USDA plants data here from 2016: https://uofi.app.box.com/v/usdaplants
+
+```{r, eval=FALSE, include=FALSE}
+# Specify URL where file is stored 
+url <- "https://uofi.app.box.com/v/usdaplants/folder/6688524249/usda_plants_20160713.csv"
+destFile <- file.path(here::here(), "analysis","data", "usda_plants_20160713.csv")
+# Apply download.file function in R
+download.file(url, destFile)
+```
+
+```{r}
+#usda_plants <- read_csv(file.path(here::here(), "analysis", "data", "usda_plants_20160713.csv"),
+                        local = locale(encoding = "latin1"))
+
+# Error in nchar(x) : invalid multibyte string, element 2
+# ^ Issue with some of the column names in the .csv? 
+```
 
@@
 
-Let’s say we’re interested in the species Xanthium strumarium and we’d like some occurrence data.
+"Let’s say we’re interested in the species Xanthium strumarium and we’d like some occurrence data."
 
@@
 ```
-The default data that is returned consists of the latitude, longitude and date collected, along with a set of attribution data.
-If we want more information on these occurrences, we just need to change the arguments:
-
-```{r}
-Xanthium_strumarium_full <- BIEN_occurrence_species(species = "Xanthium strumarium",cultivated = T,only.new.world = T,all.taxonomy = T,native.status = T,observation.type = T,political.boundaries = T)
-
-str(Xanthium_strumarium_full)
-```
-
-Let’s take a quick look at where those occurrences are.
-
-```{r}
-# Make a quick map to plot our points on
-map('world',fill=T , col= "grey", bg="light blue")
-
-# Plot the points from the full query in red
-points(cbind(Xanthium_strumarium_full$longitude,Xanthium_strumarium_full$latitude),col="red",pch=20,cex=1)
-
-# Plot the points from the default query in blue
-points(cbind(Xanthium_strumarium$longitude,Xanthium_strumarium$latitude),col="blue",pch=20,cex=1) 
-```
 
@@
 
-Let’s say we’re interested in the genus Abies, and we’d like to get an idea of how many species there are in this genus and what higher taxa it falls within.
+"Let’s say we’re interested in the genus [Abies], and we’d like to get an idea of how many species there are in this genus and what higher taxa it falls within."
 

### 2021-03-09 R/add_plant_functional_type.Rmd
try TRY next?
maybe taxize package has what we need?

@@
 ```{r}
-#usda_plants <- read_csv(file.path(here::here(), "analysis", "data", "usda_plants_20160713.csv"),
-                        local = locale(encoding = "latin1"))
+#usda_plants <- read_csv(file.path(here::here(), "analysis", "data", "usda_plants_20160713.csv"))
 
@@
 ```
+
+## TRY Database
+
+https://www.try-db.org/TryWeb/Home.php
+
+## Taxize
+
+https://www.rdocumentation.org/packages/taxize/versions/0.9.99

### 2021-03-29 R/add_plant_functional_type.Rmd
starting to add PFT labels manually 7/146 done lol

@@
 ---
-title: "Add Plant Functional Type labels"
+title: "Add Plant Functional Type (PFT) labels"
 output: html_document
@@
 
-## USDA Plants Database
+Plant functional types (PFTs) of interest: coniferous forest, deciduous forest, woody shrub, grass/herb, and bare ground.
+
+## R packages
+
+### USDA Plants Database
 
@@
 
-```{r}
+```{r, include = FALSE, eval=FALSE}
 #usda_plants <- read_csv(file.path(here::here(), "analysis", "data", "usda_plants_20160713.csv"))
@@
 
-## BIEN
+### BIEN
 
@@
 
-```{r install_load_packages_bien, message=FALSE}
+```{r install_load_packages_bien, message=FALSE, warning=FALSE}
 library(BIEN)
@@
 library(sp) # A package for spatial data
-```
-
-"Let’s say we’re interested in the species Xanthium strumarium and we’d like some occurrence data."
-
-```{r}
-Xanthium_strumarium <- BIEN_occurrence_species(species = "Xanthium strumarium")
-#View(Xanthium_strumarium)
-head(Xanthium_strumarium)
-str(Xanthium_strumarium)
 ```
@@
 #View(Abies_taxonomy)
-Abies_taxonomy[1:10,]
+Abies_taxonomy[1:3,]
 ```
 
-*Acer* genus is maple trees (deciduous trees). The "higher_plant_group" returned by the BIEN taxonomy is "flowering plants".
-
-```{r}
-Acer_taxonomy<-BIEN_taxonomy_genus(genus = "Acer")
-
-#View(Acer_taxonomy)
-Acer_taxonomy[1:10,]
-```
 *Acer rubrum* species is the red maple (deciduous trees). The "higher_plant_group" returned by the BIEN taxonomy is "flowering plants".
@@
 Acer_rubrum_taxonomy <- BIEN_taxonomy_species(species = "Acer rubrum")
-Acer_rubrum_taxonomy[1:10,]
+Acer_rubrum_taxonomy[1:3,]
 ```
-```{r}
-Artemisia_tridentata_taxonomy <- BIEN_taxonomy_species(species = "Artemisia tridentata")
-Artemisia_tridentata_taxonomy[1:10,]
-```
-```{r}
-Carpinus_caroliniana_taxonomy <- BIEN_taxonomy_species(species = "Carpinus caroliniana")
-Carpinus_caroliniana_taxonomy[1:5,]
+
+### TRY Database
+
+https://www.try-db.org/TryWeb/Home.php
+
+### Taxize
+
+https://www.rdocumentation.org/packages/taxize/versions/0.9.99
+
+## Manually add PFTs to spectral library
+
+```{r load_cleaned_spectra, message=FALSE}
+# read cleaned spectral library file 
+cleaned_spectra <- read_csv(here::here("analysis", "data", "derived_data", "cleaned_spectra.csv"))
 ```
@@
 ```{r}
-Juniperus_virginiana_taxonomy <- BIEN_taxonomy_species(species = "Juniperus virginiana")
-Juniperus_virginiana_taxonomy[1:nrow(Juniperus_virginiana_taxonomy),]
+# list unique taxon labels, add an empty columns to fill with plant functional type (PFT) and common name labels.
+# PFTs of interest: coniferous_forest, deciduous_forest, woody_shrub, grass_herb, and bare_ground.
+taxon_pft_lookup <- data.frame(genusSpecies = unique(cleaned_spectra$genusSpecies),
+                               commonName = NA,
+                               pft = NA)
+
+for(i in 1:nrow(taxon_pft_lookup)){
+  # print row and current genusSpecies label
+  print(i)
+  genusSpecies <- taxon_pft_lookup$genusSpecies[i]
+  print(genusSpecies)
+
+  if(genusSpecies == "Pseudotsuga menziesii"){
+    taxon_pft_lookup$pft[i] = "coniferous_forest"
+    taxon_pft_lookup$commonName[i] = "Douglas-fir"}
+
+  else if(genusSpecies == "Alnus rubra"){
+    taxon_pft_lookup$pft[i] = "deciduous_forest"  # deciduous broadleaf tree 
+    taxon_pft_lookup$commonName[i] = "red alder"}
+
+  else if(genusSpecies == "Vaccinium parvifolium"){ 
+    taxon_pft_lookup$pft[i] = "woody_shrub"       # deciduous Shrub
+    taxon_pft_lookup$commonName[i] = "red huckleberry"}
+
+  else if(genusSpecies == "Mahonia nervosa"){
+    taxon_pft_lookup$pft[i] = "woody_shrub"      # evergreen shrub
+    taxon_pft_lookup$commonName[i] = "Cascade barberry"}
+
+  else if(genusSpecies == "Corylus cornuta"){
+    taxon_pft_lookup$pft[i] = "woody_shrub"      # deciduous shrubby hazel
+    taxon_pft_lookup$commonName[i] = "beaked hazelnut"}
+
+  else if(genusSpecies == "Gaultheria shallon"){
+    taxon_pft_lookup$pft[i] = "woody_shrub"     # broadleaved evergreen shrub
+    taxon_pft_lookup$commonName[i] = "salal"}
+
+  else if(genusSpecies == "Acer circinatum"){
+    taxon_pft_lookup$pft[i] = "deciduous_forest"
+    taxon_pft_lookup$commonName[i] = "vine maple"}
+
+  else if(genusSpecies == ""){
+    taxon_pft_lookup$pft[i] = ""
+    taxon_pft_lookup$commonName[i] = ""}
+
+  else if(genusSpecies == ""){
+    taxon_pft_lookup$pft[i] = ""
+    taxon_pft_lookup$commonName[i] = ""}
+
+  else if(genusSpecies == ""){
+    taxon_pft_lookup$pft[i] = ""
+    taxon_pft_lookup$commonName[i] = ""}
+
+  else if(genusSpecies == ""){
+    taxon_pft_lookup$pft[i] = ""
+    taxon_pft_lookup$commonName[i] = ""}
+
+  else if(genusSpecies == ""){
+    taxon_pft_lookup$pft[i] = ""
+    taxon_pft_lookup$commonName[i] = ""}
+
+  else if(genusSpecies == ""){
+    taxon_pft_lookup$pft[i] = ""
+    taxon_pft_lookup$commonName[i] = ""}
+
+  else if(genusSpecies == ""){
+    taxon_pft_lookup$pft[i] = ""
+    taxon_pft_lookup$commonName[i] = ""}
+
+  else if(genusSpecies == ""){
+    taxon_pft_lookup$pft[i] = ""
+    taxon_pft_lookup$commonName[i] = ""}
+
+  else if(genusSpecies == ""){
+    taxon_pft_lookup$pft[i] = ""
+    taxon_pft_lookup$commonName[i] = ""}
+
+  else if(genusSpecies == ""){
+    taxon_pft_lookup$pft[i] = ""
+    taxon_pft_lookup$commonName[i] = ""}
+
+  else if(genusSpecies == ""){
+    taxon_pft_lookup$pft[i] = ""
+    taxon_pft_lookup$commonName[i] = ""}
+
+  else if(genusSpecies == ""){
+    taxon_pft_lookup$pft[i] = ""
+    taxon_pft_lookup$commonName[i] = ""}
+
+  else if(genusSpecies == ""){
+    taxon_pft_lookup$pft[i] = ""
+    taxon_pft_lookup$commonName[i] = ""}
+
+  else if(genusSpecies == ""){
+    taxon_pft_lookup$pft[i] = ""
+    taxon_pft_lookup$commonName[i] = ""}
+
+  else print("PFT unknown")
+
+  # print corresponding PFT
+  print(taxon_pft_lookup$pft[i])
+}
+
+taxon_pft_lookup$pft(taxon_pft_lookup$genusSpecies == "Pseudotsuga menziesii") <- "coniferous_forest"
 ```
-```{r}
-Quercus_laevis_taxonomy <- BIEN_taxonomy_species(species = "Quercus laevis")
-Quercus_laevis_taxonomy[1:nrow(Quercus_laevis_taxonomy),]
-```
-
-## TRY Database
-
-https://www.try-db.org/TryWeb/Home.php
-
-## Taxize
-
-https://www.rdocumentation.org/packages/taxize/versions/0.9.99

### 2021-03-29 R/add_plant_functional_type.Rmd
13 done

@@
 
+unique(cleaned_spectra$scientificName)
+
 ```{r}
@@
     taxon_pft_lookup$pft[i] = "deciduous_forest"
+                  # or woody shrub? USDA Plants shows both Shrub and Tree growth form
     taxon_pft_lookup$commonName[i] = "vine maple"}
 
-  else if(genusSpecies == ""){
-    taxon_pft_lookup$pft[i] = ""
-    taxon_pft_lookup$commonName[i] = ""}
+  else if(genusSpecies == "Frangula purshiana"){
+    taxon_pft_lookup$pft[i] = "woody_shrub" 
+    taxon_pft_lookup$commonName[i] = "Cascara buckthorn"}
 
-  else if(genusSpecies == ""){
-    taxon_pft_lookup$pft[i] = ""
-    taxon_pft_lookup$commonName[i] = ""}
+  else if(genusSpecies == "Acer rubrum"){
+    taxon_pft_lookup$pft[i] = "deciduous_forest"
+    taxon_pft_lookup$commonName[i] = "red maple"}
 
-  else if(genusSpecies == ""){
-    taxon_pft_lookup$pft[i] = ""
-    taxon_pft_lookup$commonName[i] = ""}
+  else if(genusSpecies == "Betula alleghaniensis"){
+    taxon_pft_lookup$pft[i] = "deciduous_forest"
+    taxon_pft_lookup$commonName[i] = "yellow birch"}
 
-  else if(genusSpecies == ""){
-    taxon_pft_lookup$pft[i] = ""
-    taxon_pft_lookup$commonName[i] = ""}
+  else if(genusSpecies == "Populus tremuloides"){
+    taxon_pft_lookup$pft[i] = "deciduous_forest"
+    taxon_pft_lookup$commonName[i] = "quaking aspen"}
 
-  else if(genusSpecies == ""){
-    taxon_pft_lookup$pft[i] = ""
-    taxon_pft_lookup$commonName[i] = ""}
+  else if(genusSpecies == "Tsuga canadensis"){
+    taxon_pft_lookup$pft[i] = "coniferous_forest"
+    taxon_pft_lookup$commonName[i] = "eastern hemlock"}
 
-  else if(genusSpecies == ""){
-    taxon_pft_lookup$pft[i] = ""
-    taxon_pft_lookup$commonName[i] = ""}
+  else if(genusSpecies == "Acer pensylvanicum"){
+    taxon_pft_lookup$pft[i] = "deciduous_forest"
+    taxon_pft_lookup$commonName[i] = "striped maple"}
 

### 2021-03-29 R/add_plant_functional_type.Rmd
20 done

@@
 
-  else if(genusSpecies == ""){
-    taxon_pft_lookup$pft[i] = ""
-    taxon_pft_lookup$commonName[i] = ""}
+  else if(genusSpecies == "Picea sp."){
+    taxon_pft_lookup$pft[i] = "coniferous_forest"
+    taxon_pft_lookup$commonName[i] = "spruce"}
 
-  else if(genusSpecies == ""){
-    taxon_pft_lookup$pft[i] = ""
-    taxon_pft_lookup$commonName[i] = ""}
+  else if(genusSpecies == "Fraxinus americana"){
+    taxon_pft_lookup$pft[i] = "deciduous_forest"
+    taxon_pft_lookup$commonName[i] = "white ash"}
 
-  else if(genusSpecies == ""){
-    taxon_pft_lookup$pft[i] = ""
-    taxon_pft_lookup$commonName[i] = ""}
+  else if(genusSpecies == "Betula papyrifera"){
+    taxon_pft_lookup$pft[i] = "deciduous_forest"
+    taxon_pft_lookup$commonName[i] = "paper birch"}
 
-  else if(genusSpecies == ""){
-    taxon_pft_lookup$pft[i] = ""
-    taxon_pft_lookup$commonName[i] = ""}
+  else if(genusSpecies == "Acer saccharum"){
+    taxon_pft_lookup$pft[i] = "deciduous_forest"
+    taxon_pft_lookup$commonName[i] = "sugar maple"}
 
-  else if(genusSpecies == ""){
-    taxon_pft_lookup$pft[i] = ""
-    taxon_pft_lookup$commonName[i] = ""}
+  else if(genusSpecies == "Fagus grandifolia"){
+    taxon_pft_lookup$pft[i] = "deciduous_forest"
+    taxon_pft_lookup$commonName[i] = "American beech"}
 
-  else if(genusSpecies == ""){
-    taxon_pft_lookup$pft[i] = ""
-    taxon_pft_lookup$commonName[i] = ""}
+  else if(genusSpecies == "Rhamnus davurica"){
+    taxon_pft_lookup$pft[i] = "woody_shrub"
+    taxon_pft_lookup$commonName[i] = "Dahurian buckthorn"}
 
-  else if(genusSpecies == ""){
-    taxon_pft_lookup$pft[i] = ""
-    taxon_pft_lookup$commonName[i] = ""}
+  else if(genusSpecies == "Lonicera maackii"){
+    taxon_pft_lookup$pft[i] = "woody_shrub"
+    taxon_pft_lookup$commonName[i] = "Amur honeysuckle"}
 

### 2021-03-29 R/add_plant_functional_type.Rmd
31 done

@@
 
-## R packages
-
-### USDA Plants Database
-
-USDA Plants Database R API tutorial: https://recology.info/2016/10/usda-plants-database-r/
-
-The USDA maintains a database of plant information, some of it trait data, some of it life history. Check it out at http://plants.usda.gov/java/
-
-```{r install_load_packages, message=FALSE}
-#install.packages("request")
-#library(devtools)
-#devtools::install_github("sckott/request")
-library(request)
-library(tibble)
-library(tidyverse)
-library(here)
-```
-
-```{r}
-# Set the base url
-root <- api("https://plantsdb.xyz")
-
-root %>% api_path(heartbeat)
-```
-
-^^ My results here do not match those that show up on the tutorial page above. Maybe the API is down or the URL has changed?
-
-Based on the GitHub issue discussion (https://github.com/USDA/USDA-APIs/issues/7), someone uploaded USDA plants data here from 2016: https://uofi.app.box.com/v/usdaplants
-
-```{r, eval=FALSE, include=FALSE}
-# Specify URL where file is stored 
-url <- "https://uofi.app.box.com/v/usdaplants/folder/6688524249/usda_plants_20160713.csv"
-destFile <- file.path(here::here(), "analysis","data", "usda_plants_20160713.csv")
-# Apply download.file function in R
-download.file(url, destFile)
-```
-
-```{r, include = FALSE, eval=FALSE}
-#usda_plants <- read_csv(file.path(here::here(), "analysis", "data", "usda_plants_20160713.csv"))
-
-# Error in nchar(x) : invalid multibyte string, element 2
-# ^ Issue with some of the column names in the .csv? 
-```
-
-### BIEN
-
-RBIEN tutorial: https://cmerow.github.io/RDataScience/3_3_RBIEN_tutorial.html
-
-```{r install_load_packages_bien, message=FALSE, warning=FALSE}
-library(BIEN)
-library(ape) # Package for working with phylogenies in R
-library(maps) # Useful for making quick maps of occurrences
-library(sp) # A package for spatial data
-```
-
-BIEN traits and taxonomic data
-
-"Let’s say we’re interested in the genus [Abies], and we’d like to get an idea of how many species there are in this genus and what higher taxa it falls within."
-
-*Abies* genus is fir trees (coniferous trees). The "higher_plant_group" returned by the BIEN taxonomy function is "gymnosperms (conifers)"
-
-```{r}
-Abies_taxonomy<-BIEN_taxonomy_genus(genus = "Abies")
-
-#View(Abies_taxonomy)
-Abies_taxonomy[1:3,]
-```
-
-*Acer rubrum* species is the red maple (deciduous trees). The "higher_plant_group" returned by the BIEN taxonomy is "flowering plants".
-
-```{r}
-Acer_rubrum_taxonomy <- BIEN_taxonomy_species(species = "Acer rubrum")
-Acer_rubrum_taxonomy[1:3,]
-```
-
-### TRY Database
-
-https://www.try-db.org/TryWeb/Home.php
-
-### Taxize
-
-https://www.rdocumentation.org/packages/taxize/versions/0.9.99
-
 ## Manually add PFTs to spectral library
@@
 ```
-
-unique(cleaned_spectra$scientificName)
 
@@
 
+  else if(genusSpecies == "Ailanthus altissima"){
+    taxon_pft_lookup$pft[i] = "deciduous_forest"
+    taxon_pft_lookup$commonName[i] = "tree of heaven"}
+
+  else if(genusSpecies == "Prunus serotina"){
+    taxon_pft_lookup$pft[i] = "deciduous_forest"
+    taxon_pft_lookup$commonName[i] = "black cherrt"}
+
+  else if(genusSpecies == "Celtis occidentalis"){
+    taxon_pft_lookup$pft[i] = "deciduous_forest"
+    taxon_pft_lookup$commonName[i] = "common hackberry"}
+
+  else if(genusSpecies == "Quercus alba"){
+    taxon_pft_lookup$pft[i] = "deciduous_forest"
+    taxon_pft_lookup$commonName[i] = "white oak"}
+
+  else if(genusSpecies == "Juglans nigra"){
+    taxon_pft_lookup$pft[i] = "deciduous_forest"
+    taxon_pft_lookup$commonName[i] = "black walnut"}
+
+  else if(genusSpecies == "Robinia pseudoacacia"){
+    taxon_pft_lookup$pft[i] = "deciduous_forest"
+    taxon_pft_lookup$commonName[i] = "black locust"}
+
+  else if(genusSpecies == "Pinus strobus"){
+    taxon_pft_lookup$pft[i] = "coniferous_forest"
+    taxon_pft_lookup$commonName[i] = "eastern white pine"}
+
+  else if(genusSpecies == "Carya glabra"){
+    taxon_pft_lookup$pft[i] = "deciduous_forest"
+    taxon_pft_lookup$commonName[i] = "pignut hickory"}
+
+  else if(genusSpecies == "Tetradium daniellii"){
+    taxon_pft_lookup$pft[i] = "deciduous_forest"
+    taxon_pft_lookup$commonName[i] = "bee-bee tree"}
+
+  else if(genusSpecies == "Juniperus virginiana"){
+    taxon_pft_lookup$pft[i] = "coniferous_forest"
+    taxon_pft_lookup$commonName[i] = "red cedar"}
+
+  else if(genusSpecies == "Picea mariana"){
+    taxon_pft_lookup$pft[i] = "coniferous_forest"
+    taxon_pft_lookup$commonName[i] = "black spruce"}
+
+  else if(genusSpecies == ""){
+    taxon_pft_lookup$pft[i] = ""
+    taxon_pft_lookup$commonName[i] = ""}
+
+  else if(genusSpecies == ""){
+    taxon_pft_lookup$pft[i] = ""
+    taxon_pft_lookup$commonName[i] = ""}
+
+  else if(genusSpecies == ""){
+    taxon_pft_lookup$pft[i] = ""
+    taxon_pft_lookup$commonName[i] = ""}
+
+  else if(genusSpecies == ""){
+    taxon_pft_lookup$pft[i] = ""
+    taxon_pft_lookup$commonName[i] = ""}
+
   else if(genusSpecies == ""){
@@
 ```
+
+## R packages: attempting to automate PFT labeling
+
+### USDA Plants Database
+
+USDA Plants Database R API tutorial: https://recology.info/2016/10/usda-plants-database-r/
+
+The USDA maintains a database of plant information, some of it trait data, some of it life history. Check it out at http://plants.usda.gov/java/
+
+```{r install_load_packages, message=FALSE}
+#install.packages("request")
+#library(devtools)
+#devtools::install_github("sckott/request")
+library(request)
+library(tibble)
+library(tidyverse)
+library(here)
+```
+
+```{r}
+# Set the base url
+root <- api("https://plantsdb.xyz")
+
+root %>% api_path(heartbeat)
+```
+
+^^ My results here do not match those that show up on the tutorial page above. Maybe the API is down or the URL has changed?
+
+Based on the GitHub issue discussion (https://github.com/USDA/USDA-APIs/issues/7), someone uploaded USDA plants data here from 2016: https://uofi.app.box.com/v/usdaplants
+
+```{r, eval=FALSE, include=FALSE}
+# Specify URL where file is stored 
+url <- "https://uofi.app.box.com/v/usdaplants/folder/6688524249/usda_plants_20160713.csv"
+destFile <- file.path(here::here(), "analysis","data", "usda_plants_20160713.csv")
+# Apply download.file function in R
+download.file(url, destFile)
+```
+
+```{r, include = FALSE, eval=FALSE}
+#usda_plants <- read_csv(file.path(here::here(), "analysis", "data", "usda_plants_20160713.csv"))
+
+# Error in nchar(x) : invalid multibyte string, element 2
+# ^ Issue with some of the column names in the .csv? 
+```
+
+### BIEN
+
+RBIEN tutorial: https://cmerow.github.io/RDataScience/3_3_RBIEN_tutorial.html
+
+```{r install_load_packages_bien, message=FALSE, warning=FALSE}
+library(BIEN)
+library(ape) # Package for working with phylogenies in R
+library(maps) # Useful for making quick maps of occurrences
+library(sp) # A package for spatial data
+```
+
+BIEN traits and taxonomic data
+
+"Let’s say we’re interested in the genus [Abies], and we’d like to get an idea of how many species there are in this genus and what higher taxa it falls within."
+
+*Abies* genus is fir trees (coniferous trees). The "higher_plant_group" returned by the BIEN taxonomy function is "gymnosperms (conifers)"
+
+```{r}
+Abies_taxonomy<-BIEN_taxonomy_genus(genus = "Abies")
+
+#View(Abies_taxonomy)
+Abies_taxonomy[1:3,]
+```
+
+*Acer rubrum* species is the red maple (deciduous trees). The "higher_plant_group" returned by the BIEN taxonomy is "flowering plants".
+
+```{r}
+Acer_rubrum_taxonomy <- BIEN_taxonomy_species(species = "Acer rubrum")
+Acer_rubrum_taxonomy[1:3,]
+```
+
+### TRY Database
+
+https://www.try-db.org/TryWeb/Home.php
+
+### Taxize
+
+https://www.rdocumentation.org/packages/taxize/versions/0.9.99

### 2021-03-29 R/add_plant_functional_type.Rmd
78 done. need to flag the ones that are unclear between large shrub and small tree. ask the group for ideas

@@
 
-  if(genusSpecies == "Pseudotsuga menziesii"){
+  if(genusSpecies == "Abies amabilis"){
     taxon_pft_lookup$pft[i] = "coniferous_forest"
-    taxon_pft_lookup$commonName[i] = "Douglas-fir"}
+    taxon_pft_lookup$commonName[i] = "Pacific silver fir"}
+
+  else if(genusSpecies == "Abies balsamea"){
+    taxon_pft_lookup$pft[i] = "coniferous_forest"
+    taxon_pft_lookup$commonName[i] = "balsam fir"}
+
+  else if(genusSpecies == "Abies lasiocarpa"){
+    taxon_pft_lookup$pft[i] = "coniferous_forest"
+    taxon_pft_lookup$commonName[i] = "subalpine fir"}
+
+  else if(genusSpecies == "Acacia greggii"){
+    taxon_pft_lookup$pft[i] = "woody_shrub"
+    taxon_pft_lookup$commonName[i] = "catclaw acacia"}
+
+  else if(genusSpecies == "Acer barbatum"){
+    taxon_pft_lookup$pft[i] = "deciduous_forest"
+    taxon_pft_lookup$commonName[i] = "Southern sugar maple"}
+
+  else if(genusSpecies == "Acer circinatum"){
+    taxon_pft_lookup$pft[i] = "deciduous_forest"
+              # or woody shrub? USDA Plants shows both Shrub and Tree growth form
+    taxon_pft_lookup$commonName[i] = "vine maple"}
+
+  else if(genusSpecies == "Acer negundo"){
+    taxon_pft_lookup$pft[i] = "deciduous_forest"
+    taxon_pft_lookup$commonName[i] = "boxelder"}
+
+  else if(genusSpecies == "Acer pensylvanicum"){
+    taxon_pft_lookup$pft[i] = "deciduous_forest"
+    taxon_pft_lookup$commonName[i] = "striped maple"}
+
+  else if(genusSpecies == "Acer rubrum"){
+    taxon_pft_lookup$pft[i] = "deciduous_forest"
+    taxon_pft_lookup$commonName[i] = "red maple"}
+
+  else if(genusSpecies == "Acer saccharinum"){
+    taxon_pft_lookup$pft[i] = "deciduous_forest"
+    taxon_pft_lookup$commonName[i] = "silver maple"}
+
+  else if(genusSpecies == "Acer saccharum"){
+    taxon_pft_lookup$pft[i] = "deciduous_forest"
+    taxon_pft_lookup$commonName[i] = "sugar maple"}
+
+  else if(genusSpecies == "Ailanthus altissima"){
+    taxon_pft_lookup$pft[i] = "deciduous_forest"
+    taxon_pft_lookup$commonName[i] = "tree of heaven"}
 
@@
     taxon_pft_lookup$commonName[i] = "red alder"}
+
+  else if(genusSpecies == "Aloysia wrightii"){
+    taxon_pft_lookup$pft[i] = "woody_shrub"
+    taxon_pft_lookup$commonName[i] = "Wright's beebrush"}
+
+  else if(genusSpecies == "Amelanchier laevis"){
+    taxon_pft_lookup$pft[i] = "deciduous_forest"
+    taxon_pft_lookup$commonName[i] = "Allegheny serviceberry"}
+
+  else if(genusSpecies == "Amyris elemifera"){
+    taxon_pft_lookup$pft[i] = "woody_shrub"
+    taxon_pft_lookup$commonName[i] = "sea torchwood"}
+
+  else if(genusSpecies == "Artemisia tridentata"){
+    taxon_pft_lookup$pft[i] = "woody_shrub"
+    taxon_pft_lookup$commonName[i] = "big sagebrush"}
+
+  else if(genusSpecies == "Atriplex canescens"){
+    taxon_pft_lookup$pft[i] = "woody_shrub"
+    taxon_pft_lookup$commonName[i] = "fourwing saltbush"}
+
+  else if(genusSpecies == "Atriplex confertifolia"){
+    taxon_pft_lookup$pft[i] = "woody_shrub"
+    taxon_pft_lookup$commonName[i] = "shadscale saltbush"}
+
+  else if(genusSpecies == "Betula alleghaniensis"){
+    taxon_pft_lookup$pft[i] = "deciduous_forest"
+    taxon_pft_lookup$commonName[i] = "yellow birch"}
+
+  else if(genusSpecies == "Betula lenta"){
+    taxon_pft_lookup$pft[i] = "deciduous_forest"
+    taxon_pft_lookup$commonName[i] = "sweet birch"}
+
+  else if(genusSpecies == "Betula neoalaskana"){
+    taxon_pft_lookup$pft[i] = "deciduous_forest"
+    taxon_pft_lookup$commonName[i] = "resin birch"}
+
+  else if(genusSpecies == "Betula nigra"){
+    taxon_pft_lookup$pft[i] = "deciduous_forest"
+    taxon_pft_lookup$commonName[i] = "river birch"}
+
+  else if(genusSpecies == "Betula papyrifera"){
+    taxon_pft_lookup$pft[i] = "deciduous_forest"
+    taxon_pft_lookup$commonName[i] = "paper birch"}
+
+  else if(genusSpecies == "Betula sp."){
+    taxon_pft_lookup$pft[i] = "deciduous_forest"
+    taxon_pft_lookup$commonName[i] = "birch"}
+
+  else if(genusSpecies == "Bourreria sp."){
+    taxon_pft_lookup$pft[i] = "woody_shrub"
+    taxon_pft_lookup$commonName[i] = "strongbark"}
+
+  else if(genusSpecies == "Bursera simaruba"){
+    taxon_pft_lookup$pft[i] = "deciduous_forest"  # "semi-deciduous"
+    taxon_pft_lookup$commonName[i] = "gumbo limbo"}
+
+  else if(genusSpecies == "Carpinus caroliniana"){
+    taxon_pft_lookup$pft[i] = "deciduous_forest"
+    taxon_pft_lookup$commonName[i] = "American hornbeam"}
+
+  else if(genusSpecies == "Carya aquatica"){
+    taxon_pft_lookup$pft[i] = "deciduous_forest"
+    taxon_pft_lookup$commonName[i] = "water hickory"}
+
+  else if(genusSpecies == "Carya cordiformis"){
+    taxon_pft_lookup$pft[i] = "deciduous_forest"
+    taxon_pft_lookup$commonName[i] = "bitternut hickory"}
+
+  else if(genusSpecies == "Carya glabra"){
+    taxon_pft_lookup$pft[i] = "deciduous_forest"
+    taxon_pft_lookup$commonName[i] = "pignut hickory"}
+
+  else if(genusSpecies == "Carya illinoinensis"){
+    taxon_pft_lookup$pft[i] = "deciduous_forest"
+    taxon_pft_lookup$commonName[i] = "pecan"}
+
+  else if(genusSpecies == "Carya ovalis"){
+    taxon_pft_lookup$pft[i] = "deciduous_forest"
+    taxon_pft_lookup$commonName[i] = "red hickory"}
+
+  else if(genusSpecies == "Carya ovata"){
+    taxon_pft_lookup$pft[i] = "deciduous_forest"
+    taxon_pft_lookup$commonName[i] = "shagbark hickory"}
+
+  else if(genusSpecies == "Carya tomentosa"){
+    taxon_pft_lookup$pft[i] = "deciduous_forest"
+    taxon_pft_lookup$commonName[i] = "mockernut hickory"}
+
+  else if(genusSpecies == "Celtis laevigata"){
+    taxon_pft_lookup$pft[i] = "deciduous_forest"
+    taxon_pft_lookup$commonName[i] = "sugarberry"}
+
+  else if(genusSpecies == "Celtis occidentalis"){
+    taxon_pft_lookup$pft[i] = "deciduous_forest"
+    taxon_pft_lookup$commonName[i] = "common hackberry"}
+
+  else if(genusSpecies == "Cercis canadensis"){
+    taxon_pft_lookup$pft[i] = "deciduous_forest"
+    taxon_pft_lookup$commonName[i] = "eastern redbud"}
+
+  else if(genusSpecies == "Chrysothamnus viscidiflorus"){
+    taxon_pft_lookup$pft[i] = "woody_shrub"
+    taxon_pft_lookup$commonName[i] = "yellow rabbitbrush"}
+
+  else if(genusSpecies == "Cordia collococca"){
+    taxon_pft_lookup$pft[i] = "deciduous_forest"
+    taxon_pft_lookup$commonName[i] = "red manjack"}
+
+  else if(genusSpecies == "Cornus drummondii"){
+    taxon_pft_lookup$pft[i] = "deciduous_forest"
+    taxon_pft_lookup$commonName[i] = "roughleaf dogwood"}
+
+  else if(genusSpecies == "Cornus florida"){
+    taxon_pft_lookup$pft[i] = "deciduous_forest"
+    taxon_pft_lookup$commonName[i] = "flowering dogwood"}
+
+  else if(genusSpecies == "Corylus cornuta"){
+    taxon_pft_lookup$pft[i] = "woody_shrub"      # deciduous shrubby hazel
+    taxon_pft_lookup$commonName[i] = "beaked hazelnut"}
+
+  else if(genusSpecies == "Diospyros virginiana"){
+    taxon_pft_lookup$pft[i] = "deciduous_forest"
+    taxon_pft_lookup$commonName[i] = "common persimmon"}
+
+  else if(genusSpecies == "Ephedra trifurca"){
+    taxon_pft_lookup$pft[i] = "woody_shrub"
+    taxon_pft_lookup$commonName[i] = "longleaf jointfir"}
+
+  else if(genusSpecies == "Ericameria laricifolia"){
+    taxon_pft_lookup$pft[i] = "woody_shrub"
+    taxon_pft_lookup$commonName[i] = "turpentine bush"}
+
+  else if(genusSpecies == "Exostema caribaeum"){
+    taxon_pft_lookup$pft[i] = "woody_shrub"  # "evergreen shrub or small tree" 
+    taxon_pft_lookup$commonName[i] = "Caribbean princewood"}
+
+  else if(genusSpecies == "Fagus grandifolia"){
+    taxon_pft_lookup$pft[i] = "deciduous_forest"
+    taxon_pft_lookup$commonName[i] = "American beech"}
+
+  else if(genusSpecies == "Frangula purshiana"){
+    taxon_pft_lookup$pft[i] = "woody_shrub"
+    taxon_pft_lookup$commonName[i] = "Cascara buckthorn"}
+
+  else if(genusSpecies == "Fraxinus americana"){
+    taxon_pft_lookup$pft[i] = "deciduous_forest"
+    taxon_pft_lookup$commonName[i] = "white ash"}
+
+  else if(genusSpecies == "Fraxinus nigra"){
+    taxon_pft_lookup$pft[i] = "deciduous_forest"
+    taxon_pft_lookup$commonName[i] = "black ash"}
+
+  else if(genusSpecies == "Fraxinus pennsylvanica"){
+    taxon_pft_lookup$pft[i] = "deciduous_forest"
+    taxon_pft_lookup$commonName[i] = "green ash"}
+
+  else if(genusSpecies == "Fraxinus sp."){
+    taxon_pft_lookup$pft[i] = "deciduous_forest"
+    taxon_pft_lookup$commonName[i] = "ash"}
+
+  else if(genusSpecies == "Gaultheria shallon"){
+    taxon_pft_lookup$pft[i] = "woody_shrub"     # broadleaved evergreen shrub
+    taxon_pft_lookup$commonName[i] = "salal"}
+
+  else if(genusSpecies == "Gleditsia triacanthos"){
+    taxon_pft_lookup$pft[i] = "deciduous_forest"
+    taxon_pft_lookup$commonName[i] = "honey locust"}
+
+  else if(genusSpecies == "Gutierrezia sarothrae"){
+    taxon_pft_lookup$pft[i] = "woody_shrub"
+    taxon_pft_lookup$commonName[i] = "broom snakeweed"}
+
+  else if(genusSpecies == "Halesia sp."){
+    taxon_pft_lookup$pft[i] = "deciduous_forest"  # small, deciduous, understory tree 
+    taxon_pft_lookup$commonName[i] = "silverbell"}
+
+  else if(genusSpecies == "Hamamelis virginiana"){
+    taxon_pft_lookup$pft[i] = "woody_shrub"
+    taxon_pft_lookup$commonName[i] = "American witchhazel"}
+
+  else if(genusSpecies == "Ilex montana"){
+    taxon_pft_lookup$pft[i] = "NA"
+    taxon_pft_lookup$commonName[i] = "mountain holly"}
+
+  else if(genusSpecies == ""){
+    taxon_pft_lookup$pft[i] = ""
+    taxon_pft_lookup$commonName[i] = ""}
+
+  else if(genusSpecies == ""){
+    taxon_pft_lookup$pft[i] = ""
+    taxon_pft_lookup$commonName[i] = ""}
+
+  else if(genusSpecies == ""){
+    taxon_pft_lookup$pft[i] = ""
+    taxon_pft_lookup$commonName[i] = ""}
+
+  else if(genusSpecies == "Pseudotsuga menziesii"){
+    taxon_pft_lookup$pft[i] = "coniferous_forest"
+    taxon_pft_lookup$commonName[i] = "Douglas-fir"}
 
@@
 
-  else if(genusSpecies == "Corylus cornuta"){
-    taxon_pft_lookup$pft[i] = "woody_shrub"      # deciduous shrubby hazel
-    taxon_pft_lookup$commonName[i] = "beaked hazelnut"}
-
-  else if(genusSpecies == "Gaultheria shallon"){
-    taxon_pft_lookup$pft[i] = "woody_shrub"     # broadleaved evergreen shrub
-    taxon_pft_lookup$commonName[i] = "salal"}
-
-  else if(genusSpecies == "Acer circinatum"){
-    taxon_pft_lookup$pft[i] = "deciduous_forest"
-                  # or woody shrub? USDA Plants shows both Shrub and Tree growth form
-    taxon_pft_lookup$commonName[i] = "vine maple"}
-
-  else if(genusSpecies == "Frangula purshiana"){
-    taxon_pft_lookup$pft[i] = "woody_shrub" 
-    taxon_pft_lookup$commonName[i] = "Cascara buckthorn"}
-
-  else if(genusSpecies == "Acer rubrum"){
-    taxon_pft_lookup$pft[i] = "deciduous_forest"
-    taxon_pft_lookup$commonName[i] = "red maple"}
-
-  else if(genusSpecies == "Betula alleghaniensis"){
-    taxon_pft_lookup$pft[i] = "deciduous_forest"
-    taxon_pft_lookup$commonName[i] = "yellow birch"}
-
   else if(genusSpecies == "Populus tremuloides"){
@@
 
-  else if(genusSpecies == "Acer pensylvanicum"){
-    taxon_pft_lookup$pft[i] = "deciduous_forest"
-    taxon_pft_lookup$commonName[i] = "striped maple"}
-
   else if(genusSpecies == "Picea sp."){
@@
     taxon_pft_lookup$commonName[i] = "spruce"}
-
-  else if(genusSpecies == "Fraxinus americana"){
-    taxon_pft_lookup$pft[i] = "deciduous_forest"
-    taxon_pft_lookup$commonName[i] = "white ash"}
-
-  else if(genusSpecies == "Betula papyrifera"){
-    taxon_pft_lookup$pft[i] = "deciduous_forest"
-    taxon_pft_lookup$commonName[i] = "paper birch"}
-
-  else if(genusSpecies == "Acer saccharum"){
-    taxon_pft_lookup$pft[i] = "deciduous_forest"
-    taxon_pft_lookup$commonName[i] = "sugar maple"}
-
-  else if(genusSpecies == "Fagus grandifolia"){
-    taxon_pft_lookup$pft[i] = "deciduous_forest"
-    taxon_pft_lookup$commonName[i] = "American beech"}
 
@@
 
-  else if(genusSpecies == "Ailanthus altissima"){
-    taxon_pft_lookup$pft[i] = "deciduous_forest"
-    taxon_pft_lookup$commonName[i] = "tree of heaven"}
-
   else if(genusSpecies == "Prunus serotina"){
@@
     taxon_pft_lookup$commonName[i] = "black cherrt"}
-
-  else if(genusSpecies == "Celtis occidentalis"){
-    taxon_pft_lookup$pft[i] = "deciduous_forest"
-    taxon_pft_lookup$commonName[i] = "common hackberry"}
 
@@
 
-  else if(genusSpecies == "Carya glabra"){
-    taxon_pft_lookup$pft[i] = "deciduous_forest"
-    taxon_pft_lookup$commonName[i] = "pignut hickory"}
-
   else if(genusSpecies == "Tetradium daniellii"){
@@
 
-  else if(genusSpecies == ""){
-    taxon_pft_lookup$pft[i] = ""
-    taxon_pft_lookup$commonName[i] = ""}
+  else if(genusSpecies == "Picea glauca"){
+    taxon_pft_lookup$pft[i] = "coniferous_forest"
+    taxon_pft_lookup$commonName[i] = "white spruce"}
 
-  else if(genusSpecies == ""){
-    taxon_pft_lookup$pft[i] = ""
-    taxon_pft_lookup$commonName[i] = ""}
+  else if(genusSpecies == "Ulmus crassifolia"){
+    taxon_pft_lookup$pft[i] = "deciduous_forest"
+    taxon_pft_lookup$commonName[i] = "cedar elm"}
 
-  else if(genusSpecies == ""){
-    taxon_pft_lookup$pft[i] = ""
-    taxon_pft_lookup$commonName[i] = ""}
+  else if(genusSpecies == "Quercus stellata"){
+    taxon_pft_lookup$pft[i] = "deciduous_forest"
+    taxon_pft_lookup$commonName[i] = "post oak"}
 
@@
 
-taxon_pft_lookup$pft(taxon_pft_lookup$genusSpecies == "Pseudotsuga menziesii") <- "coniferous_forest"
 ```

### 2021-03-29 R/add_plant_functional_type.Rmd
still going

@@
   else if(genusSpecies == "Acer circinatum"){
-    taxon_pft_lookup$pft[i] = "deciduous_forest"
-              # or woody shrub? USDA Plants shows both Shrub and Tree growth form
+    taxon_pft_lookup$pft[i] = "NA " ??? 
     taxon_pft_lookup$commonName[i] = "vine maple"}
@@
   else if(genusSpecies == "Amelanchier laevis"){
-    taxon_pft_lookup$pft[i] = "deciduous_forest"
+    taxon_pft_lookup$pft[i] = "???" ???
     taxon_pft_lookup$commonName[i] = "Allegheny serviceberry"}
@@
   else if(genusSpecies == "Amyris elemifera"){
-    taxon_pft_lookup$pft[i] = "woody_shrub"
+    taxon_pft_lookup$pft[i] = "???"     ???
     taxon_pft_lookup$commonName[i] = "sea torchwood"}
@@
   else if(genusSpecies == "Bourreria sp."){
-    taxon_pft_lookup$pft[i] = "woody_shrub"
+    taxon_pft_lookup$pft[i] = "???"   ???
     taxon_pft_lookup$commonName[i] = "strongbark"}
@@
   else if(genusSpecies == "Cercis canadensis"){
-    taxon_pft_lookup$pft[i] = "deciduous_forest"
+    taxon_pft_lookup$pft[i] = "???"   ???
     taxon_pft_lookup$commonName[i] = "eastern redbud"}
@@
   else if(genusSpecies == "Cornus drummondii"){
-    taxon_pft_lookup$pft[i] = "deciduous_forest"
+    taxon_pft_lookup$pft[i] = "???"    ???
     taxon_pft_lookup$commonName[i] = "roughleaf dogwood"}
@@
   else if(genusSpecies == "Cornus florida"){
-    taxon_pft_lookup$pft[i] = "deciduous_forest"
+    taxon_pft_lookup$pft[i] = "???"    ???
     taxon_pft_lookup$commonName[i] = "flowering dogwood"}
@@
   else if(genusSpecies == "Frangula purshiana"){
-    taxon_pft_lookup$pft[i] = "woody_shrub"
+    taxon_pft_lookup$pft[i] = "???"     ???
     taxon_pft_lookup$commonName[i] = "Cascara buckthorn"}
@@
   else if(genusSpecies == "Ilex montana"){
-    taxon_pft_lookup$pft[i] = "NA"
+    taxon_pft_lookup$pft[i] = "???"          ???
     taxon_pft_lookup$commonName[i] = "mountain holly"}
+
+  else if(genusSpecies == "Ilex opaca"){
+    taxon_pft_lookup$pft[i] = "woody_shrub"   ???      # broadleaved evergreen tree
+    taxon_pft_lookup$commonName[i] = "American holly"}
+
+  else if(genusSpecies == "Juglans nigra"){
+    taxon_pft_lookup$pft[i] = "deciduous_forest"
+    taxon_pft_lookup$commonName[i] = "black walnut"}
+
+  else if(genusSpecies == "Juniperus osteosperma"){
+    taxon_pft_lookup$pft[i] = "coniferous_forest"
+    taxon_pft_lookup$commonName[i] = "Utah juniper"}
+
+  else if(genusSpecies == "Juniperus virginiana"){
+    taxon_pft_lookup$pft[i] = "coniferous_forest"
+    taxon_pft_lookup$commonName[i] = "red cedar"}
+
+  else if(genusSpecies == "Larix sp."){
+    taxon_pft_lookup$pft[i] = "coniferous_forest"
+    taxon_pft_lookup$commonName[i] = "larch"}
+
+  else if(genusSpecies == "Larrea tridentata"){
+    taxon_pft_lookup$pft[i] = "woody_shrub"
+    taxon_pft_lookup$commonName[i] = "creosote bush"}
+
+  else if(genusSpecies == "Liquidambar styraciflua"){
+    taxon_pft_lookup$pft[i] = "deciduous_forest"
+    taxon_pft_lookup$commonName[i] = "sweetgum"}
+
+  else if(genusSpecies == "Liriodendron tulipifera"){
+    taxon_pft_lookup$pft[i] = "deciduous_forest"
+    taxon_pft_lookup$commonName[i] = "tuliptree"}
+
+  else if(genusSpecies == "Lonicera maackii"){
+    taxon_pft_lookup$pft[i] = "woody_shrub"
+    taxon_pft_lookup$commonName[i] = "Amur honeysuckle"}
+
+  else if(genusSpecies == "Maclura pomifera"){
+    taxon_pft_lookup$pft[i] = "deciduous_forest"   # small deciduous tree or large shrub
+    taxon_pft_lookup$commonName[i] = "Osage-orange"}
+
+  else if(genusSpecies == "Mahonia nervosa"){
+    taxon_pft_lookup$pft[i] = "woody_shrub"      # evergreen shrub
+    taxon_pft_lookup$commonName[i] = "Cascade barberry"}
+
+  else if(genusSpecies == "Mangifera indica"){
+    taxon_pft_lookup$pft[i] = "???"       ???   # broadleaf evergreen tree
+    taxon_pft_lookup$commonName[i] = "mango"}
+
+  else if(genusSpecies == "Melia azedarach"){
+    taxon_pft_lookup$pft[i] = "deciduous_forest"
+    taxon_pft_lookup$commonName[i] = "Chinaberrytree"}
+
+  else if(genusSpecies == "Morus rubra"){
+    taxon_pft_lookup$pft[i] = "deciduous_forest"
+    taxon_pft_lookup$commonName[i] = "red mulberry"}
+
+  else if(genusSpecies == "Nyssa aquatica"){
+    taxon_pft_lookup$pft[i] = "deciduous_forest"
+    taxon_pft_lookup$commonName[i] = "water tupelo"}
+
+  else if(genusSpecies == "Nyssa biflora"){
+    taxon_pft_lookup$pft[i] = "deciduous_forest"
+    taxon_pft_lookup$commonName[i] = "swamp tupelo"}
+
+  else if(genusSpecies == "Nyssa sylvatica"){
+    taxon_pft_lookup$pft[i] = "deciduous_forest"
+    taxon_pft_lookup$commonName[i] = "blackgum"}
+
+  else if(genusSpecies == ""){
+    taxon_pft_lookup$pft[i] = ""
+    taxon_pft_lookup$commonName[i] = ""}
+
+  else if(genusSpecies == ""){
+    taxon_pft_lookup$pft[i] = ""
+    taxon_pft_lookup$commonName[i] = ""}
+
+  else if(genusSpecies == ""){
+    taxon_pft_lookup$pft[i] = ""
+    taxon_pft_lookup$commonName[i] = ""}
+
+  else if(genusSpecies == ""){
+    taxon_pft_lookup$pft[i] = ""
+    taxon_pft_lookup$commonName[i] = ""}
+
+  else if(genusSpecies == ""){
+    taxon_pft_lookup$pft[i] = ""
+    taxon_pft_lookup$commonName[i] = ""}
+
+  else if(genusSpecies == ""){
+    taxon_pft_lookup$pft[i] = ""
+    taxon_pft_lookup$commonName[i] = ""}
+
+  else if(genusSpecies == ""){
+    taxon_pft_lookup$pft[i] = ""
+    taxon_pft_lookup$commonName[i] = ""}
+
+  else if(genusSpecies == ""){
+    taxon_pft_lookup$pft[i] = ""
+    taxon_pft_lookup$commonName[i] = ""}
+
+  else if(genusSpecies == ""){
+    taxon_pft_lookup$pft[i] = ""
+    taxon_pft_lookup$commonName[i] = ""}
 
@@
 
-  else if(genusSpecies == "Mahonia nervosa"){
-    taxon_pft_lookup$pft[i] = "woody_shrub"      # evergreen shrub
-    taxon_pft_lookup$commonName[i] = "Cascade barberry"}
-
   else if(genusSpecies == "Populus tremuloides"){
@@
   else if(genusSpecies == "Rhamnus davurica"){
-    taxon_pft_lookup$pft[i] = "woody_shrub"
+    taxon_pft_lookup$pft[i] = "???"              ???
     taxon_pft_lookup$commonName[i] = "Dahurian buckthorn"}
@@
   else if(genusSpecies == "Prunus serotina"){
-    taxon_pft_lookup$pft[i] = "deciduous_forest"
-    taxon_pft_lookup$commonName[i] = "black cherrt"}
+    taxon_pft_lookup$pft[i] = "???"       ???
+    taxon_pft_lookup$commonName[i] = "black cherry"}
 
@@
     taxon_pft_lookup$commonName[i] = "white oak"}
-
-  else if(genusSpecies == "Juglans nigra"){
-    taxon_pft_lookup$pft[i] = "deciduous_forest"
-    taxon_pft_lookup$commonName[i] = "black walnut"}
 
@@
     taxon_pft_lookup$commonName[i] = "bee-bee tree"}
-
-  else if(genusSpecies == "Juniperus virginiana"){
-    taxon_pft_lookup$pft[i] = "coniferous_forest"
-    taxon_pft_lookup$commonName[i] = "red cedar"}
 

### 2021-03-29 R/add_plant_functional_type.Rmd
over 95 done

@@
 knitr::opts_chunk$set(echo = TRUE)
+```
+
+```{r message=FALSE}
+library(tidyverse)
+library(here)
+library(dplyr)
 ```
@@
                                commonName = NA,
-                               pft = NA)
+                               pft = NA) %>% 
+  # arrange alphabetically
+  dplyr::arrange(genusSpecies)
 
@@
   else if(genusSpecies == "Acer circinatum"){
-    taxon_pft_lookup$pft[i] = "NA " ??? 
+    taxon_pft_lookup$pft[i] = "???" # ???
     taxon_pft_lookup$commonName[i] = "vine maple"}
@@
   else if(genusSpecies == "Amelanchier laevis"){
-    taxon_pft_lookup$pft[i] = "???" ???
+    taxon_pft_lookup$pft[i] = "???" # ???
     taxon_pft_lookup$commonName[i] = "Allegheny serviceberry"}
@@
   else if(genusSpecies == "Amyris elemifera"){
-    taxon_pft_lookup$pft[i] = "???"     ???
+    taxon_pft_lookup$pft[i] = "???"     # ???
     taxon_pft_lookup$commonName[i] = "sea torchwood"}
@@
   else if(genusSpecies == "Bourreria sp."){
-    taxon_pft_lookup$pft[i] = "???"   ???
+    taxon_pft_lookup$pft[i] = "???"   # ???
     taxon_pft_lookup$commonName[i] = "strongbark"}
@@
   else if(genusSpecies == "Cercis canadensis"){
-    taxon_pft_lookup$pft[i] = "???"   ???
+    taxon_pft_lookup$pft[i] = "???"   # ???
     taxon_pft_lookup$commonName[i] = "eastern redbud"}
@@
   else if(genusSpecies == "Cornus drummondii"){
-    taxon_pft_lookup$pft[i] = "???"    ???
+    taxon_pft_lookup$pft[i] = "???"    # ???
     taxon_pft_lookup$commonName[i] = "roughleaf dogwood"}
@@
   else if(genusSpecies == "Cornus florida"){
-    taxon_pft_lookup$pft[i] = "???"    ???
+    taxon_pft_lookup$pft[i] = "???"    # ???
     taxon_pft_lookup$commonName[i] = "flowering dogwood"}
@@
   else if(genusSpecies == "Frangula purshiana"){
-    taxon_pft_lookup$pft[i] = "???"     ???
+    taxon_pft_lookup$pft[i] = "???"     # ???
     taxon_pft_lookup$commonName[i] = "Cascara buckthorn"}
@@
   else if(genusSpecies == "Ilex montana"){
-    taxon_pft_lookup$pft[i] = "???"          ???
+    taxon_pft_lookup$pft[i] = "???"          # ???
     taxon_pft_lookup$commonName[i] = "mountain holly"}
@@
   else if(genusSpecies == "Ilex opaca"){
-    taxon_pft_lookup$pft[i] = "woody_shrub"   ???      # broadleaved evergreen tree
+    taxon_pft_lookup$pft[i] = "woody_shrub"   # ???      # broadleaved evergreen tree
     taxon_pft_lookup$commonName[i] = "American holly"}
@@
 
+ else if(genusSpecies == "Magnolia sp."){
+    taxon_pft_lookup$pft[i] = "???"
+    taxon_pft_lookup$commonName[i] = "magnolia"}
+
   else if(genusSpecies == "Mahonia nervosa"){
@@
   else if(genusSpecies == "Mangifera indica"){
-    taxon_pft_lookup$pft[i] = "???"       ???   # broadleaf evergreen tree
+    taxon_pft_lookup$pft[i] = "???"       # ???   # broadleaf evergreen tree
     taxon_pft_lookup$commonName[i] = "mango"}
@@
 
-  else if(genusSpecies == ""){
-    taxon_pft_lookup$pft[i] = ""
-    taxon_pft_lookup$commonName[i] = ""}
+  else if(genusSpecies == "Oxydendrum sp."){
+    taxon_pft_lookup$pft[i] = "deciduous_forest"
+    taxon_pft_lookup$commonName[i] = "sourwood"}
 
-  else if(genusSpecies == ""){
-    taxon_pft_lookup$pft[i] = ""
-    taxon_pft_lookup$commonName[i] = ""}
+  else if(genusSpecies == "Picea engelmannii"){
+    taxon_pft_lookup$pft[i] = "coniferous_forest"
+    taxon_pft_lookup$commonName[i] = "Engelmann spruce"}
 
-  else if(genusSpecies == ""){
-    taxon_pft_lookup$pft[i] = ""
-    taxon_pft_lookup$commonName[i] = ""}
+  else if(genusSpecies == "Picea glauca"){
+    taxon_pft_lookup$pft[i] = "coniferous_forest"
+    taxon_pft_lookup$commonName[i] = "white spruce"}
 
-  else if(genusSpecies == ""){
-    taxon_pft_lookup$pft[i] = ""
-    taxon_pft_lookup$commonName[i] = ""}
+  else if(genusSpecies == "Picea mariana"){
+    taxon_pft_lookup$pft[i] = "coniferous_forest"
+    taxon_pft_lookup$commonName[i] = "black spruce"}
 
-  else if(genusSpecies == ""){
-    taxon_pft_lookup$pft[i] = ""
-    taxon_pft_lookup$commonName[i] = ""}
+  else if(genusSpecies == "Picea sp."){
+    taxon_pft_lookup$pft[i] = "coniferous_forest"
+    taxon_pft_lookup$commonName[i] = "spruce"}
 
-  else if(genusSpecies == ""){
-    taxon_pft_lookup$pft[i] = ""
-    taxon_pft_lookup$commonName[i] = ""}
+  else if(genusSpecies == "Pinus contorta"){
+    taxon_pft_lookup$pft[i] = "coniferous_forest"
+    taxon_pft_lookup$commonName[i] = "lodgepole pine"}
+
+  else if(genusSpecies == "Pinus echinata"){
+    taxon_pft_lookup$pft[i] = "coniferous_forest"
+    taxon_pft_lookup$commonName[i] = "shortleaf pine"}
+
+  else if(genusSpecies == "Pinus flexilis"){
+    taxon_pft_lookup$pft[i] = "coniferous_forest"
+    taxon_pft_lookup$commonName[i] = "limber pine"}
+
+  else if(genusSpecies == "Pinus palustris"){
+    taxon_pft_lookup$pft[i] = "coniferous_forest"
+    taxon_pft_lookup$commonName[i] = "longleaf pine"}
+
+  else if(genusSpecies == "Pinus ponderosa"){
+    taxon_pft_lookup$pft[i] = "coniferous_forest"
+    taxon_pft_lookup$commonName[i] = "ponderosa pine"}
+
+  else if(genusSpecies == "Pinus sp."){
+    taxon_pft_lookup$pft[i] = "coniferous_forest"
+    taxon_pft_lookup$commonName[i] = "pine"}
+
+  else if(genusSpecies == "Pinus strobus"){
+    taxon_pft_lookup$pft[i] = "coniferous_forest"
+    taxon_pft_lookup$commonName[i] = "eastern white pine"}
+
+  else if(genusSpecies == "Pinus taeda"){
+    taxon_pft_lookup$pft[i] = "coniferous_forest"
+    taxon_pft_lookup$commonName[i] = "loblolly pine"}
 
@@
 
-  else if(genusSpecies == "Picea sp."){
-    taxon_pft_lookup$pft[i] = "coniferous_forest"
-    taxon_pft_lookup$commonName[i] = "spruce"}
-
   else if(genusSpecies == "Rhamnus davurica"){
-    taxon_pft_lookup$pft[i] = "???"              ???
+    taxon_pft_lookup$pft[i] = "???"              # ???
     taxon_pft_lookup$commonName[i] = "Dahurian buckthorn"}
@@
   else if(genusSpecies == "Prunus serotina"){
-    taxon_pft_lookup$pft[i] = "???"       ???
+    taxon_pft_lookup$pft[i] = "???"       # ???
     taxon_pft_lookup$commonName[i] = "black cherry"}
@@
 
-  else if(genusSpecies == "Pinus strobus"){
-    taxon_pft_lookup$pft[i] = "coniferous_forest"
-    taxon_pft_lookup$commonName[i] = "eastern white pine"}
-
   else if(genusSpecies == "Tetradium daniellii"){
@@
     taxon_pft_lookup$commonName[i] = "bee-bee tree"}
-
-  else if(genusSpecies == "Picea mariana"){
-    taxon_pft_lookup$pft[i] = "coniferous_forest"
-    taxon_pft_lookup$commonName[i] = "black spruce"}
-
-  else if(genusSpecies == "Picea glauca"){
-    taxon_pft_lookup$pft[i] = "coniferous_forest"
-    taxon_pft_lookup$commonName[i] = "white spruce"}
 
@@
     taxon_pft_lookup$commonName[i] = "post oak"}
-
-  else if(genusSpecies == ""){
-    taxon_pft_lookup$pft[i] = ""
-    taxon_pft_lookup$commonName[i] = ""}
-
-  else if(genusSpecies == ""){
-    taxon_pft_lookup$pft[i] = ""
-    taxon_pft_lookup$commonName[i] = ""}
 

### 2021-03-30 R/add_plant_functional_type.Rmd
120 done

@@
 
-  else if(genusSpecies == ""){
-    taxon_pft_lookup$pft[i] = ""
-    taxon_pft_lookup$commonName[i] = ""}
+  else if(genusSpecies == "Pisonia albida"){
+    taxon_pft_lookup$pft[i] = "???"    # Growth Form - Multiple Stem, Height, Mature - 30ft
+    taxon_pft_lookup$commonName[i] = "corcho bobo"}
+
+  else if(genusSpecies == "Platanus occidentalis"){
+    taxon_pft_lookup$pft[i] = "deciduous_forest"
+    taxon_pft_lookup$commonName[i] = "American sycamore"}
+
+  else if(genusSpecies == "Populus deltoides"){
+    taxon_pft_lookup$pft[i] = "deciduous_forest"
+    taxon_pft_lookup$commonName[i] = "eastern cottonwood"}
+
+  else if(genusSpecies == "Populus grandidentata"){
+    taxon_pft_lookup$pft[i] = "deciduous_forest"
+    taxon_pft_lookup$commonName[i] = "bigtooth aspen"}
+
+  else if(genusSpecies == "Populus sp."){
+    taxon_pft_lookup$pft[i] = "deciduous_forest"
+    taxon_pft_lookup$commonName[i] = "cottonwood"}
+
+  else if(genusSpecies == "Populus tremuloides"){
+    taxon_pft_lookup$pft[i] = "deciduous_forest"
+    taxon_pft_lookup$commonName[i] = "quaking aspen"}
+
+  else if(genusSpecies == "Prosopis glandulosa"){
+    taxon_pft_lookup$pft[i] = "???"   # shrub or tree ranging from 4 to 6m tall
+    taxon_pft_lookup$commonName[i] = "honey mesquite"}
+
+  else if(genusSpecies == "Prosopis velutina"){
+    taxon_pft_lookup$pft[i] = "???" # deciduous shrub or tree is less than 15 m
+    taxon_pft_lookup$commonName[i] = "velvet mesquite"}
+
+  else if(genusSpecies == "Prunus americana"){
+    taxon_pft_lookup$pft[i] = "???" # deciduous large shrub or small tree, reaching heights up to 15 feet
+    taxon_pft_lookup$commonName[i] = "American plum"}
+
+  else if(genusSpecies == "Prunus angustifolia"){
+    taxon_pft_lookup$pft[i] = "woody_shrub" # short, thickly branched shrub 3 to 10ft tall
+    taxon_pft_lookup$commonName[i] = "Chickasaw plum"}
+
+  else if(genusSpecies == "Prunus mexicana"){
+    taxon_pft_lookup$pft[i] = "???"
+    taxon_pft_lookup$commonName[i] = "Mexican plum"}
+
+  else if(genusSpecies == "Prunus serotina"){
+    taxon_pft_lookup$pft[i] = "???"       
+    taxon_pft_lookup$commonName[i] = "black cherry"}
+
+  else if(genusSpecies == "Pseudotsuga menziesii"){
+    taxon_pft_lookup$pft[i] = "coniferous_forest"
+    taxon_pft_lookup$commonName[i] = "Douglas-fir"}
+
+  else if(genusSpecies == "Quercus alba"){
+    taxon_pft_lookup$pft[i] = "deciduous_forest"
+    taxon_pft_lookup$commonName[i] = "white oak"}
+
+  else if(genusSpecies == "Quercus coccinea"){
+    taxon_pft_lookup$pft[i] = "deciduous_forest"
+    taxon_pft_lookup$commonName[i] = "scarlet oak"}
+
+  else if(genusSpecies == "Quercus falcata"){
+    taxon_pft_lookup$pft[i] = "deciduous_forest"
+    taxon_pft_lookup$commonName[i] = "southern red oak"}
+
+  else if(genusSpecies == "Quercus geminata"){
+    taxon_pft_lookup$pft[i] = "deciduous_forest"
+    taxon_pft_lookup$commonName[i] = "sand live oak"}
+
+  else if(genusSpecies == "Quercus hemisphaerica"){
+    taxon_pft_lookup$pft[i] = "deciduous_forest"
+    taxon_pft_lookup$commonName[i] = "Darlington oak"}
+
+  else if(genusSpecies == "Quercus laevis"){
+    taxon_pft_lookup$pft[i] = "deciduous_forest"
+    taxon_pft_lookup$commonName[i] = "turkey oak"}
+
+  else if(genusSpecies == "Quercus laurifolia"){
+    taxon_pft_lookup$pft[i] = "deciduous_forest"
+    taxon_pft_lookup$commonName[i] = "laurel oak"}
+
+  else if(genusSpecies == "Quercus lyrata"){
+    taxon_pft_lookup$pft[i] = "deciduous_forest"
+    taxon_pft_lookup$commonName[i] = "overcup oak"}
+
+  else if(genusSpecies == "Quercus macrocarpa"){
+    taxon_pft_lookup$pft[i] = "deciduous_forest"
+    taxon_pft_lookup$commonName[i] = "bur oak"}
+
+  else if(genusSpecies == "Quercus margaretta"){
+    taxon_pft_lookup$pft[i] = "deciduous_forest"
+    taxon_pft_lookup$commonName[i] = "sand post oak"}
+
+  else if(genusSpecies == "Quercus marilandica"){
+    taxon_pft_lookup$pft[i] = "deciduous_forest"
+    taxon_pft_lookup$commonName[i] = "blackjack oak"}
+
+  else if(genusSpecies == "Quercus michauxii"){
+    taxon_pft_lookup$pft[i] = "deciduous_forest"
+    taxon_pft_lookup$commonName[i] = "swamp chestnut oak"}
+
+  else if(genusSpecies == "Quercus montana"){
+    taxon_pft_lookup$pft[i] = "deciduous_forest"
+    taxon_pft_lookup$commonName[i] = "chestnut oak"}
+
+  else if(genusSpecies == "Quercus muehlenbergii"){
+    taxon_pft_lookup$pft[i] = "deciduous_forest"
+    taxon_pft_lookup$commonName[i] = "chinquapin oak"}
+
+  else if(genusSpecies == "Quercus nigra"){
+    taxon_pft_lookup$pft[i] = "deciduous_forest"
+    taxon_pft_lookup$commonName[i] = "water oak"}
+
+  else if(genusSpecies == "Quercus pagoda"){
+    taxon_pft_lookup$pft[i] = "deciduous_forest"
+    taxon_pft_lookup$commonName[i] = "cherrybark oak"}
+
+  else if(genusSpecies == "Quercus phellos"){
+    taxon_pft_lookup$pft[i] = "deciduous_forest"
+    taxon_pft_lookup$commonName[i] = "willow oak"}
+
+  else if(genusSpecies == "Quercus rubra"){
+    taxon_pft_lookup$pft[i] = "deciduous_forest"
+    taxon_pft_lookup$commonName[i] = "northern red oak"}
 
@@
 
-  else if(genusSpecies == "Pseudotsuga menziesii"){
-    taxon_pft_lookup$pft[i] = "coniferous_forest"
-    taxon_pft_lookup$commonName[i] = "Douglas-fir"}
-
   else if(genusSpecies == "Vaccinium parvifolium"){ 
@@
     taxon_pft_lookup$commonName[i] = "red huckleberry"}
-
-  else if(genusSpecies == "Populus tremuloides"){
-    taxon_pft_lookup$pft[i] = "deciduous_forest"
-    taxon_pft_lookup$commonName[i] = "quaking aspen"}
 
@@
     taxon_pft_lookup$commonName[i] = "Amur honeysuckle"}
-
-  else if(genusSpecies == "Prunus serotina"){
-    taxon_pft_lookup$pft[i] = "???"       # ???
-    taxon_pft_lookup$commonName[i] = "black cherry"}
-
-  else if(genusSpecies == "Quercus alba"){
-    taxon_pft_lookup$pft[i] = "deciduous_forest"
-    taxon_pft_lookup$commonName[i] = "white oak"}
 

### 2021-03-30 R/add_plant_functional_type.Rmd
labels added to each genusSpecies category in a lookup table. tagged uncertain PFTs as ???. once verified, need to apply the table to the spectral library

@@
 
-  else if(genusSpecies == ""){
-    taxon_pft_lookup$pft[i] = ""
-    taxon_pft_lookup$commonName[i] = ""}
+  else if(genusSpecies == "Quercus shumardii"){
+    taxon_pft_lookup$pft[i] = "deciduous_forest"
+    taxon_pft_lookup$commonName[i] = "Shumard's oak"}
 
-  else if(genusSpecies == ""){
-    taxon_pft_lookup$pft[i] = ""
-    taxon_pft_lookup$commonName[i] = ""}
+  else if(genusSpecies == "Quercus sp."){
+    taxon_pft_lookup$pft[i] = "???"  # apparently there are evergreen oak species? 
+    taxon_pft_lookup$commonName[i] = "oak"}
 
-  else if(genusSpecies == ""){
-    taxon_pft_lookup$pft[i] = ""
-    taxon_pft_lookup$commonName[i] = ""}
+  else if(genusSpecies == "Quercus stellata"){
+    taxon_pft_lookup$pft[i] = "deciduous_forest"
+    taxon_pft_lookup$commonName[i] = "post oak"}
 
-  else if(genusSpecies == ""){
-    taxon_pft_lookup$pft[i] = ""
-    taxon_pft_lookup$commonName[i] = ""}
+  else if(genusSpecies == "Quercus velutina"){
+    taxon_pft_lookup$pft[i] = "deciduous_forest"
+    taxon_pft_lookup$commonName[i] = "black oak"}
 
-  else if(genusSpecies == ""){
-    taxon_pft_lookup$pft[i] = ""
-    taxon_pft_lookup$commonName[i] = ""}
+  else if(genusSpecies == "Quercus virginiana"){
+    taxon_pft_lookup$pft[i] = "deciduous_forest"
+    taxon_pft_lookup$commonName[i] = "live oak"}
+
+  else if(genusSpecies == "Rhamnus davurica"){
+    taxon_pft_lookup$pft[i] = "???"     #  spreading shrub or small tree reaching 25 to 30 feet
+    taxon_pft_lookup$commonName[i] = "Dahurian buckthorn"}
+
+  else if(genusSpecies == "Rhus aromatica"){
+    taxon_pft_lookup$pft[i] = "woody_shrub"
+    taxon_pft_lookup$commonName[i] = "fragrant sumac"}
+
+  else if(genusSpecies == "Rhus glabra"){
+    taxon_pft_lookup$pft[i] = "woody_shrub"
+    taxon_pft_lookup$commonName[i] = "smooth sumac"}
+
+  else if(genusSpecies == "Robinia pseudoacacia"){
+    taxon_pft_lookup$pft[i] = "deciduous_forest"
+    taxon_pft_lookup$commonName[i] = "black locust"}
+
+  else if(genusSpecies == "Sassafras albidum"){
+    taxon_pft_lookup$pft[i] = "deciduous_forest" # deciduous shrub or tree. trees are 9 to 18 m
+    taxon_pft_lookup$commonName[i] = "sassafras"}
+
+  else if(genusSpecies ==  "Sassafras sp."){
+    taxon_pft_lookup$pft[i] = "deciduous_forest"
+    taxon_pft_lookup$commonName[i] = "sassafras"}
+
+  else if(genusSpecies == "Sideroxylon lanuginosum"){
+    taxon_pft_lookup$pft[i] = "???" # deciduous multi-trunked tall shrub or medium-sized tree to 48ft or more
+    taxon_pft_lookup$commonName[i] = "gum bully"}
+
+  else if(genusSpecies == "Symplocos sp."){
+    taxon_pft_lookup$pft[i] = "???"  # "deciduous or evergreen shrub or tree"
+    taxon_pft_lookup$commonName[i] = "sweetleaf"}
+
+  else if(genusSpecies == "Taxus brevifolia"){
+    taxon_pft_lookup$pft[i] = "coniferous_forest"
+    taxon_pft_lookup$commonName[i] = "Pacific yew"}
+
+  else if(genusSpecies == "Tetradium daniellii"){
+    taxon_pft_lookup$pft[i] = "deciduous_forest"
+    taxon_pft_lookup$commonName[i] = "bee-bee tree"}
+
+  else if(genusSpecies == "Triadica sebifera"){
+    taxon_pft_lookup$pft[i] = "deciduous_forest"
+    taxon_pft_lookup$commonName[i] = "Chinese tallow"}
+
+  else if(genusSpecies == "Tsuga canadensis"){
+    taxon_pft_lookup$pft[i] = "coniferous_forest"
+    taxon_pft_lookup$commonName[i] = "eastern hemlock"}
+
+  else if(genusSpecies == "Tsuga heterophylla"){
+    taxon_pft_lookup$pft[i] = "coniferous_forest"
+    taxon_pft_lookup$commonName[i] = "western hemlock"}
+
+  else if(genusSpecies == "Ulmus alata"){
+    taxon_pft_lookup$pft[i] = "deciduous_forest"
+    taxon_pft_lookup$commonName[i] = "winged elm"}
+
+  else if(genusSpecies == "Ulmus americana"){
+    taxon_pft_lookup$pft[i] = "deciduous_forest"
+    taxon_pft_lookup$commonName[i] = "American elm"}
+
+  else if(genusSpecies == "Ulmus crassifolia"){
+    taxon_pft_lookup$pft[i] = "deciduous_forest"
+    taxon_pft_lookup$commonName[i] = "cedar elm"}
+
+  else if(genusSpecies == "Ulmus rubra"){
+    taxon_pft_lookup$pft[i] = "deciduous_forest"
+    taxon_pft_lookup$commonName[i] = "slippery elm"}
+
+  else if(genusSpecies == "Ulmus sp."){
+    taxon_pft_lookup$pft[i] = "deciduous_forest"
+    taxon_pft_lookup$commonName[i] = "elm"}
 
@@
 
-  else if(genusSpecies == "Tsuga canadensis"){
-    taxon_pft_lookup$pft[i] = "coniferous_forest"
-    taxon_pft_lookup$commonName[i] = "eastern hemlock"}
+  else if(genusSpecies == "Zinnia acerosa"){
+    taxon_pft_lookup$pft[i] = "???" # forb/herb, subshrub - grass_herb?
+    taxon_pft_lookup$commonName[i] = "desert zinnia"}
 
-  else if(genusSpecies == "Rhamnus davurica"){
-    taxon_pft_lookup$pft[i] = "???"              # ???
-    taxon_pft_lookup$commonName[i] = "Dahurian buckthorn"}
-
-  else if(genusSpecies == "Lonicera maackii"){
-    taxon_pft_lookup$pft[i] = "woody_shrub"
-    taxon_pft_lookup$commonName[i] = "Amur honeysuckle"}
-
-  else if(genusSpecies == "Robinia pseudoacacia"){
-    taxon_pft_lookup$pft[i] = "deciduous_forest"
-    taxon_pft_lookup$commonName[i] = "black locust"}
-
-  else if(genusSpecies == "Tetradium daniellii"){
-    taxon_pft_lookup$pft[i] = "deciduous_forest"
-    taxon_pft_lookup$commonName[i] = "bee-bee tree"}
-
-  else if(genusSpecies == "Ulmus crassifolia"){
-    taxon_pft_lookup$pft[i] = "deciduous_forest"
-    taxon_pft_lookup$commonName[i] = "cedar elm"}
-
-  else if(genusSpecies == "Quercus stellata"){
-    taxon_pft_lookup$pft[i] = "deciduous_forest"
-    taxon_pft_lookup$commonName[i] = "post oak"}
-
+  # For taxons not described in this code, set their PFT to unknown
   else print("PFT unknown")

### 2021-03-30 R/add_plant_functional_type.Rmd
display table and write to csv

@@
 library(dplyr)
+library(kableExtra)
 ```
@@
   # print row and current genusSpecies label
-  print(i)
+  #print(i)
   genusSpecies <- taxon_pft_lookup$genusSpecies[i]
-  print(genusSpecies)
+  #print(genusSpecies)
 
@@
 
-  # For taxons not described in this code, set their PFT to unknown
-  else print("PFT unknown")
+  # For taxons not described in this code
+  #else #print("PFT unknown")
 
-  # print corresponding PFT
-  print(taxon_pft_lookup$pft[i])
 }
 
+```
+
+```{r}
+kableExtra::kable(taxon_pft_lookup) %>% 
+  kableExtra::kable_styling(bootstrap_options = "striped")
+```
+
+```{r}
+# write the lookup table to a csv 
+write_csv(x = taxon_pft_lookup,
+          file = here::here("R","output","taxon_pft_lookup.csv"))
 ```

### 2021-04-05 R/add_plant_functional_type.Rmd
working on merging growthForm data to spectra by uid

@@
 
-## Manually add PFTs to spectral library
-
 ```{r load_cleaned_spectra, message=FALSE}
@@
 ```
+
+## Use NEON growthForm observations
+
+```{r}
+library(neonUtilities)
+```
+
+```{r}
+# download woody vegetation structure data for all sites and years 
+out_dir <- here::here("analysis","data","raw_data")
+
+neonUtilities::zipsByProduct(dpID = "DP1.10098.001",
+              savepath = out_dir,
+              check.size = FALSE)
+
+stackByTable(file.path(out_dir, "filesToStack10098"), folder = TRUE)
+```
+
+```{r}
+# read the vst_apparentindividual table, which has growthForm data
+
+vst_apparentindividual <- read_csv(out_dir <- here::here("analysis","data","raw_data", "filesToStack10098",
+                                                         "stackedFiles", "vst_apparentindividual.csv"))
+
+growthForms <- vst_apparentindividual %>% 
+  select(uid, growthForm)
+
+# merge the growthForm column into the spectra data by uid 
+cleaned_spectra_merged <- merge(cleaned_spectra, growthForms, by = "uid", all.x=TRUE)
+
+```
+
+```{r}
+# summarize the downloaded growthform data to see which growth forms are assigned to each species 
+growthForms %>% 
+    distinct(uid, .keep_all = TRUE) %>% 
+    group_by(growthForm) %>% 
+    tally()
+
+# summarize the merged growthform data
+
+```
+
+## Manually add PFTs to spectral library
 

### 2021-04-05 R/add_plant_functional_type.Rmd
adam says zinnia is grass_herb
joining by individualID seems to work better than uid

@@
 growthForms <- vst_apparentindividual %>% 
-  select(uid, growthForm)
+  select(individualID, growthForm)
 
-# merge the growthForm column into the spectra data by uid 
-cleaned_spectra_merged <- merge(cleaned_spectra, growthForms, by = "uid", all.x=TRUE)
+# merge the growthForm column into the spectra data
+cleaned_spectra_merged <- right_join(cleaned_spectra, growthForms, by = "individualID")
 
@@
 # summarize the merged growthform data
-
+cleaned_spectra_merged %>% 
+  distinct(spectraID, .keep_all = TRUE) %>% 
+  group_by(growthForm) %>% 
+    tally() 
 ```
@@
   else if(genusSpecies == "Zinnia acerosa"){
-    taxon_pft_lookup$pft[i] = "???" # forb/herb, subshrub - grass_herb?
+    taxon_pft_lookup$pft[i] = "grass_herb" # forb/herb, subshrub - grass_herb?
     taxon_pft_lookup$commonName[i] = "desert zinnia"}

### 2021-04-05 R/add_plant_functional_type.Rmd
keeping height values with each spectrum. need to figure out how to simplify the growthForm values, just keep last word of each

--- 2021-03-30 R/add_plant_functional_type.Rmd
@@
 ```
+
+## Use NEON growthForm observations
+
+```{r}
+library(neonUtilities)
+```
+
+```{r}
+# download woody vegetation structure data for all sites and years 
+out_dir <- here::here("analysis","data","raw_data")
+
+neonUtilities::zipsByProduct(dpID = "DP1.10098.001",
+              savepath = out_dir,
+              check.size = FALSE)
+
+stackByTable(file.path(out_dir, "filesToStack10098"), folder = TRUE)
+```
+
+```{r}
+# read the vst_apparentindividual table, which has growthForm data
+vst_apparentindividual <- read_csv(out_dir <- here::here("analysis","data","raw_data", "filesToStack10098",
+                                                         "stackedFiles", "vst_apparentindividual.csv"))
+
+# select the growthForm and height columns
+growthForms <- vst_apparentindividual %>% 
+  select(individualID, growthForm, height)
+
+# join the growthForm and height columns into the spectra data frame
+cleaned_spectra_merged <- right_join(cleaned_spectra, growthForms, by = "individualID")
+
+# print all unique growthForm values 
+unique(cleaned_spectra_merged$growthForm)
+
+# simplify the growth form into "tree" or "shrub" 
+cleaned_spectra_merged <- cleaned_spectra_merged %>%
+
+```
+
+```{r}
+# summarize the merged growthform data
+cleaned_spectra_merged %>% 
+  distinct(spectraID, .keep_all = TRUE) %>% 
+  group_by(genusSpecies, growthForm) %>% 
+    tally() 
+```
+
+## Manually add PFTs to spectral library
 

### 2021-04-05 R/add_plant_functional_type.Rmd
extracted the last word from each growth form

@@
 
-# simplify the growth form into "tree" or "shrub" 
+# simplify the growth form into "tree", "shrub", "saplng" etc.  
 cleaned_spectra_merged <- cleaned_spectra_merged %>% 
+  rowwise() %>%
+  # parse out the last word in each entry's growthForm
+  mutate(growthForm_simple = rev(strsplit(growthForm, " ")[[1]])[1])
 
@@
   distinct(spectraID, .keep_all = TRUE) %>% 
-  group_by(genusSpecies, growthForm) %>% 
+  group_by(genusSpecies, growthForm_simple) %>% 
     tally() 

### 2021-04-05 R/add_plant_functional_type.Rmd
make parsing code more efficient?

@@
   # parse out the last word in each entry's growthForm
+  # make this code more efficient????? 
   mutate(growthForm_simple = rev(strsplit(growthForm, " ")[[1]])[1])

### 2021-04-21 R/add_plant_functional_type.Rmd
going to manually check each taxon now compared with the NEON growth form observations

@@
 
-# simplify the growth form into "tree", "shrub", "saplng" etc.  
+# simplify the growth form into "tree", "shrub", "sapling" etc.  
 cleaned_spectra_merged <- cleaned_spectra_merged %>% 
@@
   distinct(spectraID, .keep_all = TRUE) %>% 
+  group_by(genusSpecies, growthForm_simple) %>% 
+    tally()
+
+# summarize the merged growthform data
+cleaned_spectra_merged %>% 
+  distinct(spectraID, .keep_all = TRUE) %>% 
+  # keep just the tree and shrub growth forms, remove rows with "sampling" and "liana"
+  filter(growthForm_simple %in% c("tree","shrub")) %>% 
   group_by(genusSpecies, growthForm_simple) %>% 

### 2021-04-22 R/add_plant_functional_type.Rmd
up

--- 2021-04-05 R/add_plant_functional_type.Rmd
@@
 
-```{r message=FALSE}
+```{r load_packages, message=FALSE}
 library(tidyverse)
@@
 library(kableExtra)
+library(neonUtilities)
 ```
 
-Plant functional types (PFTs) of interest: coniferous forest, deciduous forest, woody shrub, grass/herb, and bare ground.
+Plant functional types (PFTs) originally proposed: coniferous forest, deciduous forest, woody shrub, grass/herb, and bare ground.
+
+Proposing to update the PFTs based on the MODIS land cover types: Evergreen Needleleaf
+Forests, Evergreen Broadleaf Forests, Deciduous Needleleaf Forests, Deciduous Broadleaf
+Forests, Shrub.
 
 ```{r load_cleaned_spectra, message=FALSE}
+# start timer 
+start_time <- Sys.time()
+
 # read cleaned spectral library file 
@@
 
-```{r}
-library(neonUtilities)
-```
-
-```{r}
+Run this code chunk to download the NEON woody vegetation structure data: 
+```{r download_neon_veg_data, warning=FALSE, message=FALSE, eval=FALSE}
 # download woody vegetation structure data for all sites and years 
@@
 
-```{r}
+```{r read_merge_simplify_growthforms}
 # read the vst_apparentindividual table, which has growthForm data
 vst_apparentindividual <- read_csv(out_dir <- here::here("analysis","data","raw_data", "filesToStack10098",
-                                                         "stackedFiles", "vst_apparentindividual.csv"))
+                                                         "stackedFiles", "vst_apparentindividual.csv"), 
+                                   col_types = cols())
 
@@
   mutate(growthForm_simple = rev(strsplit(growthForm, " ")[[1]])[1])
-
 ```
 
-```{r}
+```{r summary_table_taxon_growthforms}
 # summarize the merged growthform data
@@
     tally()
+
+# summarize the merged growthform data with "tree" and "shrub" growth forms
+cleaned_spectra_merged %>% 
+  distinct(spectraID, .keep_all = TRUE) %>% 
+  # keep just the tree and shrub growth forms, remove rows with "sampling" and "liana"
+  filter(growthForm_simple %in% c("tree","shrub")) %>% 
+  group_by(genusSpecies, growthForm_simple) %>% 
+    tally() %>% 
+  # display as a nicely formatted table
+  kableExtra::kable() %>% 
+  kableExtra::kable_styling(bootstrap_options = "striped", 
+                            full_width = F, 
+                            position = "left")
+
 ```
@@
 
-```{r}
+This code manually adds labels to each taxon (genus and species listed here) for:
+
+- common name
+- PFT (based on Macrosystems labels)
+- seasonality (evergreen or deciduous)
+- leaf type (broadleaf or needleleaf)
+- growth form (shrub or forest)
+
+```{r manually_add_pft_labels}
 # list unique taxon labels, add an empty columns to fill with plant functional type (PFT) and common name labels.
@@
                                commonName = NA,
-                               pft = NA) %>% 
+                               pft = NA,
+                               seasonality = NA,
+                               leafType = NA,
+                               growthForm = NA) %>% 
   # arrange alphabetically
@@
     taxon_pft_lookup$pft[i] = "coniferous_forest" 
+    taxon_pft_lookup$seasonality[i] = "evergreen"
+    taxon_pft_lookup$leafType[i] = "needleleaf"
+    taxon_pft_lookup$growthForm[i] = "forest"
     taxon_pft_lookup$commonName[i] = "Pacific silver fir"}
@@
     taxon_pft_lookup$pft[i] = "coniferous_forest" 
+    taxon_pft_lookup$seasonality[i] = "evergreen"
+    taxon_pft_lookup$leafType[i] = "needleleaf"
+    taxon_pft_lookup$growthForm[i] = "forest"
     taxon_pft_lookup$commonName[i] = "balsam fir"}
@@
     taxon_pft_lookup$pft[i] = "coniferous_forest"
+    taxon_pft_lookup$seasonality[i] = "evergreen"
+    taxon_pft_lookup$leafType[i] = "needleleaf"
+    taxon_pft_lookup$growthForm[i] = "forest"
     taxon_pft_lookup$commonName[i] = "subalpine fir"}
@@
     taxon_pft_lookup$pft[i] = "woody_shrub" 
+    taxon_pft_lookup$seasonality[i] = "deciduous"
+    taxon_pft_lookup$leafType[i] = ""
+    taxon_pft_lookup$growthForm[i] = "shrub"
     taxon_pft_lookup$commonName[i] = "catclaw acacia"}
@@
     taxon_pft_lookup$pft[i] = "deciduous_forest"
+    taxon_pft_lookup$seasonality[i] = "deciduous"
+    taxon_pft_lookup$leafType[i] = "broadleaf"
+    taxon_pft_lookup$growthForm[i] = "forest"
     taxon_pft_lookup$commonName[i] = "Southern sugar maple"}
@@
   else if(genusSpecies == "Acer circinatum"){
-    taxon_pft_lookup$pft[i] = "???" # ???
+    taxon_pft_lookup$pft[i] = "deciduous_forest" 
+    taxon_pft_lookup$seasonality[i] = "deciduous"
+    taxon_pft_lookup$leafType[i] = "broadleaf"
+    taxon_pft_lookup$growthForm[i] = "forest"
     taxon_pft_lookup$commonName[i] = "vine maple"}
@@
     taxon_pft_lookup$pft[i] = "deciduous_forest"
+    taxon_pft_lookup$seasonality[i] = "deciduous"
+    taxon_pft_lookup$leafType[i] = "broadleaf"
+    taxon_pft_lookup$growthForm[i] = "forest"
     taxon_pft_lookup$commonName[i] = "boxelder"}
@@
     taxon_pft_lookup$pft[i] = "deciduous_forest"
+    taxon_pft_lookup$seasonality[i] = "deciduous"
+    taxon_pft_lookup$leafType[i] = "broadleaf"
+    taxon_pft_lookup$growthForm[i] = "forest"
     taxon_pft_lookup$commonName[i] = "striped maple"}
@@
     taxon_pft_lookup$pft[i] = "deciduous_forest"
+    taxon_pft_lookup$seasonality[i] = "deciduous"
+    taxon_pft_lookup$leafType[i] = "broadleaf"
+    taxon_pft_lookup$growthForm[i] = "forest"
     taxon_pft_lookup$commonName[i] = "red maple"}
@@
     taxon_pft_lookup$pft[i] = "deciduous_forest"
+    taxon_pft_lookup$seasonality[i] = "deciduous"
+    taxon_pft_lookup$leafType[i] = "broadleaf"
+    taxon_pft_lookup$growthForm[i] = "forest"
     taxon_pft_lookup$commonName[i] = "silver maple"}
@@
     taxon_pft_lookup$pft[i] = "deciduous_forest"
+    taxon_pft_lookup$seasonality[i] = "deciduous"
+    taxon_pft_lookup$leafType[i] = "broadleaf"
+    taxon_pft_lookup$growthForm[i] = "forest"
     taxon_pft_lookup$commonName[i] = "sugar maple"}
@@
     taxon_pft_lookup$pft[i] = "deciduous_forest"
+    taxon_pft_lookup$seasonality[i] = "deciduous"
+    taxon_pft_lookup$leafType[i] = "broadleaf"
+    taxon_pft_lookup$growthForm[i] = "forest"
     taxon_pft_lookup$commonName[i] = "tree of heaven"}
@@
   else if(genusSpecies == "Alnus rubra"){
-    taxon_pft_lookup$pft[i] = "deciduous_forest"  # deciduous broadleaf tree 
+    taxon_pft_lookup$pft[i] = "deciduous_forest"
+    taxon_pft_lookup$seasonality[i] = "deciduous"
+    taxon_pft_lookup$leafType[i] = "broadleaf"
+    taxon_pft_lookup$growthForm[i] = "forest"
     taxon_pft_lookup$commonName[i] = "red alder"}
@@
     taxon_pft_lookup$pft[i] = "woody_shrub"
+    taxon_pft_lookup$seasonality[i] = "deciduous"
+    taxon_pft_lookup$leafType[i] = ""
+    taxon_pft_lookup$growthForm[i] = "shrub"
     taxon_pft_lookup$commonName[i] = "Wright's beebrush"}
@@
   else if(genusSpecies == "Amelanchier laevis"){
-    taxon_pft_lookup$pft[i] = "???" # ???
+    taxon_pft_lookup$pft[i] = "deciduous_forest"
+    taxon_pft_lookup$seasonality[i] = "deciduous"
+    taxon_pft_lookup$leafType[i] = "broadleaf"
+    taxon_pft_lookup$growthForm[i] = "forest"
     taxon_pft_lookup$commonName[i] = "Allegheny serviceberry"}
@@
   else if(genusSpecies == "Amyris elemifera"){
-    taxon_pft_lookup$pft[i] = "???"     # ???
+    taxon_pft_lookup$pft[i] = "???" # is not deciduous or coniferous forest
+    taxon_pft_lookup$seasonality[i] = "evergreen"
+    taxon_pft_lookup$leafType[i] = "broadleaf"
+    taxon_pft_lookup$growthForm[i] = "forest"
     taxon_pft_lookup$commonName[i] = "sea torchwood"}
@@
     taxon_pft_lookup$pft[i] = "woody_shrub"
+    taxon_pft_lookup$seasonality[i] = "evergreen"
+    taxon_pft_lookup$leafType[i] = ""
+    taxon_pft_lookup$growthForm[i] = "shrub"
     taxon_pft_lookup$commonName[i] = "big sagebrush"}
@@
     taxon_pft_lookup$pft[i] = "woody_shrub"
+    taxon_pft_lookup$seasonality[i] = "evergreen"
+    taxon_pft_lookup$leafType[i] = ""
+    taxon_pft_lookup$growthForm[i] = "shrub"
     taxon_pft_lookup$commonName[i] = "fourwing saltbush"}
@@
     taxon_pft_lookup$pft[i] = "woody_shrub"
+    taxon_pft_lookup$seasonality[i] = "evergreen"
+    taxon_pft_lookup$leafType[i] = ""
+    taxon_pft_lookup$growthForm[i] = "shrub"
     taxon_pft_lookup$commonName[i] = "shadscale saltbush"}
@@
     taxon_pft_lookup$pft[i] = "deciduous_forest"
+    taxon_pft_lookup$seasonality[i] = "deciduous"
+    taxon_pft_lookup$leafType[i] = "broadleaf"
+    taxon_pft_lookup$growthForm[i] = "forest"
     taxon_pft_lookup$commonName[i] = "yellow birch"}
@@
     taxon_pft_lookup$pft[i] = "deciduous_forest"
+    taxon_pft_lookup$seasonality[i] = "deciduous"
+    taxon_pft_lookup$leafType[i] = "broadleaf"
+    taxon_pft_lookup$growthForm[i] = "forest"
     taxon_pft_lookup$commonName[i] = "sweet birch"}
@@
     taxon_pft_lookup$pft[i] = "deciduous_forest"
+    taxon_pft_lookup$seasonality[i] = "deciduous"
+    taxon_pft_lookup$leafType[i] = "broadleaf"
+    taxon_pft_lookup$growthForm[i] = "forest"
     taxon_pft_lookup$commonName[i] = "resin birch"}
@@
     taxon_pft_lookup$pft[i] = "deciduous_forest"
+    taxon_pft_lookup$seasonality[i] = "deciduous"
+    taxon_pft_lookup$leafType[i] = "broadleaf"
+    taxon_pft_lookup$growthForm[i] = "forest"
     taxon_pft_lookup$commonName[i] = "river birch"}
@@
     taxon_pft_lookup$pft[i] = "deciduous_forest"
+    taxon_pft_lookup$seasonality[i] = "deciduous"
+    taxon_pft_lookup$leafType[i] = "broadleaf"
+    taxon_pft_lookup$growthForm[i] = "forest"
     taxon_pft_lookup$commonName[i] = "paper birch"}
@@
     taxon_pft_lookup$pft[i] = "deciduous_forest"
+    taxon_pft_lookup$seasonality[i] = "deciduous"
+    taxon_pft_lookup$leafType[i] = "broadleaf"
+    taxon_pft_lookup$growthForm[i] = "forest"
     taxon_pft_lookup$commonName[i] = "birch"}
@@
   else if(genusSpecies == "Bourreria sp."){
-    taxon_pft_lookup$pft[i] = "???"   # ???
+    taxon_pft_lookup$pft[i] = "???"   
+    taxon_pft_lookup$seasonality[i] = ""
+    taxon_pft_lookup$leafType[i] = ""
+    taxon_pft_lookup$growthForm[i] = "forest"
     taxon_pft_lookup$commonName[i] = "strongbark"}
@@
   else if(genusSpecies == "Bursera simaruba"){
-    taxon_pft_lookup$pft[i] = "deciduous_forest"  # "semi-deciduous"
+    taxon_pft_lookup$pft[i] = "deciduous_forest"  
+    taxon_pft_lookup$seasonality[i] = "deciduous"
+    taxon_pft_lookup$leafType[i] = ""
+    taxon_pft_lookup$growthForm[i] = "forest"
     taxon_pft_lookup$commonName[i] = "gumbo limbo"}
@@
 
-```{r}
+```{r taxon_pft_table}
 kableExtra::kable(taxon_pft_lookup) %>% 
@@
 
-```{r}
+```{r write_csv}
 # write the lookup table to a csv 
@@
 
-```{r}
+```{r usda_plants}
 # Set the base url
@@
 
-```{r, eval=FALSE, include=FALSE}
+```{r try_download, eval=FALSE, include=FALSE}
 # Specify URL where file is stored 
@@
 
-```{r install_load_packages_bien, message=FALSE, warning=FALSE}
+```{r load_packages_bien, message=FALSE, warning=FALSE}
 library(BIEN)
@@
 
-```{r}
+```{r bien_taxonomy_1}
 Abies_taxonomy<-BIEN_taxonomy_genus(genus = "Abies")
@@
 
-```{r}
+```{r bien_taxonomy_2}
 Acer_rubrum_taxonomy <- BIEN_taxonomy_species(species = "Acer rubrum")
@@
 https://www.rdocumentation.org/packages/taxize/versions/0.9.99
+
+Time the code. How long did this take? 
+```{r how_long_did_this_take}
+# stop timer
+end_time <- Sys.time()
+
+elapsed_time <- end_time - start_time
+
+print(difftime(end_time, start_time, 
+         units = c("auto", "secs", "mins", "hours",
+                   "days", "weeks")))
+```

### 2021-04-22 R/add_plant_functional_type.Rmd
adding columns for pft_MRA and pft_MODIS, renaming growth form from forest to tree to match the NEON observational data

@@
 Plant functional types (PFTs) originally proposed: coniferous forest, deciduous forest, woody shrub, grass/herb, and bare ground.
-
-Proposing to update the PFTs based on the MODIS land cover types: Evergreen Needleleaf
-Forests, Evergreen Broadleaf Forests, Deciduous Needleleaf Forests, Deciduous Broadleaf
-Forests, Shrub.
 
@@
 - common name
-- PFT (based on Macrosystems labels)
-- seasonality (evergreen or deciduous)
+- pft_MRA (based on Macrosystems proposed labels)
+    - coniferous_forest
+    - deciduous_forest
+    - woody_shrub
+- pft_MODIS (based on MODIS land cover classes)
+    - evergreen_needleleaf
+    - evergreen_broadleaf
+    - deciduous_needlelead
+    - deciduous_broadleaf
+    - shrub
+- phenology (evergreen or deciduous)
 - leaf type (broadleaf or needleleaf)
-- growth form (shrub or forest)
+- growth form (shrub or tree, based on NEON field based observations)
 
@@
 # list unique taxon labels, add an empty columns to fill with plant functional type (PFT) and common name labels.
-# PFTs of interest: coniferous_forest, deciduous_forest, woody_shrub, grass_herb, and bare_ground.
 taxon_pft_lookup <- data.frame(genusSpecies = unique(cleaned_spectra$genusSpecies),
                                commonName = NA,
-                               pft = NA,
-                               seasonality = NA,
+                               pft_MRA = NA,
+                               pft_MODIS = ,
+                               phenology = NA,
                                leafType = NA,
@@
   if(genusSpecies == "Abies amabilis"){
-    taxon_pft_lookup$pft[i] = "coniferous_forest" 
-    taxon_pft_lookup$seasonality[i] = "evergreen"
+    taxon_pft_lookup$pft_MRA[i] = "coniferous_forest" 
+    taxon_pft_lookup$pft_MODIS[i] = 
+    taxon_pft_lookup$phenology[i] = "evergreen"
     taxon_pft_lookup$leafType[i] = "needleleaf"
-    taxon_pft_lookup$growthForm[i] = "forest"
+    taxon_pft_lookup$growthForm[i] = "tree"
     taxon_pft_lookup$commonName[i] = "Pacific silver fir"}
@@
   else if(genusSpecies == "Abies balsamea"){
-    taxon_pft_lookup$pft[i] = "coniferous_forest" 
-    taxon_pft_lookup$seasonality[i] = "evergreen"
+    taxon_pft_lookup$pft_MRA[i] = "coniferous_forest" 
+    taxon_pft_lookup$phenology[i] = "evergreen"
     taxon_pft_lookup$leafType[i] = "needleleaf"
-    taxon_pft_lookup$growthForm[i] = "forest"
+    taxon_pft_lookup$growthForm[i] = "tree"
     taxon_pft_lookup$commonName[i] = "balsam fir"}
@@
   else if(genusSpecies == "Abies lasiocarpa"){
-    taxon_pft_lookup$pft[i] = "coniferous_forest"
-    taxon_pft_lookup$seasonality[i] = "evergreen"
+    taxon_pft_lookup$pft_MRA[i] = "coniferous_forest"
+    taxon_pft_lookup$phenology[i] = "evergreen"
     taxon_pft_lookup$leafType[i] = "needleleaf"
-    taxon_pft_lookup$growthForm[i] = "forest"
+    taxon_pft_lookup$growthForm[i] = "tree"
     taxon_pft_lookup$commonName[i] = "subalpine fir"}
@@
   else if(genusSpecies == "Acacia greggii"){
-    taxon_pft_lookup$pft[i] = "woody_shrub" 
-    taxon_pft_lookup$seasonality[i] = "deciduous"
+    taxon_pft_lookup$pft_MRA[i] = "woody_shrub" 
+    taxon_pft_lookup$phenology[i] = "deciduous"
     taxon_pft_lookup$leafType[i] = ""
@@
   else if(genusSpecies == "Acer barbatum"){
-    taxon_pft_lookup$pft[i] = "deciduous_forest"
-    taxon_pft_lookup$seasonality[i] = "deciduous"
+    taxon_pft_lookup$pft_MRA[i] = "deciduous_forest"
+    taxon_pft_lookup$phenology[i] = "deciduous"
     taxon_pft_lookup$leafType[i] = "broadleaf"
-    taxon_pft_lookup$growthForm[i] = "forest"
+    taxon_pft_lookup$growthForm[i] = "tree"
     taxon_pft_lookup$commonName[i] = "Southern sugar maple"}
@@
   else if(genusSpecies == "Acer circinatum"){
-    taxon_pft_lookup$pft[i] = "deciduous_forest" 
-    taxon_pft_lookup$seasonality[i] = "deciduous"
+    taxon_pft_lookup$pft_MRA[i] = "deciduous_forest" 
+    taxon_pft_lookup$phenology[i] = "deciduous"
     taxon_pft_lookup$leafType[i] = "broadleaf"
-    taxon_pft_lookup$growthForm[i] = "forest"
+    taxon_pft_lookup$growthForm[i] = "tree"
     taxon_pft_lookup$commonName[i] = "vine maple"}
@@
   else if(genusSpecies == "Acer negundo"){
-    taxon_pft_lookup$pft[i] = "deciduous_forest"
-    taxon_pft_lookup$seasonality[i] = "deciduous"
+    taxon_pft_lookup$pft_MRA[i] = "deciduous_forest"
+    taxon_pft_lookup$phenology[i] = "deciduous"
     taxon_pft_lookup$leafType[i] = "broadleaf"
-    taxon_pft_lookup$growthForm[i] = "forest"
+    taxon_pft_lookup$growthForm[i] = "tree"
     taxon_pft_lookup$commonName[i] = "boxelder"}
@@
   else if(genusSpecies == "Acer pensylvanicum"){
-    taxon_pft_lookup$pft[i] = "deciduous_forest"
-    taxon_pft_lookup$seasonality[i] = "deciduous"
+    taxon_pft_lookup$pft_MRA[i] = "deciduous_forest"
+    taxon_pft_lookup$phenology[i] = "deciduous"
     taxon_pft_lookup$leafType[i] = "broadleaf"
-    taxon_pft_lookup$growthForm[i] = "forest"
+    taxon_pft_lookup$growthForm[i] = "tree"
     taxon_pft_lookup$commonName[i] = "striped maple"}
@@
   else if(genusSpecies == "Acer rubrum"){
-    taxon_pft_lookup$pft[i] = "deciduous_forest"
-    taxon_pft_lookup$seasonality[i] = "deciduous"
+    taxon_pft_lookup$pft_MRA[i] = "deciduous_forest"
+    taxon_pft_lookup$phenology[i] = "deciduous"
     taxon_pft_lookup$leafType[i] = "broadleaf"
-    taxon_pft_lookup$growthForm[i] = "forest"
+    taxon_pft_lookup$growthForm[i] = "tree"
     taxon_pft_lookup$commonName[i] = "red maple"}
@@
   else if(genusSpecies == "Acer saccharinum"){
-    taxon_pft_lookup$pft[i] = "deciduous_forest"
-    taxon_pft_lookup$seasonality[i] = "deciduous"
+    taxon_pft_lookup$pft_MRA[i] = "deciduous_forest"
+    taxon_pft_lookup$phenology[i] = "deciduous"
     taxon_pft_lookup$leafType[i] = "broadleaf"
-    taxon_pft_lookup$growthForm[i] = "forest"
+    taxon_pft_lookup$growthForm[i] = "tree"
     taxon_pft_lookup$commonName[i] = "silver maple"}
@@
   else if(genusSpecies == "Acer saccharum"){
-    taxon_pft_lookup$pft[i] = "deciduous_forest"
-    taxon_pft_lookup$seasonality[i] = "deciduous"
+    taxon_pft_lookup$pft_MRA[i] = "deciduous_forest"
+    taxon_pft_lookup$phenology[i] = "deciduous"
     taxon_pft_lookup$leafType[i] = "broadleaf"
-    taxon_pft_lookup$growthForm[i] = "forest"
+    taxon_pft_lookup$growthForm[i] = "tree"
     taxon_pft_lookup$commonName[i] = "sugar maple"}
@@
   else if(genusSpecies == "Ailanthus altissima"){
-    taxon_pft_lookup$pft[i] = "deciduous_forest"
-    taxon_pft_lookup$seasonality[i] = "deciduous"
+    taxon_pft_lookup$pft_MRA[i] = "deciduous_forest"
+    taxon_pft_lookup$phenology[i] = "deciduous"
     taxon_pft_lookup$leafType[i] = "broadleaf"
-    taxon_pft_lookup$growthForm[i] = "forest"
+    taxon_pft_lookup$growthForm[i] = "tree"
     taxon_pft_lookup$commonName[i] = "tree of heaven"}
@@
   else if(genusSpecies == "Alnus rubra"){
-    taxon_pft_lookup$pft[i] = "deciduous_forest"
-    taxon_pft_lookup$seasonality[i] = "deciduous"
+    taxon_pft_lookup$pft_MRA[i] = "deciduous_forest"
+    taxon_pft_lookup$phenology[i] = "deciduous"
     taxon_pft_lookup$leafType[i] = "broadleaf"
-    taxon_pft_lookup$growthForm[i] = "forest"
+    taxon_pft_lookup$growthForm[i] = "tree"
     taxon_pft_lookup$commonName[i] = "red alder"}
@@
   else if(genusSpecies == "Aloysia wrightii"){
-    taxon_pft_lookup$pft[i] = "woody_shrub"
-    taxon_pft_lookup$seasonality[i] = "deciduous"
-    taxon_pft_lookup$leafType[i] = ""
+    taxon_pft_lookup$pft_MRA[i] = "woody_shrub"
+    taxon_pft_lookup$phenology[i] = "deciduous"
+    taxon_pft_lookup$leafType[i] = "NA"
     taxon_pft_lookup$growthForm[i] = "shrub"
@@
   else if(genusSpecies == "Amelanchier laevis"){
-    taxon_pft_lookup$pft[i] = "deciduous_forest"
-    taxon_pft_lookup$seasonality[i] = "deciduous"
+    taxon_pft_lookup$pft_MRA[i] = "deciduous_forest"
+    taxon_pft_lookup$phenology[i] = "deciduous"
     taxon_pft_lookup$leafType[i] = "broadleaf"
-    taxon_pft_lookup$growthForm[i] = "forest"
+    taxon_pft_lookup$growthForm[i] = "tree"
     taxon_pft_lookup$commonName[i] = "Allegheny serviceberry"}
@@
   else if(genusSpecies == "Amyris elemifera"){
-    taxon_pft_lookup$pft[i] = "???" # is not deciduous or coniferous forest
-    taxon_pft_lookup$seasonality[i] = "evergreen"
+    taxon_pft_lookup$pft_MRA[i] = "???" # is not deciduous or coniferous forest
+    taxon_pft_lookup$phenology[i] = "evergreen"
     taxon_pft_lookup$leafType[i] = "broadleaf"
-    taxon_pft_lookup$growthForm[i] = "forest"
+    taxon_pft_lookup$growthForm[i] = "tree"
     taxon_pft_lookup$commonName[i] = "sea torchwood"}
@@
   else if(genusSpecies == "Artemisia tridentata"){
-    taxon_pft_lookup$pft[i] = "woody_shrub"
-    taxon_pft_lookup$seasonality[i] = "evergreen"
-    taxon_pft_lookup$leafType[i] = ""
+    taxon_pft_lookup$pft_MRA[i] = "woody_shrub"
+    taxon_pft_lookup$phenology[i] = "evergreen"
+    taxon_pft_lookup$leafType[i] = "NA"
     taxon_pft_lookup$growthForm[i] = "shrub"
@@
   else if(genusSpecies == "Atriplex canescens"){
-    taxon_pft_lookup$pft[i] = "woody_shrub"
-    taxon_pft_lookup$seasonality[i] = "evergreen"
-    taxon_pft_lookup$leafType[i] = ""
+    taxon_pft_lookup$pft_MRA[i] = "woody_shrub"
+    taxon_pft_lookup$phenology[i] = "evergreen"
+    taxon_pft_lookup$leafType[i] = "NA"
     taxon_pft_lookup$growthForm[i] = "shrub"
@@
   else if(genusSpecies == "Atriplex confertifolia"){
-    taxon_pft_lookup$pft[i] = "woody_shrub"
-    taxon_pft_lookup$seasonality[i] = "evergreen"
-    taxon_pft_lookup$leafType[i] = ""
+    taxon_pft_lookup$pft_MRA[i] = "woody_shrub"
+    taxon_pft_lookup$phenology[i] = "evergreen"
+    taxon_pft_lookup$leafType[i] = "NA"
     taxon_pft_lookup$growthForm[i] = "shrub"
@@
   else if(genusSpecies == "Betula alleghaniensis"){
-    taxon_pft_lookup$pft[i] = "deciduous_forest"
-    taxon_pft_lookup$seasonality[i] = "deciduous"
+    taxon_pft_lookup$pft_MRA[i] = "deciduous_forest"
+    taxon_pft_lookup$phenology[i] = "deciduous"
     taxon_pft_lookup$leafType[i] = "broadleaf"
-    taxon_pft_lookup$growthForm[i] = "forest"
+    taxon_pft_lookup$growthForm[i] = "tree"
     taxon_pft_lookup$commonName[i] = "yellow birch"}
@@
   else if(genusSpecies == "Betula lenta"){
-    taxon_pft_lookup$pft[i] = "deciduous_forest"
-    taxon_pft_lookup$seasonality[i] = "deciduous"
+    taxon_pft_lookup$pft_MRA[i] = "deciduous_forest"
+    taxon_pft_lookup$phenology[i] = "deciduous"
     taxon_pft_lookup$leafType[i] = "broadleaf"
-    taxon_pft_lookup$growthForm[i] = "forest"
+    taxon_pft_lookup$growthForm[i] = "tree"
     taxon_pft_lookup$commonName[i] = "sweet birch"}
@@
   else if(genusSpecies == "Betula neoalaskana"){
-    taxon_pft_lookup$pft[i] = "deciduous_forest"
-    taxon_pft_lookup$seasonality[i] = "deciduous"
+    taxon_pft_lookup$pft_MRA[i] = "deciduous_forest"
+    taxon_pft_lookup$phenology[i] = "deciduous"
     taxon_pft_lookup$leafType[i] = "broadleaf"
-    taxon_pft_lookup$growthForm[i] = "forest"
+    taxon_pft_lookup$growthForm[i] = "tree"
     taxon_pft_lookup$commonName[i] = "resin birch"}
@@
   else if(genusSpecies == "Betula nigra"){
-    taxon_pft_lookup$pft[i] = "deciduous_forest"
-    taxon_pft_lookup$seasonality[i] = "deciduous"
+    taxon_pft_lookup$pft_MRA[i] = "deciduous_forest"
+    taxon_pft_lookup$phenology[i] = "deciduous"
     taxon_pft_lookup$leafType[i] = "broadleaf"
-    taxon_pft_lookup$growthForm[i] = "forest"
+    taxon_pft_lookup$growthForm[i] = "tree"
     taxon_pft_lookup$commonName[i] = "river birch"}
@@
   else if(genusSpecies == "Betula papyrifera"){
-    taxon_pft_lookup$pft[i] = "deciduous_forest"
-    taxon_pft_lookup$seasonality[i] = "deciduous"
+    taxon_pft_lookup$pft_MRA[i] = "deciduous_forest"
+    taxon_pft_lookup$phenology[i] = "deciduous"
     taxon_pft_lookup$leafType[i] = "broadleaf"
-    taxon_pft_lookup$growthForm[i] = "forest"
+    taxon_pft_lookup$growthForm[i] = "tree"
     taxon_pft_lookup$commonName[i] = "paper birch"}
@@
   else if(genusSpecies == "Betula sp."){
-    taxon_pft_lookup$pft[i] = "deciduous_forest"
-    taxon_pft_lookup$seasonality[i] = "deciduous"
+    taxon_pft_lookup$pft_MRA[i] = "deciduous_forest"
+    taxon_pft_lookup$phenology[i] = "deciduous"
     taxon_pft_lookup$leafType[i] = "broadleaf"
-    taxon_pft_lookup$growthForm[i] = "forest"
+    taxon_pft_lookup$growthForm[i] = "tree"
     taxon_pft_lookup$commonName[i] = "birch"}
@@
   else if(genusSpecies == "Bourreria sp."){
-    taxon_pft_lookup$pft[i] = "???"   
-    taxon_pft_lookup$seasonality[i] = ""
-    taxon_pft_lookup$leafType[i] = ""
-    taxon_pft_lookup$growthForm[i] = "forest"
+    taxon_pft_lookup$pft_MRA[i] = "NA"  # taxon class is too broad to specify 
+    taxon_pft_lookup$phenology[i] = "NA"
+    taxon_pft_lookup$leafType[i] = "NA"
+    taxon_pft_lookup$growthForm[i] = "tree"
     taxon_pft_lookup$commonName[i] = "strongbark"}
@@
   else if(genusSpecies == "Bursera simaruba"){
-    taxon_pft_lookup$pft[i] = "deciduous_forest"  
-    taxon_pft_lookup$seasonality[i] = "deciduous"
-    taxon_pft_lookup$leafType[i] = ""
-    taxon_pft_lookup$growthForm[i] = "forest"
+    taxon_pft_lookup$pft_MRA[i] = "deciduous_forest"  
+    taxon_pft_lookup$phenology[i] = "deciduous"
+    taxon_pft_lookup$leafType[i] = "broadleaf"
+    taxon_pft_lookup$growthForm[i] = "tree"
     taxon_pft_lookup$commonName[i] = "gumbo limbo"}
@@
   else if(genusSpecies == "Carpinus caroliniana"){
-    taxon_pft_lookup$pft[i] = "deciduous_forest"
+    taxon_pft_lookup$pft_MRA[i] = "deciduous_forest"
+    taxon_pft_lookup$phenology[i] = "deciduous"
+    taxon_pft_lookup$leafType[i] = "broadleaf"
+    taxon_pft_lookup$growthForm[i] = "tree"
     taxon_pft_lookup$commonName[i] = "American hornbeam"}
@@
   else if(genusSpecies == "Carya aquatica"){
-    taxon_pft_lookup$pft[i] = "deciduous_forest"
+    taxon_pft_lookup$pft_MRA[i] = "deciduous_forest"
+    taxon_pft_lookup$phenology[i] = "deciduous"
+    taxon_pft_lookup$leafType[i] = "broadleaf"
+    taxon_pft_lookup$growthForm[i] = "tree"
     taxon_pft_lookup$commonName[i] = "water hickory"}
@@
   else if(genusSpecies == "Carya cordiformis"){
-    taxon_pft_lookup$pft[i] = "deciduous_forest"
+    taxon_pft_lookup$pft_MRA[i] = "deciduous_forest"
+    taxon_pft_lookup$phenology[i] = "deciduous"
+    taxon_pft_lookup$leafType[i] = "broadleaf"
+    taxon_pft_lookup$growthForm[i] = "tree"
     taxon_pft_lookup$commonName[i] = "bitternut hickory"}
@@
   else if(genusSpecies == "Carya glabra"){
-    taxon_pft_lookup$pft[i] = "deciduous_forest"
+    taxon_pft_lookup$pft_MRA[i] = "deciduous_forest"
+    taxon_pft_lookup$phenology[i] = "deciduous"
+    taxon_pft_lookup$leafType[i] = "broadleaf"
+    taxon_pft_lookup$growthForm[i] = "tree"
     taxon_pft_lookup$commonName[i] = "pignut hickory"}
@@
   else if(genusSpecies == "Carya illinoinensis"){
-    taxon_pft_lookup$pft[i] = "deciduous_forest"
+    taxon_pft_lookup$pft_MRA[i] = "deciduous_forest"
+    taxon_pft_lookup$phenology[i] = "deciduous"
+    taxon_pft_lookup$leafType[i] = "broadleaf"
+    taxon_pft_lookup$growthForm[i] = "tree"
     taxon_pft_lookup$commonName[i] = "pecan"}
@@
   else if(genusSpecies == "Carya ovalis"){
-    taxon_pft_lookup$pft[i] = "deciduous_forest"
+    taxon_pft_lookup$pft_MRA[i] = "deciduous_forest"
+    taxon_pft_lookup$phenology[i] = "deciduous"
+    taxon_pft_lookup$leafType[i] = "broadleaf"
+    taxon_pft_lookup$growthForm[i] = "tree"
     taxon_pft_lookup$commonName[i] = "red hickory"}
@@
   else if(genusSpecies == "Carya ovata"){
-    taxon_pft_lookup$pft[i] = "deciduous_forest"
+    taxon_pft_lookup$pft_MRA[i] = "deciduous_forest"
+    taxon_pft_lookup$phenology[i] = "deciduous"
+    taxon_pft_lookup$leafType[i] = "broadleaf"
+    taxon_pft_lookup$growthForm[i] = "tree"
     taxon_pft_lookup$commonName[i] = "shagbark hickory"}
@@
   else if(genusSpecies == "Carya tomentosa"){
-    taxon_pft_lookup$pft[i] = "deciduous_forest"
+    taxon_pft_lookup$pft_MRA[i] = "deciduous_forest"
+    taxon_pft_lookup$phenology[i] = "deciduous"
+    taxon_pft_lookup$leafType[i] = "broadleaf"
+    taxon_pft_lookup$growthForm[i] = "tree"
     taxon_pft_lookup$commonName[i] = "mockernut hickory"}
@@
   else if(genusSpecies == "Celtis laevigata"){
-    taxon_pft_lookup$pft[i] = "deciduous_forest"
+    taxon_pft_lookup$pft_MRA[i] = "deciduous_forest"
+    taxon_pft_lookup$phenology[i] = "deciduous"
+    taxon_pft_lookup$leafType[i] = "broadleaf"
+    taxon_pft_lookup$growthForm[i] = "tree"
     taxon_pft_lookup$commonName[i] = "sugarberry"}
@@
   else if(genusSpecies == "Celtis occidentalis"){
-    taxon_pft_lookup$pft[i] = "deciduous_forest"
+    taxon_pft_lookup$pft_MRA[i] = "deciduous_forest"
+    taxon_pft_lookup$phenology[i] = "deciduous"
+    taxon_pft_lookup$leafType[i] = "broadleaf"
+    taxon_pft_lookup$growthForm[i] = "tree"
     taxon_pft_lookup$commonName[i] = "common hackberry"}
@@
   else if(genusSpecies == "Cercis canadensis"){
-    taxon_pft_lookup$pft[i] = "???"   # ???
+    taxon_pft_lookup$pft_MRA[i] = "deciduous_forest"
+    taxon_pft_lookup$phenology[i] = "deciduous"
+    taxon_pft_lookup$leafType[i] = "broadleaf"
+    taxon_pft_lookup$growthForm[i] = "tree"
     taxon_pft_lookup$commonName[i] = "eastern redbud"}
@@
   else if(genusSpecies == "Chrysothamnus viscidiflorus"){
-    taxon_pft_lookup$pft[i] = "woody_shrub"
+    taxon_pft_lookup$pft_MRA[i] = "woody_shrub"
+    taxon_pft_lookup$phenology[i] = "deciduous"
+    taxon_pft_lookup$leafType[i] = "NA"
+    taxon_pft_lookup$growthForm[i] = "shrub"
     taxon_pft_lookup$commonName[i] = "yellow rabbitbrush"}
@@
   else if(genusSpecies == "Cordia collococca"){
-    taxon_pft_lookup$pft[i] = "deciduous_forest"
+    taxon_pft_lookup$pft_MRA[i] = "deciduous_forest"
+    taxon_pft_lookup$phenology[i] = "deciduous"
+    taxon_pft_lookup$leafType[i] = "broadleaf"
+    taxon_pft_lookup$growthForm[i] = "tree"
     taxon_pft_lookup$commonName[i] = "red manjack"}
@@
   else if(genusSpecies == "Cornus drummondii"){
-    taxon_pft_lookup$pft[i] = "???"    # ???
+    taxon_pft_lookup$pft_MRA[i] = "woody_shrub" 
+    taxon_pft_lookup$phenology[i] = "deciduous"
+    taxon_pft_lookup$leafType[i] = "NA"
+    taxon_pft_lookup$growthForm[i] = "shrub"
     taxon_pft_lookup$commonName[i] = "roughleaf dogwood"}
@@
   else if(genusSpecies == "Cornus florida"){
-    taxon_pft_lookup$pft[i] = "???"    # ???
+    taxon_pft_lookup$pft_MRA[i] = "deciduous_forest"
+    taxon_pft_lookup$phenology[i] = "deciduous"
+    taxon_pft_lookup$leafType[i] = "broadleaf"
+    taxon_pft_lookup$growthForm[i] = "tree"
     taxon_pft_lookup$commonName[i] = "flowering dogwood"}
@@
   else if(genusSpecies == "Corylus cornuta"){
-    taxon_pft_lookup$pft[i] = "woody_shrub"      # deciduous shrubby hazel
+    taxon_pft_lookup$pft_MRA[i] = "woody_shrub" 
+    taxon_pft_lookup$phenology[i] = "deciduous"
+    taxon_pft_lookup$leafType[i] = ""
+    taxon_pft_lookup$growthForm[i] = "shrub"
     taxon_pft_lookup$commonName[i] = "beaked hazelnut"}
@@
   else if(genusSpecies == "Diospyros virginiana"){
-    taxon_pft_lookup$pft[i] = "deciduous_forest"
+    taxon_pft_lookup$pft_MRA[i] = "deciduous_forest"
+    taxon_pft_lookup$phenology[i] = "deciduous"
+    taxon_pft_lookup$leafType[i] = "broadleaf"
+    taxon_pft_lookup$growthForm[i] = "tree"
     taxon_pft_lookup$commonName[i] = "common persimmon"}
@@
   else if(genusSpecies == "Ephedra trifurca"){
-    taxon_pft_lookup$pft[i] = "woody_shrub"
+    taxon_pft_lookup$pft_MRA[i] = "woody_shrub"
+    taxon_pft_lookup$phenology[i] = "evergreen"
+    taxon_pft_lookup$leafType[i] = "NA"
+    taxon_pft_lookup$growthForm[i] = "shrub"
     taxon_pft_lookup$commonName[i] = "longleaf jointfir"}
@@
   else if(genusSpecies == "Ericameria laricifolia"){
-    taxon_pft_lookup$pft[i] = "woody_shrub"
+    taxon_pft_lookup$pft_MRA[i] = "woody_shrub"
+    taxon_pft_lookup$phenology[i] = "evergreen"
+    taxon_pft_lookup$leafType[i] = "NA"
+    taxon_pft_lookup$growthForm[i] = "shrub"
     taxon_pft_lookup$commonName[i] = "turpentine bush"}
@@
   else if(genusSpecies == "Exostema caribaeum"){
-    taxon_pft_lookup$pft[i] = "woody_shrub"  # "evergreen shrub or small tree" 
+    taxon_pft_lookup$pft_MRA[i] = "???"  # not deciduous or coniferous
+    taxon_pft_lookup$phenology[i] = "evergreen"
+    taxon_pft_lookup$leafType[i] = "broadleaf"
+    taxon_pft_lookup$growthForm[i] = "tree"
     taxon_pft_lookup$commonName[i] = "Caribbean princewood"}
@@
   else if(genusSpecies == "Fagus grandifolia"){
-    taxon_pft_lookup$pft[i] = "deciduous_forest"
+    taxon_pft_lookup$pft_MRA[i] = "deciduous_forest"
+    taxon_pft_lookup$phenology[i] = "deciduous"
+    taxon_pft_lookup$leafType[i] = "broadleaf"
+    taxon_pft_lookup$growthForm[i] = "tree"
     taxon_pft_lookup$commonName[i] = "American beech"}
@@
   else if(genusSpecies == "Frangula purshiana"){
-    taxon_pft_lookup$pft[i] = "???"     # ???
+    taxon_pft_lookup$pft_MRA[i] = "deciduous_forest"    
+    taxon_pft_lookup$phenology[i] = "deciduous"
+    taxon_pft_lookup$leafType[i] = "broadleaf"
+    taxon_pft_lookup$growthForm[i] = "tree"
     taxon_pft_lookup$commonName[i] = "Cascara buckthorn"}
@@
   else if(genusSpecies == "Fraxinus americana"){
-    taxon_pft_lookup$pft[i] = "deciduous_forest"
+    taxon_pft_lookup$pft_MRA[i] = "deciduous_forest"
+    taxon_pft_lookup$phenology[i] = "deciduous"
+    taxon_pft_lookup$leafType[i] = "broadleaf"
+    taxon_pft_lookup$growthForm[i] = "tree"
     taxon_pft_lookup$commonName[i] = "white ash"}
@@
   else if(genusSpecies == "Fraxinus nigra"){
-    taxon_pft_lookup$pft[i] = "deciduous_forest"
+    taxon_pft_lookup$pft_MRA[i] = "deciduous_forest"
+    taxon_pft_lookup$phenology[i] = "deciduous"
+    taxon_pft_lookup$leafType[i] = "broadleaf"
+    taxon_pft_lookup$growthForm[i] = "tree"
     taxon_pft_lookup$commonName[i] = "black ash"}
@@
   else if(genusSpecies == "Fraxinus pennsylvanica"){
-    taxon_pft_lookup$pft[i] = "deciduous_forest"
+    taxon_pft_lookup$pft_MRA[i] = "deciduous_forest"
+    taxon_pft_lookup$phenology[i] = "deciduous"
+    taxon_pft_lookup$leafType[i] = "broadleaf"
+    taxon_pft_lookup$growthForm[i] = "tree"
     taxon_pft_lookup$commonName[i] = "green ash"}
@@
   else if(genusSpecies == "Fraxinus sp."){
-    taxon_pft_lookup$pft[i] = "deciduous_forest"
+    taxon_pft_lookup$pft_MRA[i] = "deciduous_forest"
+    taxon_pft_lookup$phenology[i] = "deciduous"
+    taxon_pft_lookup$leafType[i] = "broadleaf"
+    taxon_pft_lookup$growthForm[i] = "tree"
     taxon_pft_lookup$commonName[i] = "ash"}
@@
   else if(genusSpecies == "Gaultheria shallon"){
-    taxon_pft_lookup$pft[i] = "woody_shrub"     # broadleaved evergreen shrub
+    taxon_pft_lookup$pft_MRA[i] = "woody_shrub"
+    taxon_pft_lookup$phenology[i] = "evergreen"
+    taxon_pft_lookup$leafType[i] = ""
+    taxon_pft_lookup$growthForm[i] = "shrub"
     taxon_pft_lookup$commonName[i] = "salal"}
@@
   else if(genusSpecies == "Gleditsia triacanthos"){
-    taxon_pft_lookup$pft[i] = "deciduous_forest"
+    taxon_pft_lookup$pft_MRA[i] = "deciduous_forest"
+    taxon_pft_lookup$phenology[i] = "deciduous"
+    taxon_pft_lookup$leafType[i] = "broadleaf"
+    taxon_pft_lookup$growthForm[i] = "tree"
     taxon_pft_lookup$commonName[i] = "honey locust"}
@@
   else if(genusSpecies == "Gutierrezia sarothrae"){
-    taxon_pft_lookup$pft[i] = "woody_shrub"
+    taxon_pft_lookup$pft_MRA[i] = "woody_shrub"
+    taxon_pft_lookup$phenology[i] = "deciduous"
+    taxon_pft_lookup$leafType[i] = "NA"
+    taxon_pft_lookup$growthForm[i] = "shrub"
     taxon_pft_lookup$commonName[i] = "broom snakeweed"}
@@
   else if(genusSpecies == "Halesia sp."){
-    taxon_pft_lookup$pft[i] = "deciduous_forest"  # small, deciduous, understory tree 
+    taxon_pft_lookup$pft_MRA[i] = "deciduous_forest"
+    taxon_pft_lookup$phenology[i] = "deciduous"
+    taxon_pft_lookup$leafType[i] = "broadleaf"
+    taxon_pft_lookup$growthForm[i] = "tree"
     taxon_pft_lookup$commonName[i] = "silverbell"}
@@
   else if(genusSpecies == "Hamamelis virginiana"){
-    taxon_pft_lookup$pft[i] = "woody_shrub"
+    taxon_pft_lookup$pft_MRA[i] = "deciduous_forest"
+    taxon_pft_lookup$phenology[i] = "deciduous"
+    taxon_pft_lookup$leafType[i] = "broadleaf"
+    taxon_pft_lookup$growthForm[i] = "tree"
     taxon_pft_lookup$commonName[i] = "American witchhazel"}
@@
   else if(genusSpecies == "Ilex montana"){
-    taxon_pft_lookup$pft[i] = "???"          # ???
+    taxon_pft_lookup$pft_MRA[i] = "deciduous_forest"
+    taxon_pft_lookup$phenology[i] = "deciduous"
+    taxon_pft_lookup$leafType[i] = "broadleaf"
+    taxon_pft_lookup$growthForm[i] = "tree"
     taxon_pft_lookup$commonName[i] = "mountain holly"}
@@
   else if(genusSpecies == "Ilex opaca"){
-    taxon_pft_lookup$pft[i] = "woody_shrub"   # ???      # broadleaved evergreen tree
+    taxon_pft_lookup$pft_MRA[i] = "???"  # not deciduous or coniferous 
+    taxon_pft_lookup$phenology[i] = "evergreen"
+    taxon_pft_lookup$leafType[i] = "broadleaf"
+    taxon_pft_lookup$growthForm[i] = "tree"
     taxon_pft_lookup$commonName[i] = "American holly"}
@@
   else if(genusSpecies == "Juglans nigra"){
-    taxon_pft_lookup$pft[i] = "deciduous_forest"
+    taxon_pft_lookup$pft_MRA[i] = "deciduous_forest"
+    taxon_pft_lookup$phenology[i] = ""
+    taxon_pft_lookup$leafType[i] = ""
+    taxon_pft_lookup$growthForm[i] = ""
     taxon_pft_lookup$commonName[i] = "black walnut"}
@@
   else if(genusSpecies == "Juniperus osteosperma"){
-    taxon_pft_lookup$pft[i] = "coniferous_forest"
+    taxon_pft_lookup$pft_MRA[i] = "coniferous_forest"
+    taxon_pft_lookup$phenology[i] = ""
+    taxon_pft_lookup$leafType[i] = ""
+    taxon_pft_lookup$growthForm[i] = ""
     taxon_pft_lookup$commonName[i] = "Utah juniper"}
@@
   else if(genusSpecies == "Juniperus virginiana"){
-    taxon_pft_lookup$pft[i] = "coniferous_forest"
+    taxon_pft_lookup$pft_MRA[i] = "coniferous_forest"
+    taxon_pft_lookup$phenology[i] = ""
+    taxon_pft_lookup$leafType[i] = ""
+    taxon_pft_lookup$growthForm[i] = ""
     taxon_pft_lookup$commonName[i] = "red cedar"}
@@
   else if(genusSpecies == "Larix sp."){
-    taxon_pft_lookup$pft[i] = "coniferous_forest"
+    taxon_pft_lookup$pft_MRA[i] = "coniferous_forest"
+    taxon_pft_lookup$phenology[i] = ""
+    taxon_pft_lookup$leafType[i] = ""
+    taxon_pft_lookup$growthForm[i] = ""
     taxon_pft_lookup$commonName[i] = "larch"}
@@
   else if(genusSpecies == "Larrea tridentata"){
-    taxon_pft_lookup$pft[i] = "woody_shrub"
+    taxon_pft_lookup$pft_MRA[i] = "woody_shrub"
+    taxon_pft_lookup$phenology[i] = ""
+    taxon_pft_lookup$leafType[i] = ""
+    taxon_pft_lookup$growthForm[i] = ""
     taxon_pft_lookup$commonName[i] = "creosote bush"}
@@
   else if(genusSpecies == "Liquidambar styraciflua"){
-    taxon_pft_lookup$pft[i] = "deciduous_forest"
+    taxon_pft_lookup$pft_MRA[i] = "deciduous_forest"
+    taxon_pft_lookup$phenology[i] = ""
+    taxon_pft_lookup$leafType[i] = ""
+    taxon_pft_lookup$growthForm[i] = ""
     taxon_pft_lookup$commonName[i] = "sweetgum"}
@@
   else if(genusSpecies == "Liriodendron tulipifera"){
-    taxon_pft_lookup$pft[i] = "deciduous_forest"
+    taxon_pft_lookup$pft_MRA[i] = "deciduous_forest"
+    taxon_pft_lookup$phenology[i] = ""
+    taxon_pft_lookup$leafType[i] = ""
+    taxon_pft_lookup$growthForm[i] = ""
     taxon_pft_lookup$commonName[i] = "tuliptree"}
@@
   else if(genusSpecies == "Lonicera maackii"){
-    taxon_pft_lookup$pft[i] = "woody_shrub"
+    taxon_pft_lookup$pft_MRA[i] = "woody_shrub"
+    taxon_pft_lookup$phenology[i] = ""
+    taxon_pft_lookup$leafType[i] = ""
+    taxon_pft_lookup$growthForm[i] = ""
     taxon_pft_lookup$commonName[i] = "Amur honeysuckle"}
@@
   else if(genusSpecies == "Maclura pomifera"){
-    taxon_pft_lookup$pft[i] = "deciduous_forest"   # small deciduous tree or large shrub
+    taxon_pft_lookup$pft_MRA[i] = "deciduous_forest"   # small deciduous tree or large shrub
+    taxon_pft_lookup$phenology[i] = ""
+    taxon_pft_lookup$leafType[i] = ""
+    taxon_pft_lookup$growthForm[i] = ""
     taxon_pft_lookup$commonName[i] = "Osage-orange"}
@@
  else if(genusSpecies == "Magnolia sp."){
-    taxon_pft_lookup$pft[i] = "???"
+    taxon_pft_lookup$pft_MRA[i] = "???"
+    taxon_pft_lookup$phenology[i] = ""
+    taxon_pft_lookup$leafType[i] = ""
+    taxon_pft_lookup$growthForm[i] = ""
     taxon_pft_lookup$commonName[i] = "magnolia"}
@@
   else if(genusSpecies == "Mahonia nervosa"){
-    taxon_pft_lookup$pft[i] = "woody_shrub"      # evergreen shrub
+    taxon_pft_lookup$pft_MRA[i] = "woody_shrub"      # evergreen shrub
+    taxon_pft_lookup$phenology[i] = ""
+    taxon_pft_lookup$leafType[i] = ""
+    taxon_pft_lookup$growthForm[i] = ""
     taxon_pft_lookup$commonName[i] = "Cascade barberry"}
@@
   else if(genusSpecies == "Mangifera indica"){
-    taxon_pft_lookup$pft[i] = "???"       # ???   # broadleaf evergreen tree
+    taxon_pft_lookup$pft_MRA[i] = "???"       # ???   # broadleaf evergreen tree
+    taxon_pft_lookup$phenology[i] = ""
+    taxon_pft_lookup$leafType[i] = ""
+    taxon_pft_lookup$growthForm[i] = ""
     taxon_pft_lookup$commonName[i] = "mango"}
@@
   else if(genusSpecies == "Melia azedarach"){
-    taxon_pft_lookup$pft[i] = "deciduous_forest"
+    taxon_pft_lookup$pft_MRA[i] = "deciduous_forest"
     taxon_pft_lookup$commonName[i] = "Chinaberrytree"}
@@
   else if(genusSpecies == "Morus rubra"){
-    taxon_pft_lookup$pft[i] = "deciduous_forest"
+    taxon_pft_lookup$pft_MRA[i] = "deciduous_forest"
     taxon_pft_lookup$commonName[i] = "red mulberry"}
@@
   else if(genusSpecies == "Nyssa aquatica"){
-    taxon_pft_lookup$pft[i] = "deciduous_forest"
+    taxon_pft_lookup$pft_MRA[i] = "deciduous_forest"
     taxon_pft_lookup$commonName[i] = "water tupelo"}
@@
   else if(genusSpecies == "Nyssa biflora"){
-    taxon_pft_lookup$pft[i] = "deciduous_forest"
+    taxon_pft_lookup$pft_MRA[i] = "deciduous_forest"
     taxon_pft_lookup$commonName[i] = "swamp tupelo"}
@@
   else if(genusSpecies == "Nyssa sylvatica"){
-    taxon_pft_lookup$pft[i] = "deciduous_forest"
+    taxon_pft_lookup$pft_MRA[i] = "deciduous_forest"
     taxon_pft_lookup$commonName[i] = "blackgum"}
@@
   else if(genusSpecies == "Oxydendrum sp."){
-    taxon_pft_lookup$pft[i] = "deciduous_forest"
+    taxon_pft_lookup$pft_MRA[i] = "deciduous_forest"
     taxon_pft_lookup$commonName[i] = "sourwood"}
@@
   else if(genusSpecies == "Picea engelmannii"){
-    taxon_pft_lookup$pft[i] = "coniferous_forest"
+    taxon_pft_lookup$pft_MRA[i] = "coniferous_forest"
     taxon_pft_lookup$commonName[i] = "Engelmann spruce"}
@@
   else if(genusSpecies == "Picea glauca"){
-    taxon_pft_lookup$pft[i] = "coniferous_forest"
+    taxon_pft_lookup$pft_MRA[i] = "coniferous_forest"
     taxon_pft_lookup$commonName[i] = "white spruce"}
@@
   else if(genusSpecies == "Picea mariana"){
-    taxon_pft_lookup$pft[i] = "coniferous_forest"
+    taxon_pft_lookup$pft_MRA[i] = "coniferous_forest"
     taxon_pft_lookup$commonName[i] = "black spruce"}
@@
   else if(genusSpecies == "Picea sp."){
-    taxon_pft_lookup$pft[i] = "coniferous_forest"
+    taxon_pft_lookup$pft_MRA[i] = "coniferous_forest"
     taxon_pft_lookup$commonName[i] = "spruce"}
@@
   else if(genusSpecies == "Pinus contorta"){
-    taxon_pft_lookup$pft[i] = "coniferous_forest"
+    taxon_pft_lookup$pft_MRA[i] = "coniferous_forest"
     taxon_pft_lookup$commonName[i] = "lodgepole pine"}
@@
   else if(genusSpecies == "Pinus echinata"){
-    taxon_pft_lookup$pft[i] = "coniferous_forest"
+    taxon_pft_lookup$pft_MRA[i] = "coniferous_forest"
     taxon_pft_lookup$commonName[i] = "shortleaf pine"}
@@
   else if(genusSpecies == "Pinus flexilis"){
-    taxon_pft_lookup$pft[i] = "coniferous_forest"
+    taxon_pft_lookup$pft_MRA[i] = "coniferous_forest"
     taxon_pft_lookup$commonName[i] = "limber pine"}
@@
   else if(genusSpecies == "Pinus palustris"){
-    taxon_pft_lookup$pft[i] = "coniferous_forest"
+    taxon_pft_lookup$pft_MRA[i] = "coniferous_forest"
     taxon_pft_lookup$commonName[i] = "longleaf pine"}
@@
   else if(genusSpecies == "Pinus ponderosa"){
-    taxon_pft_lookup$pft[i] = "coniferous_forest"
+    taxon_pft_lookup$pft_MRA[i] = "coniferous_forest"
     taxon_pft_lookup$commonName[i] = "ponderosa pine"}
@@
   else if(genusSpecies == "Pinus sp."){
-    taxon_pft_lookup$pft[i] = "coniferous_forest"
+    taxon_pft_lookup$pft_MRA[i] = "coniferous_forest"
     taxon_pft_lookup$commonName[i] = "pine"}
@@
   else if(genusSpecies == "Pinus strobus"){
-    taxon_pft_lookup$pft[i] = "coniferous_forest"
+    taxon_pft_lookup$pft_MRA[i] = "coniferous_forest"
     taxon_pft_lookup$commonName[i] = "eastern white pine"}
@@
   else if(genusSpecies == "Pinus taeda"){
-    taxon_pft_lookup$pft[i] = "coniferous_forest"
+    taxon_pft_lookup$pft_MRA[i] = "coniferous_forest"
     taxon_pft_lookup$commonName[i] = "loblolly pine"}
@@
   else if(genusSpecies == "Pisonia albida"){
-    taxon_pft_lookup$pft[i] = "???"    # Growth Form - Multiple Stem, Height, Mature - 30ft
+    taxon_pft_lookup$pft_MRA[i] = "???"    # Growth Form - Multiple Stem, Height, Mature - 30ft
     taxon_pft_lookup$commonName[i] = "corcho bobo"}
@@
   else if(genusSpecies == "Platanus occidentalis"){
-    taxon_pft_lookup$pft[i] = "deciduous_forest"
+    taxon_pft_lookup$pft_MRA[i] = "deciduous_forest"
     taxon_pft_lookup$commonName[i] = "American sycamore"}
@@
   else if(genusSpecies == "Populus deltoides"){
-    taxon_pft_lookup$pft[i] = "deciduous_forest"
+    taxon_pft_lookup$pft_MRA[i] = "deciduous_forest"
     taxon_pft_lookup$commonName[i] = "eastern cottonwood"}
@@
   else if(genusSpecies == "Populus grandidentata"){
-    taxon_pft_lookup$pft[i] = "deciduous_forest"
+    taxon_pft_lookup$pft_MRA[i] = "deciduous_forest"
     taxon_pft_lookup$commonName[i] = "bigtooth aspen"}
@@
   else if(genusSpecies == "Populus sp."){
-    taxon_pft_lookup$pft[i] = "deciduous_forest"
+    taxon_pft_lookup$pft_MRA[i] = "deciduous_forest"
     taxon_pft_lookup$commonName[i] = "cottonwood"}
@@
   else if(genusSpecies == "Populus tremuloides"){
-    taxon_pft_lookup$pft[i] = "deciduous_forest"
+    taxon_pft_lookup$pft_MRA[i] = "deciduous_forest"
     taxon_pft_lookup$commonName[i] = "quaking aspen"}
@@
   else if(genusSpecies == "Prosopis glandulosa"){
-    taxon_pft_lookup$pft[i] = "???"   # shrub or tree ranging from 4 to 6m tall
+    taxon_pft_lookup$pft_MRA[i] = "???"   # shrub or tree ranging from 4 to 6m tall
     taxon_pft_lookup$commonName[i] = "honey mesquite"}
@@
   else if(genusSpecies == "Prosopis velutina"){
-    taxon_pft_lookup$pft[i] = "???" # deciduous shrub or tree is less than 15 m
+    taxon_pft_lookup$pft_MRA[i] = "???" # deciduous shrub or tree is less than 15 m
     taxon_pft_lookup$commonName[i] = "velvet mesquite"}
@@
   else if(genusSpecies == "Prunus americana"){
-    taxon_pft_lookup$pft[i] = "???" # deciduous large shrub or small tree, reaching heights up to 15 feet
+    taxon_pft_lookup$pft_MRA[i] = "???" # deciduous large shrub or small tree, reaching heights up to 15 feet
     taxon_pft_lookup$commonName[i] = "American plum"}
@@
   else if(genusSpecies == "Prunus angustifolia"){
-    taxon_pft_lookup$pft[i] = "woody_shrub" # short, thickly branched shrub 3 to 10ft tall
+    taxon_pft_lookup$pft_MRA[i] = "woody_shrub" # short, thickly branched shrub 3 to 10ft tall
     taxon_pft_lookup$commonName[i] = "Chickasaw plum"}
@@
   else if(genusSpecies == "Prunus mexicana"){
-    taxon_pft_lookup$pft[i] = "???"
+    taxon_pft_lookup$pft_MRA[i] = "???"
     taxon_pft_lookup$commonName[i] = "Mexican plum"}
@@
   else if(genusSpecies == "Prunus serotina"){
-    taxon_pft_lookup$pft[i] = "???"       
+    taxon_pft_lookup$pft_MRA[i] = "???"       
     taxon_pft_lookup$commonName[i] = "black cherry"}
@@
   else if(genusSpecies == "Pseudotsuga menziesii"){
-    taxon_pft_lookup$pft[i] = "coniferous_forest"
+    taxon_pft_lookup$pft_MRA[i] = "coniferous_forest"
     taxon_pft_lookup$commonName[i] = "Douglas-fir"}
@@
   else if(genusSpecies == "Quercus alba"){
-    taxon_pft_lookup$pft[i] = "deciduous_forest"
+    taxon_pft_lookup$pft_MRA[i] = "deciduous_forest"
     taxon_pft_lookup$commonName[i] = "white oak"}
@@
   else if(genusSpecies == "Quercus coccinea"){
-    taxon_pft_lookup$pft[i] = "deciduous_forest"
+    taxon_pft_lookup$pft_MRA[i] = "deciduous_forest"
     taxon_pft_lookup$commonName[i] = "scarlet oak"}
@@
   else if(genusSpecies == "Quercus falcata"){
-    taxon_pft_lookup$pft[i] = "deciduous_forest"
+    taxon_pft_lookup$pft_MRA[i] = "deciduous_forest"
     taxon_pft_lookup$commonName[i] = "southern red oak"}
@@
   else if(genusSpecies == "Quercus geminata"){
-    taxon_pft_lookup$pft[i] = "deciduous_forest"
+    taxon_pft_lookup$pft_MRA[i] = "deciduous_forest"
     taxon_pft_lookup$commonName[i] = "sand live oak"}
@@
   else if(genusSpecies == "Quercus hemisphaerica"){
-    taxon_pft_lookup$pft[i] = "deciduous_forest"
+    taxon_pft_lookup$pft_MRA[i] = "deciduous_forest"
     taxon_pft_lookup$commonName[i] = "Darlington oak"}
@@
   else if(genusSpecies == "Quercus laevis"){
-    taxon_pft_lookup$pft[i] = "deciduous_forest"
+    taxon_pft_lookup$pft_MRA[i] = "deciduous_forest"
     taxon_pft_lookup$commonName[i] = "turkey oak"}
@@
   else if(genusSpecies == "Quercus laurifolia"){
-    taxon_pft_lookup$pft[i] = "deciduous_forest"
+    taxon_pft_lookup$pft_MRA[i] = "deciduous_forest"
     taxon_pft_lookup$commonName[i] = "laurel oak"}
@@
   else if(genusSpecies == "Quercus lyrata"){
-    taxon_pft_lookup$pft[i] = "deciduous_forest"
+    taxon_pft_lookup$pft_MRA[i] = "deciduous_forest"
     taxon_pft_lookup$commonName[i] = "overcup oak"}
@@
   else if(genusSpecies == "Quercus macrocarpa"){
-    taxon_pft_lookup$pft[i] = "deciduous_forest"
+    taxon_pft_lookup$pft_MRA[i] = "deciduous_forest"
     taxon_pft_lookup$commonName[i] = "bur oak"}
@@
   else if(genusSpecies == "Quercus margaretta"){
-    taxon_pft_lookup$pft[i] = "deciduous_forest"
+    taxon_pft_lookup$pft_MRA[i] = "deciduous_forest"
     taxon_pft_lookup$commonName[i] = "sand post oak"}
@@
   else if(genusSpecies == "Quercus marilandica"){
-    taxon_pft_lookup$pft[i] = "deciduous_forest"
+    taxon_pft_lookup$pft_MRA[i] = "deciduous_forest"
     taxon_pft_lookup$commonName[i] = "blackjack oak"}
@@
   else if(genusSpecies == "Quercus michauxii"){
-    taxon_pft_lookup$pft[i] = "deciduous_forest"
+    taxon_pft_lookup$pft_MRA[i] = "deciduous_forest"
     taxon_pft_lookup$commonName[i] = "swamp chestnut oak"}
@@
   else if(genusSpecies == "Quercus montana"){
-    taxon_pft_lookup$pft[i] = "deciduous_forest"
+    taxon_pft_lookup$pft_MRA[i] = "deciduous_forest"
     taxon_pft_lookup$commonName[i] = "chestnut oak"}
@@
   else if(genusSpecies == "Quercus muehlenbergii"){
-    taxon_pft_lookup$pft[i] = "deciduous_forest"
+    taxon_pft_lookup$pft_MRA[i] = "deciduous_forest"
     taxon_pft_lookup$commonName[i] = "chinquapin oak"}
@@
   else if(genusSpecies == "Quercus nigra"){
-    taxon_pft_lookup$pft[i] = "deciduous_forest"
+    taxon_pft_lookup$pft_MRA[i] = "deciduous_forest"
     taxon_pft_lookup$commonName[i] = "water oak"}
@@
   else if(genusSpecies == "Quercus pagoda"){
-    taxon_pft_lookup$pft[i] = "deciduous_forest"
+    taxon_pft_lookup$pft_MRA[i] = "deciduous_forest"
     taxon_pft_lookup$commonName[i] = "cherrybark oak"}
@@
   else if(genusSpecies == "Quercus phellos"){
-    taxon_pft_lookup$pft[i] = "deciduous_forest"
+    taxon_pft_lookup$pft_MRA[i] = "deciduous_forest"
     taxon_pft_lookup$commonName[i] = "willow oak"}
@@
   else if(genusSpecies == "Quercus rubra"){
-    taxon_pft_lookup$pft[i] = "deciduous_forest"
+    taxon_pft_lookup$pft_MRA[i] = "deciduous_forest"
     taxon_pft_lookup$commonName[i] = "northern red oak"}
@@
   else if(genusSpecies == "Quercus shumardii"){
-    taxon_pft_lookup$pft[i] = "deciduous_forest"
+    taxon_pft_lookup$pft_MRA[i] = "deciduous_forest"
     taxon_pft_lookup$commonName[i] = "Shumard's oak"}
@@
   else if(genusSpecies == "Quercus sp."){
-    taxon_pft_lookup$pft[i] = "???"  # apparently there are evergreen oak species? 
+    taxon_pft_lookup$pft_MRA[i] = "???"  # apparently there are evergreen oak species? 
     taxon_pft_lookup$commonName[i] = "oak"}
@@
   else if(genusSpecies == "Quercus stellata"){
-    taxon_pft_lookup$pft[i] = "deciduous_forest"
+    taxon_pft_lookup$pft_MRA[i] = "deciduous_forest"
     taxon_pft_lookup$commonName[i] = "post oak"}
@@
   else if(genusSpecies == "Quercus velutina"){
-    taxon_pft_lookup$pft[i] = "deciduous_forest"
+    taxon_pft_lookup$pft_MRA[i] = "deciduous_forest"
     taxon_pft_lookup$commonName[i] = "black oak"}
@@
   else if(genusSpecies == "Quercus virginiana"){
-    taxon_pft_lookup$pft[i] = "deciduous_forest"
+    taxon_pft_lookup$pft_MRA[i] = "deciduous_forest"
     taxon_pft_lookup$commonName[i] = "live oak"}
@@
   else if(genusSpecies == "Rhamnus davurica"){
-    taxon_pft_lookup$pft[i] = "???"     #  spreading shrub or small tree reaching 25 to 30 feet
+    taxon_pft_lookup$pft_MRA[i] = "???"     #  spreading shrub or small tree reaching 25 to 30 feet
     taxon_pft_lookup$commonName[i] = "Dahurian buckthorn"}
@@
   else if(genusSpecies == "Rhus aromatica"){
-    taxon_pft_lookup$pft[i] = "woody_shrub"
+    taxon_pft_lookup$pft_MRA[i] = "woody_shrub"
     taxon_pft_lookup$commonName[i] = "fragrant sumac"}
@@
   else if(genusSpecies == "Rhus glabra"){
-    taxon_pft_lookup$pft[i] = "woody_shrub"
+    taxon_pft_lookup$pft_MRA[i] = "woody_shrub"
     taxon_pft_lookup$commonName[i] = "smooth sumac"}
@@
   else if(genusSpecies == "Robinia pseudoacacia"){
-    taxon_pft_lookup$pft[i] = "deciduous_forest"
+    taxon_pft_lookup$pft_MRA[i] = "deciduous_forest"
     taxon_pft_lookup$commonName[i] = "black locust"}
@@
   else if(genusSpecies == "Sassafras albidum"){
-    taxon_pft_lookup$pft[i] = "deciduous_forest" # deciduous shrub or tree. trees are 9 to 18 m
+    taxon_pft_lookup$pft_MRA[i] = "deciduous_forest" # deciduous shrub or tree. trees are 9 to 18 m
     taxon_pft_lookup$commonName[i] = "sassafras"}
@@
   else if(genusSpecies ==  "Sassafras sp."){
-    taxon_pft_lookup$pft[i] = "deciduous_forest"
+    taxon_pft_lookup$pft_MRA[i] = "deciduous_forest"
     taxon_pft_lookup$commonName[i] = "sassafras"}
@@
   else if(genusSpecies == "Sideroxylon lanuginosum"){
-    taxon_pft_lookup$pft[i] = "???" # deciduous multi-trunked tall shrub or medium-sized tree to 48ft or more
+    taxon_pft_lookup$pft_MRA[i] = "???" # deciduous multi-trunked tall shrub or medium-sized tree to 48ft or more
     taxon_pft_lookup$commonName[i] = "gum bully"}
@@
   else if(genusSpecies == "Symplocos sp."){
-    taxon_pft_lookup$pft[i] = "???"  # "deciduous or evergreen shrub or tree"
+    taxon_pft_lookup$pft_MRA[i] = "???"  # "deciduous or evergreen shrub or tree"
     taxon_pft_lookup$commonName[i] = "sweetleaf"}
@@
   else if(genusSpecies == "Taxus brevifolia"){
-    taxon_pft_lookup$pft[i] = "coniferous_forest"
+    taxon_pft_lookup$pft_MRA[i] = "coniferous_forest"
     taxon_pft_lookup$commonName[i] = "Pacific yew"}
@@
   else if(genusSpecies == "Tetradium daniellii"){
-    taxon_pft_lookup$pft[i] = "deciduous_forest"
+    taxon_pft_lookup$pft_MRA[i] = "deciduous_forest"
     taxon_pft_lookup$commonName[i] = "bee-bee tree"}
@@
   else if(genusSpecies == "Triadica sebifera"){
-    taxon_pft_lookup$pft[i] = "deciduous_forest"
+    taxon_pft_lookup$pft_MRA[i] = "deciduous_forest"
     taxon_pft_lookup$commonName[i] = "Chinese tallow"}
@@
   else if(genusSpecies == "Tsuga canadensis"){
-    taxon_pft_lookup$pft[i] = "coniferous_forest"
+    taxon_pft_lookup$pft_MRA[i] = "coniferous_forest"
     taxon_pft_lookup$commonName[i] = "eastern hemlock"}
@@
   else if(genusSpecies == "Tsuga heterophylla"){
-    taxon_pft_lookup$pft[i] = "coniferous_forest"
+    taxon_pft_lookup$pft_MRA[i] = "coniferous_forest"
     taxon_pft_lookup$commonName[i] = "western hemlock"}
@@
   else if(genusSpecies == "Ulmus alata"){
-    taxon_pft_lookup$pft[i] = "deciduous_forest"
+    taxon_pft_lookup$pft_MRA[i] = "deciduous_forest"
     taxon_pft_lookup$commonName[i] = "winged elm"}
@@
   else if(genusSpecies == "Ulmus americana"){
-    taxon_pft_lookup$pft[i] = "deciduous_forest"
+    taxon_pft_lookup$pft_MRA[i] = "deciduous_forest"
     taxon_pft_lookup$commonName[i] = "American elm"}
@@
   else if(genusSpecies == "Ulmus crassifolia"){
-    taxon_pft_lookup$pft[i] = "deciduous_forest"
+    taxon_pft_lookup$pft_MRA[i] = "deciduous_forest"
     taxon_pft_lookup$commonName[i] = "cedar elm"}
@@
   else if(genusSpecies == "Ulmus rubra"){
-    taxon_pft_lookup$pft[i] = "deciduous_forest"
+    taxon_pft_lookup$pft_MRA[i] = "deciduous_forest"
     taxon_pft_lookup$commonName[i] = "slippery elm"}
@@
   else if(genusSpecies == "Ulmus sp."){
-    taxon_pft_lookup$pft[i] = "deciduous_forest"
+    taxon_pft_lookup$pft_MRA[i] = "deciduous_forest"
     taxon_pft_lookup$commonName[i] = "elm"}
@@
   else if(genusSpecies == "Vaccinium parvifolium"){ 
-    taxon_pft_lookup$pft[i] = "woody_shrub"       # deciduous Shrub
+    taxon_pft_lookup$pft_MRA[i] = "woody_shrub"       # deciduous Shrub
     taxon_pft_lookup$commonName[i] = "red huckleberry"}
@@
   else if(genusSpecies == "Zinnia acerosa"){
-    taxon_pft_lookup$pft[i] = "grass_herb" # forb/herb, subshrub - grass_herb?
+    taxon_pft_lookup$pft_MRA[i] = "grass_herb" # forb/herb, subshrub - grass_herb?
     taxon_pft_lookup$commonName[i] = "desert zinnia"}

### 2021-04-22 R/add_plant_functional_type.Rmd
working through taxons

@@
 - pft_MODIS (based on MODIS land cover classes)
-    - evergreen_needleleaf
-    - evergreen_broadleaf
-    - deciduous_needlelead
-    - deciduous_broadleaf
+    - evergreen-needleleaf_forest
+    - evergreen-broadleaf_forest
+    - deciduous-needleleaf_forest
+    - deciduous-broadleaf_forest
     - shrub
@@
                                pft_MRA = NA,
-                               pft_MODIS = ,
+                               pft_MODIS = NA,
                                phenology = NA,
@@
     taxon_pft_lookup$pft_MRA[i] = "coniferous_forest" 
-    taxon_pft_lookup$pft_MODIS[i] = 
+    taxon_pft_lookup$pft_MODIS[i] = "evergreen-needleleaf_forest"
     taxon_pft_lookup$phenology[i] = "evergreen"
@@
     taxon_pft_lookup$pft_MRA[i] = "coniferous_forest" 
+    taxon_pft_lookup$pft_MODIS[i] = "evergreen-needleleaf_forest"
     taxon_pft_lookup$phenology[i] = "evergreen"
@@
     taxon_pft_lookup$pft_MRA[i] = "coniferous_forest"
+    taxon_pft_lookup$pft_MODIS[i] = "evergreen-needleleaf_forest"
     taxon_pft_lookup$phenology[i] = "evergreen"
@@
     taxon_pft_lookup$pft_MRA[i] = "woody_shrub" 
-    taxon_pft_lookup$phenology[i] = "deciduous"
-    taxon_pft_lookup$leafType[i] = ""
+    taxon_pft_lookup$pft_MODIS[i] = "shrub"
+    taxon_pft_lookup$phenology[i] = "deciduous"
+    taxon_pft_lookup$leafType[i] = "NA"
     taxon_pft_lookup$growthForm[i] = "shrub"
@@
     taxon_pft_lookup$pft_MRA[i] = "deciduous_forest"
+    taxon_pft_lookup$pft_MODIS[i] = "deciduous-broadleaf_forest"
     taxon_pft_lookup$phenology[i] = "deciduous"
@@
     taxon_pft_lookup$pft_MRA[i] = "deciduous_forest" 
+    taxon_pft_lookup$pft_MODIS[i] = "deciduous-broadleaf_forest"
     taxon_pft_lookup$phenology[i] = "deciduous"
@@
     taxon_pft_lookup$pft_MRA[i] = "woody_shrub"
+    taxon_pft_lookup$pft_MODIS[i] = "shrub"
     taxon_pft_lookup$phenology[i] = "deciduous"
@@
   else if(genusSpecies == "Amyris elemifera"){
-    taxon_pft_lookup$pft_MRA[i] = "???" # is not deciduous or coniferous forest
+    taxon_pft_lookup$pft_MRA[i] = "NA" # is not deciduous or coniferous forest
+    taxon_pft_lookup$pft_MODIS[i] = "evergreen-broadleaf_forest"
     taxon_pft_lookup$phenology[i] = "evergreen"
@@
     taxon_pft_lookup$pft_MRA[i] = "woody_shrub"
+    taxon_pft_lookup$pft_MODIS[i] = "shrub"
     taxon_pft_lookup$phenology[i] = "evergreen"
@@
     taxon_pft_lookup$pft_MRA[i] = "NA"  # taxon class is too broad to specify 
+    taxon_pft_lookup$pft_MODIS[i] = "NA"
     taxon_pft_lookup$phenology[i] = "NA"
@@
     taxon_pft_lookup$pft_MRA[i] = "deciduous_forest"  
+    taxon_pft_lookup$pft_MODIS[i] = "deciduous-broadleaf_forest"
     taxon_pft_lookup$phenology[i] = "deciduous"
@@
     taxon_pft_lookup$pft_MRA[i] = "woody_shrub" 
+    taxon_pft_lookup$pft_MODIS[i] = "shrub"
     taxon_pft_lookup$phenology[i] = "deciduous"
@@
   else if(genusSpecies == "Exostema caribaeum"){
-    taxon_pft_lookup$pft_MRA[i] = "???"  # not deciduous or coniferous
+    taxon_pft_lookup$pft_MRA[i] = "NA"  # not deciduous or coniferous
+    taxon_pft_lookup$pft_MODIS[i] = "evergreen-broadleaf_forest"
     taxon_pft_lookup$phenology[i] = "evergreen"
@@
     taxon_pft_lookup$pft_MRA[i] = "deciduous_forest"    
+    taxon_pft_lookup$pft_MODIS[i] = "deciduous-broadleaf_forest"
     taxon_pft_lookup$phenology[i] = "deciduous"
@@
     taxon_pft_lookup$pft_MRA[i] = "woody_shrub"
+    taxon_pft_lookup$pft_MODIS[i] = "shrub"
     taxon_pft_lookup$phenology[i] = "evergreen"
-    taxon_pft_lookup$leafType[i] = ""
+    taxon_pft_lookup$leafType[i] = "NA"
     taxon_pft_lookup$growthForm[i] = "shrub"
@@
   else if(genusSpecies == "Ilex opaca"){
-    taxon_pft_lookup$pft_MRA[i] = "???"  # not deciduous or coniferous 
+    taxon_pft_lookup$pft_MRA[i] = "NA"  # not deciduous or coniferous 
+    taxon_pft_lookup$pft_MODIS[i] = "evergreen-broadleaf_forest"
     taxon_pft_lookup$phenology[i] = "evergreen"
@@
     taxon_pft_lookup$pft_MRA[i] = "deciduous_forest"
+    taxon_pft_lookup$pft_MODIS[i] = ""
     taxon_pft_lookup$phenology[i] = ""
@@
     taxon_pft_lookup$pft_MRA[i] = "coniferous_forest"
+    taxon_pft_lookup$pft_MODIS[i] = ""
     taxon_pft_lookup$phenology[i] = ""
@@
     taxon_pft_lookup$pft_MRA[i] = "woody_shrub"
+    taxon_pft_lookup$pft_MODIS[i] = ""
     taxon_pft_lookup$phenology[i] = ""
@@
     taxon_pft_lookup$pft_MRA[i] = "deciduous_forest"   # small deciduous tree or large shrub
+    taxon_pft_lookup$pft_MODIS[i] = ""
     taxon_pft_lookup$phenology[i] = ""
@@
  else if(genusSpecies == "Magnolia sp."){
-    taxon_pft_lookup$pft_MRA[i] = "???"
+    taxon_pft_lookup$pft_MRA[i] = "NA"
+    taxon_pft_lookup$pft_MODIS[i] = ""
     taxon_pft_lookup$phenology[i] = ""
@@
     taxon_pft_lookup$pft_MRA[i] = "woody_shrub"      # evergreen shrub
+    taxon_pft_lookup$pft_MODIS[i] = ""
     taxon_pft_lookup$phenology[i] = ""
@@
   else if(genusSpecies == "Mangifera indica"){
-    taxon_pft_lookup$pft_MRA[i] = "???"       # ???   # broadleaf evergreen tree
+    taxon_pft_lookup$pft_MRA[i] = "NA"       # ???   # broadleaf evergreen tree
+    taxon_pft_lookup$pft_MODIS[i] = ""
     taxon_pft_lookup$phenology[i] = ""
@@
     taxon_pft_lookup$pft_MRA[i] = "deciduous_forest"
+    taxon_pft_lookup$pft_MODIS[i] = ""
+    taxon_pft_lookup$phenology[i] = ""
+    taxon_pft_lookup$leafType[i] = ""
+    taxon_pft_lookup$growthForm[i] = ""
     taxon_pft_lookup$commonName[i] = "Chinaberrytree"}
@@
     taxon_pft_lookup$pft_MRA[i] = "deciduous_forest"
+    taxon_pft_lookup$pft_MODIS[i] = ""
+    taxon_pft_lookup$phenology[i] = ""
+    taxon_pft_lookup$leafType[i] = ""
+    taxon_pft_lookup$growthForm[i] = ""
     taxon_pft_lookup$commonName[i] = "red mulberry"}
@@
   else if(genusSpecies == "Pisonia albida"){
-    taxon_pft_lookup$pft_MRA[i] = "???"    # Growth Form - Multiple Stem, Height, Mature - 30ft
+    taxon_pft_lookup$pft_MRA[i] = "NA"    # Growth Form - Multiple Stem, Height, Mature - 30ft
     taxon_pft_lookup$commonName[i] = "corcho bobo"}
@@
   else if(genusSpecies == "Prosopis glandulosa"){
-    taxon_pft_lookup$pft_MRA[i] = "???"   # shrub or tree ranging from 4 to 6m tall
+    taxon_pft_lookup$pft_MRA[i] = "NA"   # shrub or tree ranging from 4 to 6m tall
     taxon_pft_lookup$commonName[i] = "honey mesquite"}
@@
   else if(genusSpecies == "Prosopis velutina"){
-    taxon_pft_lookup$pft_MRA[i] = "???" # deciduous shrub or tree is less than 15 m
+    taxon_pft_lookup$pft_MRA[i] = "NA" # deciduous shrub or tree is less than 15 m
     taxon_pft_lookup$commonName[i] = "velvet mesquite"}
@@
   else if(genusSpecies == "Prunus americana"){
-    taxon_pft_lookup$pft_MRA[i] = "???" # deciduous large shrub or small tree, reaching heights up to 15 feet
+    taxon_pft_lookup$pft_MRA[i] = "NA" # deciduous large shrub or small tree, reaching heights up to 15 feet
     taxon_pft_lookup$commonName[i] = "American plum"}
@@
   else if(genusSpecies == "Prunus mexicana"){
-    taxon_pft_lookup$pft_MRA[i] = "???"
+    taxon_pft_lookup$pft_MRA[i] = "NA"
     taxon_pft_lookup$commonName[i] = "Mexican plum"}
@@
   else if(genusSpecies == "Prunus serotina"){
-    taxon_pft_lookup$pft_MRA[i] = "???"       
+    taxon_pft_lookup$pft_MRA[i] = "NA"       
     taxon_pft_lookup$commonName[i] = "black cherry"}
@@
   else if(genusSpecies == "Quercus sp."){
-    taxon_pft_lookup$pft_MRA[i] = "???"  # apparently there are evergreen oak species? 
+    taxon_pft_lookup$pft_MRA[i] = "NA"  # apparently there are evergreen oak species? 
     taxon_pft_lookup$commonName[i] = "oak"}
@@
   else if(genusSpecies == "Rhamnus davurica"){
-    taxon_pft_lookup$pft_MRA[i] = "???"     #  spreading shrub or small tree reaching 25 to 30 feet
+    taxon_pft_lookup$pft_MRA[i] = "NA"     #  spreading shrub or small tree reaching 25 to 30 feet
     taxon_pft_lookup$commonName[i] = "Dahurian buckthorn"}
@@
   else if(genusSpecies == "Sideroxylon lanuginosum"){
-    taxon_pft_lookup$pft_MRA[i] = "???" # deciduous multi-trunked tall shrub or medium-sized tree to 48ft or more
+    taxon_pft_lookup$pft_MRA[i] = "NA" # deciduous multi-trunked tall shrub or medium-sized tree to 48ft or more
     taxon_pft_lookup$commonName[i] = "gum bully"}
@@
   else if(genusSpecies == "Symplocos sp."){
-    taxon_pft_lookup$pft_MRA[i] = "???"  # "deciduous or evergreen shrub or tree"
+    taxon_pft_lookup$pft_MRA[i] = "NA"  # "deciduous or evergreen shrub or tree"
     taxon_pft_lookup$commonName[i] = "sweetleaf"}

### 2021-04-22 R/add_plant_functional_type.Rmd
working on taxon laebls

@@
     taxon_pft_lookup$pft_MRA[i] = "deciduous_forest"
-    taxon_pft_lookup$pft_MODIS[i] = ""
-    taxon_pft_lookup$phenology[i] = ""
-    taxon_pft_lookup$leafType[i] = ""
-    taxon_pft_lookup$growthForm[i] = ""
+    taxon_pft_lookup$pft_MODIS[i] = "deciduous-broadleaf_forest"
+    taxon_pft_lookup$phenology[i] = "deciduous"
+    taxon_pft_lookup$leafType[i] = "broadleaf"
+    taxon_pft_lookup$growthForm[i] = "tree"
     taxon_pft_lookup$commonName[i] = "black walnut"}
@@
     taxon_pft_lookup$pft_MRA[i] = "coniferous_forest"
-    taxon_pft_lookup$pft_MODIS[i] = ""
-    taxon_pft_lookup$phenology[i] = ""
-    taxon_pft_lookup$leafType[i] = ""
-    taxon_pft_lookup$growthForm[i] = ""
+    taxon_pft_lookup$pft_MODIS[i] = "evergreen_needleleaf_forest"
+    taxon_pft_lookup$phenology[i] = "evergreen"
+    taxon_pft_lookup$leafType[i] = "needleleaf" # scale-like needle leaves
+    taxon_pft_lookup$growthForm[i] = "tree"
     taxon_pft_lookup$commonName[i] = "Utah juniper"}
@@
     taxon_pft_lookup$pft_MRA[i] = "coniferous_forest"
-    taxon_pft_lookup$pft_MODIS[i] = ""
-    taxon_pft_lookup$phenology[i] = ""
-    taxon_pft_lookup$leafType[i] = ""
-    taxon_pft_lookup$growthForm[i] = ""
+    taxon_pft_lookup$pft_MODIS[i] = "evergreen_needleleaf_forest"
+    taxon_pft_lookup$phenology[i] = "evergreen"
+    taxon_pft_lookup$leafType[i] = "needleleaf" # scale-like needle leaves
+    taxon_pft_lookup$growthForm[i] = "tree"
     taxon_pft_lookup$commonName[i] = "red cedar"}
@@
     taxon_pft_lookup$pft_MRA[i] = "coniferous_forest"
-    taxon_pft_lookup$pft_MODIS[i] = ""
-    taxon_pft_lookup$phenology[i] = ""
-    taxon_pft_lookup$leafType[i] = ""
-    taxon_pft_lookup$growthForm[i] = ""
+    taxon_pft_lookup$pft_MODIS[i] = "deciduous-needleleaf_forest"
+    taxon_pft_lookup$phenology[i] = "deciduous" # larch is an unusual genus with deciduous conifers
+    taxon_pft_lookup$leafType[i] = "needleleaf"
+    taxon_pft_lookup$growthForm[i] = "tree"
     taxon_pft_lookup$commonName[i] = "larch"}

### 2021-04-23 R/add_plant_functional_type.Rmd
making progress, using adam's tips about group to identify needle vs broad leaf

@@
 
+I searched for each taxon on USDA plants and determined the phenology and leaf type. 
+- shrubs that are angiosperms are broadleaf
+- group = dicot or monocot, you can say broadleaf
+- group = gymnosperm will be needleleaf
+
+Utilized the NEON vst_apparentindividual data for growthform.
+
 ```{r manually_add_pft_labels}
@@
                                leafType = NA,
-                               growthForm = NA) %>% 
+                               growthForm = NA,
+                               comments = "") %>% 
   # arrange alphabetically
@@
     taxon_pft_lookup$phenology[i] = "deciduous"
-    taxon_pft_lookup$leafType[i] = "NA"
+    taxon_pft_lookup$leafType[i] = "broadleaf"
     taxon_pft_lookup$growthForm[i] = "shrub"
@@
     taxon_pft_lookup$phenology[i] = "evergreen"
-    taxon_pft_lookup$leafType[i] = "NA"
+    taxon_pft_lookup$leafType[i] = "broadleaf"
     taxon_pft_lookup$growthForm[i] = "shrub"
@@
   else if(genusSpecies == "Bourreria sp."){
-    taxon_pft_lookup$pft_MRA[i] = "NA"  # taxon class is too broad to specify 
+    taxon_pft_lookup$pft_MRA[i] = "NA"
     taxon_pft_lookup$pft_MODIS[i] = "NA"
     taxon_pft_lookup$phenology[i] = "NA"
-    taxon_pft_lookup$leafType[i] = "NA"
-    taxon_pft_lookup$growthForm[i] = "tree"
-    taxon_pft_lookup$commonName[i] = "strongbark"}
+    taxon_pft_lookup$leafType[i] = "broadleaf"
+    taxon_pft_lookup$growthForm[i] = "tree"
+    taxon_pft_lookup$commonName[i] = "strongbark"
+    taxon_pft_lookup$commonName[i] = "varied genus"}
 
@@
     taxon_pft_lookup$phenology[i] = "evergreen"
-    taxon_pft_lookup$leafType[i] = "NA"
+    taxon_pft_lookup$leafType[i] = "needleleaf"
     taxon_pft_lookup$growthForm[i] = "shrub"
@@
     taxon_pft_lookup$phenology[i] = "evergreen"
-    taxon_pft_lookup$leafType[i] = "needleleaf" # scale-like needle leaves
-    taxon_pft_lookup$growthForm[i] = "tree"
-    taxon_pft_lookup$commonName[i] = "Utah juniper"}
+    taxon_pft_lookup$leafType[i] = "needleleaf"
+    taxon_pft_lookup$growthForm[i] = "tree"
+    taxon_pft_lookup$commonName[i] = "Utah juniper"
+    taxon_pft_lookup$comments[i] = "scale-like needle leaves"}
 
@@
     taxon_pft_lookup$phenology[i] = "evergreen"
-    taxon_pft_lookup$leafType[i] = "needleleaf" # scale-like needle leaves
-    taxon_pft_lookup$growthForm[i] = "tree"
-    taxon_pft_lookup$commonName[i] = "red cedar"}
+    taxon_pft_lookup$leafType[i] = "needleleaf" 
+    taxon_pft_lookup$growthForm[i] = "tree"
+    taxon_pft_lookup$commonName[i] = "red cedar"
+    taxon_pft_lookup$comments[i] = "scale-like needle leaves"}
 
@@
     taxon_pft_lookup$pft_MODIS[i] = "deciduous-needleleaf_forest"
-    taxon_pft_lookup$phenology[i] = "deciduous" # larch is an unusual genus with deciduous conifers
-    taxon_pft_lookup$leafType[i] = "needleleaf"
-    taxon_pft_lookup$growthForm[i] = "tree"
-    taxon_pft_lookup$commonName[i] = "larch"}
+    taxon_pft_lookup$phenology[i] = "deciduous"
+    taxon_pft_lookup$leafType[i] = "needleleaf"
+    taxon_pft_lookup$growthForm[i] = "tree"
+    taxon_pft_lookup$commonName[i] = "larch"
+    taxon_pft_lookup$comments[i] = "unusual genus with deciduous conifers"}
 
@@
     taxon_pft_lookup$pft_MRA[i] = "woody_shrub"
-    taxon_pft_lookup$pft_MODIS[i] = ""
-    taxon_pft_lookup$phenology[i] = ""
-    taxon_pft_lookup$leafType[i] = ""
-    taxon_pft_lookup$growthForm[i] = ""
+    taxon_pft_lookup$pft_MODIS[i] = "shrub"
+    taxon_pft_lookup$phenology[i] = "evergreen"
+    taxon_pft_lookup$leafType[i] = "broadleaf"
+    taxon_pft_lookup$growthForm[i] = "shrub"
     taxon_pft_lookup$commonName[i] = "creosote bush"}
@@
     taxon_pft_lookup$pft_MRA[i] = "deciduous_forest"
-    taxon_pft_lookup$pft_MODIS[i] = ""
-    taxon_pft_lookup$phenology[i] = ""
-    taxon_pft_lookup$leafType[i] = ""
-    taxon_pft_lookup$growthForm[i] = ""
+    taxon_pft_lookup$pft_MODIS[i] = "deciduous-broadleaf_forest"
+    taxon_pft_lookup$phenology[i] = "deciduous"
+    taxon_pft_lookup$leafType[i] = "broadleaf"
+    taxon_pft_lookup$growthForm[i] = "tree"
     taxon_pft_lookup$commonName[i] = "sweetgum"}
@@
     taxon_pft_lookup$pft_MRA[i] = "deciduous_forest"
-    taxon_pft_lookup$pft_MODIS[i] = ""
-    taxon_pft_lookup$phenology[i] = ""
-    taxon_pft_lookup$leafType[i] = ""
-    taxon_pft_lookup$growthForm[i] = ""
+    taxon_pft_lookup$pft_MODIS[i] = "deciduous-broadleaf_forest"
+    taxon_pft_lookup$phenology[i] = "deciduous"
+    taxon_pft_lookup$leafType[i] = "broadleaf"
+    taxon_pft_lookup$growthForm[i] = "tree"
     taxon_pft_lookup$commonName[i] = "tuliptree"}
@@
     taxon_pft_lookup$pft_MRA[i] = "woody_shrub"
-    taxon_pft_lookup$pft_MODIS[i] = ""
-    taxon_pft_lookup$phenology[i] = ""
-    taxon_pft_lookup$leafType[i] = ""
-    taxon_pft_lookup$growthForm[i] = ""
+    taxon_pft_lookup$pft_MODIS[i] = "shrub"
+    taxon_pft_lookup$phenology[i] = "deciduous"
+    taxon_pft_lookup$leafType[i] = "broadleaf"
+    taxon_pft_lookup$growthForm[i] = "shrub"
     taxon_pft_lookup$commonName[i] = "Amur honeysuckle"}
@@
   else if(genusSpecies == "Maclura pomifera"){
-    taxon_pft_lookup$pft_MRA[i] = "deciduous_forest"   # small deciduous tree or large shrub
-    taxon_pft_lookup$pft_MODIS[i] = ""
-    taxon_pft_lookup$phenology[i] = ""
-    taxon_pft_lookup$leafType[i] = ""
-    taxon_pft_lookup$growthForm[i] = ""
+    taxon_pft_lookup$pft_MRA[i] = "deciduous_forest"
+    taxon_pft_lookup$pft_MODIS[i] = "deciduous-broadleaf_forest"
+    taxon_pft_lookup$phenology[i] = "deciduous"
+    taxon_pft_lookup$leafType[i] = "broadleaf"
+    taxon_pft_lookup$growthForm[i] = "tree" 
     taxon_pft_lookup$commonName[i] = "Osage-orange"}
@@
     taxon_pft_lookup$pft_MRA[i] = "NA"
-    taxon_pft_lookup$pft_MODIS[i] = ""
-    taxon_pft_lookup$phenology[i] = ""
-    taxon_pft_lookup$leafType[i] = ""
-    taxon_pft_lookup$growthForm[i] = ""
-    taxon_pft_lookup$commonName[i] = "magnolia"}
+    taxon_pft_lookup$pft_MODIS[i] = "NA"
+    taxon_pft_lookup$phenology[i] = "NA"
+    taxon_pft_lookup$leafType[i] = "broadleaf"
+    taxon_pft_lookup$growthForm[i] = "NA"
+    taxon_pft_lookup$commonName[i] = "magnolia"
+    taxon_pft_lookup$comments[i] = "varied genus with both deciduous and evergreen trees and shrubs"}
 
   else if(genusSpecies == "Mahonia nervosa"){
-    taxon_pft_lookup$pft_MRA[i] = "woody_shrub"      # evergreen shrub
-    taxon_pft_lookup$pft_MODIS[i] = ""
-    taxon_pft_lookup$phenology[i] = ""
-    taxon_pft_lookup$leafType[i] = ""
-    taxon_pft_lookup$growthForm[i] = ""
+    taxon_pft_lookup$pft_MRA[i] = "woody_shrub"
+    taxon_pft_lookup$pft_MODIS[i] = "shrub"
+    taxon_pft_lookup$phenology[i] = "evergreen"
+    taxon_pft_lookup$leafType[i] = "broadleaf"
+    taxon_pft_lookup$growthForm[i] = "shrub"
     taxon_pft_lookup$commonName[i] = "Cascade barberry"}
@@
   else if(genusSpecies == "Mangifera indica"){
-    taxon_pft_lookup$pft_MRA[i] = "NA"       # ???   # broadleaf evergreen tree
-    taxon_pft_lookup$pft_MODIS[i] = ""
-    taxon_pft_lookup$phenology[i] = ""
-    taxon_pft_lookup$leafType[i] = ""
-    taxon_pft_lookup$growthForm[i] = ""
+    taxon_pft_lookup$pft_MRA[i] = "NA" # not coniferous or deciduous
+    taxon_pft_lookup$pft_MODIS[i] = "evergreen-broadleaf_forest"
+    taxon_pft_lookup$phenology[i] = "evergreen"
+    taxon_pft_lookup$leafType[i] = "broadleaf"
+    taxon_pft_lookup$growthForm[i] = "tree"
     taxon_pft_lookup$commonName[i] = "mango"}
@@
     taxon_pft_lookup$pft_MRA[i] = "deciduous_forest"
-    taxon_pft_lookup$pft_MODIS[i] = ""
-    taxon_pft_lookup$phenology[i] = ""
-    taxon_pft_lookup$leafType[i] = ""
-    taxon_pft_lookup$growthForm[i] = ""
+    taxon_pft_lookup$pft_MODIS[i] = "deciduous-broadleaf_forest"
+    taxon_pft_lookup$phenology[i] = "deciduous"
+    taxon_pft_lookup$leafType[i] = "broadleaf"
+    taxon_pft_lookup$growthForm[i] = "tree"
     taxon_pft_lookup$commonName[i] = "Chinaberrytree"}
@@
     taxon_pft_lookup$pft_MRA[i] = "deciduous_forest"
-    taxon_pft_lookup$pft_MODIS[i] = ""
-    taxon_pft_lookup$phenology[i] = ""
-    taxon_pft_lookup$leafType[i] = ""
-    taxon_pft_lookup$growthForm[i] = ""
+    taxon_pft_lookup$pft_MODIS[i] = "deciduous-broadleaf_forest"
+    taxon_pft_lookup$phenology[i] = "deciduous"
+    taxon_pft_lookup$leafType[i] = "broadleaf"
+    taxon_pft_lookup$growthForm[i] = "tree"
     taxon_pft_lookup$commonName[i] = "red mulberry"}
@@
     taxon_pft_lookup$pft_MRA[i] = "deciduous_forest"
+    taxon_pft_lookup$pft_MODIS[i] = "deciduous-broadleaf_forest"
+    taxon_pft_lookup$phenology[i] = "deciduous"
+    taxon_pft_lookup$leafType[i] = "broadleaf"
+    taxon_pft_lookup$growthForm[i] = "tree"
     taxon_pft_lookup$commonName[i] = "water tupelo"}
@@
     taxon_pft_lookup$pft_MRA[i] = "deciduous_forest"
+    taxon_pft_lookup$pft_MODIS[i] = "deciduous-broadleaf_forest"
+    taxon_pft_lookup$phenology[i] = "deciduous"
+    taxon_pft_lookup$leafType[i] = "broadleaf"
+    taxon_pft_lookup$growthForm[i] = "tree"
     taxon_pft_lookup$commonName[i] = "swamp tupelo"}
@@
     taxon_pft_lookup$pft_MRA[i] = "deciduous_forest"
+    taxon_pft_lookup$pft_MODIS[i] = "deciduous-broadleaf_forest"
+    taxon_pft_lookup$phenology[i] = "deciduous"
+    taxon_pft_lookup$leafType[i] = "broadleaf"
+    taxon_pft_lookup$growthForm[i] = "tree"
     taxon_pft_lookup$commonName[i] = "blackgum"}
@@
     taxon_pft_lookup$pft_MRA[i] = "deciduous_forest"
+    taxon_pft_lookup$pft_MODIS[i] = "deciduous-broadleaf_forest"
+    taxon_pft_lookup$phenology[i] = "deciduous"
+    taxon_pft_lookup$leafType[i] = "broadleaf"
+    taxon_pft_lookup$growthForm[i] = "tree"
     taxon_pft_lookup$commonName[i] = "sourwood"}
@@
     taxon_pft_lookup$pft_MRA[i] = "coniferous_forest"
+    taxon_pft_lookup$pft_MODIS[i] = "evergreen-needleleaf_forest"
+    taxon_pft_lookup$phenology[i] = "evergreen"
+    taxon_pft_lookup$leafType[i] = "needleleaf"
+    taxon_pft_lookup$growthForm[i] = "tree"
     taxon_pft_lookup$commonName[i] = "Engelmann spruce"}
@@
     taxon_pft_lookup$pft_MRA[i] = "coniferous_forest"
+    taxon_pft_lookup$pft_MODIS[i] = "evergreen-needleleaf_forest"
+    taxon_pft_lookup$phenology[i] = "evergreen"
+    taxon_pft_lookup$leafType[i] = "needleleaf"
+    taxon_pft_lookup$growthForm[i] = "tree"
     taxon_pft_lookup$commonName[i] = "white spruce"}
@@
     taxon_pft_lookup$pft_MRA[i] = "coniferous_forest"
+    taxon_pft_lookup$pft_MODIS[i] = "evergreen-needleleaf_forest"
+    taxon_pft_lookup$phenology[i] = "evergreen"
+    taxon_pft_lookup$leafType[i] = "needleleaf"
+    taxon_pft_lookup$growthForm[i] = "tree"
     taxon_pft_lookup$commonName[i] = "black spruce"}
@@
     taxon_pft_lookup$pft_MRA[i] = "coniferous_forest"
+    taxon_pft_lookup$pft_MODIS[i] = "evergreen-needleleaf_forest"
+    taxon_pft_lookup$phenology[i] = "evergreen"
+    taxon_pft_lookup$leafType[i] = "needleleaf"
+    taxon_pft_lookup$growthForm[i] = "tree"
     taxon_pft_lookup$commonName[i] = "spruce"}
@@
     taxon_pft_lookup$pft_MRA[i] = "coniferous_forest"
+    taxon_pft_lookup$pft_MODIS[i] = "evergreen-needleleaf_forest"
+    taxon_pft_lookup$phenology[i] = "evergreen"
+    taxon_pft_lookup$leafType[i] = "needleleaf"
+    taxon_pft_lookup$growthForm[i] = "tree"
     taxon_pft_lookup$commonName[i] = "lodgepole pine"}
@@
     taxon_pft_lookup$pft_MRA[i] = "coniferous_forest"
+    taxon_pft_lookup$pft_MODIS[i] = "evergreen-needleleaf_forest"
+    taxon_pft_lookup$phenology[i] = "evergreen"
+    taxon_pft_lookup$leafType[i] = "needleleaf"
+    taxon_pft_lookup$growthForm[i] = "tree"
     taxon_pft_lookup$commonName[i] = "shortleaf pine"}
@@
     taxon_pft_lookup$pft_MRA[i] = "coniferous_forest"
+    taxon_pft_lookup$pft_MODIS[i] = "evergreen-needleleaf_forest"
+    taxon_pft_lookup$phenology[i] = "evergreen"
+    taxon_pft_lookup$leafType[i] = "needleleaf"
+    taxon_pft_lookup$growthForm[i] = "tree"
     taxon_pft_lookup$commonName[i] = "limber pine"}
@@
     taxon_pft_lookup$pft_MRA[i] = "coniferous_forest"
+    taxon_pft_lookup$pft_MODIS[i] = "evergreen-needleleaf_forest"
+    taxon_pft_lookup$phenology[i] = "evergreen"
+    taxon_pft_lookup$leafType[i] = "needleleaf"
+    taxon_pft_lookup$growthForm[i] = "tree"
     taxon_pft_lookup$commonName[i] = "longleaf pine"}
@@
     taxon_pft_lookup$pft_MRA[i] = "coniferous_forest"
+    taxon_pft_lookup$pft_MODIS[i] = "evergreen-needleleaf_forest"
+    taxon_pft_lookup$phenology[i] = "evergreen"
+    taxon_pft_lookup$leafType[i] = "needleleaf"
+    taxon_pft_lookup$growthForm[i] = "tree"
     taxon_pft_lookup$commonName[i] = "ponderosa pine"}
@@
     taxon_pft_lookup$pft_MRA[i] = "coniferous_forest"
+    taxon_pft_lookup$pft_MODIS[i] = "evergreen-needleleaf_forest"
+    taxon_pft_lookup$phenology[i] = "evergreen"
+    taxon_pft_lookup$leafType[i] = "needleleaf"
+    taxon_pft_lookup$growthForm[i] = "tree"
     taxon_pft_lookup$commonName[i] = "pine"}
@@
     taxon_pft_lookup$pft_MRA[i] = "coniferous_forest"
+    taxon_pft_lookup$pft_MODIS[i] = "evergreen-needleleaf_forest"
+    taxon_pft_lookup$phenology[i] = "evergreen"
+    taxon_pft_lookup$leafType[i] = "needleleaf"
+    taxon_pft_lookup$growthForm[i] = "tree"
     taxon_pft_lookup$commonName[i] = "eastern white pine"}
@@
     taxon_pft_lookup$pft_MRA[i] = "coniferous_forest"
+    taxon_pft_lookup$pft_MODIS[i] = "evergreen-needleleaf_forest"
+    taxon_pft_lookup$phenology[i] = "evergreen"
+    taxon_pft_lookup$leafType[i] = "needleleaf"
+    taxon_pft_lookup$growthForm[i] = "tree"
     taxon_pft_lookup$commonName[i] = "loblolly pine"}
@@
   else if(genusSpecies == "Pisonia albida"){
-    taxon_pft_lookup$pft_MRA[i] = "NA"    # Growth Form - Multiple Stem, Height, Mature - 30ft
+    taxon_pft_lookup$pft_MRA[i] = "deciduous_forest"
+    taxon_pft_lookup$pft_MODIS[i] = "deciduous-broadleaf_forest"
+    taxon_pft_lookup$phenology[i] = "deciduous"
+    taxon_pft_lookup$leafType[i] = "broadleaf"
+    taxon_pft_lookup$growthForm[i] = "tree"
     taxon_pft_lookup$commonName[i] = "corcho bobo"}
@@
     taxon_pft_lookup$pft_MRA[i] = "deciduous_forest"
+    taxon_pft_lookup$pft_MODIS[i] = "deciduous-broadleaf_forest"
+    taxon_pft_lookup$phenology[i] = "deciduous"
+    taxon_pft_lookup$leafType[i] = "broadleaf"
+    taxon_pft_lookup$growthForm[i] = "tree"
     taxon_pft_lookup$commonName[i] = "American sycamore"}
@@
     taxon_pft_lookup$pft_MRA[i] = "deciduous_forest"
+    taxon_pft_lookup$pft_MODIS[i] = "deciduous-broadleaf_forest"
+    taxon_pft_lookup$phenology[i] = "deciduous"
+    taxon_pft_lookup$leafType[i] = "broadleaf"
+    taxon_pft_lookup$growthForm[i] = "tree"
     taxon_pft_lookup$commonName[i] = "eastern cottonwood"}
@@
     taxon_pft_lookup$pft_MRA[i] = "deciduous_forest"
+    taxon_pft_lookup$pft_MODIS[i] = "deciduous-broadleaf_forest"
+    taxon_pft_lookup$phenology[i] = "deciduous"
+    taxon_pft_lookup$leafType[i] = "broadleaf"
+    taxon_pft_lookup$growthForm[i] = "tree"
     taxon_pft_lookup$commonName[i] = "bigtooth aspen"}
@@
     taxon_pft_lookup$pft_MRA[i] = "deciduous_forest"
+    taxon_pft_lookup$pft_MODIS[i] = "deciduous-broadleaf_forest"
+    taxon_pft_lookup$phenology[i] = "deciduous"
+    taxon_pft_lookup$leafType[i] = "broadleaf"
+    taxon_pft_lookup$growthForm[i] = "tree"
     taxon_pft_lookup$commonName[i] = "cottonwood"}
@@
     taxon_pft_lookup$pft_MRA[i] = "deciduous_forest"
+    taxon_pft_lookup$pft_MODIS[i] = "deciduous-broadleaf_forest"
+    taxon_pft_lookup$phenology[i] = "deciduous"
+    taxon_pft_lookup$leafType[i] = "broadleaf"
+    taxon_pft_lookup$growthForm[i] = "tree"
     taxon_pft_lookup$commonName[i] = "quaking aspen"}
@@
   else if(genusSpecies == "Prosopis glandulosa"){
-    taxon_pft_lookup$pft_MRA[i] = "NA"   # shrub or tree ranging from 4 to 6m tall
+    taxon_pft_lookup$pft_MRA[i] = "woody_shrub"
+    taxon_pft_lookup$pft_MODIS[i] = "shrub"
+    taxon_pft_lookup$phenology[i] = "deciduous"
+    taxon_pft_lookup$leafType[i] = "broadleaf"
+    taxon_pft_lookup$growthForm[i] = "shrub"
     taxon_pft_lookup$commonName[i] = "honey mesquite"}
@@
   else if(genusSpecies == "Prosopis velutina"){
-    taxon_pft_lookup$pft_MRA[i] = "NA" # deciduous shrub or tree is less than 15 m
+    taxon_pft_lookup$pft_MRA[i] = "woody_shrub" 
+    taxon_pft_lookup$pft_MODIS[i] = "shrub"
+    taxon_pft_lookup$phenology[i] = "deciduous"
+    taxon_pft_lookup$leafType[i] = "broadleaf"
+    taxon_pft_lookup$growthForm[i] = "shrub"
     taxon_pft_lookup$commonName[i] = "velvet mesquite"}
@@
   else if(genusSpecies == "Prunus americana"){
-    taxon_pft_lookup$pft_MRA[i] = "NA" # deciduous large shrub or small tree, reaching heights up to 15 feet
+    taxon_pft_lookup$pft_MRA[i] = "deciduous_forest"
+    taxon_pft_lookup$pft_MODIS[i] = "deciduous-broadleaf_forest"
+    taxon_pft_lookup$phenology[i] = "deciduous"
+    taxon_pft_lookup$leafType[i] = "broadleaf"
+    taxon_pft_lookup$growthForm[i] = "tree"
     taxon_pft_lookup$commonName[i] = "American plum"}
@@
   else if(genusSpecies == "Prunus angustifolia"){
-    taxon_pft_lookup$pft_MRA[i] = "woody_shrub" # short, thickly branched shrub 3 to 10ft tall
+    taxon_pft_lookup$pft_MRA[i] = "woody_shrub" 
+    taxon_pft_lookup$pft_MODIS[i] = "shrub"
+    taxon_pft_lookup$phenology[i] = "deciduous"
+    taxon_pft_lookup$leafType[i] = "broadleaf"
+    taxon_pft_lookup$growthForm[i] = "shrub"
     taxon_pft_lookup$commonName[i] = "Chickasaw plum"}
@@
   else if(genusSpecies == "Prunus mexicana"){
-    taxon_pft_lookup$pft_MRA[i] = "NA"
+    taxon_pft_lookup$pft_MRA[i] = "deciduous_forest"
+    taxon_pft_lookup$pft_MODIS[i] = "deciduous-broadleaf_forest"
+    taxon_pft_lookup$phenology[i] = "deciduous"
+    taxon_pft_lookup$leafType[i] = "broadleaf"
+    taxon_pft_lookup$growthForm[i] = "tree"
     taxon_pft_lookup$commonName[i] = "Mexican plum"}
@@
   else if(genusSpecies == "Prunus serotina"){
-    taxon_pft_lookup$pft_MRA[i] = "NA"       
+    taxon_pft_lookup$pft_MRA[i] = "deciduous_forest"
+    taxon_pft_lookup$pft_MODIS[i] = "deciduous-broadleaf_forest"
+    taxon_pft_lookup$phenology[i] = "deciduous"
+    taxon_pft_lookup$leafType[i] = "broadleaf"
+    taxon_pft_lookup$growthForm[i] = "tree"
     taxon_pft_lookup$commonName[i] = "black cherry"}
@@
     taxon_pft_lookup$pft_MRA[i] = "coniferous_forest"
+    taxon_pft_lookup$pft_MODIS[i] = "evergreen-needleleaf_forest"
+    taxon_pft_lookup$phenology[i] = "evergreen"
+    taxon_pft_lookup$leafType[i] = "needleleaf"
+    taxon_pft_lookup$growthForm[i] = "tree"
     taxon_pft_lookup$commonName[i] = "Douglas-fir"}
@@
     taxon_pft_lookup$pft_MRA[i] = "deciduous_forest"
+    taxon_pft_lookup$pft_MODIS[i] = "deciduous-broadleaf_forest"
+    taxon_pft_lookup$phenology[i] = "deciduous"
+    taxon_pft_lookup$leafType[i] = "broadleaf"
+    taxon_pft_lookup$growthForm[i] = "tree"
     taxon_pft_lookup$commonName[i] = "white oak"}
@@
     taxon_pft_lookup$pft_MRA[i] = "deciduous_forest"
+    taxon_pft_lookup$pft_MODIS[i] = "deciduous-broadleaf_forest"
+    taxon_pft_lookup$phenology[i] = "deciduous"
+    taxon_pft_lookup$leafType[i] = "broadleaf"
+    taxon_pft_lookup$growthForm[i] = "tree"
     taxon_pft_lookup$commonName[i] = "scarlet oak"}
@@
     taxon_pft_lookup$pft_MRA[i] = "deciduous_forest"
+    taxon_pft_lookup$pft_MODIS[i] = ""
+    taxon_pft_lookup$phenology[i] = ""
+    taxon_pft_lookup$leafType[i] = ""
+    taxon_pft_lookup$growthForm[i] = ""
     taxon_pft_lookup$commonName[i] = "southern red oak"}
@@
     taxon_pft_lookup$pft_MRA[i] = "deciduous_forest"
+    taxon_pft_lookup$pft_MODIS[i] = ""
+    taxon_pft_lookup$phenology[i] = ""
+    taxon_pft_lookup$leafType[i] = ""
+    taxon_pft_lookup$growthForm[i] = ""
     taxon_pft_lookup$commonName[i] = "sand live oak"}
@@
     taxon_pft_lookup$pft_MRA[i] = "deciduous_forest"
+    taxon_pft_lookup$pft_MODIS[i] = ""
+    taxon_pft_lookup$phenology[i] = ""
+    taxon_pft_lookup$leafType[i] = ""
+    taxon_pft_lookup$growthForm[i] = ""
     taxon_pft_lookup$commonName[i] = "Darlington oak"}
@@
     taxon_pft_lookup$pft_MRA[i] = "deciduous_forest"
+    taxon_pft_lookup$pft_MODIS[i] = ""
+    taxon_pft_lookup$phenology[i] = ""
+    taxon_pft_lookup$leafType[i] = ""
+    taxon_pft_lookup$growthForm[i] = ""
     taxon_pft_lookup$commonName[i] = "turkey oak"}
@@
     taxon_pft_lookup$pft_MRA[i] = "deciduous_forest"
+    taxon_pft_lookup$pft_MODIS[i] = ""
+    taxon_pft_lookup$phenology[i] = ""
+    taxon_pft_lookup$leafType[i] = ""
+    taxon_pft_lookup$growthForm[i] = ""
     taxon_pft_lookup$commonName[i] = "laurel oak"}
@@
     taxon_pft_lookup$pft_MRA[i] = "deciduous_forest"
+    taxon_pft_lookup$pft_MODIS[i] = ""
+    taxon_pft_lookup$phenology[i] = ""
+    taxon_pft_lookup$leafType[i] = ""
+    taxon_pft_lookup$growthForm[i] = ""
     taxon_pft_lookup$commonName[i] = "overcup oak"}
@@
     taxon_pft_lookup$pft_MRA[i] = "deciduous_forest"
+    taxon_pft_lookup$pft_MODIS[i] = ""
+    taxon_pft_lookup$phenology[i] = ""
+    taxon_pft_lookup$leafType[i] = ""
+    taxon_pft_lookup$growthForm[i] = ""
     taxon_pft_lookup$commonName[i] = "bur oak"}
@@
     taxon_pft_lookup$pft_MRA[i] = "deciduous_forest"
+    taxon_pft_lookup$pft_MODIS[i] = ""
+    taxon_pft_lookup$phenology[i] = ""
+    taxon_pft_lookup$leafType[i] = ""
+    taxon_pft_lookup$growthForm[i] = ""
     taxon_pft_lookup$commonName[i] = "sand post oak"}
@@
     taxon_pft_lookup$pft_MRA[i] = "deciduous_forest"
+    taxon_pft_lookup$pft_MODIS[i] = ""
+    taxon_pft_lookup$phenology[i] = ""
+    taxon_pft_lookup$leafType[i] = ""
+    taxon_pft_lookup$growthForm[i] = ""
     taxon_pft_lookup$commonName[i] = "blackjack oak"}
@@
     taxon_pft_lookup$pft_MRA[i] = "deciduous_forest"
+    taxon_pft_lookup$pft_MODIS[i] = ""
+    taxon_pft_lookup$phenology[i] = ""
+    taxon_pft_lookup$leafType[i] = ""
+    taxon_pft_lookup$growthForm[i] = ""
     taxon_pft_lookup$commonName[i] = "swamp chestnut oak"}
@@
     taxon_pft_lookup$pft_MRA[i] = "deciduous_forest"
+    taxon_pft_lookup$pft_MODIS[i] = ""
+    taxon_pft_lookup$phenology[i] = ""
+    taxon_pft_lookup$leafType[i] = ""
+    taxon_pft_lookup$growthForm[i] = ""
     taxon_pft_lookup$commonName[i] = "chestnut oak"}
@@
     taxon_pft_lookup$pft_MRA[i] = "deciduous_forest"
+    taxon_pft_lookup$pft_MODIS[i] = ""
+    taxon_pft_lookup$phenology[i] = ""
+    taxon_pft_lookup$leafType[i] = ""
+    taxon_pft_lookup$growthForm[i] = ""
     taxon_pft_lookup$commonName[i] = "chinquapin oak"}
@@
     taxon_pft_lookup$pft_MRA[i] = "deciduous_forest"
+    taxon_pft_lookup$pft_MODIS[i] = ""
+    taxon_pft_lookup$phenology[i] = ""
+    taxon_pft_lookup$leafType[i] = ""
+    taxon_pft_lookup$growthForm[i] = ""
     taxon_pft_lookup$commonName[i] = "water oak"}
@@
     taxon_pft_lookup$pft_MRA[i] = "deciduous_forest"
+    taxon_pft_lookup$pft_MODIS[i] = ""
+    taxon_pft_lookup$phenology[i] = ""
+    taxon_pft_lookup$leafType[i] = ""
+    taxon_pft_lookup$growthForm[i] = ""
     taxon_pft_lookup$commonName[i] = "cherrybark oak"}
@@
     taxon_pft_lookup$pft_MRA[i] = "deciduous_forest"
+    taxon_pft_lookup$pft_MODIS[i] = ""
+    taxon_pft_lookup$phenology[i] = ""
+    taxon_pft_lookup$leafType[i] = ""
+    taxon_pft_lookup$growthForm[i] = ""
     taxon_pft_lookup$commonName[i] = "willow oak"}
@@
     taxon_pft_lookup$pft_MRA[i] = "deciduous_forest"
+    taxon_pft_lookup$pft_MODIS[i] = ""
+    taxon_pft_lookup$phenology[i] = ""
+    taxon_pft_lookup$leafType[i] = ""
+    taxon_pft_lookup$growthForm[i] = ""
     taxon_pft_lookup$commonName[i] = "northern red oak"}
@@
     taxon_pft_lookup$pft_MRA[i] = "deciduous_forest"
+    taxon_pft_lookup$pft_MODIS[i] = ""
+    taxon_pft_lookup$phenology[i] = ""
+    taxon_pft_lookup$leafType[i] = ""
+    taxon_pft_lookup$growthForm[i] = ""
     taxon_pft_lookup$commonName[i] = "Shumard's oak"}
@@
     taxon_pft_lookup$pft_MRA[i] = "NA"  # apparently there are evergreen oak species? 
+    taxon_pft_lookup$pft_MODIS[i] = ""
+    taxon_pft_lookup$phenology[i] = ""
+    taxon_pft_lookup$leafType[i] = ""
+    taxon_pft_lookup$growthForm[i] = ""
     taxon_pft_lookup$commonName[i] = "oak"}
@@
     taxon_pft_lookup$pft_MRA[i] = "deciduous_forest"
+    taxon_pft_lookup$pft_MODIS[i] = ""
+    taxon_pft_lookup$phenology[i] = ""
+    taxon_pft_lookup$leafType[i] = ""
+    taxon_pft_lookup$growthForm[i] = ""
     taxon_pft_lookup$commonName[i] = "post oak"}
@@
     taxon_pft_lookup$pft_MRA[i] = "deciduous_forest"
+    taxon_pft_lookup$pft_MODIS[i] = ""
+    taxon_pft_lookup$phenology[i] = ""
+    taxon_pft_lookup$leafType[i] = ""
+    taxon_pft_lookup$growthForm[i] = ""
     taxon_pft_lookup$commonName[i] = "black oak"}
@@
     taxon_pft_lookup$pft_MRA[i] = "deciduous_forest"
+    taxon_pft_lookup$pft_MODIS[i] = ""
+    taxon_pft_lookup$phenology[i] = ""
+    taxon_pft_lookup$leafType[i] = ""
+    taxon_pft_lookup$growthForm[i] = ""
     taxon_pft_lookup$commonName[i] = "live oak"}
@@
     taxon_pft_lookup$pft_MRA[i] = "NA"     #  spreading shrub or small tree reaching 25 to 30 feet
+    taxon_pft_lookup$pft_MODIS[i] = ""
+    taxon_pft_lookup$phenology[i] = ""
+    taxon_pft_lookup$leafType[i] = ""
+    taxon_pft_lookup$growthForm[i] = ""
     taxon_pft_lookup$commonName[i] = "Dahurian buckthorn"}
@@
     taxon_pft_lookup$pft_MRA[i] = "woody_shrub"
+    taxon_pft_lookup$pft_MODIS[i] = ""
+    taxon_pft_lookup$phenology[i] = ""
+    taxon_pft_lookup$leafType[i] = ""
+    taxon_pft_lookup$growthForm[i] = ""
     taxon_pft_lookup$commonName[i] = "fragrant sumac"}
@@
     taxon_pft_lookup$pft_MRA[i] = "woody_shrub"
+    taxon_pft_lookup$pft_MODIS[i] = ""
+    taxon_pft_lookup$phenology[i] = ""
+    taxon_pft_lookup$leafType[i] = ""
+    taxon_pft_lookup$growthForm[i] = ""
     taxon_pft_lookup$commonName[i] = "smooth sumac"}
@@
     taxon_pft_lookup$pft_MRA[i] = "deciduous_forest"
+    taxon_pft_lookup$pft_MODIS[i] = ""
+    taxon_pft_lookup$phenology[i] = ""
+    taxon_pft_lookup$leafType[i] = ""
+    taxon_pft_lookup$growthForm[i] = ""
     taxon_pft_lookup$commonName[i] = "black locust"}
@@
     taxon_pft_lookup$pft_MRA[i] = "deciduous_forest" # deciduous shrub or tree. trees are 9 to 18 m
+    taxon_pft_lookup$pft_MODIS[i] = ""
+    taxon_pft_lookup$phenology[i] = ""
+    taxon_pft_lookup$leafType[i] = ""
+    taxon_pft_lookup$growthForm[i] = ""
     taxon_pft_lookup$commonName[i] = "sassafras"}
@@
     taxon_pft_lookup$pft_MRA[i] = "deciduous_forest"
+    taxon_pft_lookup$pft_MODIS[i] = ""
+    taxon_pft_lookup$phenology[i] = ""
+    taxon_pft_lookup$leafType[i] = ""
+    taxon_pft_lookup$growthForm[i] = ""
     taxon_pft_lookup$commonName[i] = "sassafras"}
@@
     taxon_pft_lookup$pft_MRA[i] = "NA" # deciduous multi-trunked tall shrub or medium-sized tree to 48ft or more
+    taxon_pft_lookup$pft_MODIS[i] = ""
+    taxon_pft_lookup$phenology[i] = ""
+    taxon_pft_lookup$leafType[i] = ""
+    taxon_pft_lookup$growthForm[i] = ""
     taxon_pft_lookup$commonName[i] = "gum bully"}
@@
     taxon_pft_lookup$pft_MRA[i] = "NA"  # "deciduous or evergreen shrub or tree"
+    taxon_pft_lookup$pft_MODIS[i] = ""
+    taxon_pft_lookup$phenology[i] = ""
+    taxon_pft_lookup$leafType[i] = ""
+    taxon_pft_lookup$growthForm[i] = ""
     taxon_pft_lookup$commonName[i] = "sweetleaf"}
@@
     taxon_pft_lookup$pft_MRA[i] = "coniferous_forest"
+    taxon_pft_lookup$pft_MODIS[i] = ""
+    taxon_pft_lookup$phenology[i] = ""
+    taxon_pft_lookup$leafType[i] = ""
+    taxon_pft_lookup$growthForm[i] = ""
     taxon_pft_lookup$commonName[i] = "Pacific yew"}
@@
     taxon_pft_lookup$pft_MRA[i] = "deciduous_forest"
+    taxon_pft_lookup$pft_MODIS[i] = ""
+    taxon_pft_lookup$phenology[i] = ""
+    taxon_pft_lookup$leafType[i] = ""
+    taxon_pft_lookup$growthForm[i] = ""
     taxon_pft_lookup$commonName[i] = "bee-bee tree"}
@@
     taxon_pft_lookup$pft_MRA[i] = "deciduous_forest"
+    taxon_pft_lookup$pft_MODIS[i] = ""
+    taxon_pft_lookup$phenology[i] = ""
+    taxon_pft_lookup$leafType[i] = ""
+    taxon_pft_lookup$growthForm[i] = ""
     taxon_pft_lookup$commonName[i] = "Chinese tallow"}
@@
     taxon_pft_lookup$pft_MRA[i] = "coniferous_forest"
+    taxon_pft_lookup$pft_MODIS[i] = ""
+    taxon_pft_lookup$phenology[i] = ""
+    taxon_pft_lookup$leafType[i] = ""
+    taxon_pft_lookup$growthForm[i] = ""
     taxon_pft_lookup$commonName[i] = "eastern hemlock"}
@@
     taxon_pft_lookup$pft_MRA[i] = "coniferous_forest"
+    taxon_pft_lookup$pft_MODIS[i] = ""
+    taxon_pft_lookup$phenology[i] = ""
+    taxon_pft_lookup$leafType[i] = ""
+    taxon_pft_lookup$growthForm[i] = ""
     taxon_pft_lookup$commonName[i] = "western hemlock"}
@@
     taxon_pft_lookup$pft_MRA[i] = "deciduous_forest"
+    taxon_pft_lookup$pft_MODIS[i] = ""
+    taxon_pft_lookup$phenology[i] = ""
+    taxon_pft_lookup$leafType[i] = ""
+    taxon_pft_lookup$growthForm[i] = ""
     taxon_pft_lookup$commonName[i] = "winged elm"}
@@
     taxon_pft_lookup$pft_MRA[i] = "deciduous_forest"
+    taxon_pft_lookup$pft_MODIS[i] = ""
+    taxon_pft_lookup$phenology[i] = ""
+    taxon_pft_lookup$leafType[i] = ""
+    taxon_pft_lookup$growthForm[i] = ""
     taxon_pft_lookup$commonName[i] = "American elm"}
@@
     taxon_pft_lookup$pft_MRA[i] = "deciduous_forest"
+    taxon_pft_lookup$pft_MODIS[i] = ""
+    taxon_pft_lookup$phenology[i] = ""
+    taxon_pft_lookup$leafType[i] = ""
+    taxon_pft_lookup$growthForm[i] = ""
     taxon_pft_lookup$commonName[i] = "cedar elm"}
@@
     taxon_pft_lookup$pft_MRA[i] = "deciduous_forest"
+    taxon_pft_lookup$pft_MODIS[i] = ""
+    taxon_pft_lookup$phenology[i] = ""
+    taxon_pft_lookup$leafType[i] = ""
+    taxon_pft_lookup$growthForm[i] = ""
     taxon_pft_lookup$commonName[i] = "slippery elm"}
@@
     taxon_pft_lookup$pft_MRA[i] = "deciduous_forest"
+    taxon_pft_lookup$pft_MODIS[i] = ""
+    taxon_pft_lookup$phenology[i] = ""
+    taxon_pft_lookup$leafType[i] = ""
+    taxon_pft_lookup$growthForm[i] = ""
     taxon_pft_lookup$commonName[i] = "elm"}
@@
     taxon_pft_lookup$pft_MRA[i] = "woody_shrub"       # deciduous Shrub
+    taxon_pft_lookup$pft_MODIS[i] = ""
+    taxon_pft_lookup$phenology[i] = ""
+    taxon_pft_lookup$leafType[i] = ""
+    taxon_pft_lookup$growthForm[i] = ""
     taxon_pft_lookup$commonName[i] = "red huckleberry"}

### 2021-04-23 R/add_plant_functional_type.Rmd
finished initial labeling and checking the  col values

@@
     taxon_pft_lookup$commonName[i] = "strongbark"
-    taxon_pft_lookup$commonName[i] = "varied genus"}
+    taxon_pft_lookup$comments[i] = "varied genus"}
 
@@
     taxon_pft_lookup$pft_MRA[i] = "coniferous_forest"
-    taxon_pft_lookup$pft_MODIS[i] = "evergreen_needleleaf_forest"
+    taxon_pft_lookup$pft_MODIS[i] = "evergreen-needleleaf_forest"
     taxon_pft_lookup$phenology[i] = "evergreen"
@@
     taxon_pft_lookup$pft_MRA[i] = "deciduous_forest"
-    taxon_pft_lookup$pft_MODIS[i] = ""
-    taxon_pft_lookup$phenology[i] = ""
-    taxon_pft_lookup$leafType[i] = ""
-    taxon_pft_lookup$growthForm[i] = ""
+    taxon_pft_lookup$pft_MODIS[i] = "deciduous-broadleaf_forest"
+    taxon_pft_lookup$phenology[i] = "deciduous"
+    taxon_pft_lookup$leafType[i] = "broadleaf"
+    taxon_pft_lookup$growthForm[i] = "tree"
     taxon_pft_lookup$commonName[i] = "southern red oak"}
@@
   else if(genusSpecies == "Quercus geminata"){
-    taxon_pft_lookup$pft_MRA[i] = "deciduous_forest"
-    taxon_pft_lookup$pft_MODIS[i] = ""
-    taxon_pft_lookup$phenology[i] = ""
-    taxon_pft_lookup$leafType[i] = ""
-    taxon_pft_lookup$growthForm[i] = ""
+    taxon_pft_lookup$pft_MRA[i] = "NA"
+    taxon_pft_lookup$pft_MODIS[i] = "evergreen-broadleaf_forest"
+    taxon_pft_lookup$phenology[i] = "evergreen"
+    taxon_pft_lookup$leafType[i] = "broadleaf"
+    taxon_pft_lookup$growthForm[i] = "tree"
     taxon_pft_lookup$commonName[i] = "sand live oak"}
@@
     taxon_pft_lookup$pft_MRA[i] = "deciduous_forest"
-    taxon_pft_lookup$pft_MODIS[i] = ""
-    taxon_pft_lookup$phenology[i] = ""
-    taxon_pft_lookup$leafType[i] = ""
-    taxon_pft_lookup$growthForm[i] = ""
-    taxon_pft_lookup$commonName[i] = "Darlington oak"}
+    taxon_pft_lookup$pft_MODIS[i] = "deciduous-broadleaf_forest"
+    taxon_pft_lookup$phenology[i] = "deciduous"
+    taxon_pft_lookup$leafType[i] = "broadleaf"
+    taxon_pft_lookup$growthForm[i] = "tree"
+    taxon_pft_lookup$commonName[i] = "Darlington oak"
+    taxon_pft_lookup$comments[i] = "semi-evergreen or tardily deciduous"}
 
@@
     taxon_pft_lookup$pft_MRA[i] = "deciduous_forest"
-    taxon_pft_lookup$pft_MODIS[i] = ""
-    taxon_pft_lookup$phenology[i] = ""
-    taxon_pft_lookup$leafType[i] = ""
-    taxon_pft_lookup$growthForm[i] = ""
+    taxon_pft_lookup$pft_MODIS[i] = "deciduous-broadleaf_forest"
+    taxon_pft_lookup$phenology[i] = "deciduous"
+    taxon_pft_lookup$leafType[i] = "broadleaf"
+    taxon_pft_lookup$growthForm[i] = "tree"
     taxon_pft_lookup$commonName[i] = "turkey oak"}
@@
     taxon_pft_lookup$pft_MRA[i] = "deciduous_forest"
-    taxon_pft_lookup$pft_MODIS[i] = ""
-    taxon_pft_lookup$phenology[i] = ""
-    taxon_pft_lookup$leafType[i] = ""
-    taxon_pft_lookup$growthForm[i] = ""
+    taxon_pft_lookup$pft_MODIS[i] = "deciduous-broadleaf_forest"
+    taxon_pft_lookup$phenology[i] = "deciduous"
+    taxon_pft_lookup$leafType[i] = "broadleaf"
+    taxon_pft_lookup$growthForm[i] = "tree"
     taxon_pft_lookup$commonName[i] = "laurel oak"}
@@
     taxon_pft_lookup$pft_MRA[i] = "deciduous_forest"
-    taxon_pft_lookup$pft_MODIS[i] = ""
-    taxon_pft_lookup$phenology[i] = ""
-    taxon_pft_lookup$leafType[i] = ""
-    taxon_pft_lookup$growthForm[i] = ""
+    taxon_pft_lookup$pft_MODIS[i] = "deciduous-broadleaf_forest"
+    taxon_pft_lookup$phenology[i] = "deciduous"
+    taxon_pft_lookup$leafType[i] = "broadleaf"
+    taxon_pft_lookup$growthForm[i] = "tree"
     taxon_pft_lookup$commonName[i] = "overcup oak"}
@@
     taxon_pft_lookup$pft_MRA[i] = "deciduous_forest"
-    taxon_pft_lookup$pft_MODIS[i] = ""
-    taxon_pft_lookup$phenology[i] = ""
-    taxon_pft_lookup$leafType[i] = ""
-    taxon_pft_lookup$growthForm[i] = ""
+    taxon_pft_lookup$pft_MODIS[i] = "deciduous-broadleaf_forest"
+    taxon_pft_lookup$phenology[i] = "deciduous"
+    taxon_pft_lookup$leafType[i] = "broadleaf"
+    taxon_pft_lookup$growthForm[i] = "tree"
     taxon_pft_lookup$commonName[i] = "bur oak"}
@@
     taxon_pft_lookup$pft_MRA[i] = "deciduous_forest"
-    taxon_pft_lookup$pft_MODIS[i] = ""
-    taxon_pft_lookup$phenology[i] = ""
-    taxon_pft_lookup$leafType[i] = ""
-    taxon_pft_lookup$growthForm[i] = ""
+    taxon_pft_lookup$pft_MODIS[i] = "deciduous-broadleaf_forest"
+    taxon_pft_lookup$phenology[i] = "deciduous"
+    taxon_pft_lookup$leafType[i] = "broadleaf"
+    taxon_pft_lookup$growthForm[i] = "tree"
     taxon_pft_lookup$commonName[i] = "sand post oak"}
@@
     taxon_pft_lookup$pft_MRA[i] = "deciduous_forest"
-    taxon_pft_lookup$pft_MODIS[i] = ""
-    taxon_pft_lookup$phenology[i] = ""
-    taxon_pft_lookup$leafType[i] = ""
-    taxon_pft_lookup$growthForm[i] = ""
+    taxon_pft_lookup$pft_MODIS[i] = "deciduous-broadleaf_forest"
+    taxon_pft_lookup$phenology[i] = "deciduous"
+    taxon_pft_lookup$leafType[i] = "broadleaf"
+    taxon_pft_lookup$growthForm[i] = "tree"
     taxon_pft_lookup$commonName[i] = "blackjack oak"}
@@
     taxon_pft_lookup$pft_MRA[i] = "deciduous_forest"
-    taxon_pft_lookup$pft_MODIS[i] = ""
-    taxon_pft_lookup$phenology[i] = ""
-    taxon_pft_lookup$leafType[i] = ""
-    taxon_pft_lookup$growthForm[i] = ""
+    taxon_pft_lookup$pft_MODIS[i] = "deciduous-broadleaf_forest"
+    taxon_pft_lookup$phenology[i] = "deciduous"
+    taxon_pft_lookup$leafType[i] = "broadleaf"
+    taxon_pft_lookup$growthForm[i] = "tree"
     taxon_pft_lookup$commonName[i] = "swamp chestnut oak"}
@@
     taxon_pft_lookup$pft_MRA[i] = "deciduous_forest"
-    taxon_pft_lookup$pft_MODIS[i] = ""
-    taxon_pft_lookup$phenology[i] = ""
-    taxon_pft_lookup$leafType[i] = ""
-    taxon_pft_lookup$growthForm[i] = ""
+    taxon_pft_lookup$pft_MODIS[i] = "deciduous-broadleaf_forest"
+    taxon_pft_lookup$phenology[i] = "deciduous"
+    taxon_pft_lookup$leafType[i] = "broadleaf"
+    taxon_pft_lookup$growthForm[i] = "tree"
     taxon_pft_lookup$commonName[i] = "chestnut oak"}
@@
     taxon_pft_lookup$pft_MRA[i] = "deciduous_forest"
-    taxon_pft_lookup$pft_MODIS[i] = ""
-    taxon_pft_lookup$phenology[i] = ""
-    taxon_pft_lookup$leafType[i] = ""
-    taxon_pft_lookup$growthForm[i] = ""
+    taxon_pft_lookup$pft_MODIS[i] = "deciduous-broadleaf_forest"
+    taxon_pft_lookup$phenology[i] = "deciduous"
+    taxon_pft_lookup$leafType[i] = "broadleaf"
+    taxon_pft_lookup$growthForm[i] = "tree"
     taxon_pft_lookup$commonName[i] = "chinquapin oak"}
@@
     taxon_pft_lookup$pft_MRA[i] = "deciduous_forest"
-    taxon_pft_lookup$pft_MODIS[i] = ""
-    taxon_pft_lookup$phenology[i] = ""
-    taxon_pft_lookup$leafType[i] = ""
-    taxon_pft_lookup$growthForm[i] = ""
+    taxon_pft_lookup$pft_MODIS[i] = "deciduous-broadleaf_forest"
+    taxon_pft_lookup$phenology[i] = "deciduous"
+    taxon_pft_lookup$leafType[i] = "broadleaf"
+    taxon_pft_lookup$growthForm[i] = "tree"
     taxon_pft_lookup$commonName[i] = "water oak"}
@@
     taxon_pft_lookup$pft_MRA[i] = "deciduous_forest"
-    taxon_pft_lookup$pft_MODIS[i] = ""
-    taxon_pft_lookup$phenology[i] = ""
-    taxon_pft_lookup$leafType[i] = ""
-    taxon_pft_lookup$growthForm[i] = ""
+    taxon_pft_lookup$pft_MODIS[i] = "deciduous-broadleaf_forest"
+    taxon_pft_lookup$phenology[i] = "deciduous"
+    taxon_pft_lookup$leafType[i] = "broadleaf"
+    taxon_pft_lookup$growthForm[i] = "tree"
     taxon_pft_lookup$commonName[i] = "cherrybark oak"}
@@
     taxon_pft_lookup$pft_MRA[i] = "deciduous_forest"
-    taxon_pft_lookup$pft_MODIS[i] = ""
-    taxon_pft_lookup$phenology[i] = ""
-    taxon_pft_lookup$leafType[i] = ""
-    taxon_pft_lookup$growthForm[i] = ""
+    taxon_pft_lookup$pft_MODIS[i] = "deciduous-broadleaf_forest"
+    taxon_pft_lookup$phenology[i] = "deciduous"
+    taxon_pft_lookup$leafType[i] = "broadleaf"
+    taxon_pft_lookup$growthForm[i] = "tree"
     taxon_pft_lookup$commonName[i] = "willow oak"}
@@
     taxon_pft_lookup$pft_MRA[i] = "deciduous_forest"
-    taxon_pft_lookup$pft_MODIS[i] = ""
-    taxon_pft_lookup$phenology[i] = ""
-    taxon_pft_lookup$leafType[i] = ""
-    taxon_pft_lookup$growthForm[i] = ""
+    taxon_pft_lookup$pft_MODIS[i] = "deciduous-broadleaf_forest"
+    taxon_pft_lookup$phenology[i] = "deciduous"
+    taxon_pft_lookup$leafType[i] = "broadleaf"
+    taxon_pft_lookup$growthForm[i] = "tree"
     taxon_pft_lookup$commonName[i] = "northern red oak"}
@@
     taxon_pft_lookup$pft_MRA[i] = "deciduous_forest"
-    taxon_pft_lookup$pft_MODIS[i] = ""
-    taxon_pft_lookup$phenology[i] = ""
-    taxon_pft_lookup$leafType[i] = ""
-    taxon_pft_lookup$growthForm[i] = ""
+    taxon_pft_lookup$pft_MODIS[i] = "deciduous-broadleaf_forest"
+    taxon_pft_lookup$phenology[i] = "deciduous"
+    taxon_pft_lookup$leafType[i] = "broadleaf"
+    taxon_pft_lookup$growthForm[i] = "tree"
     taxon_pft_lookup$commonName[i] = "Shumard's oak"}
@@
   else if(genusSpecies == "Quercus sp."){
-    taxon_pft_lookup$pft_MRA[i] = "NA"  # apparently there are evergreen oak species? 
-    taxon_pft_lookup$pft_MODIS[i] = ""
-    taxon_pft_lookup$phenology[i] = ""
-    taxon_pft_lookup$leafType[i] = ""
-    taxon_pft_lookup$growthForm[i] = ""
-    taxon_pft_lookup$commonName[i] = "oak"}
+    taxon_pft_lookup$pft_MRA[i] = "NA"
+    taxon_pft_lookup$pft_MODIS[i] = "NA"
+    taxon_pft_lookup$phenology[i] = "NA"
+    taxon_pft_lookup$leafType[i] = "broadleaf"
+    taxon_pft_lookup$growthForm[i] = "tree"
+    taxon_pft_lookup$commonName[i] = "oak"
+    taxon_pft_lookup$comments[i] = "varied genus with both deciduous and evergreen species"}
 
@@
     taxon_pft_lookup$pft_MRA[i] = "deciduous_forest"
-    taxon_pft_lookup$pft_MODIS[i] = ""
-    taxon_pft_lookup$phenology[i] = ""
-    taxon_pft_lookup$leafType[i] = ""
-    taxon_pft_lookup$growthForm[i] = ""
+    taxon_pft_lookup$pft_MODIS[i] = "deciduous-broadleaf_forest"
+    taxon_pft_lookup$phenology[i] = "deciduous"
+    taxon_pft_lookup$leafType[i] = "broadleaf"
+    taxon_pft_lookup$growthForm[i] = "tree"
     taxon_pft_lookup$commonName[i] = "post oak"}
@@
     taxon_pft_lookup$pft_MRA[i] = "deciduous_forest"
-    taxon_pft_lookup$pft_MODIS[i] = ""
-    taxon_pft_lookup$phenology[i] = ""
-    taxon_pft_lookup$leafType[i] = ""
-    taxon_pft_lookup$growthForm[i] = ""
+    taxon_pft_lookup$pft_MODIS[i] = "deciduous-broadleaf_forest"
+    taxon_pft_lookup$phenology[i] = "deciduous"
+    taxon_pft_lookup$leafType[i] = "broadleaf"
+    taxon_pft_lookup$growthForm[i] = "tree"
     taxon_pft_lookup$commonName[i] = "black oak"}
@@
     taxon_pft_lookup$pft_MRA[i] = "deciduous_forest"
-    taxon_pft_lookup$pft_MODIS[i] = ""
-    taxon_pft_lookup$phenology[i] = ""
-    taxon_pft_lookup$leafType[i] = ""
-    taxon_pft_lookup$growthForm[i] = ""
-    taxon_pft_lookup$commonName[i] = "live oak"}
+    taxon_pft_lookup$pft_MODIS[i] = "deciduous-broadleaf_forest"
+    taxon_pft_lookup$phenology[i] = "deciduous"
+    taxon_pft_lookup$leafType[i] = "broadleaf"
+    taxon_pft_lookup$growthForm[i] = "tree"
+    taxon_pft_lookup$commonName[i] = "live oak"
+    taxon_pft_lookup$comments[i] = "semi-deciduous"}
 
   else if(genusSpecies == "Rhamnus davurica"){
-    taxon_pft_lookup$pft_MRA[i] = "NA"     #  spreading shrub or small tree reaching 25 to 30 feet
-    taxon_pft_lookup$pft_MODIS[i] = ""
-    taxon_pft_lookup$phenology[i] = ""
-    taxon_pft_lookup$leafType[i] = ""
-    taxon_pft_lookup$growthForm[i] = ""
+    taxon_pft_lookup$pft_MRA[i] = "woody_shrub"
+    taxon_pft_lookup$pft_MODIS[i] = "shrub"
+    taxon_pft_lookup$phenology[i] = "deciduous"
+    taxon_pft_lookup$leafType[i] = "broadleaf"
+    taxon_pft_lookup$growthForm[i] = "shrub"
     taxon_pft_lookup$commonName[i] = "Dahurian buckthorn"}
@@
     taxon_pft_lookup$pft_MRA[i] = "woody_shrub"
-    taxon_pft_lookup$pft_MODIS[i] = ""
-    taxon_pft_lookup$phenology[i] = ""
-    taxon_pft_lookup$leafType[i] = ""
-    taxon_pft_lookup$growthForm[i] = ""
+    taxon_pft_lookup$pft_MODIS[i] = "shrub"
+    taxon_pft_lookup$phenology[i] = "deciduous"
+    taxon_pft_lookup$leafType[i] = "broadleaf"
+    taxon_pft_lookup$growthForm[i] = "shrub"
     taxon_pft_lookup$commonName[i] = "fragrant sumac"}
@@
     taxon_pft_lookup$pft_MRA[i] = "woody_shrub"
-    taxon_pft_lookup$pft_MODIS[i] = ""
-    taxon_pft_lookup$phenology[i] = ""
-    taxon_pft_lookup$leafType[i] = ""
-    taxon_pft_lookup$growthForm[i] = ""
+    taxon_pft_lookup$pft_MODIS[i] = "shrub"
+    taxon_pft_lookup$phenology[i] = "deciduous"
+    taxon_pft_lookup$leafType[i] = "broadleaf"
+    taxon_pft_lookup$growthForm[i] = "shrub"
     taxon_pft_lookup$commonName[i] = "smooth sumac"}
@@
     taxon_pft_lookup$pft_MRA[i] = "deciduous_forest"
-    taxon_pft_lookup$pft_MODIS[i] = ""
-    taxon_pft_lookup$phenology[i] = ""
-    taxon_pft_lookup$leafType[i] = ""
-    taxon_pft_lookup$growthForm[i] = ""
+    taxon_pft_lookup$pft_MODIS[i] = "deciduous-broadleaf_forest"
+    taxon_pft_lookup$phenology[i] = "deciduous"
+    taxon_pft_lookup$leafType[i] = "broadleaf"
+    taxon_pft_lookup$growthForm[i] = "tree"
     taxon_pft_lookup$commonName[i] = "black locust"}
@@
   else if(genusSpecies == "Sassafras albidum"){
-    taxon_pft_lookup$pft_MRA[i] = "deciduous_forest" # deciduous shrub or tree. trees are 9 to 18 m
-    taxon_pft_lookup$pft_MODIS[i] = ""
-    taxon_pft_lookup$phenology[i] = ""
-    taxon_pft_lookup$leafType[i] = ""
-    taxon_pft_lookup$growthForm[i] = ""
+    taxon_pft_lookup$pft_MRA[i] = "deciduous_forest"
+    taxon_pft_lookup$pft_MODIS[i] = "deciduous-broadleaf_forest"
+    taxon_pft_lookup$phenology[i] = "deciduous"
+    taxon_pft_lookup$leafType[i] = "broadleaf"
+    taxon_pft_lookup$growthForm[i] = "tree"
     taxon_pft_lookup$commonName[i] = "sassafras"}
@@
     taxon_pft_lookup$pft_MRA[i] = "deciduous_forest"
-    taxon_pft_lookup$pft_MODIS[i] = ""
-    taxon_pft_lookup$phenology[i] = ""
-    taxon_pft_lookup$leafType[i] = ""
-    taxon_pft_lookup$growthForm[i] = ""
+    taxon_pft_lookup$pft_MODIS[i] = "deciduous-broadleaf_forest"
+    taxon_pft_lookup$phenology[i] = "deciduous"
+    taxon_pft_lookup$leafType[i] = "broadleaf"
+    taxon_pft_lookup$growthForm[i] = "tree"
     taxon_pft_lookup$commonName[i] = "sassafras"}
@@
   else if(genusSpecies == "Sideroxylon lanuginosum"){
-    taxon_pft_lookup$pft_MRA[i] = "NA" # deciduous multi-trunked tall shrub or medium-sized tree to 48ft or more
-    taxon_pft_lookup$pft_MODIS[i] = ""
-    taxon_pft_lookup$phenology[i] = ""
-    taxon_pft_lookup$leafType[i] = ""
-    taxon_pft_lookup$growthForm[i] = ""
-    taxon_pft_lookup$commonName[i] = "gum bully"}
+    taxon_pft_lookup$pft_MRA[i] = "woody_shrub"
+    taxon_pft_lookup$pft_MODIS[i] = "shrub"
+    taxon_pft_lookup$phenology[i] = "deciduous"
+    taxon_pft_lookup$leafType[i] = "broadleaf"
+    taxon_pft_lookup$growthForm[i] = "shrub"
+    taxon_pft_lookup$commonName[i] = "gum bully"
+    taxon_pft_lookup$comments[i] = "deciduous to semi-evergreen"}
 
   else if(genusSpecies == "Symplocos sp."){
-    taxon_pft_lookup$pft_MRA[i] = "NA"  # "deciduous or evergreen shrub or tree"
-    taxon_pft_lookup$pft_MODIS[i] = ""
-    taxon_pft_lookup$phenology[i] = ""
-    taxon_pft_lookup$leafType[i] = ""
-    taxon_pft_lookup$growthForm[i] = ""
-    taxon_pft_lookup$commonName[i] = "sweetleaf"}
+    taxon_pft_lookup$pft_MRA[i] = "NA"
+    taxon_pft_lookup$pft_MODIS[i] = "NA"
+    taxon_pft_lookup$phenology[i] = "NA"
+    taxon_pft_lookup$leafType[i] = "broadleaf"
+    taxon_pft_lookup$growthForm[i] = "tree"
+    taxon_pft_lookup$commonName[i] = "sweetleaf"
+    taxon_pft_lookup$comments[i] = "varied genus"}
 
@@
     taxon_pft_lookup$pft_MRA[i] = "coniferous_forest"
-    taxon_pft_lookup$pft_MODIS[i] = ""
-    taxon_pft_lookup$phenology[i] = ""
-    taxon_pft_lookup$leafType[i] = ""
-    taxon_pft_lookup$growthForm[i] = ""
+    taxon_pft_lookup$pft_MODIS[i] = "evergreen-needleleaf_forest"
+    taxon_pft_lookup$phenology[i] = "evergreen"
+    taxon_pft_lookup$leafType[i] = "needleleaf"
+    taxon_pft_lookup$growthForm[i] = "tree"
     taxon_pft_lookup$commonName[i] = "Pacific yew"}
@@
     taxon_pft_lookup$pft_MRA[i] = "deciduous_forest"
-    taxon_pft_lookup$pft_MODIS[i] = ""
-    taxon_pft_lookup$phenology[i] = ""
-    taxon_pft_lookup$leafType[i] = ""
-    taxon_pft_lookup$growthForm[i] = ""
+    taxon_pft_lookup$pft_MODIS[i] = "deciduous-broadleaf_forest"
+    taxon_pft_lookup$phenology[i] = "deciduous"
+    taxon_pft_lookup$leafType[i] = "broadleaf"
+    taxon_pft_lookup$growthForm[i] = "tree"
     taxon_pft_lookup$commonName[i] = "bee-bee tree"}
@@
     taxon_pft_lookup$pft_MRA[i] = "deciduous_forest"
-    taxon_pft_lookup$pft_MODIS[i] = ""
-    taxon_pft_lookup$phenology[i] = ""
-    taxon_pft_lookup$leafType[i] = ""
-    taxon_pft_lookup$growthForm[i] = ""
+    taxon_pft_lookup$pft_MODIS[i] = "deciduous-broadleaf_forest"
+    taxon_pft_lookup$phenology[i] = "deciduous"
+    taxon_pft_lookup$leafType[i] = "broadleaf"
+    taxon_pft_lookup$growthForm[i] = "tree"
     taxon_pft_lookup$commonName[i] = "Chinese tallow"}
@@
     taxon_pft_lookup$pft_MRA[i] = "coniferous_forest"
-    taxon_pft_lookup$pft_MODIS[i] = ""
-    taxon_pft_lookup$phenology[i] = ""
-    taxon_pft_lookup$leafType[i] = ""
-    taxon_pft_lookup$growthForm[i] = ""
+    taxon_pft_lookup$pft_MODIS[i] = "evergreen-needleleaf_forest"
+    taxon_pft_lookup$phenology[i] = "evergreen"
+    taxon_pft_lookup$leafType[i] = "needleleaf"
+    taxon_pft_lookup$growthForm[i] = "tree"
     taxon_pft_lookup$commonName[i] = "eastern hemlock"}
@@
     taxon_pft_lookup$pft_MRA[i] = "coniferous_forest"
-    taxon_pft_lookup$pft_MODIS[i] = ""
-    taxon_pft_lookup$phenology[i] = ""
-    taxon_pft_lookup$leafType[i] = ""
-    taxon_pft_lookup$growthForm[i] = ""
+    taxon_pft_lookup$pft_MODIS[i] = "evergreen-needleleaf_forest"
+    taxon_pft_lookup$phenology[i] = "evergreen"
+    taxon_pft_lookup$leafType[i] = "needleleaf"
+    taxon_pft_lookup$growthForm[i] = "tree"
     taxon_pft_lookup$commonName[i] = "western hemlock"}
@@
     taxon_pft_lookup$pft_MRA[i] = "deciduous_forest"
-    taxon_pft_lookup$pft_MODIS[i] = ""
-    taxon_pft_lookup$phenology[i] = ""
-    taxon_pft_lookup$leafType[i] = ""
-    taxon_pft_lookup$growthForm[i] = ""
+    taxon_pft_lookup$pft_MODIS[i] = "deciduous-broadleaf_forest"
+    taxon_pft_lookup$phenology[i] = "deciduous"
+    taxon_pft_lookup$leafType[i] = "broadleaf"
+    taxon_pft_lookup$growthForm[i] = "tree"
     taxon_pft_lookup$commonName[i] = "winged elm"}
@@
     taxon_pft_lookup$pft_MRA[i] = "deciduous_forest"
-    taxon_pft_lookup$pft_MODIS[i] = ""
-    taxon_pft_lookup$phenology[i] = ""
-    taxon_pft_lookup$leafType[i] = ""
-    taxon_pft_lookup$growthForm[i] = ""
+    taxon_pft_lookup$pft_MODIS[i] = "deciduous-broadleaf_forest"
+    taxon_pft_lookup$phenology[i] = "deciduous"
+    taxon_pft_lookup$leafType[i] = "broadleaf"
+    taxon_pft_lookup$growthForm[i] = "tree"
     taxon_pft_lookup$commonName[i] = "American elm"}
@@
     taxon_pft_lookup$pft_MRA[i] = "deciduous_forest"
-    taxon_pft_lookup$pft_MODIS[i] = ""
-    taxon_pft_lookup$phenology[i] = ""
-    taxon_pft_lookup$leafType[i] = ""
-    taxon_pft_lookup$growthForm[i] = ""
+    taxon_pft_lookup$pft_MODIS[i] = "deciduous-broadleaf_forest"
+    taxon_pft_lookup$phenology[i] = "deciduous"
+    taxon_pft_lookup$leafType[i] = "broadleaf"
+    taxon_pft_lookup$growthForm[i] = "tree"
     taxon_pft_lookup$commonName[i] = "cedar elm"}
@@
     taxon_pft_lookup$pft_MRA[i] = "deciduous_forest"
-    taxon_pft_lookup$pft_MODIS[i] = ""
-    taxon_pft_lookup$phenology[i] = ""
-    taxon_pft_lookup$leafType[i] = ""
-    taxon_pft_lookup$growthForm[i] = ""
+    taxon_pft_lookup$pft_MODIS[i] = "deciduous-broadleaf_forest"
+    taxon_pft_lookup$phenology[i] = "deciduous"
+    taxon_pft_lookup$leafType[i] = "broadleaf"
+    taxon_pft_lookup$growthForm[i] = "tree"
     taxon_pft_lookup$commonName[i] = "slippery elm"}
@@
     taxon_pft_lookup$pft_MRA[i] = "deciduous_forest"
-    taxon_pft_lookup$pft_MODIS[i] = ""
-    taxon_pft_lookup$phenology[i] = ""
-    taxon_pft_lookup$leafType[i] = ""
-    taxon_pft_lookup$growthForm[i] = ""
-    taxon_pft_lookup$commonName[i] = "elm"}
+    taxon_pft_lookup$pft_MODIS[i] = "deciduous-broadleaf_forest"
+    taxon_pft_lookup$phenology[i] = "deciduous"
+    taxon_pft_lookup$leafType[i] = "broadleaf"
+    taxon_pft_lookup$growthForm[i] = "tree"
+    taxon_pft_lookup$commonName[i] = "elm"
+    taxon_pft_lookup$comments[i] = "varied genus contains some semi-deciduous or evergreen species"}
 
   else if(genusSpecies == "Vaccinium parvifolium"){ 
-    taxon_pft_lookup$pft_MRA[i] = "woody_shrub"       # deciduous Shrub
-    taxon_pft_lookup$pft_MODIS[i] = ""
-    taxon_pft_lookup$phenology[i] = ""
-    taxon_pft_lookup$leafType[i] = ""
-    taxon_pft_lookup$growthForm[i] = ""
+    taxon_pft_lookup$pft_MRA[i] = "woody_shrub"
+    taxon_pft_lookup$pft_MODIS[i] = "shrub"
+    taxon_pft_lookup$phenology[i] = "deciduous"
+    taxon_pft_lookup$leafType[i] = "broadleaf"
+    taxon_pft_lookup$growthForm[i] = "shrub"
     taxon_pft_lookup$commonName[i] = "red huckleberry"}
@@
   else if(genusSpecies == "Zinnia acerosa"){
-    taxon_pft_lookup$pft_MRA[i] = "grass_herb" # forb/herb, subshrub - grass_herb?
-    taxon_pft_lookup$commonName[i] = "desert zinnia"}
+    taxon_pft_lookup$pft_MRA[i] = "woody_shrub"
+    taxon_pft_lookup$pft_MODIS[i] = "shrub"
+    taxon_pft_lookup$phenology[i] = "deciduous"
+    taxon_pft_lookup$leafType[i] = "broadleaf"
+    taxon_pft_lookup$growthForm[i] = "shrub"
+    taxon_pft_lookup$commonName[i] = "desert zinnia"
+    taxon_pft_lookup$comments[i] = "neon growthForm was shrub; evergreen but cold-deciduous"}
 
@@
 
+```
+
+Let's check the unique values in each column.
+
+```{r check_col_pftMRA}
+# unique values in the pft_MRA column: 
+print(unique(taxon_pft_lookup$pft_MRA))
+```
+```{r check_col_pftMODIS}
+# unique values in the pft_MODIS column: 
+print(unique(taxon_pft_lookup$pft_MODIS))
+```
+```{r check_col_phenology}
+# unique values in the phenology column: 
+print(unique(taxon_pft_lookup$phenology))
+```
+```{r check_col_leafType}
+# unique values in the leafType column: 
+print(unique(taxon_pft_lookup$leafType))
+```
+```{r check_col_growthForm}
+# unique values in the growthForm column: 
+print(unique(taxon_pft_lookup$growthForm))
+```
+
+```{r check_col_comments}
+# unique values in the comments column: 
+print(unique(taxon_pft_lookup$comments))
 ```

### 2021-04-23 R/add_plant_functional_type.Rmd
sort the modis labels

@@
 # unique values in the pft_MODIS column: 
-print(unique(taxon_pft_lookup$pft_MODIS))
+print(sort(unique(taxon_pft_lookup$pft_MODIS)))
 ```

### 2021-04-27 analysis/paper/paper.Rmd
additional reference

@@
 Methods are often performed on small scales or in simple ecosystems, and are not successful at scaling to larger areas or different forest types **(cite)**.
-Although the demand for species mapping is justified in many cases, there is also potential for both using and finding value in classifying individuals at other taxonomic levels, as explored in a recent bioacoustics study [@cramer2020chirping].
-In the plant realm, identifying taxons at the genus level had been shown to be easier than at the species level [@tuominen2018assessment].
+Although the demand for species mapping is justified in many cases, there is also potential for both using and finding value in classifying individuals at other taxonomic levels, as explored in a recent bioacoustics study [@cramer2020chirping] and plant functional type mapping work in the High Arctic [@thomson2021multiscale].
+In the plant realm, identifying taxa at the genus level had been shown to be easier than at the species level [@tuominen2018assessment].
 In their review of over 109 publications, [@surovy2019acquisition] report that species are more successfully classified into broader categories such as spruce, pine, and broadleaf, compared to identifying individual species in mixed arrangements.

### 2021-04-27 R/add_plant_functional_type.Rmd
updated descriptions. removed automation code attempts that didnt work. need to apply the LUT to add a label to the cleaned spectra

@@
 
-Plant functional types (PFTs) originally proposed: coniferous forest, deciduous forest, woody shrub, grass/herb, and bare ground.
+This script adds plant functional type labels to NEON spectra based on known scientific name.
+I explored R packages to automate this (BIEN, Taxise, USDA plants) without success, so the code below manually adds labels of interest for *Genus species* names.
+
+This code manually adds labels to each taxon (genus and species listed here) for:
+
+- **commonName**
+- **pft_MRA** (Plant Functional Type based on Macrosystems proposed labels)
+    - coniferous_forest
+    - deciduous_forest
+    - woody_shrub
+    - grass_herb
+    bare_ground
+- **pft_MODIS** (Plant Functional type based on MODIS land cover classes)
+    - evergreen-needleleaf_forest
+    - evergreen-broadleaf_forest
+    - deciduous-needleleaf_forest
+    - deciduous-broadleaf_forest
+    - shrub
+- **phenology** (evergreen or deciduous)
+- **leafType** (broadleaf or needleleaf)
+- **growthForm** (shrub or tree, based on NEON field based observations)
+
+I searched for each taxon on USDA plants and determined the phenology and leaf type. Some tips from botany friend Adam Mahood:
+
+- shrubs that are angiosperms are broadleaf
+- group = dicot or monocot, you can say broadleaf
+- group = gymnosperm will be needleleaf
+
+## Download NEON data
+
+The NEON Woody Vegetation Structure vst_apparentindividual data table contains growthform observations for each individual plant.
+Since some species can be designated as trees OR shrubs, lets' look at the NEON observations and use whichever growth form is most often observed for each taxon.
 
@@
 ```
-
-## Use NEON growthForm observations
 
@@
                             position = "left")
-
 ```
@@
 ## Manually add PFTs to spectral library
-
-This code manually adds labels to each taxon (genus and species listed here) for:
-
-- common name
-- pft_MRA (based on Macrosystems proposed labels)
-    - coniferous_forest
-    - deciduous_forest
-    - woody_shrub
-- pft_MODIS (based on MODIS land cover classes)
-    - evergreen-needleleaf_forest
-    - evergreen-broadleaf_forest
-    - deciduous-needleleaf_forest
-    - deciduous-broadleaf_forest
-    - shrub
-- phenology (evergreen or deciduous)
-- leaf type (broadleaf or needleleaf)
-- growth form (shrub or tree, based on NEON field based observations)
-
-I searched for each taxon on USDA plants and determined the phenology and leaf type. 
-- shrubs that are angiosperms are broadleaf
-- group = dicot or monocot, you can say broadleaf
-- group = gymnosperm will be needleleaf
-
-Utilized the NEON vst_apparentindividual data for growthform.
 
@@
 
+## Add PFT label to each spectrum
+
+```{r}
+
+```
+
 ## R packages: attempting to automate PFT labeling
@@
 
-```{r install_load_packages, message=FALSE}
-#install.packages("request")
-#library(devtools)
-#devtools::install_github("sckott/request")
-library(request)
-library(tibble)
-library(tidyverse)
-library(here)
-```
-
-```{r usda_plants}
-# Set the base url
-root <- api("https://plantsdb.xyz")
-
-root %>% api_path(heartbeat)
-```
-
-^^ My results here do not match those that show up on the tutorial page above. Maybe the API is down or the URL has changed?
-
-Based on the GitHub issue discussion (https://github.com/USDA/USDA-APIs/issues/7), someone uploaded USDA plants data here from 2016: https://uofi.app.box.com/v/usdaplants
-
-```{r try_download, eval=FALSE, include=FALSE}
-# Specify URL where file is stored 
-url <- "https://uofi.app.box.com/v/usdaplants/folder/6688524249/usda_plants_20160713.csv"
-destFile <- file.path(here::here(), "analysis","data", "usda_plants_20160713.csv")
-# Apply download.file function in R
-download.file(url, destFile)
-```
-
-```{r, include = FALSE, eval=FALSE}
-#usda_plants <- read_csv(file.path(here::here(), "analysis", "data", "usda_plants_20160713.csv"))
-
-# Error in nchar(x) : invalid multibyte string, element 2
-# ^ Issue with some of the column names in the .csv? 
-```
-
 ### BIEN
@@
 RBIEN tutorial: https://cmerow.github.io/RDataScience/3_3_RBIEN_tutorial.html
-
-```{r load_packages_bien, message=FALSE, warning=FALSE}
-library(BIEN)
-library(ape) # Package for working with phylogenies in R
-library(maps) # Useful for making quick maps of occurrences
-library(sp) # A package for spatial data
-```
-
-BIEN traits and taxonomic data
-
-"Let’s say we’re interested in the genus [Abies], and we’d like to get an idea of how many species there are in this genus and what higher taxa it falls within."
-
-*Abies* genus is fir trees (coniferous trees). The "higher_plant_group" returned by the BIEN taxonomy function is "gymnosperms (conifers)"
-
-```{r bien_taxonomy_1}
-Abies_taxonomy<-BIEN_taxonomy_genus(genus = "Abies")
-
-#View(Abies_taxonomy)
-Abies_taxonomy[1:3,]
-```
-
-*Acer rubrum* species is the red maple (deciduous trees). The "higher_plant_group" returned by the BIEN taxonomy is "flowering plants".
-
-```{r bien_taxonomy_2}
-Acer_rubrum_taxonomy <- BIEN_taxonomy_species(species = "Acer rubrum")
-Acer_rubrum_taxonomy[1:3,]
-```
 

### 2021-04-27 R/add_plant_functional_type.Rmd
added PFt labels to cleaned spectra

@@
 
-```{r}
-
+```{r add_pft_labels}
+# select the columns to merge with all the spectra
+pft_labels <- taxon_pft_lookup %>% 
+  select(genusSpecies, commonName, pft_MRA, pft_MODIS, phenology, leafType, growthForm)
+
+# join the PFT labels into the spectral data frame
+cleaned_spectra_pft <- right_join(cleaned_spectra, pft_labels, by = "genusSpecies")
+
+# write to a csv
+write_csv(x = cleaned_spectra_pft,
+          file = here::here("analysis", "data","derived_data", "cleaned_spectra_pft.csv"))
 ```
@@
 
-The USDA maintains a database of plant information, some of it trait data, some of it life history. Check it out at http://plants.usda.gov/java/
+The USDA maintains a database of plant information, some of it trait data, some of it life history: http://plants.usda.gov/java/
 

### 2021-04-27 analysis/paper/paper.Rmd
another ref

--- 2020-11-12 analysis/paper/paper.Rmd
@@
 Methods are often performed on small scales or in simple ecosystems, and are not successful at scaling to larger areas or different forest types **(cite)**.
-Although the demand for species mapping is justified in many cases, there is also potential for both using and finding value in classifying individuals at other taxonomic levels, as explored in a recent bioacoustics study [@cramer2020chirping].
-In the plant realm, identifying taxons at the genus level had been shown to be easier than at the species level [@tuominen2018assessment].
+Although the demand for species mapping is justified in many cases, there is also potential for both using and finding value in classifying individuals at other taxonomic levels, as explored in a recent bioacoustics study [@cramer2020chirping], plant functional type mapping work in the High Arctic [@thomson2021multiscale], and [@seeland2019image].
+In the plant realm, identifying taxa at higher taxonomic levels has been found to be more accurate than at lower taxonomic levels.
+[@tuominen2018assessment] found that classification accuracy was higher at the genus level than at the species level.
+[@seeland2019image] found that classification accuracy increased with broader taxonomic level from species to genus to family, even for species excluded from the training process. 
 In their review of over 109 publications, [@surovy2019acquisition] report that species are more successfully classified into broader categories such as spruce, pine, and broadleaf, compared to identifying individual species in mixed arrangements.

### 2021-04-27 analysis/paper/paper.Rmd
added ref and summary of the PFT labeling process

@@
 These wavelength ranges feature noisy reflectance data due to high water absorption [@bajorski2011].
+
+We manually added plant functional type (PFT) along with additional morphological and taxonomic labels to the spectra.
+We used the [USDA Plants Database](https://plants.sc.egov.usda.gov/) to identify each taxon's phenology (evergreen or deciduous), leaf type (needleleaf or broadleaf), and taxonomic family classification.
+We used the NEON Woody Vegetation Structure data to determine the growth form (shrub or tree) most commonly assigned to each taxon class, since some taxa can exist as trees or shrubs.
+The PFT labels were inspired by the MODIS Land Cover classes [@cover1999modis]: evergreen needleleaf forest, evergreen broadleaf forest, deciduous needleleaf forest, deciduous broadleaf forest, shrub, grass, and bare ground.
 

### 2021-04-27 R/add_plant_functional_type.Rmd
before adding family column
updated PFT table with family column

@@
 taxon_pft_lookup <- data.frame(genusSpecies = unique(cleaned_spectra$genusSpecies),
+                               family = NA,
                                commonName = NA,
@@
     taxon_pft_lookup$growthForm[i] = "tree"
-    taxon_pft_lookup$commonName[i] = "Pacific silver fir"}
+    taxon_pft_lookup$commonName[i] = "Pacific silver fir"
+    taxon_pft_lookup$family[i] = "Pinaceae"}
 
@@
     taxon_pft_lookup$growthForm[i] = "tree"
-    taxon_pft_lookup$commonName[i] = "balsam fir"}
+    taxon_pft_lookup$commonName[i] = "balsam fir"
+    taxon_pft_lookup$family[i] = "Pinaceae"}
 
@@
     taxon_pft_lookup$growthForm[i] = "tree"
-    taxon_pft_lookup$commonName[i] = "subalpine fir"}
+    taxon_pft_lookup$commonName[i] = "subalpine fir"
+    taxon_pft_lookup$family[i] = "Pinaceae"}
 
@@
     taxon_pft_lookup$growthForm[i] = "shrub"
-    taxon_pft_lookup$commonName[i] = "catclaw acacia"}
+    taxon_pft_lookup$commonName[i] = "catclaw acacia"
+    taxon_pft_lookup$family[i] = "Aceraceae"}
 
@@
     taxon_pft_lookup$growthForm[i] = "tree"
-    taxon_pft_lookup$commonName[i] = "Southern sugar maple"}
+    taxon_pft_lookup$commonName[i] = "Southern sugar maple"
+    taxon_pft_lookup$family[i] = "Aceraceae"}
 
@@
     taxon_pft_lookup$growthForm[i] = "tree"
-    taxon_pft_lookup$commonName[i] = "vine maple"}
+    taxon_pft_lookup$commonName[i] = "vine maple"
+    taxon_pft_lookup$family[i] = "Aceraceae"}
 
@@
     taxon_pft_lookup$growthForm[i] = "tree"
-    taxon_pft_lookup$commonName[i] = "boxelder"}
+    taxon_pft_lookup$commonName[i] = "boxelder"
+    taxon_pft_lookup$family[i] = "Aceraceae"}
 
@@
     taxon_pft_lookup$growthForm[i] = "tree"
-    taxon_pft_lookup$commonName[i] = "striped maple"}
+    taxon_pft_lookup$commonName[i] = "striped maple"
+    taxon_pft_lookup$family[i] = "Aceraceae"}
 
@@
     taxon_pft_lookup$growthForm[i] = "tree"
-    taxon_pft_lookup$commonName[i] = "red maple"}
+    taxon_pft_lookup$commonName[i] = "red maple"
+    taxon_pft_lookup$family[i] = "Aceraceae"}
 
@@
     taxon_pft_lookup$growthForm[i] = "tree"
-    taxon_pft_lookup$commonName[i] = "silver maple"}
+    taxon_pft_lookup$commonName[i] = "silver maple"
+    taxon_pft_lookup$family[i] = "Aceraceae"}
 
@@
     taxon_pft_lookup$growthForm[i] = "tree"
-    taxon_pft_lookup$commonName[i] = "sugar maple"}
+    taxon_pft_lookup$commonName[i] = "sugar maple"
+    taxon_pft_lookup$family[i] = "Aceraceae"}
 
@@
     taxon_pft_lookup$growthForm[i] = "tree"
-    taxon_pft_lookup$commonName[i] = "tree of heaven"}
+    taxon_pft_lookup$commonName[i] = "tree of heaven"
+    taxon_pft_lookup$family[i] = "Simaroubaceae"}
 
@@
     taxon_pft_lookup$growthForm[i] = "tree"
-    taxon_pft_lookup$commonName[i] = "red alder"}
+    taxon_pft_lookup$commonName[i] = "red alder"
+    taxon_pft_lookup$family[i] = "Betulaceae"}
 
@@
     taxon_pft_lookup$growthForm[i] = "shrub"
-    taxon_pft_lookup$commonName[i] = "Wright's beebrush"}
+    taxon_pft_lookup$commonName[i] = "Wright's beebrush"
+    taxon_pft_lookup$family[i] = "Verbenaceae"}
 
@@
     taxon_pft_lookup$growthForm[i] = "tree"
-    taxon_pft_lookup$commonName[i] = "Allegheny serviceberry"}
+    taxon_pft_lookup$commonName[i] = "Allegheny serviceberry"
+    taxon_pft_lookup$family[i] = "Rosaceae"}
 
@@
     taxon_pft_lookup$growthForm[i] = "tree"
-    taxon_pft_lookup$commonName[i] = "sea torchwood"}
+    taxon_pft_lookup$commonName[i] = "sea torchwood"
+    taxon_pft_lookup$family[i] = "Rutaceae"}
 
@@
     taxon_pft_lookup$growthForm[i] = "shrub"
-    taxon_pft_lookup$commonName[i] = "big sagebrush"}
+    taxon_pft_lookup$commonName[i] = "big sagebrush"
+    taxon_pft_lookup$family[i] = "Asteraceae"}
 
@@
     taxon_pft_lookup$growthForm[i] = "shrub"
-    taxon_pft_lookup$commonName[i] = "fourwing saltbush"}
+    taxon_pft_lookup$commonName[i] = "fourwing saltbush"
+    taxon_pft_lookup$family[i] = "Chenopodiaceae"}
 
@@
     taxon_pft_lookup$growthForm[i] = "shrub"
-    taxon_pft_lookup$commonName[i] = "shadscale saltbush"}
+    taxon_pft_lookup$commonName[i] = "shadscale saltbush"
+    taxon_pft_lookup$family[i] = "Chenopodiaceae"}
 
@@
     taxon_pft_lookup$growthForm[i] = "tree"
-    taxon_pft_lookup$commonName[i] = "yellow birch"}
+    taxon_pft_lookup$commonName[i] = "yellow birch"
+    taxon_pft_lookup$family[i] = "Betulaceae"}
 
@@
     taxon_pft_lookup$growthForm[i] = "tree"
-    taxon_pft_lookup$commonName[i] = "sweet birch"}
+    taxon_pft_lookup$commonName[i] = "sweet birch"
+    taxon_pft_lookup$family[i] = "Betulaceae"}
 
@@
     taxon_pft_lookup$growthForm[i] = "tree"
-    taxon_pft_lookup$commonName[i] = "resin birch"}
+    taxon_pft_lookup$commonName[i] = "resin birch"
+    taxon_pft_lookup$family[i] = "Betulaceae"}
 
@@
     taxon_pft_lookup$growthForm[i] = "tree"
-    taxon_pft_lookup$commonName[i] = "river birch"}
+    taxon_pft_lookup$commonName[i] = "river birch"
+    taxon_pft_lookup$family[i] = "Betulaceae"}
 
@@
     taxon_pft_lookup$growthForm[i] = "tree"
-    taxon_pft_lookup$commonName[i] = "paper birch"}
+    taxon_pft_lookup$commonName[i] = "paper birch"
+    taxon_pft_lookup$family[i] = "Betulaceae"}
 
@@
     taxon_pft_lookup$growthForm[i] = "tree"
-    taxon_pft_lookup$commonName[i] = "birch"}
+    taxon_pft_lookup$commonName[i] = "birch"
+    taxon_pft_lookup$family[i] = "Betulaceae"}
 
@@
     taxon_pft_lookup$commonName[i] = "strongbark"
+    taxon_pft_lookup$family[i] = "Boraginaceae"
     taxon_pft_lookup$comments[i] = "varied genus"}
@@
     taxon_pft_lookup$growthForm[i] = "tree"
-    taxon_pft_lookup$commonName[i] = "gumbo limbo"}
+    taxon_pft_lookup$commonName[i] = "gumbo limbo"
+    taxon_pft_lookup$family[i] = "Burseraceae"}
 
@@
     taxon_pft_lookup$growthForm[i] = "tree"
-    taxon_pft_lookup$commonName[i] = "American hornbeam"}
+    taxon_pft_lookup$commonName[i] = "American hornbeam"
+    taxon_pft_lookup$family[i] = "Betulaceae"}
 
@@
     taxon_pft_lookup$growthForm[i] = "tree"
-    taxon_pft_lookup$commonName[i] = "water hickory"}
+    taxon_pft_lookup$commonName[i] = "water hickory"
+    taxon_pft_lookup$family[i] = "Juglandaceae"}
 
@@
     taxon_pft_lookup$growthForm[i] = "tree"
-    taxon_pft_lookup$commonName[i] = "bitternut hickory"}
+    taxon_pft_lookup$commonName[i] = "bitternut hickory"
+    taxon_pft_lookup$family[i] = "Juglandaceae"}
 
@@
     taxon_pft_lookup$growthForm[i] = "tree"
-    taxon_pft_lookup$commonName[i] = "pignut hickory"}
+    taxon_pft_lookup$commonName[i] = "pignut hickory"
+    taxon_pft_lookup$family[i] = "Juglandaceae"}
 
@@
     taxon_pft_lookup$growthForm[i] = "tree"
-    taxon_pft_lookup$commonName[i] = "pecan"}
+    taxon_pft_lookup$commonName[i] = "pecan"
+    taxon_pft_lookup$family[i] = "Juglandaceae"}
 
@@
     taxon_pft_lookup$growthForm[i] = "tree"
-    taxon_pft_lookup$commonName[i] = "red hickory"}
+    taxon_pft_lookup$commonName[i] = "red hickory"
+    taxon_pft_lookup$family[i] = "Juglandaceae"}
 
@@
     taxon_pft_lookup$growthForm[i] = "tree"
-    taxon_pft_lookup$commonName[i] = "shagbark hickory"}
+    taxon_pft_lookup$commonName[i] = "shagbark hickory"
+    taxon_pft_lookup$family[i] = "Juglandaceae"}
 
@@
     taxon_pft_lookup$growthForm[i] = "tree"
-    taxon_pft_lookup$commonName[i] = "mockernut hickory"}
+    taxon_pft_lookup$commonName[i] = "mockernut hickory"
+    taxon_pft_lookup$family[i] = "Juglandaceae"}
 
@@
     taxon_pft_lookup$growthForm[i] = "tree"
-    taxon_pft_lookup$commonName[i] = "sugarberry"}
+    taxon_pft_lookup$commonName[i] = "sugarberry"
+    taxon_pft_lookup$family[i] = "Ulmaceae"}
 
@@
     taxon_pft_lookup$growthForm[i] = "tree"
-    taxon_pft_lookup$commonName[i] = "common hackberry"}
+    taxon_pft_lookup$commonName[i] = "common hackberry"
+    taxon_pft_lookup$family[i] = "Ulmaceae"}
 
@@
     taxon_pft_lookup$growthForm[i] = "tree"
-    taxon_pft_lookup$commonName[i] = "eastern redbud"}
+    taxon_pft_lookup$commonName[i] = "eastern redbud"
+    taxon_pft_lookup$family[i] = "Fabaceae"}
 
@@
     taxon_pft_lookup$growthForm[i] = "shrub"
-    taxon_pft_lookup$commonName[i] = "yellow rabbitbrush"}
+    taxon_pft_lookup$commonName[i] = "yellow rabbitbrush"
+    taxon_pft_lookup$family[i] = "Asteraceae"}
 
@@
     taxon_pft_lookup$growthForm[i] = "tree"
-    taxon_pft_lookup$commonName[i] = "red manjack"}
+    taxon_pft_lookup$commonName[i] = "red manjack"
+    taxon_pft_lookup$family[i] = "Boraginaceae"}
 
@@
     taxon_pft_lookup$growthForm[i] = "shrub"
-    taxon_pft_lookup$commonName[i] = "roughleaf dogwood"}
+    taxon_pft_lookup$commonName[i] = "roughleaf dogwood"
+    taxon_pft_lookup$family[i] = "Cornaceae"}
 
@@
     taxon_pft_lookup$growthForm[i] = "tree"
-    taxon_pft_lookup$commonName[i] = "flowering dogwood"}
+    taxon_pft_lookup$commonName[i] = "flowering dogwood"
+    taxon_pft_lookup$family[i] = "Cornaceae"}
 
@@
     taxon_pft_lookup$growthForm[i] = "shrub"
-    taxon_pft_lookup$commonName[i] = "beaked hazelnut"}
+    taxon_pft_lookup$commonName[i] = "beaked hazelnut"
+    taxon_pft_lookup$family[i] = "Betulaceae"}
 
@@
     taxon_pft_lookup$growthForm[i] = "tree"
-    taxon_pft_lookup$commonName[i] = "common persimmon"}
+    taxon_pft_lookup$commonName[i] = "common persimmon"
+    taxon_pft_lookup$family[i] = "Ebenaceae"}
 
@@
     taxon_pft_lookup$growthForm[i] = "shrub"
-    taxon_pft_lookup$commonName[i] = "longleaf jointfir"}
+    taxon_pft_lookup$commonName[i] = "longleaf jointfir"
+    taxon_pft_lookup$family[i] = "Ephedraceae"}
 
@@
     taxon_pft_lookup$growthForm[i] = "shrub"
-    taxon_pft_lookup$commonName[i] = "turpentine bush"}
+    taxon_pft_lookup$commonName[i] = "turpentine bush"
+    taxon_pft_lookup$family[i] = "Asteraceae"}
 
@@
     taxon_pft_lookup$growthForm[i] = "tree"
-    taxon_pft_lookup$commonName[i] = "Caribbean princewood"}
+    taxon_pft_lookup$commonName[i] = "Caribbean princewood"
+    taxon_pft_lookup$family[i] = "Rubiaceae"}
 
@@
     taxon_pft_lookup$growthForm[i] = "tree"
-    taxon_pft_lookup$commonName[i] = "American beech"}
+    taxon_pft_lookup$commonName[i] = "American beech"
+    taxon_pft_lookup$family[i] = "Fagaceae"}
 
@@
     taxon_pft_lookup$growthForm[i] = "tree"
-    taxon_pft_lookup$commonName[i] = "Cascara buckthorn"}
+    taxon_pft_lookup$commonName[i] = "Cascara buckthorn"
+    taxon_pft_lookup$family[i] = "Rhamnaceae"}
 
@@
     taxon_pft_lookup$growthForm[i] = "tree"
-    taxon_pft_lookup$commonName[i] = "white ash"}
+    taxon_pft_lookup$commonName[i] = "white ash"
+    taxon_pft_lookup$family[i] = "Oleaceae"}
 
@@
     taxon_pft_lookup$growthForm[i] = "tree"
-    taxon_pft_lookup$commonName[i] = "black ash"}
+    taxon_pft_lookup$commonName[i] = "black ash"
+    taxon_pft_lookup$family[i] = "Oleaceae"}
 
@@
     taxon_pft_lookup$growthForm[i] = "tree"
-    taxon_pft_lookup$commonName[i] = "green ash"}
+    taxon_pft_lookup$commonName[i] = "green ash"
+    taxon_pft_lookup$family[i] = "Oleaceae"}
 
@@
     taxon_pft_lookup$growthForm[i] = "tree"
-    taxon_pft_lookup$commonName[i] = "ash"}
+    taxon_pft_lookup$commonName[i] = "ash"
+    taxon_pft_lookup$family[i] = "Oleaceae"}
 
@@
     taxon_pft_lookup$growthForm[i] = "shrub"
-    taxon_pft_lookup$commonName[i] = "salal"}
+    taxon_pft_lookup$commonName[i] = "salal"
+    taxon_pft_lookup$family[i] = "Ericaceae"}
 
@@
     taxon_pft_lookup$growthForm[i] = "tree"
-    taxon_pft_lookup$commonName[i] = "honey locust"}
+    taxon_pft_lookup$commonName[i] = "honey locust"
+    taxon_pft_lookup$family[i] = "Fabaceae"}
 
@@
     taxon_pft_lookup$growthForm[i] = "shrub"
-    taxon_pft_lookup$commonName[i] = "broom snakeweed"}
+    taxon_pft_lookup$commonName[i] = "broom snakeweed"
+    taxon_pft_lookup$family[i] = "Asteraceae"}
 
@@
     taxon_pft_lookup$growthForm[i] = "tree"
-    taxon_pft_lookup$commonName[i] = "silverbell"}
+    taxon_pft_lookup$commonName[i] = "silverbell"
+    taxon_pft_lookup$family[i] = "Styracaceae"}
 
@@
     taxon_pft_lookup$growthForm[i] = "tree"
-    taxon_pft_lookup$commonName[i] = "American witchhazel"}
+    taxon_pft_lookup$commonName[i] = "American witchhazel"
+    taxon_pft_lookup$family[i] = "Hamamelidaceae"}
 
@@
     taxon_pft_lookup$growthForm[i] = "tree"
-    taxon_pft_lookup$commonName[i] = "mountain holly"}
+    taxon_pft_lookup$commonName[i] = "mountain holly"
+    taxon_pft_lookup$family[i] = "Aquifoliaceae"}
 
@@
     taxon_pft_lookup$growthForm[i] = "tree"
-    taxon_pft_lookup$commonName[i] = "American holly"}
+    taxon_pft_lookup$commonName[i] = "American holly"
+    taxon_pft_lookup$family[i] = "Aquifoliaceae"}
 
@@
     taxon_pft_lookup$growthForm[i] = "tree"
-    taxon_pft_lookup$commonName[i] = "black walnut"}
+    taxon_pft_lookup$commonName[i] = "black walnut"
+    taxon_pft_lookup$family[i] = "Juglandaceae"}
 
@@
     taxon_pft_lookup$commonName[i] = "Utah juniper"
+    taxon_pft_lookup$family[i] = "Cupressaceae"
     taxon_pft_lookup$comments[i] = "scale-like needle leaves"}
@@
     taxon_pft_lookup$commonName[i] = "red cedar"
+    taxon_pft_lookup$family[i] = "Cupressaceae"
     taxon_pft_lookup$comments[i] = "scale-like needle leaves"}
@@
     taxon_pft_lookup$commonName[i] = "larch"
+    taxon_pft_lookup$family[i] = "Pinaceae"
     taxon_pft_lookup$comments[i] = "unusual genus with deciduous conifers"}
@@
     taxon_pft_lookup$growthForm[i] = "shrub"
-    taxon_pft_lookup$commonName[i] = "creosote bush"}
+    taxon_pft_lookup$commonName[i] = "creosote bush"
+    taxon_pft_lookup$family[i] = "Zygophyllaceae"}
 
@@
     taxon_pft_lookup$growthForm[i] = "tree"
-    taxon_pft_lookup$commonName[i] = "sweetgum"}
+    taxon_pft_lookup$commonName[i] = "sweetgum"
+    taxon_pft_lookup$family[i] = "Hamamelidaceae"}
 
@@
     taxon_pft_lookup$growthForm[i] = "tree"
-    taxon_pft_lookup$commonName[i] = "tuliptree"}
+    taxon_pft_lookup$commonName[i] = "tuliptree"
+    taxon_pft_lookup$family[i] = "Magnoliaceae"}
 
@@
     taxon_pft_lookup$growthForm[i] = "shrub"
-    taxon_pft_lookup$commonName[i] = "Amur honeysuckle"}
+    taxon_pft_lookup$commonName[i] = "Amur honeysuckle"
+    taxon_pft_lookup$family[i] = "Caprifoliaceae"}
 
@@
     taxon_pft_lookup$growthForm[i] = "tree" 
-    taxon_pft_lookup$commonName[i] = "Osage-orange"}
+    taxon_pft_lookup$commonName[i] = "Osage-orange"
+    taxon_pft_lookup$family[i] = "Moraceae"}
 
@@
     taxon_pft_lookup$commonName[i] = "magnolia"
+    taxon_pft_lookup$family[i] = "Magnoliaceae"
     taxon_pft_lookup$comments[i] = "varied genus with both deciduous and evergreen trees and shrubs"}
@@
     taxon_pft_lookup$growthForm[i] = "shrub"
-    taxon_pft_lookup$commonName[i] = "Cascade barberry"}
+    taxon_pft_lookup$commonName[i] = "Cascade barberry"
+    taxon_pft_lookup$family[i] = "Berberidaceae"}
 
@@
     taxon_pft_lookup$growthForm[i] = "tree"
-    taxon_pft_lookup$commonName[i] = "mango"}
+    taxon_pft_lookup$commonName[i] = "mango"
+    taxon_pft_lookup$family[i] = "Anacardiaceae"}
 
@@
     taxon_pft_lookup$growthForm[i] = "tree"
-    taxon_pft_lookup$commonName[i] = "Chinaberrytree"}
+    taxon_pft_lookup$commonName[i] = "Chinaberrytree"
+    taxon_pft_lookup$family[i] = "Meliaceae"}
 
@@
     taxon_pft_lookup$growthForm[i] = "tree"
-    taxon_pft_lookup$commonName[i] = "red mulberry"}
+    taxon_pft_lookup$commonName[i] = "red mulberry"
+    taxon_pft_lookup$family[i] = "Moraceae"}
 
@@
     taxon_pft_lookup$growthForm[i] = "tree"
-    taxon_pft_lookup$commonName[i] = "water tupelo"}
+    taxon_pft_lookup$commonName[i] = "water tupelo"
+    taxon_pft_lookup$family[i] = "Cornaceae"}
 
@@
     taxon_pft_lookup$growthForm[i] = "tree"
-    taxon_pft_lookup$commonName[i] = "swamp tupelo"}
+    taxon_pft_lookup$commonName[i] = "swamp tupelo"
+    taxon_pft_lookup$family[i] = "Cornaceae"}
 
@@
     taxon_pft_lookup$growthForm[i] = "tree"
-    taxon_pft_lookup$commonName[i] = "blackgum"}
+    taxon_pft_lookup$commonName[i] = "blackgum"
+    taxon_pft_lookup$family[i] = "Cornaceae"}
 
@@
     taxon_pft_lookup$growthForm[i] = "tree"
-    taxon_pft_lookup$commonName[i] = "sourwood"}
+    taxon_pft_lookup$commonName[i] = "sourwood"
+    taxon_pft_lookup$family[i] = "Ericaceae"}
 
@@
     taxon_pft_lookup$growthForm[i] = "tree"
-    taxon_pft_lookup$commonName[i] = "Engelmann spruce"}
+    taxon_pft_lookup$commonName[i] = "Engelmann spruce"
+    taxon_pft_lookup$family[i] = "Picea mariana"}
 
@@
     taxon_pft_lookup$growthForm[i] = "tree"
-    taxon_pft_lookup$commonName[i] = "white spruce"}
+    taxon_pft_lookup$commonName[i] = "white spruce"
+    taxon_pft_lookup$family[i] = "Picea mariana"}
 
@@
     taxon_pft_lookup$growthForm[i] = "tree"
-    taxon_pft_lookup$commonName[i] = "black spruce"}
+    taxon_pft_lookup$commonName[i] = "black spruce"
+    taxon_pft_lookup$family[i] = "Picea mariana"}
 
@@
     taxon_pft_lookup$growthForm[i] = "tree"
-    taxon_pft_lookup$commonName[i] = "spruce"}
+    taxon_pft_lookup$commonName[i] = "spruce"
+    taxon_pft_lookup$family[i] = "Pinaceae"}
 
@@
     taxon_pft_lookup$growthForm[i] = "tree"
-    taxon_pft_lookup$commonName[i] = "lodgepole pine"}
+    taxon_pft_lookup$commonName[i] = "lodgepole pine"
+    taxon_pft_lookup$family[i] = "Pinaceae"}
 
@@
     taxon_pft_lookup$growthForm[i] = "tree"
-    taxon_pft_lookup$commonName[i] = "shortleaf pine"}
+    taxon_pft_lookup$commonName[i] = "shortleaf pine"
+    taxon_pft_lookup$family[i] = "Pinaceae"}
 
@@
     taxon_pft_lookup$growthForm[i] = "tree"
-    taxon_pft_lookup$commonName[i] = "limber pine"}
+    taxon_pft_lookup$commonName[i] = "limber pine"
+    taxon_pft_lookup$family[i] = "Pinaceae"}
 
@@
     taxon_pft_lookup$growthForm[i] = "tree"
-    taxon_pft_lookup$commonName[i] = "longleaf pine"}
+    taxon_pft_lookup$commonName[i] = "longleaf pine"
+    taxon_pft_lookup$family[i] = "Pinaceae"}
 
@@
     taxon_pft_lookup$growthForm[i] = "tree"
-    taxon_pft_lookup$commonName[i] = "ponderosa pine"}
+    taxon_pft_lookup$commonName[i] = "ponderosa pine"
+    taxon_pft_lookup$family[i] = "Pinaceae"}
 
@@
     taxon_pft_lookup$growthForm[i] = "tree"
-    taxon_pft_lookup$commonName[i] = "pine"}
+    taxon_pft_lookup$commonName[i] = "pine"
+    taxon_pft_lookup$family[i] = "Pinaceae"}
 
@@
     taxon_pft_lookup$growthForm[i] = "tree"
-    taxon_pft_lookup$commonName[i] = "eastern white pine"}
+    taxon_pft_lookup$commonName[i] = "eastern white pine"
+    taxon_pft_lookup$family[i] = "Pinaceae"}
 
@@
     taxon_pft_lookup$growthForm[i] = "tree"
-    taxon_pft_lookup$commonName[i] = "loblolly pine"}
+    taxon_pft_lookup$commonName[i] = "loblolly pine"
+    taxon_pft_lookup$family[i] = "Pinaceae"}
 
@@
     taxon_pft_lookup$growthForm[i] = "tree"
-    taxon_pft_lookup$commonName[i] = "corcho bobo"}
+    taxon_pft_lookup$commonName[i] = "corcho bobo"
+    taxon_pft_lookup$family[i] = "Nyctaginaceae"}
 
@@
     taxon_pft_lookup$growthForm[i] = "tree"
-    taxon_pft_lookup$commonName[i] = "American sycamore"}
+    taxon_pft_lookup$commonName[i] = "American sycamore"
+    taxon_pft_lookup$family[i] = "Platanaceae"}
 
@@
     taxon_pft_lookup$growthForm[i] = "tree"
-    taxon_pft_lookup$commonName[i] = "eastern cottonwood"}
+    taxon_pft_lookup$commonName[i] = "eastern cottonwood"
+    taxon_pft_lookup$family[i] = "Salicaceae"}
 
@@
     taxon_pft_lookup$growthForm[i] = "tree"
-    taxon_pft_lookup$commonName[i] = "bigtooth aspen"}
+    taxon_pft_lookup$commonName[i] = "bigtooth aspen"
+    taxon_pft_lookup$family[i] = "Salicaceae"}
 
@@
     taxon_pft_lookup$growthForm[i] = "tree"
-    taxon_pft_lookup$commonName[i] = "cottonwood"}
+    taxon_pft_lookup$commonName[i] = "cottonwood"
+    taxon_pft_lookup$family[i] = "Salicaceae"}
 
@@
     taxon_pft_lookup$growthForm[i] = "tree"
-    taxon_pft_lookup$commonName[i] = "quaking aspen"}
+    taxon_pft_lookup$commonName[i] = "quaking aspen"
+    taxon_pft_lookup$family[i] = "Salicaceae"}
 
@@
     taxon_pft_lookup$growthForm[i] = "shrub"
-    taxon_pft_lookup$commonName[i] = "honey mesquite"}
+    taxon_pft_lookup$commonName[i] = "honey mesquite"
+    taxon_pft_lookup$family[i] = "Fabaceae"}
 
@@
     taxon_pft_lookup$growthForm[i] = "shrub"
-    taxon_pft_lookup$commonName[i] = "velvet mesquite"}
+    taxon_pft_lookup$commonName[i] = "velvet mesquite"
+    taxon_pft_lookup$family[i] = "Fabaceae"}
 
@@
     taxon_pft_lookup$growthForm[i] = "tree"
-    taxon_pft_lookup$commonName[i] = "American plum"}
+    taxon_pft_lookup$commonName[i] = "American plum"
+    taxon_pft_lookup$family[i] = "Rosaceae"}
 
@@
     taxon_pft_lookup$growthForm[i] = "shrub"
-    taxon_pft_lookup$commonName[i] = "Chickasaw plum"}
+    taxon_pft_lookup$commonName[i] = "Chickasaw plum"
+    taxon_pft_lookup$family[i] = "Rosaceae"}
 
@@
     taxon_pft_lookup$growthForm[i] = "tree"
-    taxon_pft_lookup$commonName[i] = "Mexican plum"}
+    taxon_pft_lookup$commonName[i] = "Mexican plum"
+    taxon_pft_lookup$family[i] = "Rosaceae"}
 
@@
     taxon_pft_lookup$growthForm[i] = "tree"
-    taxon_pft_lookup$commonName[i] = "black cherry"}
+    taxon_pft_lookup$commonName[i] = "black cherry"
+    taxon_pft_lookup$family[i] = "Rosaceae"}
 
@@
     taxon_pft_lookup$growthForm[i] = "tree"
-    taxon_pft_lookup$commonName[i] = "Douglas-fir"}
+    taxon_pft_lookup$commonName[i] = "Douglas-fir"
+    taxon_pft_lookup$family[i] = "Pinaceae"}
 
@@
     taxon_pft_lookup$growthForm[i] = "tree"
-    taxon_pft_lookup$commonName[i] = "white oak"}
+    taxon_pft_lookup$commonName[i] = "white oak"
+    taxon_pft_lookup$family[i] = "Fagaceae"}
 
@@
     taxon_pft_lookup$growthForm[i] = "tree"
-    taxon_pft_lookup$commonName[i] = "scarlet oak"}
+    taxon_pft_lookup$commonName[i] = "scarlet oak"
+    taxon_pft_lookup$family[i] = "Fagaceae"}
 
@@
     taxon_pft_lookup$growthForm[i] = "tree"
-    taxon_pft_lookup$commonName[i] = "southern red oak"}
+    taxon_pft_lookup$commonName[i] = "southern red oak"
+    taxon_pft_lookup$family[i] = "Fagaceae"}
 
@@
     taxon_pft_lookup$growthForm[i] = "tree"
-    taxon_pft_lookup$commonName[i] = "sand live oak"}
+    taxon_pft_lookup$commonName[i] = "sand live oak"
+    taxon_pft_lookup$family[i] = "Fagaceae"}
 
@@
     taxon_pft_lookup$commonName[i] = "Darlington oak"
-    taxon_pft_lookup$comments[i] = "semi-evergreen or tardily deciduous"}
+    taxon_pft_lookup$comments[i] = "semi-evergreen or tardily deciduous"
+    taxon_pft_lookup$family[i] = "Fagaceae"}
 
@@
     taxon_pft_lookup$growthForm[i] = "tree"
-    taxon_pft_lookup$commonName[i] = "turkey oak"}
+    taxon_pft_lookup$commonName[i] = "turkey oak"
+    taxon_pft_lookup$family[i] = "Fagaceae"}
 
@@
     taxon_pft_lookup$growthForm[i] = "tree"
-    taxon_pft_lookup$commonName[i] = "laurel oak"}
+    taxon_pft_lookup$commonName[i] = "laurel oak"
+    taxon_pft_lookup$family[i] = "Fagaceae"}
 
@@
     taxon_pft_lookup$growthForm[i] = "tree"
-    taxon_pft_lookup$commonName[i] = "overcup oak"}
+    taxon_pft_lookup$commonName[i] = "overcup oak"
+    taxon_pft_lookup$family[i] = "Fagaceae"}
 
@@
     taxon_pft_lookup$growthForm[i] = "tree"
-    taxon_pft_lookup$commonName[i] = "bur oak"}
+    taxon_pft_lookup$commonName[i] = "bur oak"
+    taxon_pft_lookup$family[i] = "Fagaceae"}
 
@@
     taxon_pft_lookup$growthForm[i] = "tree"
-    taxon_pft_lookup$commonName[i] = "sand post oak"}
+    taxon_pft_lookup$commonName[i] = "sand post oak"
+    taxon_pft_lookup$family[i] = "Fagaceae"}
 
@@
     taxon_pft_lookup$growthForm[i] = "tree"
-    taxon_pft_lookup$commonName[i] = "blackjack oak"}
+    taxon_pft_lookup$commonName[i] = "blackjack oak"
+    taxon_pft_lookup$family[i] = "Fagaceae"}
 
@@
     taxon_pft_lookup$growthForm[i] = "tree"
-    taxon_pft_lookup$commonName[i] = "swamp chestnut oak"}
+    taxon_pft_lookup$commonName[i] = "swamp chestnut oak"
+    taxon_pft_lookup$family[i] = "Fagaceae"}
 
@@
     taxon_pft_lookup$growthForm[i] = "tree"
-    taxon_pft_lookup$commonName[i] = "chestnut oak"}
+    taxon_pft_lookup$commonName[i] = "chestnut oak"
+    taxon_pft_lookup$family[i] = "Fagaceae"}
 
@@
     taxon_pft_lookup$growthForm[i] = "tree"
-    taxon_pft_lookup$commonName[i] = "chinquapin oak"}
+    taxon_pft_lookup$commonName[i] = "chinquapin oak"
+    taxon_pft_lookup$family[i] = "Fagaceae"}
 
@@
     taxon_pft_lookup$growthForm[i] = "tree"
-    taxon_pft_lookup$commonName[i] = "water oak"}
+    taxon_pft_lookup$commonName[i] = "water oak"
+    taxon_pft_lookup$family[i] = "Fagaceae"}
 
@@
     taxon_pft_lookup$growthForm[i] = "tree"
-    taxon_pft_lookup$commonName[i] = "cherrybark oak"}
+    taxon_pft_lookup$commonName[i] = "cherrybark oak"
+    taxon_pft_lookup$family[i] = "Fagaceae"}
 
@@
     taxon_pft_lookup$growthForm[i] = "tree"
-    taxon_pft_lookup$commonName[i] = "willow oak"}
+    taxon_pft_lookup$commonName[i] = "willow oak"
+    taxon_pft_lookup$family[i] = "Fagaceae"}
 
@@
     taxon_pft_lookup$growthForm[i] = "tree"
-    taxon_pft_lookup$commonName[i] = "northern red oak"}
+    taxon_pft_lookup$commonName[i] = "northern red oak"
+    taxon_pft_lookup$family[i] = "Fagaceae"}
 
@@
     taxon_pft_lookup$growthForm[i] = "tree"
-    taxon_pft_lookup$commonName[i] = "Shumard's oak"}
+    taxon_pft_lookup$commonName[i] = "Shumard's oak"
+    taxon_pft_lookup$family[i] = "Fagaceae"}
 
@@
     taxon_pft_lookup$commonName[i] = "oak"
+    taxon_pft_lookup$family[i] = "Fagaceae"
     taxon_pft_lookup$comments[i] = "varied genus with both deciduous and evergreen species"}
@@
     taxon_pft_lookup$growthForm[i] = "tree"
-    taxon_pft_lookup$commonName[i] = "post oak"}
+    taxon_pft_lookup$commonName[i] = "post oak"
+    taxon_pft_lookup$family[i] = "Fagaceae"}
 
@@
     taxon_pft_lookup$growthForm[i] = "tree"
-    taxon_pft_lookup$commonName[i] = "black oak"}
+    taxon_pft_lookup$commonName[i] = "black oak"
+    taxon_pft_lookup$family[i] = "Fagaceae"}
 
@@
     taxon_pft_lookup$commonName[i] = "live oak"
-    taxon_pft_lookup$comments[i] = "semi-deciduous"}
+    taxon_pft_lookup$comments[i] = "semi-deciduous"
+    taxon_pft_lookup$family[i] = "Fagaceae"}
 
@@
     taxon_pft_lookup$growthForm[i] = "shrub"
-    taxon_pft_lookup$commonName[i] = "Dahurian buckthorn"}
+    taxon_pft_lookup$commonName[i] = "Dahurian buckthorn"
+    taxon_pft_lookup$family[i] = "Rhamnaceae"}
 
@@
     taxon_pft_lookup$growthForm[i] = "shrub"
-    taxon_pft_lookup$commonName[i] = "fragrant sumac"}
+    taxon_pft_lookup$commonName[i] = "fragrant sumac"
+    taxon_pft_lookup$family[i] = "Anacardiaceae"}
 
@@
     taxon_pft_lookup$growthForm[i] = "shrub"
-    taxon_pft_lookup$commonName[i] = "smooth sumac"}
+    taxon_pft_lookup$commonName[i] = "smooth sumac"
+    taxon_pft_lookup$family[i] = "Anacardiaceae"}
 
@@
     taxon_pft_lookup$growthForm[i] = "tree"
-    taxon_pft_lookup$commonName[i] = "black locust"}
+    taxon_pft_lookup$commonName[i] = "black locust"
+    taxon_pft_lookup$family[i] = "Fabaceae"}
 
@@
     taxon_pft_lookup$growthForm[i] = "tree"
-    taxon_pft_lookup$commonName[i] = "sassafras"}
+    taxon_pft_lookup$commonName[i] = "sassafras"
+    taxon_pft_lookup$family[i] = "Lauraceae"}
 
@@
     taxon_pft_lookup$commonName[i] = "gum bully"
-    taxon_pft_lookup$comments[i] = "deciduous to semi-evergreen"}
+    taxon_pft_lookup$comments[i] = "deciduous to semi-evergreen"
+    taxon_pft_lookup$family[i] = "Sapotaceae"}
 
@@
     taxon_pft_lookup$commonName[i] = "sweetleaf"
+    taxon_pft_lookup$family[i] = "Symplocaceae"
     taxon_pft_lookup$comments[i] = "varied genus"}
@@
     taxon_pft_lookup$growthForm[i] = "tree"
-    taxon_pft_lookup$commonName[i] = "Pacific yew"}
+    taxon_pft_lookup$commonName[i] = "Pacific yew"
+    taxon_pft_lookup$family[i] = "Taxaceae"}
 
@@
     taxon_pft_lookup$growthForm[i] = "tree"
-    taxon_pft_lookup$commonName[i] = "bee-bee tree"}
+    taxon_pft_lookup$commonName[i] = "bee-bee tree"
+    taxon_pft_lookup$family[i] = "Rutaceae"}
 
@@
     taxon_pft_lookup$growthForm[i] = "tree"
-    taxon_pft_lookup$commonName[i] = "Chinese tallow"}
+    taxon_pft_lookup$commonName[i] = "Chinese tallow"
+    taxon_pft_lookup$family[i] = "Euphorbiaceae"}
 
@@
     taxon_pft_lookup$growthForm[i] = "tree"
-    taxon_pft_lookup$commonName[i] = "eastern hemlock"}
+    taxon_pft_lookup$commonName[i] = "eastern hemlock"
+    taxon_pft_lookup$family[i] = "Pinaceae"}
 
@@
     taxon_pft_lookup$growthForm[i] = "tree"
-    taxon_pft_lookup$commonName[i] = "western hemlock"}
+    taxon_pft_lookup$commonName[i] = "western hemlock"
+    taxon_pft_lookup$family[i] = "Pinaceae"}
 
@@
     taxon_pft_lookup$growthForm[i] = "tree"
-    taxon_pft_lookup$commonName[i] = "winged elm"}
+    taxon_pft_lookup$commonName[i] = "winged elm"
+    taxon_pft_lookup$family[i] = "Ulmaceae"}
 
@@
     taxon_pft_lookup$growthForm[i] = "tree"
-    taxon_pft_lookup$commonName[i] = "American elm"}
+    taxon_pft_lookup$commonName[i] = "American elm"
+    taxon_pft_lookup$family[i] = "Ulmaceae"}
 
@@
     taxon_pft_lookup$growthForm[i] = "tree"
-    taxon_pft_lookup$commonName[i] = "cedar elm"}
+    taxon_pft_lookup$commonName[i] = "cedar elm"
+    taxon_pft_lookup$family[i] = "Ulmaceae"}
 
@@
     taxon_pft_lookup$growthForm[i] = "tree"
-    taxon_pft_lookup$commonName[i] = "slippery elm"}
+    taxon_pft_lookup$commonName[i] = "slippery elm"
+    taxon_pft_lookup$family[i] = "Ulmaceae"}
 
@@
     taxon_pft_lookup$commonName[i] = "elm"
-    taxon_pft_lookup$comments[i] = "varied genus contains some semi-deciduous or evergreen species"}
+    taxon_pft_lookup$comments[i] = "varied genus contains some semi-deciduous or evergreen species"
+    taxon_pft_lookup$family[i] = "Ulmaceae"}
 
@@
     taxon_pft_lookup$growthForm[i] = "shrub"
-    taxon_pft_lookup$commonName[i] = "red huckleberry"}
+    taxon_pft_lookup$commonName[i] = "red huckleberry"
+    taxon_pft_lookup$family[i] = "Ericaceae"}
 
@@
     taxon_pft_lookup$commonName[i] = "desert zinnia"
+    taxon_pft_lookup$family[i] = "Asteraceae"
     taxon_pft_lookup$comments[i] = "neon growthForm was shrub; evergreen but cold-deciduous"}
@@
 
+```{r check_col_family}
+# unique values in the family column: 
+print(unique(taxon_pft_lookup$family))
+```
+
 ```{r taxon_pft_table}
@@
 write_csv(x = taxon_pft_lookup,
-          file = here::here("R","output","taxon_pft_lookup.csv"))
+          file = here::here("analysis", "data", "derived_data","taxon_pft_lookup.csv"))
 ```
@@
 pft_labels <- taxon_pft_lookup %>% 
-  select(genusSpecies, commonName, pft_MRA, pft_MODIS, phenology, leafType, growthForm)
+  select(genusSpecies, commonName, pft_MRA, pft_MODIS, phenology, leafType, growthForm, family)
 
@@
 cleaned_spectra_pft <- right_join(cleaned_spectra, pft_labels, by = "genusSpecies")
+
+# reorder the columns 
+cleaned_spectra_pft <- cleaned_spectra_pft %>% 
+  dplyr::relocate(scientificName, family, genus, species, genusSpecies, 
+                  commonName, growthForm, phenology, leafType, 
+                  pft_MRA, pft_MODIS, .after = last_col())
 
